### 2018-08-04 contest/abstract.Rmd
Started an abstract.  It is in the contest folder

---
title: "On the Surface: Analyzing the effect of court type in tennis grand slams"
author: "The Competent Ones"
date: "August 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.  The surface type may substantially effect play including ball speed, height, and spin as well as player speed and agility.  It is also believed that play style and practice habits may contribute to different results across surface types.  For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.  On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to produce better results, but perhaps has suffered on grass as a consequence.

In this analysis, we examine the result of top 100 players in grand slams from 2013-2018 across the four different surfaces.  We create a hierarchical model with fixed and random effects in order to predict the number of points won in a match.  We take into consideration player-specific effects, nationality (which is thought to have an effect on playstyle), sex, ranking, ELO, and game statistics.

We assess the fit of our model using (MSE, BIC??) in addition to "common sense" factors such as the fact that Rafael Nadal should in some way be indicated as a superior clay court player by our model.  We compare the results of top 100 players across slams as well as look into the results of Nadal, Federer, and S. Williams.

### 2018-08-05 contest/abstract.Rmd

@@
 Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
-The surface type may substantially effect play including ball speed, height, and spin as well as player speed and agility.
+The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.
 It is also believed that play style and practice habits may contribute to different results across surface types.
 For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.
-On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to produce better results, but perhaps has suffered on grass as a consequence.
+On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.
 
-In this analysis, we examine the result of top 100 players in grand slams from 2013-2018 across the four different surfaces.
-We create a hierarchical model with fixed and random effects in order to predict the number of points won in a match.
-We take into consideration player-specific effects, nationality (which is thought to have an effect on playstyle), sex, ranking, ELO, and game statistics.
+In this analysis, we examine the result of the top 100 players in grand slams from 2013-2018 across the four different surfaces.
+We create a hierarchical model with fixed and random effects to predict the number of points won in a match.
+We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.
 
-We assess the fit of our model using (MSE, BIC??) in addition to "common sense" factors such as the fact that Rafael Nadal should in some way be indicated as a superior clay court player by our model.
-We compare the results of top 100 players across slams as well as look into the results of Nadal, Federer, and S.
+We assess the fit of our model using standard statistical techniques (e.g.
+MSE, AIC, BIC, residual diagnostics) in addition to "common knowledge" factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).
+We compare the results of top 100 players across grand slams to examine the effect of court surface.
+We also provide an in-depth analysis of Nadal, Federer, and S.
 Williams.

### 2018-08-05 contest/eda/eda.Rmd
Started some EDA.  At least Fed, Nadal, and SWilliams are in point by point.  We may want to use ATP/WTA match data otherwise

---
title: "EDA"
author: "The Competent Ones"
date: "August 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(devtools)
library(deuce)
library(knitr)
devtools::load_all("..") # Loads courtsports package
```

```{r data-extraction	}
data("point_by_point")
pbp = tbl_df(point_by_point)
pbp$wim = grepl("Wimbledon", pbp$tny_name)
pbp$french = grepl("FrenchOpen", pbp$tny_name)
pbp$aus = grepl("AustralianOpen", pbp$tny_name)
pbp$us = grepl("USOpen", pbp$tny_name)
pbp$qual = grepl("Qualifying", pbp$tny_name)

# Create data for Grand Slam finals
gsf = filter(pbp, !qual & (wim | french | us | aus))
```
```{r numbers}
dim(gsf)
colnames(gsf)
kable(table(gsf$tny_name))

any(grepl("Serena", gsf$server1))
any(grepl("Nadal", gsf$server1))
any(grepl("Federer", gsf$server1))

## Opponents of our 3
## Serena
gsf$server2[grepl("Serena", gsf$server1)]
## Nadal
gsf$server2[grepl("Nadal", gsf$server1)]
## Fed
gsf$server2[grepl("Federer", gsf$server1)]
```

### 2018-08-10 contest/eda/eda.Rmd
Looked at some EDA with atp_matches

@@
 ```
+
+## Looking at ATP (Men's) matches
+
+```{r}
+data(atp_matches)
+
+gs <- atp_matches %>% filter(tourney_level == "Grand Slams" & year >= 2013)
+sort(unique(gs$tourney_name))
+colnames(gs)
+
+## Add new column of points won
+gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon)
+gs <- gs %>% mutate(l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
+gs <- gs %>% mutate(w_morepts = w_pointswon >= l_pointswon)
+n_correct <- sum(gs$w_morepts, na.rm = TRUE)
+na_pt_indices <- which(is.na(gs$w_morepts))
+n_correct / (nrow(gs) - length(na_pt_indices)) # more points won tells us who won >95% of the time
+gs[na_pt_indices,]  ## All NA are straight up retirements so not really NA
+incorrect_inds <- which(!gs$w_morepts)
+table(gs[incorrect_inds, "tourney_name"])
+table(gs[incorrect_inds, "Retirement"])
+```
+
+Australian Open with most wins with loser winning more points than winner.	Of all the matches where the winners won fewer points than the losers, only 10 were retirements.
+Which means you are more likely to quit when you are down (10 night before).

### 2018-08-17 contest/eda/eda.Rmd
made some hsitograms

--- 2018-08-05 contest/eda/eda.Rmd
@@
 ```
+
+## Looking at ATP (Men's) matches
+
+```{r}
+data(atp_matches)
+
+gs <- atp_matches %>% filter(tourney_level == "Grand Slams" & year >= 2013)
+sort(unique(gs$tourney_name))
+colnames(gs)
+
+## Add new column of points won
+gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon)
+gs <- gs %>% mutate(l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
+gs <- gs %>% mutate(w_morepts = w_pointswon >= l_pointswon)
+n_correct <- sum(gs$w_morepts, na.rm = TRUE)
+na_pt_indices <- which(is.na(gs$w_morepts))
+n_correct / (nrow(gs) - length(na_pt_indices)) # more points won tells us who won >95% of the time
+gs[na_pt_indices,]  ## All NA are straight up retirements so not really NA
+incorrect_inds <- which(!gs$w_morepts)
+table(gs[incorrect_inds, "tourney_name"])
+table(gs[incorrect_inds, "Retirement"])
+
+```
+
+Australian Open with most wins with loser winning more points than winner.	Of all the matches where the winners won fewer points than the losers, only 10 were retirements.
+Which means you are more likely to quit when you are down (10 night before).
+
+```{r data-cleaning}
+library(forcats)
+library(stringr)
+## Combine US Open
+gs$tournament <- factor(gs$tourney_name)
+gs$tournament <- gs$tournament %>% fct_collapse("US Open" = c("US Open", "Us Open"))
+
+## Extract te number of sets in each match
+no_ret <- gsub(" RET", "", gs$score)
+set_scores <- str_split(no_ret, " ")
+n_sets <- sapply(set_scores, length)
+gs$n_sets <- n_sets
+
+```
+```{r some-plots}
+library(ggplot2)
+library(reshape2)
+
+g <- ggplot(data = gs, aes(x = w_pointswon)) + geom_histogram()
+g
+
+g2 <- ggplot(data = gs, aes(x = w_pointswon)) +
+    geom_histogram(aes(y=..density..)) +
+    facet_grid(~tournament)
+g2
+
+##
+bp <- ggplot(data = gs, aes(x = n_sets)) +
+    geom_bar(aes(y = ..prop..)) +
+    facet_grid(~tournament)
+bp
+
+g3 <- ggplot(data = gs, aes(x = w_pointswon)) +
+    geom_histogram(aes(y=..density..)) +
+    facet_grid(n_sets~tournament)
+g3
+
+```

### 2018-08-28 contest/eda/eda.Rmd
More EDA and thoughts

@@
 ```
+```{r}
+## Moving to boxplots
+g2 <- ggplot(data = gs,
+             aes(x = tournament, y= minutes,
+                            group = tournament, col = tournament)) +
+    geom_boxplot(size = 1)  + facet_grid(~n_sets)
+g2
+
+```
+It appears Wimbledon has the shortest matches.
+	## Retirements by court?
+
+```{r}
+## More boxplots
+g2 <- ggplot(data = gs,
+             aes(x = winner_ioc,
+                            group = tournament, fill = tournament)) +
+    geom_bar(aes(y=..prop..), size = 1)  + facet_grid(~tournament) + 
+    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip()
+g2
+
+## RG has least amount of double faults
+```
+
++ Find games won and lost
++ Find service games won/lost
++ Look at retirements by court
++ Look at wins by country by court
++ Find number of tiebreaks by court
++ Need to normalize wins by country by n_players in tournament (but both are interesting)
++ Look at gs point by point
++ Need to get French Open 2018 results
++ Get Women's data and combine it
++ Look atGS_point_by_point.  Is it complete?  Is it complete for rafa, serena, fed?
++ Facet by round
++ Do above but instead predict winner games, or winner sets, loser games, or loser sets
++ get hex colors of GS

### 2018-09-07 contest/surface-type/paper-tennis.Rmd
Data ready to go!

---
title: "Analyzing the Effect of Surface in Tennis Grand Slams"
author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
date: "September 7, 2018"
output: pdf_document
bibliography: master.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(deuce)
library(dplyr)
library(ggplot2)
library(forcats)
library(stringr)
```

```{r data-load, include = FALSE}
## Load WTA and ATP matches from deuce package
data(atp_matches)
data(wta_matches)
wta_matches$league <- "WTA"
atp_matches$league <- "ATP"

## Combining WTA and ATP
combined_tennis <- rbind(wta_matches, atp_matches)

## Subsetting from 2013 through 2017 inclusive, all the grand slams
gs <- combined_tennis %>% filter(tourney_level == "Grand Slams" & year >= 2013 & year <= 2017)
## Get rid of women's qualifiers
gs <- gs %>% filter(!(round %in% c("Q1", "Q2", "Q3", "Q4")))
dim(gs)

saveRDS(gs, "../../data/gs_2013_2017.rda")

```

```{r data-cleaning, include = FALSE}
## Add new columns of winner points won, loser points won, and whether the winner had more points
gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon)
gs <- gs %>% mutate(l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
gs <- gs %>% mutate(w_morepts = w_pointswon >= l_pointswon)

## Extract number of sets in each match
no_ret <- gsub(" RET", "", gs$score)
set_scores <- str_split(no_ret, " ")
n_sets <- sapply(set_scores, length)
gs$n_sets <- n_sets

## Data cleaning the tournament names
gs$tournament <- factor(gs$tourney_name)
gs$tournament <- gs$tournament %>% fct_collapse("US Open" = c("US Open", "Us Open"), 
                                                "French Open" = c("French Open", "Roland Garros"))
```

```{r games-won, include = FALSE, warning = FALSE}
## Extract games won for winner and loser
score_list <- strsplit(gs$score, " ")
games_won <- sapply(score_list, strsplit, split = "-|\\(")
gs$w_gameswon <- sapply(games_won, function(list){
  sum(as.integer(sapply(list, "[", 1)), na.rm = TRUE)
})
gs$l_gameswon <- sapply(games_won, function(list){
  sum(as.integer(sapply(list, "[", 2)), na.rm = TRUE)
})

```

```{r sets, include = FALSE, warning = FALSE}
## Extract number of sets for each match
no_ret <- gsub(" RET", "", gs$score)
set_scores <- str_split(no_ret, " ")
n_sets <- sapply(set_scores, length)
gs$n_sets <- n_sets

```

```{r bp-lost, include = FALSE}
gs$w_bplost <- gs$w_bpFaced - gs$w_bpSaved
gs$l_bplost <- gs$l_bpFaced - gs$l_bpSaved
```

## Introduction
Example reference[@cdc-measles2018]

## Data

## Methods

## Results

## Discussion

## References

### 2018-09-07 contest/surface-type/paper-tennis.Rmd
Minor formatting issues

@@
 author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
-date: "September 7, 2018"
-output: pdf_document
+date: '`r format(Sys.time(), "%B %d, %Y")`'
+output:
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: no
+abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
+The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.
+It is also believed that play style and practice habits may contribute to different results across surface types.
+For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.
+On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.
+In this analysis, we examine the result of the top 100 players in grand slams from 2013-2017 across the four different surfaces.
+We create a hierarchical model with fixed and random effects to predict the number of points won in a match.
+We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.
+We assess the fit of our model using standard statistical techniques (e.g.
+MSE, AIC, BIC, residual diagnostics) in addition to `common knowledge' factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).
+We compare the results of top 100 players across grand slams to examine the effect of court surface.
+We also provide an in-depth analysis of Nadal, Federer, and S.
+Williams."
 bibliography: master.bib

### 2018-09-07 contest/surface-type/paper-tennis.tex
Minor formatting issues

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Analyzing the Effect of Surface in Tennis Grand Slams},
            pdfauthor={Kayla Frisoli, Shannon Gallagher, and Amanda Luby},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 07, 2018}

\begin{document}
\maketitle
\begin{abstract}
Tennis grand slams consist of the Australian Open, French Open,
Wimbledon, and US Open, which are played on hard (Plexicushion), clay,
grass, and hard (DecoTurf) courts, respectively. The surface type may
substantially impact ball speed, height, and spin as well as player
speed and agility. It is also believed that play style and practice
habits may contribute to different results across surface types. For
example, Rafael Nadal is thought to be the best clay court player of all
time whereas Roger Federer is particularly known for dominance at
Wimbledon. On the women's side, Serena Williams once struggled on clay
courts but has seemingly transformed her style to perform better on clay
courts, but has perhaps suffered on grass as a consequence. In this
analysis, we examine the result of the top 100 players in grand slams
from 2013-2017 across the four different surfaces. We create a
hierarchical model with fixed and random effects to predict the number
of points won in a match. We take into consideration player-specific
effects, nationality (which is thought to have an effect on play style),
sex, ranking, ELO, and game statistics. We assess the fit of our model
using standard statistical techniques (e.g.~MSE, AIC, BIC, residual
diagnostics) in addition to `common knowledge' factors (for instance,
Rafael Nadal should be indicated as a superior clay court player by the
model). We compare the results of top 100 players across grand slams to
examine the effect of court surface. We also provide an in-depth
analysis of Nadal, Federer, and S. Williams.
\end{abstract}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Example reference(Centers for Disease Control and Prevention 2018)

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-cdc-measles2018}{}%
Centers for Disease Control and Prevention. 2018. ``Measles History.''
Available online at \url{https://www.cdc.gov/measles/about}.

\end{document}

### 2018-09-07 contest/surface-type/paper-tennis.Rmd
Adding some format stuff

@@
 library(stringr)
+library(knitr)
+library(kableExtra)
 ```
@@
 ## Data
+The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues (ATP and WTA).
+Each match has `r ncol(gs)` attributes, many of which are redundant.
+We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
+Additionally, we take into account the number of sets in a match, the surface type, minutes played, and round of the tournament.
+A subset of the data is shown in Table \ref{tab:data}.
+
+```{r tab-data, include = TRUE, echo = FALSE}
+set.seed(19)
+s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
+rand_inds <- sample(s_id, 5)
+cols <- c("tourney_name", "year", "winner_ioc", "w_pointswon", "winner_name", "winner_rank", "l_pointswon", "loser_name")
+knitr::kable(gs[rand_inds, cols], format = "latex",
+             row.names = FALSE, col.names = c("Tournament", "Year", "W. Country", "W. Points", "Winner", "W. Rank", "L. Points", "Loser"),
+             caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
+             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
+```
 

### 2018-09-07 contest/surface-type/paper-tennis.tex
Adding some format stuff

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage[table]{xcolor}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+
 \begin{document}
@@
 
+The data consists of 5080 matches split evenly over the four grand slams
+and the two leagues (ATP and WTA). Each match has 82 attributes, many of
+which are redundant. We focus on the following attributes for both the
+winner and loser of the match: games won, points won, retirement, break
+points faced, break points saved, aces, country of origin, and player
+attributes. Additionally, we take into account the number of sets in a
+match, the surface type, minutes played, and round of the tournament. A
+subset of the data is shown in Table \ref{tab:data}.
+
+\rowcolors{2}{gray!6}{white}
+\begin{table}
+
+\caption{\label{tab:tab-data}\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.}
+\centering
+\begin{tabular}[t]{lrlrlrrl}
+\hiderowcolors
+\toprule
+Tournament & Year & W. Country & W. Points & Winner & W. Rank & L. Points & Loser\\
+\midrule
+\showrowcolors
+US Open & 2014 & USA & 58 & Serena Williams & 1 & 31 & Taylor Townsend\\
+US Open & 2013 & ESP & 136 & Rafael Nadal & 2 & 112 & Philipp Kohlschreiber\\
+Australian Open & 2015 & ESP & 91 & Rafael Nadal & 3 & 51 & Mikhail Youzhny\\
+US Open & 2013 & USA & 65 & Serena Williams & 1 & 41 & Yaroslava Shvedova\\
+Australian Open & 2017 & USA & 73 & Serena Williams & 2 & 60 & Lucie Safarova\\
+\bottomrule
+\end{tabular}
+\end{table}
+\rowcolors{2}{white}{white}
+
 \hypertarget{methods}{%

### 2018-09-07 contest/eda/eda.Rmd
kayla small edits

@@
 date: "August 4, 2018"
-output: html_document
+output: 
+  html_document:
+    toc:  true
+    toc_float:  true
+    code_folding: hide
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(message = FALSE)
+knitr::opts_chunk$set(warning = FALSE)
+
 ```
@@
 library(knitr)
-devtools::load_all("..") # Loads courtsports package
+# devtools::load_all("..") # Loads courtsports package
 ```
@@
 ```
+
 ```{r numbers}

### 2018-09-07 contest/surface-type/paper-tennis.Rmd
kayla small edits; remove saveds

@@
 
-saveRDS(gs, "../../data/gs_2013_2017.rda")
+# saveRDS(gs, "../../data/gs_2013_2017.rda")
 
@@
 ## Add new columns of winner points won, loser points won, and whether the winner had more points
-gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon)
-gs <- gs %>% mutate(l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
-gs <- gs %>% mutate(w_morepts = w_pointswon >= l_pointswon)
+gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon,
+                    l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
 
@@
 
+```{r}
+remove_from_loser <- names(gs)[grep("^w|^W", names(gs))] %>%
+  .[-which(. == "winner_rank")]
+
+remove_from_winner <- names(gs)[grep("^l|^L", names(gs))] %>%
+  .[-which(. %in% c("loser_rank", "league"))]
+
+gs_l <- dplyr::select(gs, -remove_from_loser) %>%
+  rename(opponent_rank = winner_rank)
+gs_w <- dplyr::select(gs, -remove_from_winner)  %>%
+  rename(opponent_rank = loser_rank)
+
+names(gs_w) <- gsub("^winner_|^W|^w_", "", names(gs_w))
+names(gs_l) <- gsub("^loser_|^L|^l_", "", names(gs_l))
+
+player_gs <- bind_rows(gs_w, gs_l)
+
+# saveRDS(player_gs, "../../data/player_gs.rda")
+```
+
 ## Introduction

### 2018-09-08 contest/surface-type/paper-tennis.Rmd
Fixed games won and added data cleaning to package

--- 2018-09-07 contest/surface-type/paper-tennis.Rmd
@@
 library(kableExtra)
+
+## To change into library(courtsports)
+devtools::load_all("../..")
 ```
 
-```{r data-load, include = FALSE}
+```{r data-cleaning, include = FALSE}
 ## Load WTA and ATP matches from deuce package
@@
 data(wta_matches)
-wta_matches$league <- "WTA"
-atp_matches$league <- "ATP"
 
-## Combining WTA and ATP
-combined_tennis <- rbind(wta_matches, atp_matches)
+## Clean data
+gs <- clean_matches_data(atp_matches = atp_matches,
+                         wta_matches = wta_matches,
+                         start_year = 2013,
+                         end_year = 2017,
+                         include_quals = FALSE,
+                         level = "Grand Slams")
 
-## Subsetting from 2013 through 2017 inclusive, all the grand slams
-gs <- combined_tennis %>% filter(tourney_level == "Grand Slams" & year >= 2013 & year <= 2017)
-## Get rid of women's qualifiers
-gs <- gs %>% filter(!(round %in% c("Q1", "Q2", "Q3", "Q4")))
-dim(gs)
+## Extract sets, points, and games
+gs <- extract_sgp(gs)
 
-saveRDS(gs, "../../data/gs_2013_2017.rda")
+# devtools::use_data(gs, overwrite = TRUE)
 
-```
+## Make into player data
+gs_players <- matches_to_player_data(gs,
+                                     w_opp_vars = "loser_rank",
+                                     l_opp_vars = "winner_rank",
+                                     opp_var_names = "opponent_rank")
 
-```{r data-cleaning, include = FALSE}
-## Add new columns of winner points won, loser points won, and whether the winner had more points
-gs <- gs %>% mutate(w_pointswon = w_1stWon + w_2ndWon + l_svpt - l_1stWon - l_2ndWon)
-gs <- gs %>% mutate(l_pointswon = l_1stWon + l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
-gs <- gs %>% mutate(w_morepts = w_pointswon >= l_pointswon)
+# devtools::use_data(gs_players, overwrite = TRUE)
 
-## Extract number of sets in each match
-no_ret <- gsub(" RET", "", gs$score)
-set_scores <- str_split(no_ret, " ")
-n_sets <- sapply(set_scores, length)
-gs$n_sets <- n_sets
-
-## Data cleaning the tournament names
-gs$tournament <- factor(gs$tourney_name)
-gs$tournament <- gs$tournament %>% fct_collapse("US Open" = c("US Open", "Us Open"), 
-                                                "French Open" = c("French Open", "Roland Garros"))
-```
-
-```{r games-won, include = FALSE, warning = FALSE}
-## Extract games won for winner and loser
-score_list <- strsplit(gs$score, " ")
-games_won <- sapply(score_list, strsplit, split = "-|\\(")
-gs$w_gameswon <- sapply(games_won, function(list){
-  sum(as.integer(sapply(list, "[", 1)), na.rm = TRUE)
-})
-gs$l_gameswon <- sapply(games_won, function(list){
-  sum(as.integer(sapply(list, "[", 2)), na.rm = TRUE)
-})
-
-```
-
-```{r sets, include = FALSE, warning = FALSE}
-## Extract number of sets for each match
-no_ret <- gsub(" RET", "", gs$score)
-set_scores <- str_split(no_ret, " ")
-n_sets <- sapply(set_scores, length)
-gs$n_sets <- n_sets
-
-```
-
-```{r bp-lost, include = FALSE}
-gs$w_bplost <- gs$w_bpFaced - gs$w_bpSaved
-gs$l_bplost <- gs$l_bpFaced - gs$l_bpSaved
 ```

### 2018-09-08 R/clean-data.R

@@ -1,1 +1,155 @@
-## Hello
+#' Clean the matches data from deuce atp_maches and wta_matches
+#'
+#' @param atp_matches atp_matches object from R package deuce (Men's).  Default is NULL
+#' @param wta_matches atp_matches object from R package deuce (Women's).  Default is NULL
+#' @param start_year start year (earliest is 1968)
+#' @param end_year end year (latest available is 2017)
+#' @param include_quals logical to include qualifying rounds.  Default is FALSE
+#' @param level level of tournament, corresponding to column from atp_matches/wta_matches.  Default is "Grand Slams"
+#' @return cleaned and subsetted data set, combined if there is both an atp_matches and wta_data set.  A new column called "league" is added and has value "ATP" or "WTA".  We also clean the tourney_name making a new column called tournament
+#' @importFrom magrittr %>%
+clean_matches_data <- function(atp_matches = NULL, wta_matches = NULL,
+                               start_year,
+                               end_year,
+                               include_quals = FALSE,
+                               level = "Grand Slams"){
+
+    ## Combine the two and add a "league" variable
+    matches <- NULL
+    if(is.null(atp_matches) & is.null(wta_matches)) {
+        stop("Both atp_matches and wta_matches are NULL")
+    }
+    if(!is.null(atp_matches)) {
+        atp_matches$league <- "ATP"
+    }
+    if(!is.null(wta_matches)){
+        wta_matches$league <- "WTA"
+    }
+    matches <- rbind(wta_matches, atp_matches)
+
+    ## Subset to proper years
+    matches <- matches  %>% dplyr::filter(tourney_level == level &
+                                   year >= start_year & year <= end_year)
+
+    ## Get rid of qualifiers if appropriate
+    if(!include_quals){
+        matches <- matches %>% dplyr::filter(!(round %in% c("Q1", "Q2", "Q3", "Q4")))
+    }
+
+    matches$tournament <- factor(matches$tourney_name) %>%
+        forcats::fct_collapse("US Open" = c("US Open", "Us Open"),
+                     "French Open" = c("French Open", "Roland Garros"))
+
+    return(matches)
+
+}
+
+
+#' Extract set, games, and points for winner and loser
+#'
+#' @param matches object from deuce package or clean_matches_data.
[…]

### 2018-09-08 R/clean-data.R
Made game extracting function more readable and fixed set extraction

@@ -51,7 +51,6 @@ clean_matches_data <- function(atp_matches = NULL, wta_matches = NULL,
 #' @return matches with extra columns w_pointswon, l_pointswon, w_gameswon, l_gameswon, w_setswon, l_setswon
 extract_sgp<- function(matches){
 
-    browser()
     ## Points
 
     matches <- matches %>% dplyr::mutate(
@@ -61,46 +60,35 @@ extract_sgp<- function(matches){
                                           l_2ndWon + w_svpt - w_1stWon - w_2ndWon)
 
     ## Games
-    score_list <- strsplit(matches$score, " ")
-    games_won <- lapply(score_list, strsplit, split = "-|\\(")
-    matches$w_gameswon <- sapply(games_won, function(list){
-      win_list <- lapply(list, "[", 1)
-      win_list <- unlist(win_list)
-      win_list <- gsub("RET|W\\/O", "0", win_list)  ## Walkover or retirement
-      if(any(is.na(win_list)))  win_list <-  0
-      sum(as.integer(win_list))
+    ## Get rid of retirements and walkover, the leading space, and any tiebreaker scores
+    score_no_retire <- gsub("RET|W\\/O| RET| W\\/O|\\([0-9]+\\)", "", matches$score)
+    score_list <- strsplit(score_no_retire, " ")
+    matches$w_gameswon <- sapply(score_list, function(vec){
+      sum(as.integer(gsub("\\-[0-9]+$","", vec)))
     })
 
 
-    matches$l_gameswon <- sapply(games_won, function(list){
-      win_list <- lapply(list, "[", 2)
-      win_list <- unlist(win_list)
-      win_list <- gsub("RET|W\\/O", "0", win_list)  ## Walkover or retirement
-      if(any(is.na(win_list)))  win_list <-  0
-      sum(as.integer(win_list))
+    matches$l_gameswon <- sapply(score_list, function(vec){
+      sum(as.integer(gsub("^[0-9]+\\-","", vec)))
     })
+
     ## Sets
-
     ## a set is won if they scored more games in the set than opponent
-    matches$w_setswon <-  sapply(games_won, function(list){
-        sum(sapply(list, function(vec){
-            if(is.null(dim(vec))){
-               return((as.integer(vec[1]) - as.integer(vec[2])) > 0)
-            } else {
-                return((as.integer(vec[,1]) - as.integer(vec[, 2])) > 0)
-            }
-        }), na.rm = TRUE)
[…]

### 2018-09-09 contest/surface-type/paper-tennis.tex
Fixed games won and added data cleaning to package
started intro and expanded data section.  Updated README

@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{5}
 % Redefines (sub)paragraphs to behave more like sections
@@
   \postdate{\par}
-    \date{September 07, 2018}
+    \date{September 08, 2018}
 
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\hypertarget{sec:iintro}{%
+\section{Introduction}\label{sec:iintro}}
 
-Example reference(Centers for Disease Control and Prevention 2018)
+Rafael Nadal is known as the ``King of Clay'' in tennis, having won 11
+out of his current 17 grand slams titles at the French Open, which is
+played on a clay surface {[}cite{]}. In contrast, his rival Roger
+Federer has won his most grand slam titles (8 out of 20) at Wimbledon,
+which is played on grass. On the women's side, Serena Williams has been
+dominant both on hard court (7 titles at the Australian Open and 6 at US
+Open) and grass (7 at Wimbledon). This trend extends to other top
+players, who seem to have better results at some grand slams than
+others. More broadly, it seems that country of origin has an interaction
+effect with court type. For example, Spanish players seem to excel on
+clay courts and Americans have great success at Wimbledon despite grass
+courts not being of wide use in the USA. It also worth questioning
+whether the US Open and Australian Open should be grouped together as
+hard courts despite having different surface compositions {[}cite{]}. In
+this paper, we analyze the results of grand slam players from 2013-2017,
+and we
 
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Determine if and how court surface effects players by implementation
+  of a series of nested hierarchical models
+\item
+  Examine how Nadal, Federer, and Williams' play differs by surface
+\item
+  Assess whether we can group the two hard court surfaces together.
+\end{enumerate}
+
+As to issue (1) the hiearchical modelling, we use models similar to
+{[}cite{]} and {[}cite{]}. More specifically, we model the players'
+expected points in a match based on the player's own characteristics,
+the court/tournament effects, and the oponent's ranking. We use a
+Bayesian method because in {[}cite,cite{]} they had success in similar
+situations.
+
+For issue (2) the player analysis of Nadal, Federer, and Williams, we
+examine whether our model passes the ``common sense''" tests (e.g.~these
+players should win often when contrasting them to opponents) like in
+{[}ventura something{]} and whether these players do have surface
+apparent effects.
+
+Finally, for issue (3), we use clustering methods {[}cite{]} in order to
+determine which court surface types are more similar to one another.
+
+Readers may object that we are looking at difference between grand
+slams, which each have their own time period, weather conditions, play
+time conditions, and ``home court effects'' instead of differences in
+surfaces alone. However, (1) grand slam data is the most readily
+available and most complete which makes it the best choice for
+modelling, (2) we adjust for these confounders where we can, and (3)
+analyzing the difference in the grand slams is still useful as they are
+considered to be the most prestigious events in tennis.
+
+The rest of this paper is organized as follows. In
+\protect\hyperlink{sec:data}{Section Data} we describe our grand slam
+tennis data. In \protect\hyperlink{sec:eda}{Section Early Data Analysis}
+we examine the data at a high level and use clustering whether to
+determine how the courts differ from one another. In
+\protect\hyperlink{sec:methods}{Section Methods} we describe our
+hierarchical models we use to determine difference in court surfaces. In
+\protect\hyperlink{sec:results}{Section Results} we describe the results
+of our modelling and also examine the play of Nadal, Federer, and
+Williams. Finally in \protect\hyperlink{discussion}{Section Discussion},
+we discuss future work and extensions or our model.
+
+\hypertarget{sec:data-eda}{%
+\section{Data and EDA}\label{sec:data-eda}}
+
+\hypertarget{sec:data}{%
+\subsection{Data}\label{sec:data}}
 
 The data consists of 5080 matches split evenly over the four grand slams
-and the two leagues (ATP and WTA). Each match has 82 attributes, many of
-which are redundant. We focus on the following attributes for both the
-winner and loser of the match: games won, points won, retirement, break
-points faced, break points saved, aces, country of origin, and player
-attributes. Additionally, we take into account the number of sets in a
-match, the surface type, minutes played, and round of the tournament. A
-subset of the data is shown in Table \ref{tab:data}.
+and the two leagues: ATP (men's) and WTA (women's). Each match has 80
+attributes, many of which are redundant. We focus on the following
+attributes for both the winner and loser of the match: games won, points
+won, retirement, break points faced, break points saved, aces, country
+of origin, and player attributes. Additionally, we take into account the
+number of sets in a match, the surface type, minutes played, and round
+of the tournament. A subset of the data is shown in Table
+\ref{tab:data}.
 
@@
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+The data is obtained from Jeff Sackman's open website {[}cite{]} via the
+\texttt{R} package \texttt{deuce} {[}cite{]}. All steps of our analysis
+from collection to dissemination are freely available online and are
+compiled into an \texttt{R} package
+\url{http://github.com/shannong19/courtsports}. The paper itself may be
+generated from the file
+\url{https://github.com/shannong19/courtsports/blob/master/contest/surface-type/paper-tennis.Rmd}.
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\hypertarget{sec:eda}{%
+\subsection{Early Data Analysis}\label{sec:eda}}
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\hypertarget{sec:methods}{%
+\section{Methods}\label{sec:methods}}
 
-\hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
+\hypertarget{sec:results}{%
+\section{Results}\label{sec:results}}
 
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-cdc-measles2018}{}%
-Centers for Disease Control and Prevention. 2018. ``Measles History.''
-Available online at \url{https://www.cdc.gov/measles/about}.
+\hypertarget{sec:discussion}{%
+\section{Discussion}\label{sec:discussion}}
+
+\hypertarget{sec:refs}{%
+\section{References}\label{sec:refs}}
 

### 2018-09-09 contest/graphics/graphics.Rmd
add graphics

--- 2018-09-08 contest/surface-type/paper-tennis.Rmd
@@
 ---
-title: "Analyzing the Effect of Surface in Tennis Grand Slams"
+title: "Graphics"
 author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
-date: '`r format(Sys.time(), "%B %d, %Y")`'
 output:
@@
     keep_tex: yes
-    number_sections: no
-abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
-The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.
-It is also believed that play style and practice habits may contribute to different results across surface types.
-For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.
-On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.
-In this analysis, we examine the result of the top 100 players in grand slams from 2013-2017 across the four different surfaces.
-We create a hierarchical model with fixed and random effects to predict the number of points won in a match.
-We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.
-We assess the fit of our model using standard statistical techniques (e.g.
-MSE, AIC, BIC, residual diagnostics) in addition to `common knowledge' factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).
-We compare the results of top 100 players across grand slams to examine the effect of court surface.
-We also provide an in-depth analysis of Nadal, Federer, and S.
-Williams."
-bibliography: master.bib
+    number_sections: yes
 ---
@@
 
-# devtools::use_data(gs, overwrite = TRUE)
-
 ## Make into player data
@@
 
-# devtools::use_data(gs_players, overwrite = TRUE)
-
 ```
-
-## Introduction
-Example reference[@cdc-measles2018]
-
-## Data
-The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues (ATP and WTA).
-Each match has `r ncol(gs)` attributes, many of which are redundant.
-We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
-Additionally, we take into account the number of sets in a match, the surface type, minutes played, and round of the tournament.
-A subset of the data is shown in Table \ref{tab:data}.
-
-```{r tab-data, include = TRUE, echo = FALSE}
-set.seed(19)
-s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
-rand_inds <- sample(s_id, 5)
-cols <- c("tourney_name", "year", "winner_ioc", "w_pointswon", "winner_name", "winner_rank", "l_pointswon", "loser_name")
-knitr::kable(gs[rand_inds, cols], format = "latex",
-             row.names = FALSE, col.names = c("Tournament", "Year", "W. Country", "W. Points", "Winner", "W. Rank", "L. Points", "Loser"),
-             caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
-```
-
-## Methods
-
-## Results
-
-## Discussion
-
-## References

### 2018-09-09 R/clean-data.R
trying to ignore caches take 2

@@ -1,143 +1,0 @@
-#' Clean the matches data from deuce atp_maches and wta_matches
-#'
-#' @param atp_matches atp_matches object from R package deuce (Men's).  Default is NULL
-#' @param wta_matches atp_matches object from R package deuce (Women's).  Default is NULL
-#' @param start_year start year (earliest is 1968)
-#' @param end_year end year (latest available is 2017)
-#' @param include_quals logical to include qualifying rounds.  Default is FALSE
-#' @param level level of tournament, corresponding to column from atp_matches/wta_matches.  Default is "Grand Slams"
-#' @return cleaned and subsetted data set, combined if there is both an atp_matches and wta_data set.  A new column called "league" is added and has value "ATP" or "WTA".  We also clean the tourney_name making a new column called tournament
-#' @importFrom magrittr %>%
-clean_matches_data <- function(atp_matches = NULL, wta_matches = NULL,
-                               start_year,
-                               end_year,
-                               include_quals = FALSE,
-                               level = "Grand Slams"){
-
-    ## Combine the two and add a "league" variable
-    matches <- NULL
-    if(is.null(atp_matches) & is.null(wta_matches)) {
-        stop("Both atp_matches and wta_matches are NULL")
-    }
-    if(!is.null(atp_matches)) {
-        atp_matches$league <- "ATP"
-    }
-    if(!is.null(wta_matches)){
-        wta_matches$league <- "WTA"
-    }
-    matches <- rbind(wta_matches, atp_matches)
-
-    ## Subset to proper years
-    matches <- matches  %>% dplyr::filter(tourney_level == level &
-                                   year >= start_year & year <= end_year)
-
-    ## Get rid of qualifiers if appropriate
-    if(!include_quals){
-        matches <- matches %>% dplyr::filter(!(round %in% c("Q1", "Q2", "Q3", "Q4")))
-    }
-
-    matches$tournament <- factor(matches$tourney_name) %>%
-        forcats::fct_collapse("US Open" = c("US Open", "Us Open"),
-                     "French Open" = c("French Open", "Roland Garros"))
-
-    return(matches)
-
-}
-
-
-#' Extract set, games, and points for winner and loser
-#'
-#' @param matches object from deuce package or clean_matches_data.
[…]

### 2018-09-09 contest/eda/makeGSF.R

@@ -1,16 +1,0 @@
-library(deuce)
-library(dplyr)
-
-data("point_by_point")
-pbp = tbl_df(point_by_point)
-pbp$wim = grepl("Wimbledon", pbp$tny_name)
-pbp$french = grepl("FrenchOpen", pbp$tny_name)
-pbp$aus = grepl("AustralianOpen", pbp$tny_name)
-pbp$us = grepl("USOpen", pbp$tny_name)
-pbp$qual = grepl("Qualifying", pbp$tny_name)
-
-# Create data for Grand Slam finals
-gsf = select(pbp, c("tny_name", "wim", "french", "us", "aus", "qual")) %>%
-  filter(., wim | french | us | aus & !qual)
-
-save(gsf, file = "GrandSlamFinals.RData")

### 2018-09-09 R/clean-data.R
added did_win column to gs_players

@@ -1,0 +1,145 @@
+#' Clean the matches data from deuce atp_maches and wta_matches
+#'
+#' @param atp_matches atp_matches object from R package deuce (Men's).  Default is NULL
+#' @param wta_matches atp_matches object from R package deuce (Women's).  Default is NULL
+#' @param start_year start year (earliest is 1968)
+#' @param end_year end year (latest available is 2017)
+#' @param include_quals logical to include qualifying rounds.  Default is FALSE
+#' @param level level of tournament, corresponding to column from atp_matches/wta_matches.  Default is "Grand Slams"
+#' @return cleaned and subsetted data set, combined if there is both an atp_matches and wta_data set.  A new column called "league" is added and has value "ATP" or "WTA".  We also clean the tourney_name making a new column called tournament
+#' @importFrom magrittr %>%
+clean_matches_data <- function(atp_matches = NULL, wta_matches = NULL,
+                               start_year,
+                               end_year,
+                               include_quals = FALSE,
+                               level = "Grand Slams"){
+
+    ## Combine the two and add a "league" variable
+    matches <- NULL
+    if(is.null(atp_matches) & is.null(wta_matches)) {
+        stop("Both atp_matches and wta_matches are NULL")
+    }
+    if(!is.null(atp_matches)) {
+        atp_matches$league <- "ATP"
+    }
+    if(!is.null(wta_matches)){
+        wta_matches$league <- "WTA"
+    }
+    matches <- rbind(wta_matches, atp_matches)
+
+    ## Subset to proper years
+    matches <- matches  %>% dplyr::filter(tourney_level == level &
+                                   year >= start_year & year <= end_year)
+
+    ## Get rid of qualifiers if appropriate
+    if(!include_quals){
+        matches <- matches %>% dplyr::filter(!(round %in% c("Q1", "Q2", "Q3", "Q4")))
+    }
+
+    matches$tournament <- factor(matches$tourney_name) %>%
+        forcats::fct_collapse("US Open" = c("US Open", "Us Open"),
+                     "French Open" = c("French Open", "Roland Garros"))
+
+    return(matches)
+
+}
+
+
+#' Extract set, games, and points for winner and loser
+#'
+#' @param matches object from deuce package or clean_matches_data.
[…]

### 2018-09-10 contest/graphics/graphics.Rmd
initial stages of graphics
add spain vs tournament and home court advantage graphics

@@
     fig_caption: yes
-    keep_tex: yes
     number_sections: yes
+    toc: true
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 knitr::opts_chunk$set(cache = TRUE)
@@
 library(kableExtra)
+library(pander)
 
@@
 ```
+
+# Spain vs. tournament
+
+## Is being a winner from Spain independent of tournament?
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+gs <- mutate(gs, 
+             winner_ioc_spain = ifelse(winner_ioc == "ESP", "Spain", "Other"))
+
+props <- gs %>% group_by(tournament, winner_ioc_spain) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = winner_ioc_spain)) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c( "#FFC400", "#C60B1E")) + 
+  labs(x = " ", y = "Proportion", title = "Spanish players win the French Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, winner_ioc_spain == "Spain"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4) + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r sample, echo=FALSE, results='asis'}
+library(xtable)
+
+spain_vs_tourn <- with(gs, table(tournament, winner_ioc_spain))
+spain_vs_tourn_chisq <- chisq.test(spain_vs_tourn)
+
+t1 <- kable(spain_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(spain_vs_tourn_chisq$expected, format = "latex", booktabs = TRUE)
+
+cat(c("\\begin{table}[!htb]
+    \\begin{minipage}{.5\\linewidth}
+      \\caption{Actual table}
+      \\centering",
+        t1,
+    "\\end{minipage}%
+    \\begin{minipage}{.5\\linewidth}
+      \\centering
+        \\caption{Expected table}",
+        t2,
+    "\\end{minipage} 
+\\end{table}"
+))  
+```
+
+```{r}
+spain_vs_tourn_chisq %>% 
+  pander()
+```
+
+\newpage
+
+## But do Spaniards enter the French open more than they do other tournaments?
+
+We look at players in the *first round* of all tournaments.
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+mini.d <- bind_rows(select(filter(gs, round == "R128"), tournament, winner_ioc) %>% rename(ioc = winner_ioc),
+            select(filter(gs, round == "R128"), tournament, loser_ioc) %>% rename(ioc = loser_ioc)) %>%
+  mutate(ioc_spain = ifelse(ioc == "ESP", "Spain", "Other"))
+
+props <- mini.d %>% group_by(tournament, ioc_spain) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = ioc_spain)) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c( "#FFC400", "#C60B1E")) + 
+  labs(x = " ", y = "Proportion",
+       title = "Spanish players enter the French Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, ioc_spain == "Spain"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4)  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+with(mini.d, table(tournament, ioc_spain)) %>%
+  chisq.test() %>%
+  pander()
+```
+
+Yes, but not significantly more.
+
+# Home court advantage
+
+## France
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+gs <- mutate(gs, 
+             winner_ioc_france = ifelse(winner_ioc == "FRA", "France", "Other"))
+
+props <- gs %>% group_by(tournament, winner_ioc_france) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(winner_ioc_france, levels = c("Other", "France")))) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c("#F42A41", "#00209F")) + 
+  labs(x = " ", y = "Proportion", title = "French players win the French Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, winner_ioc_france == "France"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4)  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r, echo=FALSE, results='asis'}
+library(xtable)
+
+france_vs_tourn <- with(gs, table(tournament, winner_ioc_france))
+
+france_vs_tourn_chisq <- chisq.test(france_vs_tourn)
+
+t1 <- kable(france_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(france_vs_tourn_chisq$expected, format = "latex", booktabs = TRUE)
+
+cat(c("\\begin{table}[!htb]
+    \\begin{minipage}{.5\\linewidth}
+      \\caption{Actual table}
+      \\centering",
+        t1,
+    "\\end{minipage}%
+    \\begin{minipage}{.5\\linewidth}
+      \\centering
+        \\caption{Expected table}",
+        t2,
+    "\\end{minipage} 
+\\end{table}"
+))  
+```
+
+```{r}
+france_vs_tourn_chisq %>% 
+  pander()
+```
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+mini.d <- mini.d %>%
+  mutate(ioc_france = ifelse(ioc == "FRA", "France", "Other"))
+
+props <- mini.d %>% group_by(tournament, ioc_france) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(ioc_france, levels = c("Other", "France")))) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c("#F42A41", "#00209F")) + 
+  labs(x = " ", y = "Proportion",
+       title = "French players enter the French Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, ioc_france == "France"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4)  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+with(mini.d, table(tournament, ioc_france)) %>%
+  chisq.test() %>%
+  pander()
+```
+
+*Yes, significantly.*
+
+\newpage
+
+## United States
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+gs <- mutate(gs, 
+             winner_ioc_usa = ifelse(winner_ioc == "USA", "USA", "Other"))
+
+props <- gs %>% group_by(tournament, winner_ioc_usa) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(winner_ioc_usa, levels = c("Other", "USA")))) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c( "#002868", "#BF0A30")) + 
+  labs(x = " ", y = "Proportion", title = "American players win the U.S. Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, winner_ioc_usa == "USA"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4,
+            color="white")  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r, echo=FALSE, results='asis'}
+usa_vs_tourn <- with(gs, table(tournament, winner_ioc_usa))
+
+usa_vs_tourn_chisq <- chisq.test(usa_vs_tourn)
+
+t1 <- kable(usa_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(usa_vs_tourn_chisq$expected, format = "latex", booktabs = TRUE)
+
+# cat(c("\\begin{table}[!htb]
+#     \\begin{minipage}{.5\\linewidth}
+#       \\caption{Actual table}
+#       \\centering",
+#         t1,
+#     "\\end{minipage}%
+#     \\begin{minipage}{.5\\linewidth}
+#       \\centering
+#         \\caption{Expected table}",
+#         t2,
+#     "\\end{minipage} 
+# \\end{table}"
+# ))  
+```
+
+```{r}
+usa_vs_tourn_chisq %>% 
+  pander()
+```
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+mini.d <- mini.d %>%
+  mutate(ioc_usa = ifelse(ioc == "USA", "USA", "Other"))
+
+props <- mini.d %>% group_by(tournament, ioc_usa) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(ioc_usa, levels = c("Other", "USA")))) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c( "#002868", "#BF0A30")) + 
+  labs(x = " ", y = "Proportion",
+       title = "American players enter the U.S. Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, ioc_usa == "USA"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4,
+            color="white")  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+with(mini.d, table(tournament, ioc_usa)) %>%
+  chisq.test() %>%
+  pander()
+```
+
+## Australia
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+gs <- mutate(gs, 
+             winner_ioc_Australia = ifelse(winner_ioc == "AUS", "Australia", "Other"))
+
+props <- gs %>% group_by(tournament, winner_ioc_Australia) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(winner_ioc_Australia, levels = c("Other", "Australia")))) +
+        geom_bar(stat = "identity", color="grey") +
+        scale_fill_manual("Country of origin",
+                          values = c("#FFFFFF", "#012169")) + 
+  labs(x = " ", y = "Proportion", title = "Australian players win the Australian Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, winner_ioc_Australia == "Australia"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4)  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r, echo=FALSE, results='asis'}
+Australia_vs_tourn <- with(gs, table(tournament, winner_ioc_Australia))
+
+Australia_vs_tourn_chisq <- chisq.test(Australia_vs_tourn)
+
+t1 <- kable(Australia_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(Australia_vs_tourn_chisq$expected, format = "latex", booktabs = TRUE)
+
+cat(c("\\begin{table}[!htb]
+    \\begin{minipage}{.5\\linewidth}
+      \\caption{Actual table}
+      \\centering",
+        t1,
+    "\\end{minipage}%
+    \\begin{minipage}{.5\\linewidth}
+      \\centering
+        \\caption{Expected table}",
+        t2,
+    "\\end{minipage} 
+\\end{table}"
+))  
+```
+
+```{r}
+Australia_vs_tourn_chisq %>% 
+  pander()
+```
+
+```{r, fig.height=3.2, fig.width=6, fig.align='center'}
+
+mini.d <- mini.d %>%
+  mutate(ioc_Australia = ifelse(ioc == "AUS", "Australia", "Other"))
+
+props <- mini.d %>% group_by(tournament, ioc_Australia) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = factor(ioc_Australia, levels = c("Other", "Australia")))) +
+        geom_bar(stat = "identity", color="grey") +
+        scale_fill_manual("Country of origin",
+                          values = c("#FFFFFF", "#012169")) + 
+  labs(x = " ", y = "Proportion",
+       title = "Australian players enter the Australian Open\nmore often than other tournaments") + 
+  geom_text(data = filter(props, ioc_Australia == "Australia"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct*100, 2),"%")),
+                       size=4)  + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+with(mini.d, table(tournament, ioc_Australia)) %>%
+  chisq.test() %>%
+  pander()
+```
+
+# Conclusions
+
+* Spaniards win the French Open more often than they win other tournaments, and this difference is significant.
+They play in the French Open more often than they play in other tournaments, but not significantly.
+
+* French win the French Open more often than they win other tournaments, but this difference is (barely) insignificant.
+They play in the French Open more often than they play in other tournaments, at a significantly higher rate.
+
+* Americans win the US Open more often than they win other tournaments, and this difference is significant.
+But, they play in the US Open more often than they play in other tournaments, at a significantly higher rate.
+
+* Australians win the US Open more often than they win other tournaments, and this difference is significant.
+But, they play in the Australian Open more often than they play in other tournaments, at a significantly higher rate.

### 2018-09-10 contest/surface-type/paper-tennis.Rmd
add toc

@@
     keep_tex: yes
-    number_sections: no
+    number_sections: yes
+    toc: true
 abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
@@
 
-## Introduction
-Example reference[@cdc-measles2018]
+# Introduction {#sec:iintro}
+Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open, which is played on a clay surface [cite].
+In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon, which is played on grass.
+On the women's side, Serena Williams has been dominant both on hard court (7 titles at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
+This trend extends to other top players, who seem to have better results at some grand slams than others.
+More broadly, it seems that country of origin has an interaction effect with court type.
+For example, Spanish players seem to excel on clay courts and Americans have great success at Wimbledon despite grass courts not being of wide use in the USA.
+It also worth questioning whether the US Open and Australian Open should be grouped together as hard courts despite having different surface compositions [cite].
+In this paper, we analyze the results of grand slam players from 2013-2017,  and we
 
-## Data
-The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues (ATP and WTA).
+1. Determine if and how court surface effects players by implementation of a series of nested hierarchical models
+
+2. Examine how Nadal, Federer, and Williams' play differs by surface
+
+3. Assess whether we can group the two hard court surfaces together.
+
+As to issue (1) the hiearchical modelling, we use models similar to [cite] and [cite].
+More specifically, we model the players' expected points in a match based on the player's own characteristics, the court/tournament effects, and the oponent's ranking.
+We use a Bayesian method because in [cite,cite] they had success in similar situations.
+
+For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense"" tests (e.g.
+these players should win often when contrasting them to opponents) like in [ventura something] and whether these players do have surface apparent effects.
+
+Finally, for issue (3), we use clustering methods [cite] in order to determine which court surface types are more similar to one another.
+
+Readers may object that we are looking at difference between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.
+However, (1) grand slam data is the most readily available and most complete which makes it the best choice for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.
+
+The rest of this paper is organized as follows.
+In [Section Data](#sec:data) we describe our grand slam tennis data.
+In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.
+In [Section Methods](#sec:methods) we describe our hierarchical models we use to determine difference in court surfaces.
+In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
+Finally in [Section Discussion](#discussion), we discuss future work and extensions or our model.
+
+# Data and EDA {#sec:data-eda}
+
+## Data {#sec:data}
+The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
 Each match has `r ncol(gs)` attributes, many of which are redundant.
@@
 
-## Methods
+The data is obtained from Jeff Sackman's open website [cite] via the `R` package `deuce` [cite].
+All steps of our analysis from collection to dissemination are freely available online and are compiled into an `R` package \url{http://github.com/shannong19/courtsports}.
 
-## Results
+## Early Data Analysis {#sec:eda}
 
-## Discussion
+# Methods {#sec:methods}
 
-## References
+# Results {#sec:results}
+
+# Discussion {#sec:discussion}
+
+# References {#sec:refs}

### 2018-09-10 contest/graphics/graphics.Rmd
Merging

--- 2018-09-09 contest/graphics/graphics.Rmd
@@
 ```
+
+## Is being a winner from Spain independent of tournament?
+
+```{r, fig.height=5, fig.width=8, fig.align='center'}
+
+gs <- mutate(gs, 
+             winner_ioc_spain = ifelse(winner_ioc == "ESP", "Spain", "Other"))
+
+props <- gs %>% group_by(tournament, winner_ioc_spain) %>%
+        tally() %>%
+        group_by(tournament) %>%
+        mutate(pct = n / sum(n))
+
+ggplot(props, aes(tournament, pct, fill = winner_ioc_spain)) +
+        geom_bar(stat = "identity") +
+        scale_fill_manual("Country of origin",
+                          values = c( "#FFC400", "#C60B1E")) + 
+  labs(x = " ", y = "Proportion", title = "Spanish players win the French Open more often than other tournaments") + 
+  geom_text(data = filter(props, winner_ioc_spain == "Spain"), 
+            aes(tournament, y = pct + .05, label = paste0(round(pct, 2),"%")),
+                       size=4)
+
+```
+
+```{r sample, echo=FALSE, results='asis'}
+library(xtable)
+
+spain_vs_tourn <- with(gs, table(tournament, winner_ioc_spain))
+
+t1 <- kable(spain_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(chisq.test(spain_vs_tourn)$expected, format = "latex", booktabs = TRUE)
+
+cat(c("\\begin{table}[!htb]
+    \\begin{minipage}{.5\\linewidth}
+      \\caption{Actual table}
+      \\centering",
+        t1,
+    "\\end{minipage}%
+    \\begin{minipage}{.5\\linewidth}
+      \\centering
+        \\caption{Expected table}",
+        t2,
+    "\\end{minipage} 
+\\end{table}"
+))  
+```
+
+```{r}
+chisq.test(spain_vs_tourn) %>% 
+  pander()
+```

### 2018-09-10 contest/surface-type/paper-tennis.Rmd
Merging

@@
     number_sections: yes
-    toc: true
 abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.

### 2018-09-10 contest/surface-type/paper-tennis.tex
Merging

@@
   \postdate{\par}
-    \date{September 08, 2018}
+    \date{September 09, 2018}
 
@@
 
-\hypertarget{sec:iintro}{%
-\section{Introduction}\label{sec:iintro}}
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
+\section{Introduction}\label{sec:iintro}
 
@@
 
-\hypertarget{sec:data-eda}{%
-\section{Data and EDA}\label{sec:data-eda}}
+\section{Data and EDA}\label{sec:data-eda}
 
-\hypertarget{sec:data}{%
-\subsection{Data}\label{sec:data}}
+\hypertarget{sec:data}{\subsection{Data}\label{sec:data}}
 
@@
 \rowcolors{2}{gray!6}{white}
+
 \begin{table}
@@
 \end{table}
+
 \rowcolors{2}{white}{white}
@@
 compiled into an \texttt{R} package
-\url{http://github.com/shannong19/courtsports}. The paper itself may be
-generated from the file
-\url{https://github.com/shannong19/courtsports/blob/master/contest/surface-type/paper-tennis.Rmd}.
+\url{http://github.com/shannong19/courtsports}.
 
-\hypertarget{sec:eda}{%
-\subsection{Early Data Analysis}\label{sec:eda}}
+\hypertarget{sec:eda}{\subsection{Early Data Analysis}\label{sec:eda}}
 
-\hypertarget{sec:methods}{%
-\section{Methods}\label{sec:methods}}
+\hypertarget{sec:methods}{\section{Methods}\label{sec:methods}}
 
-\hypertarget{sec:results}{%
-\section{Results}\label{sec:results}}
+\hypertarget{sec:results}{\section{Results}\label{sec:results}}
 
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
+\section{Discussion}\label{sec:discussion}
 
-\hypertarget{sec:refs}{%
-\section{References}\label{sec:refs}}
+\section{References}\label{sec:refs}
 

### 2018-09-10 contest/graphics/graphics.Rmd
finished spain vs. tournamnet and home court advantage

--- 2018-09-10 contest/graphics/graphics.Rmd
@@
     fig_caption: yes
+    keep_tex: yes
     number_sections: yes
-    toc: true
 ---
@@
 ```{r, fig.height=3.2, fig.width=6, fig.align='center'}
-
 gs <- mutate(gs, 
@@
                        size=4) + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
-
 ```
@@
 spain_vs_tourn <- with(gs, table(tournament, winner_ioc_spain))
+
+t1 <- kable(spain_vs_tourn, format = "latex", booktabs = TRUE)
+t2 <- kable(chisq.test(spain_vs_tourn)$expected, format = "latex", booktabs = TRUE)
 spain_vs_tourn_chisq <- chisq.test(spain_vs_tourn)

### 2018-09-10 contest/model-fitting/model-fitting.Rmd
added model-fitting, deleted old data

---
title: "Model Fitting"
author: "Amanda Luby, Shannon Gallagher, Kayla Frisoli"
date: "9/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rstan)
library(shinystan)
library(brms)
```

### Modeling notes:

* Did not include 'height' in models due to missingness in data
* Used brms package to call Stan (multilevel modeling-specific, uses syntax similar to lme4 if we want to switch to maximum-likelihood based estimation)
* Each model takes 15-20 minutes to fit (3 models in document)

### Model 1:

No effect for player; fixed effect for IOC, tournament, etc.

```{r}
load(file = "../../data/gs_players.rda")
load("../../data/gs.rda")

gs_players$name_int = as.integer(as.factor(gs_players$name))
gs_players$hand_fac = as.factor(gs_players$hand)
gs_players$ioc_fac = as.factor(gs_players$ioc)

basemod = brm(pointswon ~ ioc_fac + hand_fac + tournament + round + rank + opponent_rank + minutes, data = gs_players, family = gaussian())
```

### Model 2:

Group-level effect and tournament effect for IOC; fixed effect for other vars (still no individual-level effects)

```{r}
country_mod = brm(pointswon ~ hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |ioc_fac), data = gs_players, family = gaussian())
```

### Model 2:

Group-level tournament effect for player; fixed effect for IOC, tournament, etc.

```{r}
ind_mod = brm(pointswon ~ ioc_fac + hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |name_int), data = gs_players, family = gaussian())
```

### Model Comparison (stats methods)

(Note: for full MCMC diagnostics, run `launch_shinystan(MOD_NAME))` )

In models that don't include individual effects (basemod and country_mod), the estimated effect of "Wimbledon" on points won is quite high compared to the other models. (Could this be because better players are more likely to play in Wimbledon??) This effect starts to even out after including individual effects.

```{r}
plot(marginal_effects(basemod, effects = "tournament"))
plot(marginal_effects(country_mod, effects = "tournament"))
plot(marginal_effects(ind_mod, effects = "tournament"))
```

### Model Comparison (tennis knowledge)

Checks:
* Serena should be good on hard surface
* Nadal should be good on clay 
* Federer should be good on grass

```{r}

```

### 2018-09-10 contest/graphics/graphics.Rmd
add graphs for points and games won

@@
 title: "Graphics"
-author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
+author: "We have no names"
 output:
@@
     keep_tex: yes
-    number_sections: yes
+    number_sections: no
 ---
@@
 knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(warning=FALSE)
+knitr::opts_chunk$set(message = FALSE)
 knitr::opts_chunk$set(cache = TRUE)
@@
 library(pander)
+library(gridExtra)
 
@@
                                      opp_var_names = "opponent_rank")
+
+```
+
+# Univariate distributions
+
+## Points won
+
+```{r, fig.height=4, fig.width=12}
+p1 <- ggplot(gs_players) + geom_histogram(aes(pointswon), bins=40) + 
+  labs(x = "Winner points")
+
+p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
+  labs(x = "Winner points")
+
+p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
+  labs(x = "Winner points")
+
+grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
+
+```
+
+```{r, fig.height=3.5, fig.width=10, fig.align='center'}
+
+p1 <- ggplot(gs) + geom_point(aes(w_pointswon, l_pointswon)) + 
+  geom_abline(slope=1, intercept=0, color="red") + 
+  labs(x = "Winner points", y = "Loser points")
+
+p2 <- ggplot(gs) + geom_point(aes(w_pointswon, l_pointswon, color = league), alpha=.3) + 
+    scale_color_manual(values=c("#339966", "#336699")) +
+  geom_abline(slope=1, intercept=0, color="red") + 
+  geom_smooth(aes(w_pointswon, l_pointswon, color = league), method='lm') + 
+  labs(x = "Winner points", y = "Loser points")
+
+grid.arrange(p1, p2, ncol=2, widths=c(.48, .52))
+
+```
+
+## Games won
+
+```{r, fig.height=4, fig.width=10}
+
+p1 <- ggplot(gs_players) + geom_histogram(aes(gameswon), bins=30)+ 
+  labs(x = "Winner games")
+
+p2 <- ggplot(gs_players) + geom_density(aes(gameswon, color = tournament))+ 
+  labs(x = "Winner games")
+
+p3 <- ggplot(gs_players) + geom_density(aes(gameswon, color = league))+ 
+  labs(x = "Winner games")
+
+grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
+
+```
+
+```{r, fig.height=4, fig.width=8, fig.align='center'}
+
+p1 <- ggplot(gs) + geom_point(aes(w_gameswon, l_gameswon)) + 
+  geom_abline(slope=1, intercept=0, color="red") + 
+  labs(x = "Winner games", y = "Loser games")
+
+p2 <- ggplot(gs) + geom_point(aes(w_gameswon, l_gameswon, color = league), alpha=.3) + 
+    scale_color_manual(values=c("#339966", "#336699")) +
+  geom_abline(slope=1, intercept=0, color="red") + 
+  geom_smooth(aes(w_gameswon, l_gameswon, color = league), method='lm')  + 
+  labs(x = "Winner games", y = "Loser games")
+
+grid.arrange(p1, p2, ncol=2, widths=c(.48, .52))
 
@@
 But, they play in the Australian Open more often than they play in other tournaments, at a significantly higher rate.
+
+# MDS
+
+```{r}
+gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced) %>%
+na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced, tournament) %>%
+na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+```

### 2018-09-10 contest/surface-type/paper-tennis.Rmd
Added citations and did some more research which I added to intro

@@
     number_sections: yes
+header-includes:
+  - \usepackage{hyperref}
 abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
@@
 # Introduction {#sec:iintro}
-Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open, which is played on a clay surface [cite].
+Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open, which is played on a clay surface [@bbc2018].
 In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon, which is played on grass.
@@
 For example, Spanish players seem to excel on clay courts and Americans have great success at Wimbledon despite grass courts not being of wide use in the USA.
-It also worth questioning whether the US Open and Australian Open should be grouped together as hard courts despite having different surface compositions [cite].
+It also worth questioning whether the US Open and Australian Open should be grouped together as hard courts despite having different surface compositions [@theage2007].
 In this paper, we analyze the results of grand slam players from 2013-2017,  and we
@@
 
-As to issue (1) the hiearchical modelling, we use models similar to [cite] and [cite].
+As to issue (1) quantifying the effect of court surface on players, there has not been much written about with regards to tennis.
+There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
+[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another.
+Other sports analyses do take into account surface type such as grass vs.
+turf in soccer and football.
+Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
+
+We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
+Both of those models had success using hierarhcical Bayesian models, which we employ in our own models.
 More specifically, we model the players' expected points in a match based on the player's own characteristics, the court/tournament effects, and the oponent's ranking.
-We use a Bayesian method because in [cite,cite] they had success in similar situations.
 
-For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense"" tests (e.g.
+HERE
+For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests (e.g.
 these players should win often when contrasting them to opponents) like in [ventura something] and whether these players do have surface apparent effects.
+START HERE GOAT [@kovalchik2016]
 

### 2018-09-10 contest/surface-type/paper-tennis.tex
Added citations and did some more research which I added to intro

--- 2018-09-09 contest/surface-type/paper-tennis.tex
@@
   \postdate{\par}
-    \date{September 08, 2018}
+    \date{September 10, 2018}
 
@@
 \usepackage{makecell}
+
+\usepackage{hyperref}
 
@@
 out of his current 17 grand slams titles at the French Open, which is
-played on a clay surface {[}cite{]}. In contrast, his rival Roger
+played on a clay surface (Jurejko 2018). In contrast, his rival Roger
 Federer has won his most grand slam titles (8 out of 20) at Wimbledon,
@@
 whether the US Open and Australian Open should be grouped together as
-hard courts despite having different surface compositions {[}cite{]}. In
-this paper, we analyze the results of grand slam players from 2013-2017,
-and we
+hard courts despite having different surface compositions (Paxinos
+2007). In this paper, we analyze the results of grand slam players from
+2013-2017, and we
 
@@
 
-As to issue (1) the hiearchical modelling, we use models similar to
-{[}cite{]} and {[}cite{]}. More specifically, we model the players'
-expected points in a match based on the player's own characteristics,
-the court/tournament effects, and the oponent's ranking. We use a
-Bayesian method because in {[}cite,cite{]} they had success in similar
-situations.
+As to issue (1) quantifying the effect of court surface on players,
+there has not been much written about with regards to tennis. There are
+materials available in the literature for forecasting the outcome of
+tennis matches (Klaassen and Magnus 2003; Newton and Keller 2005; McHale
+and Morton 2011) or for assessing whether points within a match are
+independent and identically distributed (Klaassen and Magnus 2001).
+(Knottenbelt, Spanias, and Madurska 2012) do take into account surface
+in their model but do not compare the results of one surface to another.
+Other sports analyses do take into account surface type such as grass
+vs.~turf in soccer and football. Results from these studies show that
+surface type does have an effect on the game, either directly or
+indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains et al.
+(2010);{]}.
 
-For issue (2) the player analysis of Nadal, Federer, and Williams, we
-examine whether our model passes the ``common sense''" tests (e.g.~these
-players should win often when contrasting them to opponents) like in
-{[}ventura something{]} and whether these players do have surface
-apparent effects.
+We use models that take into account both individual and group effects
+such as in the Gaussian-process player production basketball model or
+predicting individual soccer performance (Page, Barney, and McGuire
+2013; Egidi and Gabry 2018). Both of those models had success using
+hierarhcical Bayesian models, which we employ in our own models. More
+specifically, we model the players' expected points in a match based on
+the player's own characteristics, the court/tournament effects, and the
+oponent's ranking.
+
+HERE For issue (2) the player analysis of Nadal, Federer, and Williams,
+we examine whether our model passes the ``common sense'' tests
+(e.g.~these players should win often when contrasting them to opponents)
+like in {[}ventura something{]} and whether these players do have
+surface apparent effects. START HERE GOAT (Kovalchik 2016)
 
@@
 compiled into an \texttt{R} package
-\url{http://github.com/shannong19/courtsports}. The paper itself may be
-generated from the file
-\url{https://github.com/shannong19/courtsports/blob/master/contest/surface-type/paper-tennis.Rmd}.
+\url{http://github.com/shannong19/courtsports}.
 
@@
 \hypertarget{sec:refs}{%
-\section{References}\label{sec:refs}}
+\section*{References}\label{sec:refs}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-andersson2008}{}%
+Andersson, Helena, Björn Ekblom, and Peter Krustrup. 2008. ``Elite
+Football on Artificial Turf Versus Natural Grass: Movement Patterns,
+Technical Standards, and Player Impressions'' 26 (February): 113--22.
+
+\leavevmode\hypertarget{ref-egidi2018}{}%
+Egidi, Leonardo, and Jonah Gabry. 2018. ``Bayesian Hierarchical Models
+for Predicting Individual Performance in Soccer.'' \emph{Journal of
+Quantitative Analysis in Sports} 14 (3). De Gruyter: 143--57.
+
+\leavevmode\hypertarget{ref-gains2010}{}%
+Gains, Graydon L, Andy N Swedenhjelm, Jerry L Mayhew, H Michael Bird,
+and Jeremy J Houser. 2010. ``Comparison of Speed and Agility Performance
+of College Football Players on Field Turf and Natural Grass.'' \emph{The
+Journal of Strength \& Conditioning Research} 24 (10). LWW: 2613--7.
+
+\leavevmode\hypertarget{ref-bbc2018}{}%
+Jurejko, Jonathan. 2018. ``French Open 2018: Why does 'King of Clay'
+Rafael Nadal reign supreme?'' Edited by BBC Sport at Roland Garros.
+\url{https://www.bbc.com/sport/tennis/44385223}.
+
+\leavevmode\hypertarget{ref-klaassen2003}{}%
+Klaassen, Franc J.G.M., and Jan R. Magnus. 2003. ``Forecasting the
+Winner of a Tennis Match.'' \emph{European Journal of Operational
+Research} 148 (2): 257--67.
+\url{https://doi.org/https://doi.org/10.1016/S0377-2217(02)00682-3}.
+
+\leavevmode\hypertarget{ref-klaassen2001}{}%
+Klaassen, Franc J.G.M., and Jan R Magnus. 2001. ``Are Points in Tennis
+Independent and Identically Distributed? Evidence from a Dynamic Binary
+Panel Data Model.'' \emph{Journal of the American Statistical
+Association} 96 (454). Taylor \& Francis: 500--509.
+\url{https://doi.org/10.1198/016214501753168217}.
+
+\leavevmode\hypertarget{ref-knottenbelt2012}{}%
+Knottenbelt, William J., Demetris Spanias, and Agnieszka M. Madurska.
+2012. ``A Common-Opponent Stochastic Model for Predicting the Outcome of
+Professional Tennis Matches.'' \emph{Computers \& Mathematics with
+Applications} 64 (12): 3820--7.
+\url{https://doi.org/https://doi.org/10.1016/j.camwa.2012.03.005}.
+
+\leavevmode\hypertarget{ref-kovalchik2016}{}%
+Kovalchik, Stephanie Ann. 2016. ``Searching for the Goat of Tennis Win
+Prediction.'' \emph{Journal of Quantitative Analysis in Sports} 12 (3).
+De Gruyter: 127--38.
+
+\leavevmode\hypertarget{ref-mchale2011}{}%
+McHale, Ian, and Alex Morton. 2011. ``A Bradley-Terry Type Model for
+Forecasting Tennis Match Results.'' \emph{International Journal of
+Forecasting} 27 (2): 619--30.
+\url{https://doi.org/https://doi.org/10.1016/j.ijforecast.2010.04.004}.
+
+\leavevmode\hypertarget{ref-newton2005}{}%
+Newton, Paul K, and Joseph B Keller. 2005. ``Probability of Winning at
+Tennis I. Theory and Data.'' \emph{Studies in Applied Mathematics} 114
+(3). Wiley Online Library: 241--69.
+
+\leavevmode\hypertarget{ref-page2013}{}%
+Page, Garritt L, Bradley J Barney, and Aaron T McGuire. 2013. ``Effect
+of Position, Usage Rate, and Per Game Minutes Played on Nba Player
+Production Curves.'' \emph{Journal of Quantitative Analysis in Sports} 9
+(4). De Gruyter: 337--45.
+
+\leavevmode\hypertarget{ref-theage2007}{}%
+Paxinos, Stathi. 2007. ``Australian Open court surface is speeding up.''
+Edited by The Age.
+\url{https://www.theage.com.au/sport/tennis/australian-open-court-surface-is-speeding-up-20071120-ge6chj.html}.
 

### 2018-09-11 contest/model-fitting/model-comparisons.Rmd
please work

---
title: "Model Comparisons"
author: "Amanda Luby, Shannon Gallagher, Kayla Frisoli"
date: "9/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(dplyr)
load("stan-model-fits.RData")
load(file = "../../data/gs_players.rda")
load("../../data/gs.rda")

gs_players$name_int = as.integer(as.factor(gs_players$name))
gs_players$hand_fac = as.factor(gs_players$hand)
gs_players$ioc_fac = as.factor(gs_players$ioc)
gs_players$atp = gs_players$league == "ATP"
gs_players$year_fac = as.factor(gs_players$year)
```

### Model Fit

* Took minutes out because it was disproportionately missing for women players (and missing data gets thrown out)
* Individual effects for tournament don't really seem to make a difference - mostly wrapped up in other variables

```{r}
linmod = lm(pointswon ~ ioc_fac +  tournament + round + rank + opponent_rank + year_fac + atp, data = gs_players)
country_effects = lmer(pointswon ~ round + rank + opponent_rank + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players)
ind_effects = lmer(pointswon ~  ioc_fac + round + rank + opponent_rank + year_fac + atp + (0 + tournament |name_int), data = gs_players)

par(mfrow=c(1,3))
plot(residuals(linmod)~fitted(linmod), pch =19, cex = .5, main = "Model 1",
     xlab = "Fitted", ylab = "Residuals")
plot(residuals(country_effects) ~ fitted(country_effects), pch =19, cex = .5, main = "Model 2",
     xlab = "Fitted", ylab = "Residuals")
plot(residuals(ind_effects) ~ fitted(ind_effects), pch =19, cex = .5, main = "Model 3",
     xlab = "Fitted", ylab = "Residuals")

```

```{r}
extractAIC(linmod)
extractAIC(country_effects)
extractAIC(ind_effects)
```

### Model Comparison (tennis knowledge)

```{r}
superstars = gs_players %>% 
  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer")

par(mfrow = c(1,3))
plot(predict(linmod, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
plot(predict(country_effects, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
plot(predict(ind_effects, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
```

Sanity Checks:

* Serena should be good on hard surface
* Nadal should be good on clay 
* Federer should be good on grass

```{r}

```

### 2018-09-11 contest/model-fitting/model-fitting.Rmd
please work

@@
 ind_mod = brm(pointswon ~ ioc_fac + hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |name_int), data = gs_players, family = gaussian())
+ind_lmer = lmer(pointswon ~ ioc_fac + hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |name_int), data = gs_players)
 ```
@@
 ```{r}
+
+save(basemod, country_mod, ind_mod, file = "stan-model-fits.RData")
 plot(marginal_effects(basemod, effects = "tournament"))
@@
 ```
-
-### Model Comparison (tennis knowledge)
-
-Checks:
-* Serena should be good on hard surface
-* Nadal should be good on clay 
-* Federer should be good on grass
-
-```{r}
-
-```

### 2018-09-12 contest/graphics/graphics.Rmd
graphics updates

@@
 
-```{r, fig.height=4, fig.width=10}
+```{r, fig.height=4, fig.width=12}
 
@@
 
-```{r, fig.height=4, fig.width=8, fig.align='center'}
+```{r, fig.height=3.5, fig.width=10, fig.align='center'}
 
@@
 
+# Pairs
+
+```{r, fig.height=12, fig.width=12}
+library(GGally)
+
+vars <- c("ht", "age", "rank", "opponent_rank", "ace", "df", "svpt", "1stIn", "SvGms", "bpSaved",
+          "bpFaced", "pointswon", "gameswon")
+
+ggpairs(gs_players,
+        columns = which(names(gs_players) %in% vars),
+        mapping=ggplot2::aes(colour = factor(did_win)))
+
+```
+
+## Follow-up
+
+```{r, fig.height=3, fig.width=4}
+ggplot(gs_players) + geom_point(aes(log(ht+1), log(ace+1), color = factor(did_win)))
+```
+
 # MDS
@@
 ```
+
+```{r}
+gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn,
+                w_bpSaved, w_bpFaced) %>%
+na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- select(gs, w_ace, w_df, w_svpt, w_1stIn,
+                w_bpSaved, w_bpFaced, tournament) %>%
+na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+```

### 2018-09-12 contest/surface-type/paper-tennis.Rmd
added references

@@
 As to issue (1) quantifying the effect of court surface on players, there has not been much written about with regards to tennis.
-There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
+There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
 [@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another.
@@
 
-HERE
-For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests (e.g.
-these players should win often when contrasting them to opponents) like in [ventura something] and whether these players do have surface apparent effects.
-START HERE GOAT [@kovalchik2016]
+For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests  like how the models in [thomas2013] show that commonly well known hockey players also have high status in the model.
+We also examine whether these players do have surface apparent effects.
+Few academic papers have been written about Nadal, Federer, or Williams.
+One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon showed that Federer was too heavily favored by bookmakers [@leitner2009].
+One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].
 
@@
 
-Readers may object that we are looking at difference between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.
-However, (1) grand slam data is the most readily available and most complete which makes it the best choice for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.
+Readers may object that we are looking at differences between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.
+However, (1) grand slam data is the most readily available and most complete which makes it the best choice at the moment for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.
 
@@
 ```{r tab-data, include = TRUE, echo = FALSE}
-set.seed(19)
+w_id <- which(gs$winner_name == "Serena Williams")[1:2]
+f_id <- which(gs$winner_name == "Roger Federer")[1:2]
+n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
 s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
-rand_inds <- sample(s_id, 5)
-cols <- c("tourney_name", "year", "winner_ioc", "w_pointswon", "winner_name", "winner_rank", "l_pointswon", "loser_name")
-knitr::kable(gs[rand_inds, cols], format = "latex",
-             row.names = FALSE, col.names = c("Tournament", "Year", "W. Country", "W. Points", "Winner", "W. Rank", "L. Points", "Loser"),
+cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
+knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
+             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
              caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",

### 2018-09-12 contest/surface-type/paper-tennis.tex
added references

@@
   \postdate{\par}
-    \date{September 10, 2018}
+    \date{September 12, 2018}
 
@@
 tennis matches (Klaassen and Magnus 2003; Newton and Keller 2005; McHale
-and Morton 2011) or for assessing whether points within a match are
-independent and identically distributed (Klaassen and Magnus 2001).
-(Knottenbelt, Spanias, and Madurska 2012) do take into account surface
-in their model but do not compare the results of one surface to another.
-Other sports analyses do take into account surface type such as grass
-vs.~turf in soccer and football. Results from these studies show that
-surface type does have an effect on the game, either directly or
-indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains et al.
-(2010);{]}.
+and Morton 2011; Kovalchik 2016){]} or for assessing whether points
+within a match are independent and identically distributed (Klaassen and
+Magnus 2001). (Knottenbelt, Spanias, and Madurska 2012) do take into
+account surface in their model but do not compare the results of one
+surface to another. Other sports analyses do take into account surface
+type such as grass vs.~turf in soccer and football. Results from these
+studies show that surface type does have an effect on the game, either
+directly or indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains
+et al. (2010);{]}.
 
@@
 
-HERE For issue (2) the player analysis of Nadal, Federer, and Williams,
-we examine whether our model passes the ``common sense'' tests
-(e.g.~these players should win often when contrasting them to opponents)
-like in {[}ventura something{]} and whether these players do have
-surface apparent effects. START HERE GOAT (Kovalchik 2016)
+For issue (2) the player analysis of Nadal, Federer, and Williams, we
+examine whether our model passes the ``common sense'' tests like how the
+models in {[}thomas2013{]} show that commonly well known hockey players
+also have high status in the model. We also examine whether these
+players do have surface apparent effects. Few academic papers have been
+written about Nadal, Federer, or Williams. One paper studies Federer's
+odds of winning when Nadal suddenly withdrew from Wimbledon showed that
+Federer was too heavily favored by bookmakers (Leitner, Zeileis, and
+Hornik 2009). One analysis of Williams shows how she has gotten better
+with age, even past the point when other greats began to decline, but
+the study does not look at surface type (Morris 2015).
 
@@
 
-Readers may object that we are looking at difference between grand
+Readers may object that we are looking at differences between grand
 slams, which each have their own time period, weather conditions, play
@@
 surfaces alone. However, (1) grand slam data is the most readily
-available and most complete which makes it the best choice for
-modelling, (2) we adjust for these confounders where we can, and (3)
+available and most complete which makes it the best choice at the moment
+for modelling, (2) we adjust for these confounders where we can, and (3)
 analyzing the difference in the grand slams is still useful as they are
@@
 \centering
-\begin{tabular}[t]{lrlrlrrl}
+\begin{tabular}[t]{llrlrrrr}
 \hiderowcolors
 \toprule
-Tournament & Year & W. Country & W. Points & Winner & W. Rank & L. Points & Loser\\
+Winner & Tournament & Year & W. IOC & W. Points & W. Rank & L. Points & L. Rank\\
 \midrule
 \showrowcolors
-US Open & 2014 & USA & 58 & Serena Williams & 1 & 31 & Taylor Townsend\\
-US Open & 2013 & ESP & 136 & Rafael Nadal & 2 & 112 & Philipp Kohlschreiber\\
-Australian Open & 2015 & ESP & 91 & Rafael Nadal & 3 & 51 & Mikhail Youzhny\\
-US Open & 2013 & USA & 65 & Serena Williams & 1 & 41 & Yaroslava Shvedova\\
-Australian Open & 2017 & USA & 73 & Serena Williams & 2 & 60 & Lucie Safarova\\
+Serena Williams & Australian Open & 2013 & USA & 52 & 3 & 18 & 110\\
+Serena Williams & Australian Open & 2013 & USA & 70 & 3 & 41 & 112\\
+Roger Federer & Australian Open & 2013 & SUI & 95 & 2 & 63 & 46\\
+Roger Federer & Australian Open & 2013 & SUI & 111 & 2 & 86 & 40\\
+Rafael Nadal & Roland Garros & 2013 & ESP & 140 & 4 & 115 & 59\\
+Rafael Nadal & Roland Garros & 2013 & ESP & 113 & 4 & 90 & 35\\
 \bottomrule
@@
 
+\leavevmode\hypertarget{ref-leitner2009}{}%
+Leitner, Christoph, Achim Zeileis, and Kurt Hornik. 2009. ``Is Federer
+Stronger in a Tournamentwithout Nadal? An Evaluation of Odds and
+Seedings for Wimbledon 2009.'' \emph{Austrian Journal of Statistics} 38
+(4): 277--86.
+
 \leavevmode\hypertarget{ref-mchale2011}{}%
@@
 \url{https://doi.org/https://doi.org/10.1016/j.ijforecast.2010.04.004}.
+
+\leavevmode\hypertarget{ref-five2015}{}%
+Morris, Benjamin. 2015. ``Serena Williams and the Difference Between
+All-Time Great and Greatest of All Time.'' Edited by FiveThirtyEight.
+\url{https://fivethirtyeight.com/features/serena-williams-and-the-difference-between-all-time-great-and-greatest-of-all-time/}.
 

### 2018-09-13 contest/surface-type/paper-tennis.Rmd
added gs partial point by point data.  3066 out of 5000+ matches

@@
 In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon, which is played on grass.
-On the women's side, Serena Williams has been dominant both on hard court (7 titles at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
+On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
 This trend extends to other top players, who seem to have better results at some grand slams than others.
@@
 
-For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests  like how the models in [thomas2013] show that commonly well known hockey players also have high status in the model.
+For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests  like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
 We also examine whether these players do have surface apparent effects.
@@
 
-Finally, for issue (3), we use clustering methods [cite] in order to determine which court surface types are more similar to one another.
+Finally, for issue (3), we use clustering methods in order to determine which court surface types are more similar to one another.
 
@@
 We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
-Additionally, we take into account the number of sets in a match, the surface type, minutes played, and round of the tournament.
+Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.
 A subset of the data is shown in Table \ref{tab:data}.
@@
 
-The data is obtained from Jeff Sackman's open website [cite] via the `R` package `deuce` [cite].
-All steps of our analysis from collection to dissemination are freely available online and are compiled into an `R` package \url{http://github.com/shannong19/courtsports}.
+The data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.
 
@@
 # References {#sec:refs}
+
+\footnotesize

### 2018-09-13 contest/surface-type/paper-tennis.tex
added gs partial point by point data.  3066 out of 5000+ matches

@@
 Federer has won his most grand slam titles (8 out of 20) at Wimbledon,
-which is played on grass. On the women's side, Serena Williams has been
-dominant both on hard court (7 titles at the Australian Open and 6 at US
-Open) and grass (7 at Wimbledon). This trend extends to other top
-players, who seem to have better results at some grand slams than
-others. More broadly, it seems that country of origin has an interaction
-effect with court type. For example, Spanish players seem to excel on
-clay courts and Americans have great success at Wimbledon despite grass
-courts not being of wide use in the USA. It also worth questioning
-whether the US Open and Australian Open should be grouped together as
-hard courts despite having different surface compositions (Paxinos
-2007). In this paper, we analyze the results of grand slam players from
-2013-2017, and we
+which is played on grass. On the women's side, Serena Williams, winner
+of 24 grand slam titles, has been dominant both on hard court (7 titles
+at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
+This trend extends to other top players, who seem to have better results
+at some grand slams than others. More broadly, it seems that country of
+origin has an interaction effect with court type. For example, Spanish
+players seem to excel on clay courts and Americans have great success at
+Wimbledon despite grass courts not being of wide use in the USA. It also
+worth questioning whether the US Open and Australian Open should be
+grouped together as hard courts despite having different surface
+compositions (Paxinos 2007). In this paper, we analyze the results of
+grand slam players from 2013-2017, and we
 
@@
 examine whether our model passes the ``common sense'' tests like how the
-models in {[}thomas2013{]} show that commonly well known hockey players
-also have high status in the model. We also examine whether these
-players do have surface apparent effects. Few academic papers have been
-written about Nadal, Federer, or Williams. One paper studies Federer's
-odds of winning when Nadal suddenly withdrew from Wimbledon showed that
-Federer was too heavily favored by bookmakers (Leitner, Zeileis, and
-Hornik 2009). One analysis of Williams shows how she has gotten better
-with age, even past the point when other greats began to decline, but
-the study does not look at surface type (Morris 2015).
+models in (Thomas et al. 2013) show that commonly well known hockey
+players also have high status in the model. We also examine whether
+these players do have surface apparent effects. Few academic papers have
+been written about Nadal, Federer, or Williams. One paper studies
+Federer's odds of winning when Nadal suddenly withdrew from Wimbledon
+showed that Federer was too heavily favored by bookmakers (Leitner,
+Zeileis, and Hornik 2009). One analysis of Williams shows how she has
+gotten better with age, even past the point when other greats began to
+decline, but the study does not look at surface type (Morris 2015).
 
-Finally, for issue (3), we use clustering methods {[}cite{]} in order to
-determine which court surface types are more similar to one another.
+Finally, for issue (3), we use clustering methods in order to determine
+which court surface types are more similar to one another.
 
@@
 
-The data is obtained from Jeff Sackman's open website {[}cite{]} via the
-\texttt{R} package \texttt{deuce} {[}cite{]}. All steps of our analysis
-from collection to dissemination are freely available online and are
-compiled into an \texttt{R} package
-\url{http://github.com/shannong19/courtsports}.
+The data is obtained from Jeff Sackmann's open website via the
+\texttt{R} package \texttt{deuce} (Sackmann 2018; Kovalchik 2017). All
+steps of our analysis from collection to dissemination are freely
+available online.
 
@@
 \hypertarget{sec:refs}{%
-\section*{References}\label{sec:refs}}
-\addcontentsline{toc}{section}{References}
+\section{References}\label{sec:refs}}
+
+\footnotesize
 
@@
 
+\leavevmode\hypertarget{ref-deuce2017}{}%
+Kovalchik, Stephanie. 2017. \emph{Deuce: Resources for Analysis of
+Professional Tennis Data}.
+
 \leavevmode\hypertarget{ref-kovalchik2016}{}%
@@
 
+\leavevmode\hypertarget{ref-sackmann2018}{}%
+Sackmann, Jeff. 2018. ``Tennis Data Repositories.''
+\url{https://github.com/JeffSackmann}.
+
+\leavevmode\hypertarget{ref-thomas2013}{}%
+Thomas, A. C., Samuel L. Ventura, Shane T. Jensen, and Stephen Ma. 2013.
+``COMPETING Process Hazard Function Models for Player Ratings in Ice
+Hockey.'' \emph{The Annals of Applied Statistics} 7 (3). Institute of
+Mathematical Statistics: 1497--1524.
+\url{http://www.jstor.org/stable/23566482}.
+
 \end{document}

### 2018-09-13 contest/point-by-point/gspbp.R

@@ -1,0 +1,98 @@
+## SKG
+## September 13, 2018
+## Happy grand opening tepper quad
+## Analyzing grand slam point by point
+
+library(deuce)
+library(dplyr)
+library(forcats)
+
+data(gs_point_by_point)
+
+dim(gs_point_by_point)
+summary(gs_point_by_point)
+
+## too long
+gs_pbp <- gs_point_by_point
+na_col_inds <- colSums(is.na(gs_pbp))
+complete_cols <- which(na_col_inds == 0)
+names(gs_pbp)[-complete_cols]
+gs_complete <- gs_pbp %>% dplyr::select(names(gs_pbp[complete_cols])) %>%
+    filter(year >= 2013 & year <= 2017)
+gs_complete$tournament <- factor(gs_complete$slam) %>% 
+    forcats::fct_collapse(
+                 "US Open" = c("usopen"),
+                 "French Open" = c("frenchopen"),
+                 "Wimbledon" = c("wimbledon"),
+                 "Australian Open" = c("ausopen"))
+
+
+
+#######################
+
+## Mutating into match stats
+match_stats <- gs_complete %>% group_by(match_id, player1, player2, Tour,  
+                                       slam, year, tournament) %>%
+    dplyr::summarize(ave_serve_speed_p1 = mean(Speed_MPH[PointServer == 1]),
+                     ave_serve_speed_p2 = mean(Speed_MPH[PointServer == 2]),
+                     n_aces_p1 = sum(P1Ace),
+                     n_aces_p2 = sum(P2Ace),
+                     n_winners_p1 = sum(P1Winner),
+                     n_winners_p2 = sum(P2Winner),
+                     n_netpt_wpct_p1 = sum(P1NetPointWon) / sum(P1NetPoint),
+                     n_netpt_wpct_p2 = sum(P1NetPointWon) / sum(P1NetPoint),
+                     n_bp_wpct_p1 = sum(P1BreakPointWon) / sum(P1BreakPoint),
+                     n_bp_wpct_p1 = sum(P2BreakPointWon) / sum(P2BreakPoint),
+                     n_netpt_wpct_p2 = sum(P1NetPointWon) / sum(P1NetPoint),
+                     n_ue_p1 = sum(P1UnfErr),
+                     n_ue_p2 = sum(P2UnfErr),
+                     n_sv_wpct_p1 = sum((PointServer == 1) * (PointWinner == 1 ) )/
+                         sum(PointServer == 1),
[…]

### 2018-09-14 contest/point-by-point/gspbp.R
Changed partial pbp

@@ -6,6 +6,7 @@
 library(deuce)
 library(dplyr)
 library(forcats)
+library(ggplot2)
 
 data(gs_point_by_point)
 
@@ -93,6 +94,70 @@ gs_join <- left_join(gs_sub, match_stats, by = c("tournament", "year", "a1", "a2
 head(gs_join[, c("player1", "winner_name", "player2", "loser_name")])
 gs_partial_pbp <- na.omit(gs_join)
 
+gs_partial_pbp$Week <- gs_partial_pbp$round %>%
+    forcats::fct_collapse("Week 1" = c("R128","R32", "R64" ),
+                          "Week 2" = c("R16", "QF", "SF", "F"))
+
 dim(gs_partial_pbp)
 
 devtools::use_data(gs_partial_pbp, overwrite = TRUE)
+
+
+
+
+## get court
+
+court <- gs_partial_pbp %>% group_by(tournament, Tour, year,
+                                     week) %>%
+    summarize(n_retire = sum(Retirement),
+              ave_sets = mean(w_setswon + l_setswon),
+              ave_aces = mean(n_aces_p1 + n_aces_p2),
+              ave_ue = mean(n_ue_p1 + n_ue_p2),
+              ave_winners = mean(n_winners_p1 + n_winners_p2),
+              med_rank = median(c(loser_rank, winner_rank)),
+              n_obs = length(loser_rank)
+)
+
+
+
+
+ggplot(court, aes(y=ave_sets, group = tournament, col = tournament)) +
+              geom_boxplot() + 
+    facet_grid(week~Tour)
+
+ggplot(court, aes(y=ave_aces, group = tournament, col = tournament)) +
+              geom_boxplot() + 
+              facet_grid(~Tour)
+
+ggplot(court, aes(y=ave_ue, group = tournament, col = tournament)) +
+              geom_boxplot() + 
+              facet_grid(~Tour)
+
+ggplot(court, aes(y=ave_ue / ave_sets, group = tournament,
+                  col = tournament)) +
+    geom_boxplot() + 
+    facet_grid(~Tour)
+
+ggplot(court, aes(y=n_retire, group = tournament,
+                  col = tournament)) +
+    geom_boxplot() + 
+    facet_grid(~Tour)
+
+ggplot(court, aes(x = year,y=ave_sets, col = tournament)) +
+    geom_point() + facet_grid(~Tour)
+
+
[…]

### 2018-09-14 contest/model-fitting/model-comparisons.Rmd

@@
 library(dplyr)
-load("stan-model-fits.RData")
+library(kableExtra)
+#load("stan-model-fits.RData")
 load(file = "../../data/gs_players.rda")
@@
 gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+
 ```
@@
 * Took minutes out because it was disproportionately missing for women players (and missing data gets thrown out)
-* Individual effects for tournament don't really seem to make a difference - mostly wrapped up in other variables
+* Initially began by modeling points won in the match, but found that this was not the best approach because more points won does not necessarily correspond to a better performance.
+
+Variables considered as predictors:
+
+* `ioc_fac`: Country that the player represents
+* `tournament`: Which Grand Slam the match was played at
+* `late_round`: Indicator for whether the match occured in Round of 16 or later
+* `rank`: Rank of player at the time of the match
+    + Included on log scale 
+* `opponent_rank`: Rank of opponent at the time of the match
+    + Included on log scale
+* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
+* `atp`: Indicator that the match was played in the ATP league instead of the WTF league
+
+Models considered:
+
+* No mixed effects
+    + `nocourt_logistic`: $\text{logit}(\pi_i) = \beta X$, $X$ does not include `tournament`
+        + If this model fits well, court surface does not have an effect on winning probability
+    + `base_logistic`: $\text{logit}(\pi_i) = \beta X$
+        + If this model fits better than the mixed-effects models, performance does not depend on individual-level or IOC-level 
+* IOC-level effect
+    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $beta_{T[i]} = \alpha_i C_i$
 
 ```{r}
-linmod = lm(pointswon ~ ioc_fac +  tournament + round + rank + opponent_rank + year_fac + atp, data = gs_players)
-country_effects = lmer(pointswon ~ round + rank + opponent_rank + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players)
-ind_effects = lmer(pointswon ~  ioc_fac + round + rank + opponent_rank + year_fac + atp + (0 + tournament |name_int), data = gs_players)
+nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
+ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
+ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
 
-par(mfrow=c(1,3))
-plot(residuals(linmod)~fitted(linmod), pch =19, cex = .5, main = "Model 1",
-     xlab = "Fitted", ylab = "Residuals")
-plot(residuals(country_effects) ~ fitted(country_effects), pch =19, cex = .5, main = "Model 2",
-     xlab = "Fitted", ylab = "Residuals")
-plot(residuals(ind_effects) ~ fitted(ind_effects), pch =19, cex = .5, main = "Model 3",
-     xlab = "Fitted", ylab = "Residuals")
+model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
 
+for(mod in 1:nrow(model.sums)){
+  name = model_names[mod]
+  model.sums$aic[mod] = extractAIC(get(name))[2]
+  model.sums$edf[mod] = extractAIC(get(name))[1]
+}
+
+kable(model.sums)
 ```
 
-```{r}
-extractAIC(linmod)
-extractAIC(country_effects)
-extractAIC(ind_effects)
-```
+Based on AIC, the `ind_logistic_noioc` model performed best after accounting for the number of variables in the model.
 
@@
 * Federer should be good on grass
-
-```{r}
-
-```

### 2018-09-14 contest/model-fitting/model-comparisons.Rmd
updated model results in model-comparisons doc

@@
 library(kableExtra)
+library(cowplot)
+library(broom)
+library(tidyr)
 #load("stan-model-fits.RData")
@@
 * IOC-level effect
-    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $beta_{T[i]} = \alpha_i C_i$
+    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i C_i$
+* Player-level effect
+    + `ind_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i P_i$
+    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC, $\beta_{T[i]} = \alpha_i P_i$
+    + `ind_int_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{0[i]}$, where $X$ does not include IOC, $\beta_{0[i]} = \alpha_i$
+    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{Y[i]}$, where $X$ does not include IOC, $\beta_{Y[i]} = \alpha_i Y_i$
 
-```{r}
+```{r, cache = TRUE}
 nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
@@
 model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
-
+model.ests = data.frame()
+ran.effects = data.frame()
 for(mod in 1:nrow(model.sums)){
@@
   model.sums$edf[mod] = extractAIC(get(name))[1]
+  betas = tidy(get(name))
+  betas$model = name
+  if("group" %in% colnames(betas)){
+  model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+    select(., -"group"))
+  }
+  else model.ests = rbind(model.ests, betas)
 }
@@
 
-### Model Comparison (tennis knowledge)
+```{r}
+fixed.effects.plot = model.ests %>%
+  filter(., !grepl("ioc", term)) %>%
+  ggplot(., aes(x = term, y = exp(estimate), col = model, ymin = exp(estimate - 2*std.error), ymax = exp(estimate + 2*std.error))) +
+  geom_errorbar(width = .7, position = position_dodge()) +
+  theme_minimal() + 
+  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
+  scale_color_brewer(palette = "Dark2") + 
+  ggtitle("Fixed effects")
+
+```
 
 ```{r}
-superstars = gs_players %>% 
-  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer")
+ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
+ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
 
-par(mfrow = c(1,3))
-plot(predict(linmod, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
-plot(predict(country_effects, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
-plot(predict(ind_effects, newdata = superstars) ~ superstars$pointswon, col = as.factor(superstars$name))
+# 486, 504, 526
+indices = c(486, 504, 526)
+superstars = data.frame(
+  name = rep(c("Serena", "Nadal", "Federer"), 2),
+  model = c(rep("ind_logistic", 3), rep("ind_logistic_noioc",3)),
+  `Australian Open` = c(ranef(ind_logistic)$name_int[indices,1], ranef(ind_logistic_noioc)$name_int[indices,1]),
+  `French Open` = c(ranef(ind_logistic)$name_int[indices,2], ranef(ind_logistic_noioc)$name_int[indices,2]),
+  `US Open` = c(ranef(ind_logistic)$name_int[indices,3], ranef(ind_logistic_noioc)$name_int[indices,3]),
+  `Wimbledon` = c(ranef(ind_logistic)$name_int[indices,4], ranef(ind_logistic_noioc)$name_int[indices,4])
+)
+
+superstars = superstars %>%
+  gather(., tournament, effect, -c("name", "model"))
+
+random.effects.plot = ran_tourn %>%
+  gather(., tourn, estimate, -model) %>%
+  mutate(., tournament = gsub("tournament", "", tourn)) %>%
+  ggplot(., aes(x = tournament, y = exp(estimate), fill = model)) +
+  geom_violin() + 
+  theme_minimal() +
+  scale_fill_brewer(palette = "Dark2") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  ggtitle("Random effects")
+
+plot_grid(fixed.effects.plot, random.effects.plot)
+```
+
+Looking at the `ind_logistic_noioc` random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
+
+```{r}
+pairs(ranef(ind_logistic_noioc)$name_int, pch = 21)
+```
+
+If we look at our big three, their individual effects are not very informative (especially after taking uncertainty into consideration).
+
+```{r}
+name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
+name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
+as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
+  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
+  ggplot(., aes(y = name, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament, scales = "free_x") + 
+  geom_errorbarh(height = .5)
 ```

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
Put data in R package format and put pbp data in form of gs

@@
 ```{r data-cleaning, include = FALSE}
-## Load WTA and ATP matches from deuce package
-data(atp_matches)
-data(wta_matches)
-
-## Clean data
-gs <- clean_matches_data(atp_matches = atp_matches,
-                         wta_matches = wta_matches,
-                         start_year = 2013,
-                         end_year = 2017,
-                         include_quals = FALSE,
-                         level = "Grand Slams")
-
-## Extract sets, points, and games
-gs <- extract_sgp(gs)
-
-# devtools::use_data(gs, overwrite = TRUE)
-
-## Make into player data
-gs_players <- matches_to_player_data(gs,
-                                     w_opp_vars = "loser_rank",
-                                     l_opp_vars = "winner_rank",
-                                     opp_var_names = "opponent_rank")
-
-# devtools::use_data(gs_players, overwrite = TRUE)
+data(gs)
+data(gs_players)
+data(gs_partial)
+data(gs_partial_players)
 

### 2018-09-14 R/clean-data.R

@@ -119,20 +119,20 @@ matches_to_player_data <- function(matches,
     matches_w$did_win <- 1 # yes did win
 
     ## Get rid of winner attributes for the loser except for l_opp_vars
-    remove_from_loser <- names(matches)[grep("^w|^W", names(matches))] %>%
-        .[-which(. == l_opp_vars)]
+    remove_from_loser <- names(matches)[grep("^w|^W", names(matches))]
+    remove_from_loser <- which(remove_from_loser == l_opp_vars)
     rename_w_inds <- which(names(matches) %in% l_opp_vars)
     names(matches_l)[rename_w_inds] <- opp_var_names
 
     ## Get rid of loser attributes for the winner except for w_opp_vars
-    remove_from_winner <- names(matches)[grep("^l|^L", names(matches))] %>%
-        .[-which(. %in% c(w_opp_vars, "league"))]
+    remove_from_winner <- names(matches)[grep("^l|^L", names(matches))] 
+    remove_from_winner <- which(remove_from_winner %in% c(w_opp_vars, "league"))
     rename_l_inds <- which(names(matches) %in% w_opp_vars)
     names(matches_w)[rename_l_inds] <- opp_var_names
 
     ## Filter and rename variables
-    matches_l <- dplyr::select(matches_l, -remove_from_loser)
-    matches_w <- dplyr::select(matches_w, -remove_from_winner)
+    matches_l <- matches_l[, -remove_from_loser]
+    matches_w <- matches_w[,-remove_from_winner]
     names(matches_w) <- gsub("^winner_|^W|^w_", "", names(matches_w))
     names(matches_l) <- gsub("^loser_|^L|^l_", "", names(matches_l))
 
@@ -142,4 +142,127 @@ matches_to_player_data <- function(matches,
     return(player_matches)
 
 
+}
+
+
+#' Summarize the point by point data
+#'
+#' @param pbp point by point data from deuce package
+#' @return summary of point by point data in terms of unique matches
+#' @importFrom magrittr %>%
+summarize_pbp <- function(pbp,
+                          start_year = 2013, end_year = 2017){
+
+    ## Omit non-complete columns
+    na_col_inds <- colSums(is.na(pbp))
+    complete_cols <- which(na_col_inds == 0)
+    pbp <- pbp %>% dplyr::select(names(pbp[complete_cols])) %>%
+        filter(year >= start_year& year <= end_year)
+
+    ## Change tournament to match the names of gs data
[…]

### 2018-09-14 contest/point-by-point/gspbp.R

@@ -160,4 +160,41 @@ table(gs_partial_pbp$year, gs_partial_pbp$round,
 ## 1 mens us open, 1 aus 2014, 2013
 
 ###
-nadal <- gs_partial_pbp %>% 
+nadal_wins <- gs_partial_pbp %>% filter(winner_name == "Rafael Nadal")  %>%
+    group_by(tournament) %>%
+    summarize(n_wins = sum(winner_name == "Rafael Nadal"),
+              n_matches = length(winner_name),
+              n_retire = sum(Retirement),
+              ave_sets = mean(w_setswon + l_setswon),
+              ave_aces = mean(n_aces_p1 + n_aces_p2),
+              ave_ue = mean(n_ue_p1 + n_ue_p2),
+              ave_winners = mean(n_winners_p1 + n_winners_p2),
+              med_rank = median(c(loser_rank, winner_rank)),
+              n_obs = length(loser_rank)
+              )
+nadal_wins$did_win <- 1
+
+nadal_losses <- gs_partial_pbp %>% filter(loser_name == "Rafael Nadal")  %>%
+    group_by(tournament) %>%
+    summarize(n_wins = sum(winner_name == "Rafael Nadal"),
+              n_matches = length(winner_name),
+              n_retire = sum(Retirement),
+              ave_sets = mean(w_setswon + l_setswon),
+              ave_aces = mean(n_aces_p1 + n_aces_p2),
+              ave_ue = mean(n_ue_p1 + n_ue_p2),
+              ave_winners = mean(n_winners_p1 + n_winners_p2),
+              med_rank = median(c(loser_rank, winner_rank)),
+              n_obs = length(loser_rank)
+              )
+head(nadal)
+
+
+## How to combine winner_name and loser_name with player_1 and player2
+## if (player_1 == winner_name)
+## new columns:
+## {variable_name_w = variable_name_p1}
+## variable_name_l = variable_name_p2
+## else
+## variable_name_w = variable_name_p2
+## variable_name_l = variable_name_p1
+

### 2018-09-14 data-raw/make-gs-data.R

@@ -1,0 +1,71 @@
+## Script to make data used in our analysis, specifically
+##
+## 1. gs.rda
+## 2. gs_players.rda
+## 3. gs_pbp.rda
+## 4. gs_partial.rda
+
+devtools::load_all("../") ## Load court sports functions 
+library(deuce)
+library(dplyr)
+
+
+## Load WTA and ATP matches from deuce package
+data(atp_matches)
+data(wta_matches)
+
+## Clean data
+gs <- clean_matches_data(atp_matches = atp_matches,
+                         wta_matches = wta_matches,
+                         start_year = 2013,
+                         end_year = 2017,
+                         include_quals = FALSE,
+                         level = "Grand Slams")
+
+ 
+
+## Extract sets, points, and games
+gs <- extract_sgp(gs)
+
+devtools::use_data(gs, overwrite = TRUE)
+
+## Make into player data
+gs_players <- matches_to_player_data(gs,
+                                     w_opp_vars = "loser_rank",
+                                     l_opp_vars = "winner_rank",
+                                     opp_var_names = "opponent_rank")
+
+devtools::use_data(gs_players, overwrite = TRUE)
+
+###############################################
+
+### pbp data
+data(gs_point_by_point)
+
+
+
+gs_pbp <- summarize_pbp(gs_point_by_point,
+                        start_year = 2013,
+                        end_year = 2017)
+
+devtools::use_data(gs_pbp, overwrite = TRUE)
+
+data(gs)
+
+## Join gs_pbp and gs together
+
+gs_join <- join_gs_pbp(gs_pbp, gs)
+
+gs_partial <- combine_fields(gs_join)
+
+
+head(gs_partial[, c("winner_name", "loser_name", "player1", "player2", "w_n_sv")])
+
+devtools::use_data(gs_partial, overwrite = TRUE)
+
+## gs_partial_players
+devtools::load_all("../") ## Load court sports functions
+data(gs_partial)
+gs_partial_players <- matches_to_player_data(gs_partial)
+
+devtools::use_data(gs_partial_players, overwrite = TRUE)

### 2018-09-14 R/clean-data.R
saved data as data frames instead of tibbles

@@ -112,6 +112,7 @@ matches_to_player_data <- function(matches,
                                    l_opp_vars = "winner_rank",
                                    opp_var_names = "opponent_rank"){
 
+
     ## Copy data frames
     matches_l <- matches
     matches_w <- matches
@@ -120,19 +121,19 @@ matches_to_player_data <- function(matches,
 
     ## Get rid of winner attributes for the loser except for l_opp_vars
     remove_from_loser <- names(matches)[grep("^w|^W", names(matches))]
-    remove_from_loser <- which(remove_from_loser == l_opp_vars)
+    remove_from_loser <- remove_from_loser[-which(remove_from_loser == l_opp_vars)]
     rename_w_inds <- which(names(matches) %in% l_opp_vars)
     names(matches_l)[rename_w_inds] <- opp_var_names
 
     ## Get rid of loser attributes for the winner except for w_opp_vars
-    remove_from_winner <- names(matches)[grep("^l|^L", names(matches))] 
-    remove_from_winner <- which(remove_from_winner %in% c(w_opp_vars, "league"))
+    remove_from_winner <- names(matches)[grep("^l|^L", names(matches))]
+    remove_from_winner <- remove_from_winner[-which(remove_from_winner %in% c(w_opp_vars, "league"))]
     rename_l_inds <- which(names(matches) %in% w_opp_vars)
     names(matches_w)[rename_l_inds] <- opp_var_names
 
     ## Filter and rename variables
-    matches_l <- matches_l[, -remove_from_loser]
-    matches_w <- matches_w[,-remove_from_winner]
+    matches_l <- matches_l %>% dplyr::select(-remove_from_loser)
+    matches_w <- matches_w %>% dplyr::select(-remove_from_winner)
     names(matches_w) <- gsub("^winner_|^W|^w_", "", names(matches_w))
     names(matches_l) <- gsub("^loser_|^L|^l_", "", names(matches_l))
 
@@ -160,7 +161,7 @@ summarize_pbp <- function(pbp,
         filter(year >= start_year& year <= end_year)
 
     ## Change tournament to match the names of gs data
-    pbp$tournament <- factor(pbp$slam) %>% 
+    pbp$tournament <- factor(pbp$slam) %>%
         forcats::fct_collapse(
                      "US Open" = c("usopen"),
                      "French Open" = c("frenchopen"),
@@ -169,7 +170,7 @@ summarize_pbp <- function(pbp,
 
 
[…]

### 2018-09-14 data-raw/make-gs-data.R

@@ -5,7 +5,7 @@
 ## 3. gs_pbp.rda
 ## 4. gs_partial.rda
 
-devtools::load_all("../") ## Load court sports functions 
+devtools::load_all("../") ## Load court sports functions
 library(deuce)
 library(dplyr)
 
@@ -22,7 +22,7 @@ gs <- clean_matches_data(atp_matches = atp_matches,
                          include_quals = FALSE,
                          level = "Grand Slams")
 
- 
+
 
 ## Extract sets, points, and games
 gs <- extract_sgp(gs)
@@ -44,9 +44,9 @@ data(gs_point_by_point)
 
 
 
-gs_pbp <- summarize_pbp(gs_point_by_point,
+gs_pbp <- as.data.frame(summarize_pbp(gs_point_by_point,
                         start_year = 2013,
-                        end_year = 2017)
+                        end_year = 2017))
 
 devtools::use_data(gs_pbp, overwrite = TRUE)
 
@@ -56,7 +56,7 @@ data(gs)
 
 gs_join <- join_gs_pbp(gs_pbp, gs)
 
-gs_partial <- combine_fields(gs_join)
+gs_partial <- as.data.frame(combine_fields(gs_join))
 
 
 head(gs_partial[, c("winner_name", "loser_name", "player1", "player2", "w_n_sv")])

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
add points scored and home town advantage eda; more coming

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(warning=FALSE)
+knitr::opts_chunk$set(message = FALSE)
 knitr::opts_chunk$set(cache = TRUE)
@@
 library(kableExtra)
+library(gridExtra)
+library(grid)
 
@@
 
+```
+
+```{r graphing-parameters, include = FALSE}
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+        family="serif"),
+        plot.title = element_text(hjust = 0.5, size=16))
+
+tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
+league_colors <- c("#0C2340", "#902BA3")
+win_colors <- c("#336699", "#339966")
 ```
@@
 
+### Examining the distribution of points earned
+
+We first examine the distibution of points earned per match.
+We find that the distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
+As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
+Also unsuprisingly, the winners of the match tended to earn more points than the losers.
+
+```{r, fig.height=5, fig.width=12}
+
+# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
+
+p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
+  labs(x = "Points") + my_theme +
+  scale_color_manual("Tournament", values=tournament_colors)
+
+p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
+  labs(x = "Points") +
+  scale_color_manual("League", values=league_colors) +
+  my_theme
+
+# grid.arrange( p2, p3, ncol=2,
+#              top = textGrob("Winning points by tournament and league",
+#                             gp=gpar(fontsize=20)))
+
+p1 <- ggplot(gs_players) + geom_density(aes(pointswon, color = factor(did_win))) + 
+  labs(x = "Points") +
+  scale_color_manual("Winner?", values=win_colors) +
+  my_theme +
+  facet_wrap(~league, ncol=1)
+
+grid.arrange(p1, p2, ncol=2,
+             top = textGrob("Distribution of points per match",
+                            gp=gpar(fontsize=20, fontfamily="serif")),
+              layout_matrix = rbind(c(NA, 1),
+                                    c(2, 1),
+                                    c(NA, 1)),
+             heights=c(1, 3, 1))
+
+```
+
+### Home court advantage
+
+It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
+In our data we find this to be true (i.e.
+French players win the French open more than French players win other slams).
+But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a particular country play in "their" tournament more often than they play in other tournaments.
+We also find this to be true in our data for France, The United States, and Australia (i.e.
+the proportion of French players in the French Open is greater than the proportion of French players in other slams).
+
+Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments.
+If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon.
+The same would be true for Australia and the US.
+But, we see that this isn't the case.
+After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
+
+```{r, fig.height=4, fig.width=12}
+france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+us_wins <- filter(gs_players, ioc == "USA", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+p1<- ggplot(france_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "French players",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
+
+p2 <- ggplot(us_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "American players",
+       x = "", y = "Win proportion") +  
+  scale_color_manual(guide=FALSE, values=tournament_colors)  + my_theme
+
+p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "Australian players",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(" ", values=tournament_colors) + my_theme
+
+grid.arrange(p1,p2, p3, nrow=1, widths = c(.29, .29, .42),
+             top = textGrob("Home court advantage debunked: win proportion by country",
+                            gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size as the tournament goes on",
+                           # x = 0,
+                           # y = 0.5,
+                           just = "left"))
+
+```
+
 # Methods {#sec:methods}

### 2018-09-14 contest/graphics/graphics.Rmd
update grpahics file; eval=FALSE some stuff

@@
 library(gridExtra)
+library(grid)
 
@@
 ```{r data-cleaning, include = FALSE}
-## Load WTA and ATP matches from deuce package
-data(atp_matches)
-data(wta_matches)
+data("gs")
+data("gs_players")
+data("gs_partial_pbp")
 
-## Clean data
-gs <- clean_matches_data(atp_matches = atp_matches,
-                         wta_matches = wta_matches,
-                         start_year = 2013,
-                         end_year = 2017,
-                         include_quals = FALSE,
-                         level = "Grand Slams")
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+        family="serif"),
+        plot.title = element_text(hjust = 0.5, size=16))
 
-## Extract sets, points, and games
-gs <- extract_sgp(gs)
+tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
 
-## Make into player data
-gs_players <- matches_to_player_data(gs,
-                                     w_opp_vars = "loser_rank",
-                                     l_opp_vars = "winner_rank",
-                                     opp_var_names = "opponent_rank")
+league_colors <- c("#0C2340", "#902BA3")
+
+win_colors <- c("#336699", "#339966")
 
@@
 
-# Univariate distributions
+## Examining the distribution of points earned
 
-## Points won
+We first examine the distibution of points earned per match.
+We find that the distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
+As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
+Also unsuprisingly, the winners of the match tended to earn more points than the losers.
+
+```{r, fig.height=5, fig.width=12}
+
+# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
+
+p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
+  labs(x = "Points") + my_theme +
+  scale_color_manual("Tournament", values=tournament_colors)
+
+p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
+  labs(x = "Points") +
+  scale_color_manual("League", values=league_colors) +
+  my_theme
+
+# grid.arrange( p2, p3, ncol=2,
+#              top = textGrob("Winning points by tournament and league",
+#                             gp=gpar(fontsize=20)))
+
+p1 <- ggplot(gs_players) + geom_density(aes(pointswon, color = factor(did_win))) + 
+  labs(x = "Points") +
+  scale_color_manual("Winner?", values=win_colors) +
+  my_theme +
+  facet_wrap(~league, ncol=1)
+
+grid.arrange(p1, p2, ncol=2,
+             top = textGrob("Distribution of points per match",
+                            gp=gpar(fontsize=20, fontfamily="serif")),
+              layout_matrix = rbind(c(NA, 1),
+                                    c(2, 1),
+                                    c(NA, 1)),
+             heights=c(1, 3, 1))
+
+```
+
+## Home court advantage
+
+It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
+In our data we find this to be true (i.e.
+French players win the French open more than French players win other slams).
+But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a particular country play in "their" tournament more often than they play in other tournaments.
+We also find this to be true in our data for France, The United States, and Australia (i.e.
+the proportion of French players in the French Open is greater than the proportion of French players in other slams).
+
+Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments.
+If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon.
+The same would be true for Australia and the US.
+But, we see that this isn't the case.
+After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
 
 ```{r, fig.height=4, fig.width=12}
-p1 <- ggplot(gs_players) + geom_histogram(aes(pointswon), bins=40) + 
-  labs(x = "Winner points")
+france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
 
+us_wins <- filter(gs_players, ioc == "USA", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+p1<- ggplot(france_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "French players",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
+
+p2 <- ggplot(us_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "American players",
+       x = "", y = "Win proportion") +  
+  scale_color_manual(guide=FALSE, values=tournament_colors)  + my_theme
+
+p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "Australian players",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(" ", values=tournament_colors) + my_theme
+
+grid.arrange(p1,p2, p3, nrow=1, widths = c(.29, .29, .42),
+             top = textGrob("Home court advantage debunked: win proportion by country",
+                            gp=gpar(fontsize=20)),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           # x = 0,
+                           # y = 0.5,
+                           just = "left"))
+
+```
+
+## Spaniards on clay
+
+```{r}
+filter(gs_players, round == "R128", ioc == "ESP") %>% 
+  group_by(tournament) %>% 
+  summarise(avg_ranking = mean(rank))
+
+spain_ranks <- filter(gs_players, ioc == "ESP") %>% 
+  group_by(tournament, round) %>% 
+  summarise(avg_ranking = mean(rank))
+
+ggplot(spain_ranks) + 
+  geom_line(aes(round, avg_ranking, color=tournament, group=tournament)) + 
+  labs(title = "Average ranking among Spaniards")
+
+```
+
+# Points won
+
+```{r, fig.height=4, fig.width=12}
 p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
-  labs(x = "Winner points")
+  labs(x = "Winner points") + my_theme +
+  scale_color_manual("Tournament", values=tournament_colors)
 
 p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
-  labs(x = "Winner points")
+  labs(x = "Winner points") +
+  scale_color_manual("League", values=league_colors) +
+  my_theme
 
-grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
-
+# grid.arrange( p2, p3, ncol=2,
+#              top = textGrob("Winning points by tournament and league",
+#                             gp=gpar(fontsize=20)))
 ```
@@
 grid.arrange(p1, p2, ncol=2, widths=c(.48, .52))
+
+```
+
+```{r, fig.height=4, fig.width=12}
+ggplot(filter(gs_players, league == "WTA", round == "R128")) + 
+  geom_point(aes(rank, pointswon, color=factor(did_win)), alpha=.5) + 
+  labs(x = "Rank", y = "Points", title="Round 128")
+
+```
+
+```{r, fig.height=4, fig.width=12}
+ggplot(filter(gs)) + 
+  geom_point(aes(loser_rank, winner_rank))
 
@@
 Yes, but not significantly more.
+
+## Spain proportion of wins across the tournaments
+
+```{r}
+spain_wins <- filter(gs_players, ioc == "ESP") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line(cex=2) + 
+  geom_point() +
+  labs(title = "Proportion of wins among Spaniards") +
+  geom_errorbar(aes(ymin = win_prop - 2*win_prop_se,
+                    ymax = win_prop + 2*win_prop_se),
+                width=.2,
+                 position=position_dodge(0.05))
+
+```
+
+# Player rankings over time
+
+```{r}
+
+# filter(gs_players, round == "R128") %>%
+#   select(tourney_start_date, rank, )
+#   ggplot(aes(x = tourney_start_date, y = rank)) + geom_line(group = name)
+#   
+filter(gs_players, round == "R128") %>%
+  rename(player_name = name) %>%
+  ggplot(aes(x = tourney_start_date, y = rank, color=player_name)) + 
+  geom_line(aes(group = factor(player_name))) + guides(color=FALSE, group=FALSE)
+
+```
 
@@
 
-```{r, fig.height=12, fig.width=12}
+```{r, fig.height=12, fig.width=12, eval=FALSE}
 library(GGally)
@@
 
-```{r}
+```{r, eval=FALSE}
 gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
@@
 ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+
+gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced, winner_rank) %>%
+na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced, winner_rank, tournament) %>%
+na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+
 ```
 
-```{r}
+```{r, eval=FALSE}
+gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced) %>%
+na.omit()
+
+hclust_results <- hclust(dist(scale(gs_num)))
+
+# plot(hclust_results)
+
+colors_to_use <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced, tournament) %>%
+  na.omit() %>% pull(tournament) %>% as.numeric()
+
+library(dendextend)
+dend <- as.dendrogram(hclust_results)
+colors_to_use <- colors_to_use[order.dendrogram(dend)]
+
+labels_colors(dend) <- colors_to_use
+plot(dend, main = "A color for every tournament")
+
+mds$tournament <- select(gs, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon,
+                w_bpSaved, w_bpFaced, tournament) %>%
+na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+```
+
+```{r, eval=FALSE}
 gs_num <- select(gs, w_ace, w_df, w_svpt, w_1stIn,
@@
 ```
+
+# MDS
+
+```{r, eval=FALSE}
+# winner_rank, loser_rank, 
+gs_num <- select(ungroup(gs_partial_pbp), winner_age,
+                 w_pointswon, w_gameswon, w_setswon, l_pointswon, l_gameswon,
+                 l_setswon, ave_serve_speed_p1,
+                 ave_serve_speed_p2, n_aces_p1, n_aces_p2, n_winners_p1, n_winners_p2,
+                 n_netpt_wpct_p1, n_netpt_wpct_p2, n_bp_wpct_p1, n_ue_p1, n_ue_p2,
+                 n_sv_wpct_p1, n_sv_wpct_p2) %>% 
+  na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- select(ungroup(gs_partial_pbp), winner_age,
+                 w_pointswon, w_gameswon, w_setswon, l_pointswon, l_gameswon,
+                 l_setswon, ave_serve_speed_p1,
+                 ave_serve_speed_p2, n_aces_p1, n_aces_p2, n_winners_p1, n_winners_p2,
+                 n_netpt_wpct_p1, n_netpt_wpct_p2, n_bp_wpct_p1, n_ue_p1, n_ue_p2,
+                 n_sv_wpct_p1, n_sv_wpct_p2, tournament) %>% 
+  na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+
+gs_num <- select(ungroup(gs_partial_pbp), ave_serve_speed_p1,
+                 ave_serve_speed_p2, n_aces_p1, n_aces_p2, n_winners_p1, n_winners_p2,
+                 n_netpt_wpct_p1, n_netpt_wpct_p2, n_bp_wpct_p1, n_ue_p1, n_ue_p2,
+                 n_sv_wpct_p1, n_sv_wpct_p2) %>% 
+  na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- select(ungroup(gs_partial_pbp), ave_serve_speed_p1,
+                 ave_serve_speed_p2, n_aces_p1, n_aces_p2, n_winners_p1, n_winners_p2,
+                 n_netpt_wpct_p1, n_netpt_wpct_p2, n_bp_wpct_p1, n_ue_p1, n_ue_p2,
+                 n_sv_wpct_p1, n_sv_wpct_p2, tournament) %>% 
+  na.omit() %>% pull( tournament)
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_text()
+
+gs_num <- ungroup(filter(gs_partial_pbp, Tour == "atp")) %>%
+    mutate(aces = n_aces_p1 + n_aces_p2) %>%
+  select(n_winners_p1, n_winners_p2, n_ue_p1, n_ue_p2, aces) %>%
+  na.omit()
+
+mds <- cmdscale(scale(dist(gs_num)), k = 2) %>%
+  as.data.frame()  # default column names will be V1, V2
+
+mds$tournament <- ungroup(filter(gs_partial_pbp, Tour == "atp")) %>%
+    mutate(aces = n_aces_p1 + n_aces_p2) %>%
+  select(n_winners_p1, n_winners_p2, n_ue_p1, n_ue_p2, aces, tournament) %>% 
+  na.omit() %>% pull( tournament)
+
+mds2 <- bind_cols(mds, ungroup(filter(gs_partial_pbp, Tour == "atp")))
+
+# mds$tournament[mds$tournament == "US Open"] <- "Australian Open"
+
+ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + geom_point()
+
+filter(mds2, V1 > 5)
+
+filter(mds2, V2 > 2)
+
+```
+
+```

### 2018-09-14 contest/graphics/graphics.Rmd
add spainiards clay and update graphics

@@
                            # y = 0.5,
-                           just = "left"))
+                           just = "left",
+                           gp=gpar(fontsize=20, fontfamily="serif")))
 
@@
 
-```{r}
-filter(gs_players, round == "R128", ioc == "ESP") %>% 
-  group_by(tournament) %>% 
-  summarise(avg_ranking = mean(rank))
+It is also commonly thought that Spaniards play better on clay.
+We are intersted in whether Spaniards win the French Open more than they win in other tournaments.
+It does appear that Spaniards are winning the French Open more than they are winning other tournaments.
+But, this result is not significant.
+In addition, the ranking of Spaniards in the French Open is, on average, higher than other tournmanents, which may help explain this common misconseption, although this result is not significant either.
 
-spain_ranks <- filter(gs_players, ioc == "ESP") %>% 
+```{r, fig.height=5, fig.width=12}
+
+spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
   group_by(tournament, round) %>% 
-  summarise(avg_ranking = mean(rank))
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
+            avg_ranking = mean(rank),
+            rank_se = sd(rank)/sqrt(n()))
 
-ggplot(spain_ranks) + 
-  geom_line(aes(round, avg_ranking, color=tournament, group=tournament)) + 
-  labs(title = "Average ranking among Spaniards")
+p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  labs(title = "Win proportion among Spaniards",
+       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme  
+  # geom_point() + 
+  # labs(title = "Proportion of wins among Spaniards") +
+  # geom_errorbar(aes(ymin = win_prop - win_prop_se,
+  #                   ymax = win_prop + win_prop_se),
+  #               width=.2,
+  #                position=position_dodge(0.05))
+
+p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament)) + 
+  geom_line() + 
+  labs(title = "Average ranking among Spaniards",
+       x = " ",
+       y = "Average rank") + 
+  scale_color_manual(" ", values=tournament_colors) + my_theme 
+# +
+#   geom_errorbar(aes(ymin = avg_ranking - 1.96*rank_se,
+#                     ymax = avg_ranking + 1.96*rank_se),
+#                 width=.2,
+#                  position=position_dodge(0.05))
+
+grid.arrange(p1,p2, nrow=1, widths = c(.44, .56),
+             top = textGrob("Spaniards perform better on Clay: debunked",
+                            gp=gpar(fontsize=20)),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           # x = 0,
+                           # y = 0.5,
+                           just = "left",
+                           gp=gpar(fontfamily="serif")))
 
@@
 Yes, but not significantly more.
-
-## Spain proportion of wins across the tournaments
-
-```{r}
-spain_wins <- filter(gs_players, ioc == "ESP") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line(cex=2) + 
-  geom_point() +
-  labs(title = "Proportion of wins among Spaniards") +
-  geom_errorbar(aes(ymin = win_prop - 2*win_prop_se,
-                    ymax = win_prop + 2*win_prop_se),
-                width=.2,
-                 position=position_dodge(0.05))
-
-```
 

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
add spainiards clay and update graphics

@@
 
+## Spaniards on clay
+
+It is also commonly thought that Spaniards play better on clay.
+We are intersted in whether Spaniards win the French Open more than they win in other tournaments.
+It does appear that Spaniards are winning the French Open more than they are winning other tournaments.
+But, this result is not significant.
+In addition, the ranking of Spaniards in the French Open is, on average, higher than other tournmanents, which may help explain this common misconseption, although this result is not significant either.
+
+```{r, fig.height=5, fig.width=12}
+
+spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
+            avg_ranking = mean(rank),
+            rank_se = sd(rank)/sqrt(n()))
+
+p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  labs(title = "Win proportion among Spaniards",
+       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
+       x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme  
+  # geom_point() + 
+  # labs(title = "Proportion of wins among Spaniards") +
+  # geom_errorbar(aes(ymin = win_prop - win_prop_se,
+  #                   ymax = win_prop + win_prop_se),
+  #               width=.2,
+  #                position=position_dodge(0.05))
+
+p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament)) + 
+  geom_line() + 
+  labs(title = "Average ranking among Spaniards",
+       x = " ",
+       y = "Average rank") + 
+  scale_color_manual(" ", values=tournament_colors) + my_theme 
+# +
+#   geom_errorbar(aes(ymin = avg_ranking - 1.96*rank_se,
+#                     ymax = avg_ranking + 1.96*rank_se),
+#                 width=.2,
+#                  position=position_dodge(0.05))
+
+grid.arrange(p1,p2, nrow=1, widths = c(.44, .56),
+             top = textGrob("Spaniards perform better on Clay: debunked",
+                            gp=gpar(fontsize=20)),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           # x = 0,
+                           # y = 0.5,
+                           just = "left",
+                           gp=gpar(fontfamily="serif")))
+
+```
+
 # Methods {#sec:methods}

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
spell check lol

@@
 We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
-Both of those models had success using hierarhcical Bayesian models, which we employ in our own models.
-More specifically, we model the players' expected points in a match based on the player's own characteristics, the court/tournament effects, and the oponent's ranking.
+Both of those models had success using hierarchical Bayesian models, which we employ in our own models.
+More specifically, we model the players' expected points in a match based on the player's own characteristics, the court/tournament effects, and the opponent's ranking.
 
@@
 
-We first examine the distibution of points earned per match.
+We first examine the distribution of points earned per match.
 We find that the distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
 As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
-Also unsuprisingly, the winners of the match tended to earn more points than the losers.
+Also unsurprisingly, the winners of the match tended to earn more points than the losers.
 
@@
 It is also commonly thought that Spaniards play better on clay.
-We are intersted in whether Spaniards win the French Open more than they win in other tournaments.
+We are interested in whether Spaniards win the French Open more than they win in other tournaments.
 It does appear that Spaniards are winning the French Open more than they are winning other tournaments.
 But, this result is not significant.
-In addition, the ranking of Spaniards in the French Open is, on average, higher than other tournmanents, which may help explain this common misconseption, although this result is not significant either.
+In addition, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, which may help explain this common misconception, although this result is not significant either.
 

### 2018-09-14 contest/model-fitting/model-comparisons.Rmd
added data, writeup in model-fitting

@@
 gs_players$late_round = gs_players$round >= "R16"
-
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
 ```
@@
     + `ind_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i P_i$
-    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC, $\beta_{T[i]} = \alpha_i P_i$
+    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC and $T$ represents indicators for tournament, $\beta_{T[i]} = \alpha_i P_i$
     + `ind_int_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{0[i]}$, where $X$ does not include IOC, $\beta_{0[i]} = \alpha_i$
-    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{Y[i]}$, where $X$ does not include IOC, $\beta_{Y[i]} = \alpha_i Y_i$
+    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{D[i]}$, where $X$ does not include IOC and $D$ represents indicators for years, $\beta_{Y[i]} = \alpha_i Y_i$
 
 ```{r, cache = TRUE}
+set.seed(091418)
 nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
@@
 ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
+ind_noioc_rank = glmer(did_win ~ late_round + seeded + opponent_seeded + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+```
 
+```{r}
 model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
@@
 
-# 486, 504, 526
-indices = c(486, 504, 526)
-superstars = data.frame(
-  name = rep(c("Serena", "Nadal", "Federer"), 2),
-  model = c(rep("ind_logistic", 3), rep("ind_logistic_noioc",3)),
-  `Australian Open` = c(ranef(ind_logistic)$name_int[indices,1], ranef(ind_logistic_noioc)$name_int[indices,1]),
-  `French Open` = c(ranef(ind_logistic)$name_int[indices,2], ranef(ind_logistic_noioc)$name_int[indices,2]),
-  `US Open` = c(ranef(ind_logistic)$name_int[indices,3], ranef(ind_logistic_noioc)$name_int[indices,3]),
-  `Wimbledon` = c(ranef(ind_logistic)$name_int[indices,4], ranef(ind_logistic_noioc)$name_int[indices,4])
-)
-
 superstars = superstars %>%
@@
   facet_wrap(~tournament, scales = "free_x") + 
-  geom_errorbarh(height = .5)
+  geom_errorbarh(height = .5) + 
+  xlab("Random Effect") + 
+  ylab("")
+```
+
+Rank and opponent's rank clearly have the largest effect on predicting the winner.
+Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
+Including rank as an indicator (seeded/not-seeded) substantially decreases the fit of the model, even when individual effects are included.
+We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, and net points won -- and see if individual differences are detectable.
+
+```{r}
+load("../../data/gs_partial_players.rda")
+gs_partial_players$atp = gs_partial_players$Tour == "atp"
+gs_partial_players$year_fac = as.factor(gs_partial_players$year)
+gs_partial_players$late_round = gs_partial_players$round >= "R16"
+gs_partial_players$seeded = gs_partial_players$rank <= 32
+gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+
+aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Aces")
+
+winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Winners")
+
+nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Net Points Won")
+
+ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Unforced Errors")
+
+plot_grid(aces_ranef, winners_ranef, nets_ranef, ue_ranef)
 ```

### 2018-09-14 contest/model-fitting/model-fitting.Rmd
added data, writeup in model-fitting

--- 2018-09-10 contest/model-fitting/model-fitting.Rmd
@@
 date: "9/10/2018"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+library(lme4)
+library(dplyr)
+library(kableExtra)
+library(cowplot)
+library(broom)
+library(tidyr)
 
-library(rstan)
-library(shinystan)
-library(brms)
-```
-
-### Modeling notes:
-
-* Did not include 'height' in models due to missingness in data
-* Used brms package to call Stan (multilevel modeling-specific, uses syntax similar to lme4 if we want to switch to maximum-likelihood based estimation)
-* Each model takes 15-20 minutes to fit (3 models in document)
-
-### Model 1:
-
-No effect for player; fixed effect for IOC, tournament, etc.
-
-```{r}
-load(file = "../../data/gs_players.rda")
-load("../../data/gs.rda")
-
+load("../../data/gs_players.rda")
+load("../../data/gs_partial_players.rda")
+load("../../data/nocourt_logistic.rda")
+load("../../data/base_logistic.rda")
+load("../../data/country_logistic.rda")
+load("../../data/ind_int_logistic.rda")
+load("../../data/ind_logistic.rda")
+load("../../data/ind_logistic_noioc.rda")
+load("../../data/ind_year_logistic.rda")
 gs_players$name_int = as.integer(as.factor(gs_players$name))
@@
 gs_players$ioc_fac = as.factor(gs_players$ioc)
-
-basemod = brm(pointswon ~ ioc_fac + hand_fac + tournament + round + rank + opponent_rank + minutes, data = gs_players, family = gaussian())
+gs_players$atp = gs_players$league == "ATP"
+gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
 ```
 
-### Model 2:
+We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance.
+For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match.
+Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
+For these reasons, we chose to model whether a player won the match, with the following predictors:
 
-Group-level effect and tournament effect for IOC; fixed effect for other vars (still no individual-level effects)
+* `ioc_fac`: Country that the player represents
+* `tournament`: Which Grand Slam the match was played at
+* `late_round`: Indicator for whether the match occured in Round of 16 or later
+* `rank`: Rank of player at the time of the match
+    + Included on log scale 
+* `opponent_rank`: Rank of opponent at the time of the match
+    + Included on log scale
+* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
+* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.
+
+We will use a multilevel logistic regression framework.
+This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects.
+Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
+It also provides a way to examine individual player effects and assess, for instance, whether Nadal is more likely to win at Wimbledon after accounting for other variables.
+We considered the following models:
+
+* No mixed effects
+    + `nocourt_logistic`: $\text{logit}(\pi_i) = \beta X$, $X$ does not include `tournament`
+        + If this model fits well, court surface does not have an effect on winning probability
+    + `base_logistic`: $\text{logit}(\pi_i) = \beta X$
+        + If this model fits better than the mixed-effects models, performance does not depend on individual-level or IOC-level 
+* IOC-level effect
+    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i C_i$
+* Player-level effect
+    + `ind_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i P_i$
+        * Includes both a fixed-effect term for IOC and a player-level effect for each grand slam
+    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC and $T$ represents indicators for tournament, $\beta_{T[i]} = \alpha_i P_i$
+        * Simpler than the `ind_logistic` model, since we exclude IOC effects.
+    + `ind_int_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{0[i]}$, where $X$ does not include IOC, $\beta_{0[i]} = \alpha_i$
+        * Similar to `ind_logistic_noioc`, but includes a single player-level effect rather than a player-level effect for each grand slam. 
+    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{D[i]}$, where $X$ does not include IOC and $D$ represents indicators for years, $\beta_{Y[i]} = \alpha_i Y_i$
+        * Rather than including a player-level effect for grand slam, includes a player-level effect for each year in the dataset
+
+```{r, eval = FALSE}
+set.seed(091418)
+nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
+ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
+ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
+```
+
+Based on AIC, the `ind_logistic_noioc` model performed best after accounting for the number of variables in the model.
 
 ```{r}
-country_mod = brm(pointswon ~ hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |ioc_fac), data = gs_players, family = gaussian())
+model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
+model.ests = data.frame()
+ran.effects = data.frame()
+for(mod in 1:nrow(model.sums)){
+  name = model_names[mod]
+  model.sums$aic[mod] = extractAIC(get(name))[2]
+  model.sums$edf[mod] = extractAIC(get(name))[1]
+  betas = tidy(get(name))
+  betas$model = name
+  if("group" %in% colnames(betas)){
+  model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+    select(., -"group"))
+  }
+  else model.ests = rbind(model.ests, betas)
+}
+
+kable(model.sums)
 ```
 
-### Model 2:
+```{r}
+fixed.effects.plot = model.ests %>%
+  filter(., !grepl("ioc", term)) %>%
+  ggplot(., aes(x = term, y = exp(estimate), col = model, ymin = exp(estimate - 2*std.error), ymax = exp(estimate + 2*std.error))) +
+  geom_errorbar(width = .7, position = position_dodge()) +
+  theme_minimal() + 
+  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
+  scale_color_brewer(palette = "Dark2") + 
+  ggtitle("Fixed effects") +
+  theme(legend.position = "bottom")
 
-Group-level tournament effect for player; fixed effect for IOC, tournament, etc.
+```
 
 ```{r}
-ind_mod = brm(pointswon ~ ioc_fac + hand_fac + round + rank + opponent_rank + minutes + (1 + tournament |name_int), data = gs_players, family = gaussian())
+ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
+ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
+
+random.effects.plot = ran_tourn %>%
+  gather(., tourn, estimate, -model) %>%
+  mutate(., tournament = gsub("tournament", "", tourn)) %>%
+  ggplot(., aes(x = tournament, y = exp(estimate), fill = model)) +
+  geom_violin() + 
+  theme_minimal() +
+  scale_fill_brewer(palette = "Dark2") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  ggtitle("Random effects") +
+  theme(legend.position = "bottom")
+
+plot_grid(fixed.effects.plot, random.effects.plot)
 ```
 
-### Model Comparison (stats methods)
-
-(Note: for full MCMC diagnostics, run `launch_shinystan(MOD_NAME))` )
-
-In models that don't include individual effects (basemod and country_mod), the estimated effect of "Wimbledon" on points won is quite high compared to the other models.
-(Could this be because better players are more likely to play in Wimbledon??) This effect starts to even out after including individual effects.
+Looking at the `ind_logistic_noioc` random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
 
 ```{r}
-plot(marginal_effects(basemod, effects = "tournament"))
-plot(marginal_effects(country_mod, effects = "tournament"))
-plot(marginal_effects(ind_mod, effects = "tournament"))
+pairs(ranef(ind_logistic_noioc)$name_int, pch = 21)
 ```
 
-### Model Comparison (tennis knowledge)
-
-Checks:
-* Serena should be good on hard surface
-* Nadal should be good on clay 
-* Federer should be good on grass
+If we look at our big three, their individual effects are not very informative (especially after taking uncertainty into consideration).
 
 ```{r}
+name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
+name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
+as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
+  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
+  ggplot(., aes(y = name, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament, scales = "free_x") + 
+  geom_errorbarh(height = .5) + 
+  xlab("Random Effect") + 
+  ylab("")
+```
 
+Rank and opponent's rank clearly have the largest effect on predicting the winner.
+Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
+Including rank as an indicator (seeded/not-seeded) substantially decreases the fit of the model, even when individual effects are included.
+We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, and net points won -- and see if individual differences are detectable.
+
+```{r}
+load("../../data/gs_partial_players.rda")
+gs_partial_players$atp = gs_partial_players$Tour == "atp"
+gs_partial_players$year_fac = as.factor(gs_partial_players$year)
+gs_partial_players$late_round = gs_partial_players$round >= "R16"
+gs_partial_players$seeded = gs_partial_players$rank <= 32
+gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+
+aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Aces")
+
+winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Winners")
+
+nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Net Points Won")
+
+ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(y = grp, x = condval, xmin = condval- 2*condsd, xmax = condval + 2*condsd)) +
+  facet_wrap(~tournament) + 
+  geom_errorbarh(height = .5) +
+  xlab("Random Effect") + 
+  ylab("") +
+  ggtitle("Unforced Errors")
+
+plot_grid(aces_ranef, winners_ranef, nets_ranef, ue_ranef)
 ```

### 2018-09-14 contest/model-fitting/fit-models-and-save.R

@@ -1,0 +1,26 @@
+load("gs_players.rda")
+load("gs_partial_players.rda")
+set.seed(091418)
+nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
+ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
+ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+
+devtools::use_data(nocourt_logistic,
+                   base_logistic,
+                   country_logistic,
+                   ind_logistic,
+                   ind_logistic_noioc,
+                   ind_int_logistic,
+                   ind_year_logistic,
+                   n_aces_mod,
+                   n_winners_mod,
+                   n_net_mod,
+                   n_ue_mod)

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
spell check lol

@@
 title: "Analyzing the Effect of Surface in Tennis Grand Slams"
-author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
+author: "###########"
 date: '`r format(Sys.time(), "%B %d, %Y")`'
@@
 
-We first examine the distribution of points earned per match.
-We find that the distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
+We first examine the distribution of points earned per match to assess normality.
+The distribution of points earned is approximately normal.
+This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
 As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
@@
 
-```{r, fig.height=4, fig.width=12}
+```{r, fig.height=4, fig.width=14, fig.align='center'}
 france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
 aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
   group_by(tournament, round) %>% 
@@
        x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
+
+p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "British players",
+       x = "", y = "Win proportion") + 
   scale_color_manual(" ", values=tournament_colors) + my_theme
 
-grid.arrange(p1,p2, p3, nrow=1, widths = c(.29, .29, .42),
-             top = textGrob("Home court advantage debunked: win proportion by country",
+grid.arrange(p3, p1, p2,  p4, nrow=1, widths = c(.22, .22, .22, .34),
+             top = textGrob("Home court advantage: debunked",
                             gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size as the tournament goes on",
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                            # x = 0,
                            # y = 0.5,
-                           just = "left"))
+                           just = "left",
+                           gp=gpar(fontfamily="serif")))
 
@@
 
-```{r, fig.height=5, fig.width=12}
+```{r, fig.height=4, fig.width=12, fig.align='center'}
 
@@
   scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
-  # geom_point() + 
-  # labs(title = "Proportion of wins among Spaniards") +
-  # geom_errorbar(aes(ymin = win_prop - win_prop_se,
-  #                   ymax = win_prop + win_prop_se),
-  #               width=.2,
-  #                position=position_dodge(0.05))
 
@@
 
-grid.arrange(p1,p2, nrow=1, widths = c(.44, .56),
+grid.arrange(p1, p2, nrow=1, widths = c(.44, .56),
              top = textGrob("Spaniards perform better on Clay: debunked",
-                            gp=gpar(fontsize=20)),
+                            gp=gpar(fontsize=20, fontfamily="serif")),
              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
tall players and toc

@@
     number_sections: yes
+    toc: true
 header-includes:
@@
 library(grid)
+library(ggrepel)
 
@@
 
+## Tall players and aces
+
+Why the actual EFF won't fig.width work?
+
+```{r, fig.width=10, fig.height=8, fig.align='center'}
+p_atp <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
+  geom_jitter() + 
+  scale_color_manual(guide=FALSE, values = win_colors, labels = c("No", "Yes")) + 
+  geom_smooth(method = "lm") + 
+  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
+                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
+                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label = name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "ATP")+
+  my_theme
+
+p_wta <- gs_players %>% filter(league == "WTA") %>%
+ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
+  geom_jitter() + 
+  scale_color_manual("Winner?", values = win_colors, labels = c("No", "Yes")) + 
+  geom_smooth(method = "lm") + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "WTA")+
+  my_theme + theme(legend.position="bottom")
+
+grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
+             top = textGrob("Taller players are better servers: confirmed",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
+
+```
+
 # Methods {#sec:methods}

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
dreaded merge

--- 2018-09-14 contest/surface-type/paper-tennis.Rmd
@@
 library(dplyr)
+library(tidyr)
 library(ggplot2)
@@
 library(grid)
-
+library(MASS)
 ## To change into library(courtsports)
 devtools::load_all("../..")
+
 ```
@@
 
-# Results {#sec:results}
+## Examining individual players
+```{r table}
+data(gs_partial_players)
+
+top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
+matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
+  summarize(matches_played = n())
+matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
+  summarize(matches_played = sum(matches_played))
+tab <- spread(matches_played_slam, key = name, value = matches_played)
+total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
+colnames(total_df) <- colnames(tab)
+tab <- rbind(tab, total_df)
+
+knitr::kable(tab, format = "latex",
+               row.names = FALSE, col.names = c("Tournament", "Nadal", "Federer", "Williams"),
+               caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
+               booktabs = TRUE) %>% kable_styling(latex_options = "striped") %>%
+    row_spec(5, bold = TRUE, background = "lightgray")
+```
+
+In Table \ref{tab:three-gs-counts}, we display the number of matches Nadal, Federer, and Williams have played from 2013-2017.
+Over that time span, Nadal won 6 grand slams, Federer won 1, Williams won 8.
+Despite this, Federer played more total matches in Nadal.
+All three players were absent for exactly three slams during this time period due to external factors (Nadal: (1 AO, 0 FO, 1 Wim, 1 USO), Federer: (0 AO, 2 FO, 0 Wim, 1 USO), Serena: (0 AO, 1 FO, 1 Wim, 1 USO)).
+Not unexpectedly, Nadal has the most wins on clay (`r tab[2, 2]`), Federer has the most wins at Wimbledon (`r tab[4, 3]`) and Williams the most on hardcourt (`r tab[1, 4]` at AO and `r tab[4, 4]` at USO).
+Despite having played more matches than Nadal, Federer only has 1 grand slam to show compared to Nadal's six, which indicates that Federer made it deeper into tournaments on average but had difficulties winning the championships.
+For the WTA, Williams had her second most successful five-year span at the grand slams over this time period, winning 8.
+
+We fit individual models for these three individuals (subsetting the data to their matches only), estimating the percent of points won in a match using stepwise regression.
+The lower model is the  percent of points regressed on the opponent rank and indicator variables for court type with the FO as the reference variable.
+The upper model is the lower model with the additional variables of winner to unforced error ratio, average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
+We do not display the full best fit models here, but they are available online.
+
+ADD CIS!!
+
+## Nadal
+
+```{r nadal-model}
+
+step_model_n <- model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
+
+std_res_n <- rstandard(step_model_n)
+
+gn <- ggplot(lm(step_model_n$formula, data = step_model_n$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Nadal",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+coefs_n <- coefficients(step_model_n)
+```
+
+Looking at Nadal individually, the model with the lowest AIC has positive coefficients opponent rank, percent of aces, and percent of break points won.
+It has negative coefficients for AO, USO, and Wim.
+compared to FO and for percent of net points won.
+There are no interaction effects in this model.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids} (along with the other linear regression diagnostics plots not shown here) shows that the model is a good fit for predicting the percent of points won.
+For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)`$ less at the AO and $`r signif(abs(coefs_n[4]), 2)`$ less at Wim compared to the FO; to increase by $`r signif(abs(coefs_n[5]), 2)`$ for a one unit increase in opponent rank; a `r .01 * signif(abs(coefs_n[7]), 2)`\% increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)`$ for a .01 increase in percent of net points won.
+As such, we see clear evidence Nadal performs better at the FO compared to the other grand slams.
+
+## Federer
+```{r federer-model}
+
+step_model_f <- model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
+
+std_res_f <- rstandard(step_model_f)
+
+gf <- ggplot(lm(step_model_f$formula, data = step_model_f$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Federer",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+coefs_f <- coefficients(step_model_f)
+```
+
+For Federer, the model with the lowest AIC has negative coefficients AO, FO, and USO.
+vs Wim., and W/UE,  It has positive coefficients for opponent rank, percent of break points won, and all the interaction terms: court and W/UE.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.
+For the significant variabless ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)`$ less at AO compared to the Wim.; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`$ less at FO compared to the Wim.; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`$\% less at USO compared to the Wim.;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)`$ for a one unit increase in W/UE at AO, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)`$ for a one unit increase in W/UE at FO, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)`$ for a one unit increase in W/UE at USO.
+It is clear that Federer performs best at Wimbledon.
+
+## Williams
+```{r williams-model}
+
+step_model_w <- model_individual(player_name = "Serena Williams", data = gs_partial_players,
+                                 start_lower = TRUE, ref = "Australian Open")
+
+std_res_w <- rstandard(step_model_w)
+
+gw <- ggplot(lm(step_model_w$formula, data = step_model_w$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Williams",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+coefs_w <- coefficients(step_model_w)
+```
+
+```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}"}
+
+grid.arrange(gn, gf, gw, ncol = 3, 
+              top = textGrob("Standardized Residuals vs. Predicted % Points Won",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
+
+```
+
+For Williams, the model with the lowest AIC is the largest model of the three selected 22 coefficients.
+Since Williams only has 59 observations, we believe this model is overfitting and so caution against inference with this model.
+It should also be noted that no women's point by point data was recorded for 2015 so that year is excluded from this model.
+The coefficients are court, opponent rank, W/UE, average serve speed, percent aces, percent break points won, percent net points won, and interaction effects with court and W/UE, court and average service speed, court and percent aces, and court and break points won.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal.
+For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)`$for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2)`$ for a one unit increase in W/UE at Wim.
+compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)`$ for a .01 increase in percent of break points won at the French Open.
+
+For this, we do not see that Williams is dominant on the hard courts compared to the other courts.
+Rather, she has good results on all the courts.
+Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points than at the Australian Open.
 

### 2018-09-15 contest/surface-type/paper-tennis.tex
dreaded merge

@@
   \postdate{\par}
-    \date{September 12, 2018}
+    \date{September 15, 2018}
 
@@
 2013; Egidi and Gabry 2018). Both of those models had success using
-hierarhcical Bayesian models, which we employ in our own models. More
+hierarchical Bayesian models, which we employ in our own models. More
 specifically, we model the players' expected points in a match based on
 the player's own characteristics, the court/tournament effects, and the
-oponent's ranking.
+opponent's ranking.
 
@@
 of origin, and player attributes. Additionally, we take into account the
-number of sets in a match, the surface type, minutes played, and round
-of the tournament. A subset of the data is shown in Table
-\ref{tab:data}.
+number of sets in a match, the surface type, and round of the
+tournament. A subset of the data is shown in Table \ref{tab:data}.
 
@@
 
+\hypertarget{examining-the-distribution-of-points-earned}{%
+\subsubsection{Examining the distribution of points
+earned}\label{examining-the-distribution-of-points-earned}}
+
+We first examine the distribution of points earned per match. We find
+that the distribution is similar across tournament, with Wimbledon
+differing slightly from the other grand slams. As expected, there are
+more points earned in the WTA than the ATP due to the differing numbers
+of games played. Also unsurprisingly, the winners of the match tended to
+earn more points than the losers.
+
+\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-1-1.pdf}
+
+\hypertarget{home-court-advantage}{%
+\subsubsection{Home court advantage}\label{home-court-advantage}}
+
+It is commonly thought that there is a home court advantage in grand
+slam games (SOURCE). In our data we find this to be true (i.e.~French
+players win the French open more than French players win other slams).
+But, we also know that the home team is given preference for wild card
+bids (SOURCE) so potentially citizens of a particular country play in
+``their'' tournament more often than they play in other tournaments. We
+also find this to be true in our data for France, The United States, and
+Australia (i.e.~the proportion of French players in the French Open is
+greater than the proportion of French players in other slams).
+
+Therefore, we want to see how the proportion of wins for the home
+country changes across the different tournaments. If there was really a
+home court advantage, the proportion of French wins each round would be
+higher at the French Open than the Australian Open, the US Open, and
+Wimbledon. The same would be true for Australia and the US. But, we see
+that this isn't the case. After accounting for the number of players
+from each country, we don't find a home court advantage in the grand
+slam.
+
+\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-2-1.pdf}
+
+\hypertarget{spaniards-on-clay}{%
+\subsection{Spaniards on clay}\label{spaniards-on-clay}}
+
+It is also commonly thought that Spaniards play better on clay. We are
+interested in whether Spaniards win the French Open more than they win
+in other tournaments. It does appear that Spaniards are winning the
+French Open more than they are winning other tournaments. But, this
+result is not significant. In addition, the ranking of Spaniards in the
+French Open is, on average, higher than other tournaments, which may
+help explain this common misconception, although this result is not
+significant either.
+
+\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-3-1.pdf}
+
 \hypertarget{sec:methods}{%
@@
 
-\hypertarget{sec:results}{%
-\section{Results}\label{sec:results}}
+\hypertarget{examining-individual-players}{%
+\subsection{Examining individual
+players}\label{examining-individual-players}}
+
+\rowcolors{2}{gray!6}{white}
+\begin{table}
+
+\caption{\label{tab:table}\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.}
+\centering
+\begin{tabular}[t]{lrrr}
+\hiderowcolors
+\toprule
+Tournament & Nadal & Federer & Williams\\
+\midrule
+\showrowcolors
+Australian Open & 20 & 28 & 30\\
+French Open & 29 & 14 & 23\\
+US Open & 21 & 22 & 26\\
+Wimbledon & 11 & 29 & 21\\
+\rowcolor{lightgray}  \textbf{Total} & \textbf{81} & \textbf{93} & \textbf{100}\\
+\bottomrule
+\end{tabular}
+\end{table}
+\rowcolors{2}{white}{white}
+
+In Table \ref{tab:three-gs-counts}, we display the number of matches
+Nadal, Federer, and Williams have played from 2013-2017. Over that time
+span, Nadal won 6 grand slams, Federer won 1, Williams won 8. Despite
+this, Federer played more total matches in Nadal. All three players were
+absent for exactly three slams during this time period due to external
+factors (Nadal: (1 AO, 0 FO, 1 Wim, 1 USO), Federer: (0 AO, 2 FO, 0 Wim,
+1 USO), Serena: (0 AO, 1 FO, 1 Wim, 1 USO)). Not unexpectedly, Nadal has
+the most wins on clay (29), Federer has the most wins at Wimbledon (29)
+and Williams the most on hardcourt (30 at AO and 21 at USO). Despite
+having played more matches than Nadal, Federer only has 1 grand slam to
+show compared to Nadal's six, which indicates that Federer made it
+deeper into tournaments on average but had difficulties winning the
+championships. For the WTA, Williams had her second most successful
+five-year span at the grand slams over this time period, winning 8.
+
+We fit individual models for these three individuals (subsetting the
+data to their matches only), estimating the percent of points won in a
+match using stepwise regression. The lower model is the percent of
+points regressed on the opponent rank and indicator variables for court
+type with the FO as the reference variable. The upper model is the lower
+model with the additional variables of winner to unforced error ratio,
+average serve speed, percent aces, percent break points, percent net
+points won, and their interaction effects with court type. We do not
+display the full best fit models here, but they are available online.
+
+ADD CIS!!
+
+\hypertarget{nadal}{%
+\subsection{Nadal}\label{nadal}}
+
+Looking at Nadal individually, the model with the lowest AIC has
+positive coefficients opponent rank, percent of aces, and percent of
+break points won. It has negative coefficients for AO, USO, and Wim.
+compared to FO and for percent of net points won. There are no
+interaction effects in this model.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}
+(along with the other linear regression diagnostics plots not shown
+here) shows that the model is a good fit for predicting the percent of
+points won. For the significant variables (\(\alpha\) = 95\%), while
+adjusting for the other variables, we expect percent of points won to be
+\(0.034\) less at the AO and \(0.046\) less at Wim compared to the FO;
+to increase by \(1.5\times 10^{-4}\) for a one unit increase in opponent
+rank; a 0.0012\% increase for a .01 increase in percent of break points
+won; and to decrease by \(8.8\times 10^{-4}\) for a .01 increase in
+percent of net points won. As such, we see clear evidence Nadal performs
+better at the FO compared to the other grand slams.
+
+\hypertarget{federer}{%
+\subsection{Federer}\label{federer}}
+
+For Federer, the model with the lowest AIC has negative coefficients AO,
+FO, and USO. vs Wim., and W/UE, It has positive coefficients for
+opponent rank, percent of break points won, and all the interaction
+terms: court and W/UE.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}
+shows that the model is a fair fit for predicting the percent of points
+won, but almost seem to see two clusters of residuals splitting with
+predicted point percentage of about 0.63. For the significant variabless
+(\(\alpha\) = 95\%), while adjusting for the other variables, we expect
+percent of points won to be \(0.04\) less at AO compared to the Wim.;
+\(0.058\) less at FO compared to the Wim.; \(0.007\)\% less at USO
+compared to the Wim.; to decrease by \(0.022\) for a one unit increase
+in W/UE at AO, \(0.073\) for a one unit increase in W/UE at FO, and
+\(0.041\) for a one unit increase in W/UE at USO. It is clear that
+Federer performs best at Wimbledon.
+
+\hypertarget{williams}{%
+\subsection{Williams}\label{williams}}
+
+\begin{figure}
+\centering
+\includegraphics{paper-tennis_files/figure-latex/resid plots-1.pdf}
+\caption{\label{fig:individ-resids}Standardized residuals vs percent of
+points won best fit model from forward-backwards stepwise regression for
+Nadal's, Federer's, and William's, respectively.\}}
+\end{figure}
+
+For Williams, the model with the lowest AIC is the largest model of the
+three selected 22 coefficients. Since Williams only has 59 observations,
+we believe this model is overfitting and so caution against inference
+with this model. It should also be noted that no women's point by point
+data was recorded for 2015 so that year is excluded from this model. The
+coefficients are court, opponent rank, W/UE, average serve speed,
+percent aces, percent break points won, percent net points won, and
+interaction effects with court and W/UE, court and average service
+speed, court and percent aces, and court and break points won.
+
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}
+shows that the model is a fair fit for predicting the percent of points
+won, but we tend to overestimate the percent of points won in comparison
+with Nadal. For the significant variables (\(\alpha\) = 95\%), while
+adjusting for the other variables, we expect percent of points won to
+increase by \(4.8\times 10^{-4}\) for a 1 unit increase in oponent rank;
+to increase by \(0.034\) for a one unit increase in W/UE at FO, to
+increase by \(0.092\)for a one unit increase in W/UE at USO, and to
+increase by \(0.0038\) for a one unit increase in W/UE at Wim. compared
+to AO; to increase by \(2\times 10^{-4}\) for a .01 increase in percent
+of aces at the French Open; and to increase by \(0.0024\) for a .01
+increase in percent of break points won at the French Open.
+
+For this, we do not see that Williams is dominant on the hard courts
+compared to the other courts. Rather, she has good results on all the
+courts. Her significant interaction effects show that if she focuses on
+acing her opponent and capitalizing on breakpoints, then she has a
+better chance at winning more points than at the Australian Open.
 

### 2018-09-15 R/clean-data.R

@@ -113,6 +113,7 @@ matches_to_player_data <- function(matches,
                                    opp_var_names = "opponent_rank"){
 
 
+    matches$total_points <- matches$w_pointswon + matches$l_pointswon
     ## Copy data frames
     matches_l <- matches
     matches_w <- matches

### 2018-09-15 contest/surface-type/paper-tennis.Rmd

@@
 library(MASS)
+library(ggrepel)
+
 ## To change into library(courtsports)
@@
                            gp=gpar(fontfamily="serif")))
+
+```
+
+## Tall players and aces
+
+Why the actual EFF won't fig.width work?
+
+```{r, fig.width=10, fig.height=8, fig.align='center'}
+p_atp <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
+  geom_jitter() + 
+  scale_color_manual(guide=FALSE, values = win_colors, labels = c("No", "Yes")) + 
+  geom_smooth(method = "lm") + 
+  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
+                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
+                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label = name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "ATP")+
+  my_theme
+
+p_wta <- gs_players %>% filter(league == "WTA") %>%
+ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
+  geom_jitter() + 
+  scale_color_manual("Winner?", values = win_colors, labels = c("No", "Yes")) + 
+  geom_smooth(method = "lm") + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "WTA")+
+  my_theme + theme(legend.position="bottom")
+
+grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
+             top = textGrob("Taller players are better servers: confirmed",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
 

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
updates

@@
 
-## Nadal
+### Nadal
 
@@
 
-## Federer
+### Federer
 ```{r federer-model}
@@
 
-## Williams
+### Williams
 ```{r williams-model}
@@
 
+## Mixed-effects models for all grand slams
+
+### Logistic regression for wins
+
+We will use a multilevel logistic regression framework.
+This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects.
+Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
+It also provides a way to examine individual player effects and assess, for instance, whether Federer is more likely to win at Wimbledon after accounting for other variables.
+
+We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance.
+For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match.
+Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
+For these reasons, we chose to model whether a player won the match, with the following predictors:
+
+* `ioc_fac`: Country that the player represents
+* `tournament`: Which Grand Slam the match was played at
+* `late_round`: Indicator for whether the match occured in Round of 16 or later
+* `rank`: Rank of player at the time of the match
+    + Included on log scale 
+* `opponent_rank`: Rank of opponent at the time of the match
+    + Included on log scale
+* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
+* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.
+
+(short discussion of models fit, show plots for final model)
+
+### Linear regression for other stats
+
+Rank and opponent's rank clearly have the largest effect on predicting the winner.
+Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
+Including rank as an indicator (seeded/not-seeded) substantially decreases the fit of the model, even when individual effects are included.
+We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, and net points won -- and see if individual differences are detectable.
+
+(some plots that show a) Serena/Federer are stronger servers, b) Federer makes fewer unforced errors at Wimbledon, c) something about Nadal)
+
 # Discussion {#sec:discussion}

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
merge

--- 2018-09-15 contest/surface-type/paper-tennis.Rmd
@@
 ---
-title: "Analyzing the Effect of Surface in Tennis Grand Slams"
-author: "###########"
+title: "Opening up the court (surface) in tennis grand slams"
 date: '`r format(Sys.time(), "%B %d, %Y")`'
@@
              caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
+             booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
 ```
@@
 ## Examining individual players
-```{r table}
+
+```{r table, fig.pos = 'H'}
 data(gs_partial_players)
@@
                caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
-               booktabs = TRUE) %>% kable_styling(latex_options = "striped") %>%
+               booktabs = TRUE) %>% kable_styling(latex_options = "striped", "HOLD_position") %>%
     row_spec(5, bold = TRUE, background = "lightgray")
@@
 
-We fit individual models for these three individuals (subsetting the data to their matches only), estimating the percent of points won in a match using stepwise regression.
+We fit individual linear models for these three individuals (subsetting the data to their matches only), estimating the percent of points won in a match using stepwise regression.
 The lower model is the  percent of points regressed on the opponent rank and indicator variables for court type with the FO as the reference variable.
-The upper model is the lower model with the additional variables of winner to unforced error ratio, average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
+The upper model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
 We do not display the full best fit models here, but they are available online.
+We do report the effects ($\pm 2 \cdot$ Std.
+Err.) of significant variables and their interaction terms, adjusting for the other covariates.
+Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically, AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.
+For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.
+For Williams, we also use forwards-backwards stepwise regression but begin at the low model since she has more missing data.
 
-ADD CIS!!
-
-## Nadal
+## Nadal: King of Clay: confirmed
 
@@
 coefs_n <- coefficients(step_model_n)
+cov_n <- vcov(step_model_n)
 ```
@@
 The standardized residuals plot shown in Figure \ref{fig:individ-resids} (along with the other linear regression diagnostics plots not shown here) shows that the model is a good fit for predicting the percent of points won.
-For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)`$ less at the AO and $`r signif(abs(coefs_n[4]), 2)`$ less at Wim compared to the FO; to increase by $`r signif(abs(coefs_n[5]), 2)`$ for a one unit increase in opponent rank; a `r .01 * signif(abs(coefs_n[7]), 2)`\% increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)`$ for a .01 increase in percent of net points won.
+For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)` \pm `r signif(2 * sqrt(cov_n[2,2]), 2)`$ less at the AO and $`r signif(abs(coefs_n[4]), 2)` \pm `r signif(2 * sqrt(cov_n[4,4]), 2)`$ less at Wim compared to the FO; to increase by $`r signif(abs(coefs_n[5]), 2)`\pm `r signif(2 * sqrt(cov_n[5,5]), 2)`$ for a one unit increase in opponent rank; a $`r .01 * signif(abs(coefs_n[7]), 2)`\pm `r signif(.02 * sqrt(cov_n[7,7]), 2)`$ increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)` \pm `r signif(.02 * sqrt(cov_n[8,8]), 2)`$ for a .01 increase in percent of net points won.
 As such, we see clear evidence Nadal performs better at the FO compared to the other grand slams.
 
-## Federer
+## Federer: Wimbledon extraordinaire: debunked
 ```{r federer-model}
@@
 coefs_f <- coefficients(step_model_f)
+cov_f <- vcov(step_model_f)
 ```
@@
 
-The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.
-For the significant variabless ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)`$ less at AO compared to the Wim.; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`$ less at FO compared to the Wim.; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`$\% less at USO compared to the Wim.;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)`$ for a one unit increase in W/UE at AO, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)`$ for a one unit increase in W/UE at FO, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)`$ for a one unit increase in W/UE at USO.
-It is clear that Federer performs best at Wimbledon.
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fairly good fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.
+For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)` \pm `r signif(2 * sqrt(cov_f[2,2] + cov_f[8,8] + 2 * cov_f[2,8]), 2)`$ less at AO compared to the Wim.; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[3,3] + cov_f[9,9] + 2 * cov_f[3,9]), 2)`$ less at FO compared to the Wim.; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[4,4] + cov_f[9,9] + 2 * cov_f[4,9]), 2)`$ less at USO compared to the Wim.;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[8,8] + 2 * cov_f[6,8]), 2)`$ for a one unit increase in W/UE at AO, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[9,9] + 2 * cov_f[6,9]), 2)`$ for a one unit increase in W/UE at FO, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[10,10] + 2 * cov_f[6,10]), 2)`$ for a one unit increase in W/UE at USO.
+This is strong evidence that Federer performs best at Wimbledon but possibly does better at the USO, depending on his W/UE ratio.
 
-## Williams
+## Williams: hard court magician only: debunked
 ```{r williams-model}
@@
 
+cov_w <- vcov(step_model_w)
 coefs_w <- coefficients(step_model_w)
@@
 The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal.
-For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)`$for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2)`$ for a one unit increase in W/UE at Wim.
-compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)`$ for a .01 increase in percent of break points won at the French Open.
+For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wim.
+compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the French Open.
 

### 2018-09-15 contest/surface-type/paper-tennis.tex
merge

@@
 \hypersetup{unicode=true,
-            pdftitle={Analyzing the Effect of Surface in Tennis Grand Slams},
-            pdfauthor={Kayla Frisoli, Shannon Gallagher, and Amanda Luby},
+            pdftitle={Opening up the court (surface) in tennis grand slams},
             pdfborder={0 0 0},
@@
 
-  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
+  \title{Opening up the court (surface) in tennis grand slams}
     \pretitle{\vspace{\droptitle}\centering\huge}
   \posttitle{\par}
-    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
+    \author{}
+    \preauthor{}\postauthor{}
       \predate{\centering\large\emph}
@@
 
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \hypertarget{sec:iintro}{%
@@
 
-We first examine the distribution of points earned per match. We find
-that the distribution is similar across tournament, with Wimbledon
-differing slightly from the other grand slams. As expected, there are
-more points earned in the WTA than the ATP due to the differing numbers
-of games played. Also unsurprisingly, the winners of the match tended to
-earn more points than the losers.
+We first examine the distribution of points earned per match to assess
+normality. The distribution of points earned is approximately normal.
+This distribution is similar across tournament, with Wimbledon differing
+slightly from the other grand slams. As expected, there are more points
+earned in the WTA than the ATP due to the differing numbers of games
+played. Also unsurprisingly, the winners of the match tended to earn
+more points than the losers.
 
@@
 
-\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-2-1.pdf}
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-2-1} \end{center}
 
@@
 
-\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-3-1.pdf}
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\hypertarget{tall-players-and-aces}{%
+\subsection{Tall players and aces}\label{tall-players-and-aces}}
+
+Why the actual EFF won't fig.width work?
+
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-4-1} \end{center}
 
@@
 
-We fit individual models for these three individuals (subsetting the
-data to their matches only), estimating the percent of points won in a
-match using stepwise regression. The lower model is the percent of
+We fit individual linear models for these three individuals (subsetting
+the data to their matches only), estimating the percent of points won in
+a match using stepwise regression. The lower model is the percent of
 points regressed on the opponent rank and indicator variables for court
 type with the FO as the reference variable. The upper model is the lower
-model with the additional variables of winner to unforced error ratio,
-average serve speed, percent aces, percent break points, percent net
-points won, and their interaction effects with court type. We do not
-display the full best fit models here, but they are available online.
+model with the additional variables of winner to unforced error ratio
+(W/UE), average serve speed, percent aces, percent break points, percent
+net points won, and their interaction effects with court type. We do not
+display the full best fit models here, but they are available online. We
+do report the effects (\(\pm 2 \cdot\) Std. Err.) of significant
+variables and their interaction terms, adjusting for the other
+covariates. Since we postulated that Nadal is best on clay, Federer on
+grass, and Williams on hard court, specificically, AO, we use the FO,
+Wim., and AO, respectively, as the reference variable in the model. For
+Nadal and Federer we use forward-backwards stepwise regression, using
+AIC as the criterion, beginning with the full model. For Williams, we
+also use forwards-backwards stepwise regression but begin at the low
+model since she has more missing data.
 
-ADD CIS!!
-
-\hypertarget{nadal}{%
-\subsection{Nadal}\label{nadal}}
+\hypertarget{nadal-king-of-clay-confirmed}{%
+\subsection{Nadal: King of Clay:
+confirmed}\label{nadal-king-of-clay-confirmed}}
 
@@
 adjusting for the other variables, we expect percent of points won to be
-\(0.034\) less at the AO and \(0.046\) less at Wim compared to the FO;
-to increase by \(1.5\times 10^{-4}\) for a one unit increase in opponent
-rank; a 0.0012\% increase for a .01 increase in percent of break points
-won; and to decrease by \(8.8\times 10^{-4}\) for a .01 increase in
+\(0.034 \pm 0.028\) less at the AO and \(0.046 \pm 0.033\) less at Wim
+compared to the FO; to increase by
+\(1.5\times 10^{-4}\pm 1.3\times 10^{-4}\) for a one unit increase in
+opponent rank; a \(0.0012\pm 5.6\times 10^{-4}\) increase for a .01
+increase in percent of break points won; and to decrease by
+\(8.8\times 10^{-4} \pm 5.5\times 10^{-4}\) for a .01 increase in
 percent of net points won. As such, we see clear evidence Nadal performs
@@
 
-\hypertarget{federer}{%
-\subsection{Federer}\label{federer}}
+\hypertarget{federer-wimbledon-extraordinaire-debunked}{%
+\subsection{Federer: Wimbledon extraordinaire:
+debunked}\label{federer-wimbledon-extraordinaire-debunked}}
 
@@
 The standardized residuals plot shown in Figure \ref{fig:individ-resids}
-shows that the model is a fair fit for predicting the percent of points
-won, but almost seem to see two clusters of residuals splitting with
-predicted point percentage of about 0.63. For the significant variabless
-(\(\alpha\) = 95\%), while adjusting for the other variables, we expect
-percent of points won to be \(0.04\) less at AO compared to the Wim.;
-\(0.058\) less at FO compared to the Wim.; \(0.007\)\% less at USO
-compared to the Wim.; to decrease by \(0.022\) for a one unit increase
-in W/UE at AO, \(0.073\) for a one unit increase in W/UE at FO, and
-\(0.041\) for a one unit increase in W/UE at USO. It is clear that
-Federer performs best at Wimbledon.
+shows that the model is a fairly good fit for predicting the percent of
+points won, but almost seem to see two clusters of residuals splitting
+with predicted point percentage of about 0.63. For the significant
+variables (\(\alpha\) = 95\%), while adjusting for the other variables,
+we expect percent of points won to be \(0.04 \pm 0.034\) less at AO
+compared to the Wim.; \(0.058\pm 0.046\) less at FO compared to the
+Wim.; \(0.007\pm 0.093\) less at USO compared to the Wim.; to decrease
+by \(0.022 \pm 0.019\) for a one unit increase in W/UE at AO,
+\(0.073 \pm 0.066\) for a one unit increase in W/UE at FO, and
+\(0.041 \pm 0.033\) for a one unit increase in W/UE at USO. This is
+strong evidence that Federer performs best at Wimbledon but possibly
+does better at the USO, depending on his W/UE ratio.
 
-\hypertarget{williams}{%
-\subsection{Williams}\label{williams}}
+\hypertarget{williams-hard-court-magician-only-debunked}{%
+\subsection{Williams: hard court magician only:
+debunked}\label{williams-hard-court-magician-only-debunked}}
 
@@
 won, but we tend to overestimate the percent of points won in comparison
-with Nadal. For the significant variables (\(\alpha\) = 95\%), while
+with Nadal. For the significant variables (\(\alpha\) = 95\(\%\)), while
 adjusting for the other variables, we expect percent of points won to
-increase by \(4.8\times 10^{-4}\) for a 1 unit increase in oponent rank;
-to increase by \(0.034\) for a one unit increase in W/UE at FO, to
-increase by \(0.092\)for a one unit increase in W/UE at USO, and to
-increase by \(0.0038\) for a one unit increase in W/UE at Wim. compared
-to AO; to increase by \(2\times 10^{-4}\) for a .01 increase in percent
-of aces at the French Open; and to increase by \(0.0024\) for a .01
-increase in percent of break points won at the French Open.
+increase by \(4.8\times 10^{-4} \pm 3.2\times 10^{-4}\) for a 1 unit
+increase in oponent rank; to increase by \(0.034 \pm 0.02\) for a one
+unit increase in W/UE at FO, to increase by \(0.092 \pm 0.045\) for a
+one unit increase in W/UE at USO, and to increase by \(0.0038\pm 0.066\)
+for a one unit increase in W/UE at Wim. compared to AO; to increase by
+\(2\times 10^{-4}\pm 1.6\times 10^{-4}\) for a .01 increase in percent
+of aces at the French Open; and to increase by \(0.0024 \pm 0.0011\) for
+a .01 increase in percent of break points won at the French Open.
 

### 2018-09-15 contest/surface-type/paper-tennis.Rmd

@@
 ## Federer: Wimbledon extraordinaire: debunked
+
 ```{r federer-model}
@@
 ## Williams: hard court magician only: debunked
+
 ```{r williams-model}

### 2018-09-15 contest/surface-type/paper-tennis.tex
hopefulyl this gets overwritten

--- 2018-09-13 contest/surface-type/paper-tennis.tex
@@
             pdftitle={Analyzing the Effect of Surface in Tennis Grand Slams},
-            pdfauthor={Kayla Frisoli, Shannon Gallagher, and Amanda Luby},
             pdfborder={0 0 0},
@@
   \posttitle{\par}
-    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
+    \author{}
+    \preauthor{}\postauthor{}
       \predate{\centering\large\emph}
   \postdate{\par}
-    \date{September 12, 2018}
+    \date{September 14, 2018}
 
@@
 tennis matches (Klaassen and Magnus 2003; Newton and Keller 2005; McHale
-and Morton 2011; Kovalchik 2016){]} or for assessing whether points
-within a match are independent and identically distributed (Klaassen and
-Magnus 2001). (Knottenbelt, Spanias, and Madurska 2012) do take into
-account surface in their model but do not compare the results of one
-surface to another. Other sports analyses do take into account surface
-type such as grass vs.~turf in soccer and football. Results from these
-studies show that surface type does have an effect on the game, either
-directly or indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains
-et al. (2010);{]}.
+and Morton 2011; S. A. Kovalchik 2016){]} or for assessing whether
+points within a match are independent and identically distributed
+(Klaassen and Magnus 2001). (Knottenbelt, Spanias, and Madurska 2012) do
+take into account surface in their model but do not compare the results
+of one surface to another. Other sports analyses do take into account
+surface type such as grass vs.~turf in soccer and football. Results from
+these studies show that surface type does have an effect on the game,
+either directly or indirectly {[}Andersson, Ekblom, and Krustrup (2008);
+Gains et al. (2010);{]}.
 
@@
 of origin, and player attributes. Additionally, we take into account the
-number of sets in a match, the surface type, minutes played, and round
-of the tournament. A subset of the data is shown in Table
-\ref{tab:data}.
+number of sets in a match, the surface type, and round of the
+tournament. A subset of the data is shown in Table \ref{tab:data}.
 
 \rowcolors{2}{gray!6}{white}
+
 \begin{table}
@@
 The data is obtained from Jeff Sackmann's open website via the
-\texttt{R} package \texttt{deuce} (Sackmann 2018; Kovalchik 2017). All
-steps of our analysis from collection to dissemination are freely
+\texttt{R} package \texttt{deuce} (Sackmann 2018; S. Kovalchik 2017).
+All steps of our analysis from collection to dissemination are freely
 available online.
 
-\hypertarget{sec:eda}{%
-\subsection{Early Data Analysis}\label{sec:eda}}
+\hypertarget{sec:eda}{\subsection{Early Data Analysis}\label{sec:eda}}
 
-\hypertarget{sec:methods}{%
-\section{Methods}\label{sec:methods}}
+\subsubsection{Examining the distribution of points
+earned}\label{examining-the-distribution-of-points-earned}
 
-\hypertarget{sec:results}{%
-\section{Results}\label{sec:results}}
+We first examine the distribution of points earned per match to assess
+normality. The distribution of points earned is approximately normal.
+This distribution is similar across tournament, with Wimbledon differing
+slightly from the other grand slams. As expected, there are more points
+earned in the WTA than the ATP due to the differing numbers of games
+played. Also unsurprisingly, the winners of the match tended to earn
+more points than the losers.
 
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
+\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-1-1.pdf}
 
-\hypertarget{sec:refs}{%
-\section{References}\label{sec:refs}}
+\subsubsection{Home court advantage}\label{home-court-advantage}
+
+It is commonly thought that there is a home court advantage in grand
+slam games (SOURCE). In our data we find this to be true (i.e.~French
+players win the French open more than French players win other slams).
+But, we also know that the home team is given preference for wild card
+bids (SOURCE) so potentially citizens of a particular country play in
+``their'' tournament more often than they play in other tournaments. We
+also find this to be true in our data for France, The United States, and
+Australia (i.e.~the proportion of French players in the French Open is
+greater than the proportion of French players in other slams).
+
+Therefore, we want to see how the proportion of wins for the home
+country changes across the different tournaments. If there was really a
+home court advantage, the proportion of French wins each round would be
+higher at the French Open than the Australian Open, the US Open, and
+Wimbledon. The same would be true for Australia and the US. But, we see
+that this isn't the case. After accounting for the number of players
+from each country, we don't find a home court advantage in the grand
+slam.
+
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
+\subsection{Spaniards on clay}\label{spaniards-on-clay}
+
+It is also commonly thought that Spaniards play better on clay. We are
+interested in whether Spaniards win the French Open more than they win
+in other tournaments. It does appear that Spaniards are winning the
+French Open more than they are winning other tournaments. But, this
+result is not significant. In addition, the ranking of Spaniards in the
+French Open is, on average, higher than other tournaments, which may
+help explain this common misconception, although this result is not
+significant either.
+
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\subsection{Tall players and aces}\label{tall-players-and-aces}
+
+Why the actual EFF won't fig.width work?
+
+\begin{center}\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\hypertarget{sec:methods}{\section{Methods}\label{sec:methods}}
+
+\hypertarget{sec:results}{\section{Results}\label{sec:results}}
+
+\section{Discussion}\label{sec:discussion}
+
+\section{References}\label{sec:refs}
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-andersson2008}{}%
+\hypertarget{ref-andersson2008}{}
 Andersson, Helena, Björn Ekblom, and Peter Krustrup. 2008. ``Elite
@@
 
-\leavevmode\hypertarget{ref-egidi2018}{}%
+\hypertarget{ref-egidi2018}{}
 Egidi, Leonardo, and Jonah Gabry. 2018. ``Bayesian Hierarchical Models
@@
 
-\leavevmode\hypertarget{ref-gains2010}{}%
+\hypertarget{ref-gains2010}{}
 Gains, Graydon L, Andy N Swedenhjelm, Jerry L Mayhew, H Michael Bird,
@@
 
-\leavevmode\hypertarget{ref-bbc2018}{}%
+\hypertarget{ref-bbc2018}{}
 Jurejko, Jonathan. 2018. ``French Open 2018: Why does 'King of Clay'
@@
 
-\leavevmode\hypertarget{ref-klaassen2003}{}%
-Klaassen, Franc J.G.M., and Jan R. Magnus. 2003. ``Forecasting the
-Winner of a Tennis Match.'' \emph{European Journal of Operational
-Research} 148 (2): 257--67.
-\url{https://doi.org/https://doi.org/10.1016/S0377-2217(02)00682-3}.
-
-\leavevmode\hypertarget{ref-klaassen2001}{}%
+\hypertarget{ref-klaassen2001}{}
 Klaassen, Franc J.G.M., and Jan R Magnus. 2001. ``Are Points in Tennis
@@
 Association} 96 (454). Taylor \& Francis: 500--509.
-\url{https://doi.org/10.1198/016214501753168217}.
+doi:\href{https://doi.org/10.1198/016214501753168217}{10.1198/016214501753168217}.
 
-\leavevmode\hypertarget{ref-knottenbelt2012}{}%
+\hypertarget{ref-klaassen2003}{}
+Klaassen, Franc J.G.M., and Jan R. Magnus. 2003. ``Forecasting the
+Winner of a Tennis Match.'' \emph{European Journal of Operational
+Research} 148 (2): 257--67.
+doi:\href{https://doi.org/https://doi.org/10.1016/S0377-2217(02)00682-3}{https://doi.org/10.1016/S0377-2217(02)00682-3}.
+
+\hypertarget{ref-knottenbelt2012}{}
 Knottenbelt, William J., Demetris Spanias, and Agnieszka M. Madurska.
@@
 Applications} 64 (12): 3820--7.
-\url{https://doi.org/https://doi.org/10.1016/j.camwa.2012.03.005}.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.camwa.2012.03.005}{https://doi.org/10.1016/j.camwa.2012.03.005}.
 
-\leavevmode\hypertarget{ref-deuce2017}{}%
+\hypertarget{ref-deuce2017}{}
 Kovalchik, Stephanie. 2017. \emph{Deuce: Resources for Analysis of
@@
 
-\leavevmode\hypertarget{ref-kovalchik2016}{}%
+\hypertarget{ref-kovalchik2016}{}
 Kovalchik, Stephanie Ann. 2016. ``Searching for the Goat of Tennis Win
@@
 
-\leavevmode\hypertarget{ref-leitner2009}{}%
+\hypertarget{ref-leitner2009}{}
 Leitner, Christoph, Achim Zeileis, and Kurt Hornik. 2009. ``Is Federer
@@
 
-\leavevmode\hypertarget{ref-mchale2011}{}%
+\hypertarget{ref-mchale2011}{}
 McHale, Ian, and Alex Morton. 2011. ``A Bradley-Terry Type Model for
@@
 Forecasting} 27 (2): 619--30.
-\url{https://doi.org/https://doi.org/10.1016/j.ijforecast.2010.04.004}.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.ijforecast.2010.04.004}{https://doi.org/10.1016/j.ijforecast.2010.04.004}.
 
-\leavevmode\hypertarget{ref-five2015}{}%
+\hypertarget{ref-five2015}{}
 Morris, Benjamin. 2015. ``Serena Williams and the Difference Between
@@
 
-\leavevmode\hypertarget{ref-newton2005}{}%
+\hypertarget{ref-newton2005}{}
 Newton, Paul K, and Joseph B Keller. 2005. ``Probability of Winning at
@@
 
-\leavevmode\hypertarget{ref-page2013}{}%
+\hypertarget{ref-page2013}{}
 Page, Garritt L, Bradley J Barney, and Aaron T McGuire. 2013. ``Effect
@@
 
-\leavevmode\hypertarget{ref-theage2007}{}%
+\hypertarget{ref-theage2007}{}
 Paxinos, Stathi. 2007. ``Australian Open court surface is speeding up.''
@@
 
-\leavevmode\hypertarget{ref-sackmann2018}{}%
+\hypertarget{ref-sackmann2018}{}
 Sackmann, Jeff. 2018. ``Tennis Data Repositories.''
@@
 
-\leavevmode\hypertarget{ref-thomas2013}{}%
+\hypertarget{ref-thomas2013}{}
 Thomas, A. C., Samuel L. Ventura, Shane T. Jensen, and Stephen Ma. 2013.

### 2018-09-15 contest/model-fitting/model-fitting.Rmd

@@
 Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
-It also provides a way to examine individual player effects and assess, for instance, whether Nadal is more likely to win at Wimbledon after accounting for other variables.
+It also provides a way to examine individual player effects and assess, for instance, whether Federer is more likely to win at Wimbledon after accounting for other variables.
 We considered the following models:
@@
 
-plot_grid(aces_ranef, winners_ranef, nets_ranef, ue_ranef)
+aces_ranef
+winners_ranef
+nets_ranef
+ue_ranef
 ```

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
most recent paper edits

--- 2018-09-15 contest/surface-type/paper-tennis.Rmd
@@
 library(ggrepel)
+library(broom)
+library(lme4)
+library(cowplot)
 
@@
 
+## Mixed-effects models for all grand slams
+
+We use a mixed-effects modeling approach to examine the grand slam results from all players from 2013-2017.
+This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects.
+Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
+It also provides a way to examine individual player effects.
+
+### Logistic regression for wins
+
+We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance.
+For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match.
+Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
+For these reasons, we chose to model whether a player won the match using a mixed-effects logistic regression approach, with the following predictors:
+
+* `ioc_fac`: Country that the player represents
+* `tournament`: Which Grand Slam the match was played at
+* `late_round`: Indicator for whether the match occured in Round of 16 or later
+* `rank`: Rank of player at the time of the match
+    + Included on log scale 
+* `opponent_rank`: Rank of opponent at the time of the match
+    + Included on log scale
+* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
+* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.
+
+We considered the following models:
+
+* No mixed effects
+    + `nocourt_logistic`: $\text{logit}(\pi_i) = \beta X$, $X$ does not include `tournament`
+        + If this model fits well, court surface does not have an effect on winning probability
+    + `base_logistic`: $\text{logit}(\pi_i) = \beta X$
+        + If this model fits better than the mixed-effects models, performance does not depend on individual-level or IOC-level 
+* IOC-level effect
+    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i C_i$
+* Player-level effect
+    + `ind_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i P_i$
+        * Includes both a fixed-effect term for IOC and a player-level effect for each grand slam
+    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC and $T$ represents indicators for tournament, $\beta_{T[i]} = \alpha_i P_i$
+        * Simpler than the `ind_logistic` model, since we exclude IOC effects.
+    + `ind_int_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{0[i]}$, where $X$ does not include IOC, $\beta_{0[i]} = \alpha_i$
+        * Similar to `ind_logistic_noioc`, but includes a single player-level effect rather than a player-level effect for each grand slam. 
+    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{D[i]}$, where $X$ does not include IOC and $D$ represents indicators for years, $\beta_{Y[i]} = \alpha_i Y_i$
+        * Rather than including a player-level effect for grand slam, includes a player-level effect for each year in the dataset
+
+```{r}
+load("../../data/gs_players.rda")
+load("../../data/gs_partial_players.rda")
+load("../../data/nocourt_logistic.rda")
+load("../../data/base_logistic.rda")
+load("../../data/country_logistic.rda")
+load("../../data/ind_int_logistic.rda")
+load("../../data/ind_logistic.rda")
+load("../../data/ind_logistic_noioc.rda")
+load("../../data/ind_year_logistic.rda")
+```
+
+Based on AIC, the `ind_logistic_noioc` model performed best after accounting for the number of variables in the model.
+
+```{r, warning = FALSE}
+model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
+model.ests = data.frame()
+ran.effects = data.frame()
+for(mod in 1:nrow(model.sums)){
+  name = model_names[mod]
+  model.sums$aic[mod] = extractAIC(get(name))[2]
+  model.sums$edf[mod] = extractAIC(get(name))[1]
+  betas = tidy(get(name))
+  betas$model = name
+  if("group" %in% colnames(betas)){
+  model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+    dplyr::select(., -c("group")))
+  }
+  else model.ests = rbind(model.ests, betas)
+}
+
+knitr::kable(model.sums, format = "latex",
+             row.names = FALSE, col.names = c("Model", "AIC", "EDF"),
+             caption = "\\label{model.sums}AIC summary of the seven logistic regression models fitted.
+Including individual effects for each grand slam, without including any country effects, leads to the best model fit according to AIC.",
+             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
+```
+
+If we examine the fixed effects for all seven models, and the random effects for the `ind_logistic` and `ind_logistic_noioc`, we see that the model structure chosen does not substantially change the parameter estimates.
+Across all models, `rank` and `opponent_rank` appear to best explain the probability of winning the match.
+Other fixed effects in the models, like whether it was a late round in the grand slam, the year, or ATP/WTF, were not found to be significantly different than zero.
+
+```{r}
+fixed.effects.plot = model.ests %>%
+  filter(., !grepl("ioc", term)) %>%
+  ggplot(., aes(x = term, y = exp(estimate), col = model, ymin = exp(estimate - 2*std.error), ymax = exp(estimate + 2*std.error))) +
+  geom_errorbar(width = .7, position = position_dodge()) +
+  theme_minimal() + 
+  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
+  scale_color_brewer(palette = "Dark2") + 
+  ggtitle("Fixed effects") +
+  theme(legend.position = "bottom")
+
+ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
+ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
+
+random.effects.plot = ran_tourn %>%
+  gather(., tourn, estimate, -model) %>%
+  mutate(., tournament = gsub("tournament", "", tourn)) %>%
+  ggplot(., aes(x = tournament, y = exp(estimate), fill = model)) +
+  geom_violin() + 
+  theme_minimal() +
+  scale_fill_brewer(palette = "Dark2") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  ggtitle("Random effects") +
+  theme(legend.position = "bottom")
+
+plot_grid(fixed.effects.plot, random.effects.plot)
+```
+
+Looking at the `ind_logistic_noioc` random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
+
+```{r}
+knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
+             row.names = TRUE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
+             caption = "Correlation matrix for random effects",
+             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
+```
+
+If we look at our big three, their individual effects are not very informative (especially after taking uncertainty into consideration).
+
+```{r}
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
+name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
+as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
+  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
+  ggplot(., aes(x = name, y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, color = tournament)) +
+  geom_errorbar(width = .5, position = "dodge") + 
+  xlab("") + 
+  ylab("Random Effect") + 
+  scale_color_brewer(palette = "Dark2") +
+  coord_flip()
+```
+
+### Linear regression for other stats
+
+Rank and opponent's rank clearly have the largest effect on predicting the winner.
+Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
+We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, net points won, and unforced errors -- and see if individual differences are detectable.
+
+(some plots that show a) Serena/Federer are stronger servers, b) Federer makes fewer unforced errors at Wimbledon, c) something about Nadal)
+
 ## Examining individual players
@@
 
-## Mixed-effects models for all grand slams
-
-### Logistic regression for wins
-
-We will use a multilevel logistic regression framework.
-This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects.
-Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
-It also provides a way to examine individual player effects and assess, for instance, whether Federer is more likely to win at Wimbledon after accounting for other variables.
-
-We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance.
-For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match.
-Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
-For these reasons, we chose to model whether a player won the match, with the following predictors:
-
-* `ioc_fac`: Country that the player represents
-* `tournament`: Which Grand Slam the match was played at
-* `late_round`: Indicator for whether the match occured in Round of 16 or later
-* `rank`: Rank of player at the time of the match
-    + Included on log scale 
-* `opponent_rank`: Rank of opponent at the time of the match
-    + Included on log scale
-* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
-* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.
-
-(short discussion of models fit, show plots for final model)
-
-### Linear regression for other stats
-
-Rank and opponent's rank clearly have the largest effect on predicting the winner.
-Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
-Including rank as an indicator (seeded/not-seeded) substantially decreases the fit of the model, even when individual effects are included.
-We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, and net points won -- and see if individual differences are detectable.
-
-(some plots that show a) Serena/Federer are stronger servers, b) Federer makes fewer unforced errors at Wimbledon, c) something about Nadal)
-
 # Discussion {#sec:discussion}

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
merge

--- 2018-09-15 contest/surface-type/paper-tennis.Rmd
@@
 ## Data {#sec:data}
-The data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
+
+The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
 Each match has `r ncol(gs)` attributes, many of which are redundant.
@@
 
-The data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.
+The secondary data point by point data for grand slam matches.
+In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.
+There is also tournament info such as court, year, and time start.
+Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.
+However, these additional attributes are only available for about 1/10 of the data.
+
+We aggregate the point by point data into match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
+We then join the point by point match data with the grand slam data to find the score of the match.
+We do this by matching on tournament, year, player one, and player two.
+This final data set consists of 3066 observations (compared to the 5080 in the original grand slams).
+The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)` and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
+The same trend is true for the loser rank.
+This means that it is more likely for better players to have court by court data, which may effect our analysis.
+Additionally, we found that no women's point by point data was recorded for 2015.
+
+All data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.
 

### 2018-09-15 contest/surface-type/paper-tennis.tex
merge

--- 2018-09-15 contest/surface-type/paper-tennis.tex
@@
 
-The data consists of 5080 matches split evenly over the four grand slams
-and the two leagues: ATP (men's) and WTA (women's). Each match has 80
-attributes, many of which are redundant. We focus on the following
-attributes for both the winner and loser of the match: games won, points
-won, retirement, break points faced, break points saved, aces, country
-of origin, and player attributes. Additionally, we take into account the
-number of sets in a match, the surface type, and round of the
-tournament. A subset of the data is shown in Table \ref{tab:data}.
+The primary data consists of 5080 matches split evenly over the four
+grand slams and the two leagues: ATP (men's) and WTA (women's). Each
+match has 80 attributes, many of which are redundant. We focus on the
+following attributes for both the winner and loser of the match: games
+won, points won, retirement, break points faced, break points saved,
+aces, country of origin, and player attributes. Additionally, we take
+into account the number of sets in a match, the surface type, and round
+of the tournament. A subset of the data is shown in Table
+\ref{tab:data}.
 
@@
 
-The data is obtained from Jeff Sackmann's open website via the
+The secondary data point by point data for grand slam matches. In this
+data set, each row is a point in a match with details on who won the
+point, whether a player had a forced or unforced error, winner, ace, or
+net point win, and serve speed. There is also tournament info such as
+court, year, and time start. Additional attributes include rally length,
+winner and final score of the match, retirement, and minutes played.
+However, these additional attributes are only available for about 1/10
+of the data.
+
+We aggregate the point by point data into match data, summing the total
+number of errors, winners, aces, net points, and taking the average
+service speed for each player in the match. We then join the point by
+point match data with the grand slam data to find the score of the
+match. We do this by matching on tournament, year, player one, and
+player two. This final data set consists of 3066 observations (compared
+to the 5080 in the original grand slams). The median rank of the winners
+for the original data is 28 and the median rank for the winners of the
+partial data is 21. The same trend is true for the loser rank. This
+means that it is more likely for better players to have court by court
+data, which may effect our analysis. Additionally, we found that no
+women's point by point data was recorded for 2015.
+
+All data is obtained from Jeff Sackmann's open website via the
 \texttt{R} package \texttt{deuce} (Sackmann 2018; Kovalchik 2017). All
@@
 
+\hypertarget{mixed-effects-models-for-all-grand-slams}{%
+\subsection{Mixed-effects models for all grand
+slams}\label{mixed-effects-models-for-all-grand-slams}}
+
+\hypertarget{logistic-regression-for-wins}{%
+\subsubsection{Logistic regression for
+wins}\label{logistic-regression-for-wins}}
+
+We will use a multilevel logistic regression framework. This allows us
+to include fixed effects, which remain the same for each player, as well
+as include player-level effects. Since players appear multiple times in
+the data, the observations are not independent and including a
+player-level effect allows us to account for this dependence. It also
+provides a way to examine individual player effects and assess, for
+instance, whether Federer is more likely to win at Wimbledon after
+accounting for other variables.
+
+We initially planned to use points won by an individual as our outcome
+variable, but found that it may not be an ideal measure of player
+performance. For instance, players may score few points in a match due
+to a poor performance, but they may also score few (relative) points if
+they win a short match. Modeling the number of points won is also
+complicated by the difference between length of matches for men (best of
+5 sets) and women (best of 3 sets). For these reasons, we chose to model
+whether a player won the match, with the following predictors:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{ioc\_fac}: Country that the player represents
+\item
+  \texttt{tournament}: Which Grand Slam the match was played at
+\item
+  \texttt{late\_round}: Indicator for whether the match occured in Round
+  of 16 or later
+\item
+  \texttt{rank}: Rank of player at the time of the match
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Included on log scale
+  \end{itemize}
+\item
+  \texttt{opponent\_rank}: Rank of opponent at the time of the match
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Included on log scale
+  \end{itemize}
+\item
+  \texttt{year}: Factor variable with a level for each year included in
+  the dataset (2013-2017)
+\item
+  \texttt{atp}: Indicator that the match was played in the ATP league
+  instead of the WTF league.
+\end{itemize}
+
+(short discussion of models fit, show plots for final model)
+
+\hypertarget{linear-regression-for-other-stats}{%
+\subsubsection{Linear regression for other
+stats}\label{linear-regression-for-other-stats}}
+
+Rank and opponent's rank clearly have the largest effect on predicting
+the winner. Looking at the individual effects for Serena, Nadal, and
+Federer; they actually fall near the average for all four grand slams.
+Including rank as an indicator (seeded/not-seeded) substantially
+decreases the fit of the model, even when individual effects are
+included. We now look at a similar model, but instead of using `win' as
+the outcome variable, we'll look at game statistics that may be more
+sensitive to court surface -- aces, winners, and net points won -- and
+see if individual differences are detectable.
+
+(some plots that show a) Serena/Federer are stronger servers, b) Federer
+makes fewer unforced errors at Wimbledon, c) something about Nadal)
+
 \hypertarget{sec:discussion}{%

### 2018-09-15 contest/model-fitting/fit-models-and-save.R
amanda changes

@@ -1,5 +1,15 @@
 load("gs_players.rda")
 load("gs_partial_players.rda")
+
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+gs_players$hand_fac = as.factor(gs_players$hand)
+gs_players$ioc_fac = as.factor(gs_players$ioc)
+gs_players$atp = gs_players$league == "ATP"
+gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
+
 set.seed(091418)
 nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
 base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
Rewrote intro, deleted abstract, and touched up some stuff

@@
     number_sections: yes
-    toc: true
 header-includes:
   - \usepackage{hyperref}
-abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.
-The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.
-It is also believed that play style and practice habits may contribute to different results across surface types.
-For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.
-On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.
-In this analysis, we examine the result of the top 100 players in grand slams from 2013-2017 across the four different surfaces.
-We create a hierarchical model with fixed and random effects to predict the number of points won in a match.
-We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.
-We assess the fit of our model using standard statistical techniques (e.g.
-MSE, AIC, BIC, residual diagnostics) in addition to `common knowledge' factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).
-We compare the results of top 100 players across grand slams to examine the effect of court surface.
-We also provide an in-depth analysis of Nadal, Federer, and S.
-Williams."
+abstract: "IDK writing is hard"
 bibliography: master.bib
@@
 knitr::opts_chunk$set(message = FALSE)
-knitr::opts_chunk$set(cache = TRUE)
 library(deuce)
@@
 
-# Introduction {#sec:iintro}
-Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open, which is played on a clay surface [@bbc2018].
-In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon, which is played on grass.
-On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
-This trend extends to other top players, who seem to have better results at some grand slams than others.
-More broadly, it seems that country of origin has an interaction effect with court type.
-For example, Spanish players seem to excel on clay courts and Americans have great success at Wimbledon despite grass courts not being of wide use in the USA.
-It also worth questioning whether the US Open and Australian Open should be grouped together as hard courts despite having different surface compositions [@theage2007].
-In this paper, we analyze the results of grand slam players from 2013-2017,  and we
+```{r modelling, eval = FALSE}
+## HIERARCHICAL MODELLING
+## NOT RUN
+## WARNING: Takes about 1 HOUR
 
-1. Determine if and how court surface effects players by implementation of a series of nested hierarchical models
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+gs_players$hand_fac = as.factor(gs_players$hand)
+gs_players$ioc_fac = as.factor(gs_players$ioc)
+gs_players$atp = gs_players$league == "ATP"
+gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
 
-2. Examine how Nadal, Federer, and Williams' play differs by surface
+set.seed(091418)
+nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
+country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
+ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
+ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
+ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
 
-3. Assess whether we can group the two hard court surfaces together.
+```
 
-As to issue (1) quantifying the effect of court surface on players, there has not been much written about with regards to tennis.
+# Introduction {#sec:intro}
+
+Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open (FO), which is played on a clay surface [@bbc2018].
+In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon (Wim.), which is played on grass.
+On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open (AO) and 6 at US Open (USO)) and grass (7 at Wim.).
+These three players have different playstyles and origins, and they are legends in the sport of tennis.
+However, their success at grand slams seems to differ among the tournaments, which leads to question why that may be the case.
+
+The four grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different starting dates (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
+As a feature unique to tennis, all four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
+It is often thought that top players achieve better results on some surfaces than others.
+For example, there is the "home court advantage:" do players from the country of the grand slam tournament perform better than at other grand slams?
+Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
+
+ In this paper, we analyze the results of grand slam players from 2013-2017, and we analyze the surface effect on players at the four grand slams, first broadly at a high level and then narrowing down to player effects.
+More specifically,
+
+1. Determine if results in terms of wins/losses and point distribution differ across the four surfaces
+
+2. Examine more fine grained attributes such as winners, aces, and unforced errors across the four surfaces
+
+3. Assess how individuals differ across the four surfaces, with special focus on Nadal, Federer, and Williams.
+
+As to issue (1) quantifying the high-level effect of court surface on players, there has not been much written about with regards to tennis.
 There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another.
+[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
 Other sports analyses do take into account surface type such as grass vs.
@@
 
+For issue (2) we examine the match attributes more unique to tennis such as winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
+Specifically, serving is thought to be of greater importance at Wimbledon since the ball travels faster on grass.
+In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].
+
+Finally, for issue (3) the player analyses, we take two approaches.
+On one hand, we use a hierarchical model with fixed and random effects taking into account player and tournament attributes using the complete data set and to fairly compare players to one another.
+This allows us to leverage tournament information to players without many matches played and to help account for noise.
+On the other hand, since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
+In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
+We also examine whether these players do have surface apparent effects.
+
 We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
-Both of those models had success using hierarchical Bayesian models, which we employ in our own models.
-More specifically, we model the players' expected points in a match based on the player's own characteristics, the court/tournament effects, and the opponent's ranking.
+For our hierarchical model across all players, we model whether a player or lost a match based on player attributes along with opponent rank.
+For the individual models, we instead model percent of points won which can better show how dominant a player is in a match.
+For model selection, we use AIC to compare models from one another [@wasserman2004].
+For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.
+However, when the data becomes available, we would like to test these models on the 2018 grand slam results.
 
-For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests  like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
-We also examine whether these players do have surface apparent effects.
 Few academic papers have been written about Nadal, Federer, or Williams.
-One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon showed that Federer was too heavily favored by bookmakers [@leitner2009].
+One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers [@leitner2009].
 One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].
-
-Finally, for issue (3), we use clustering methods in order to determine which court surface types are more similar to one another.
 
@@
 In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.
-In [Section Methods](#sec:methods) we describe our hierarchical models we use to determine difference in court surfaces.
+In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.
 In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
-Finally in [Section Discussion](#discussion), we discuss future work and extensions or our model.
+Finally in [Section Discussion](#discussion), we discuss future work and extensions or our models.
 
@@
 The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
+There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two, leagues, and 7 rounds.
 Each match has `r ncol(gs)` attributes, many of which are redundant.
@@
              row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
-             caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
+             caption = "\\label{tab:data}Example of the complete grand slam data.
+It includes winner and loser attributes, match attributes, and tournament attributes.
+Not all attributes are shown here.",
              booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
@@
 
-The secondary data point by point data for grand slam matches.
+The secondary data is partial point by point data for grand slam matches.
 In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.
-There is also tournament info such as court, year, and time start.
+There is also tournament info such as surface, year, and time start.
 Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.
-However, these additional attributes are only available for about 1/10 of the data.
+However, these additional attributes are only available for about 1/8 of the data.
 
-We aggregate the point by point data into match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
-We then join the point by point match data with the grand slam data to find the score of the match.
-We do this by matching on tournament, year, player one, and player two.
-This final data set consists of 3066 observations (compared to the 5080 in the original grand slams).
-The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)` and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
+We aggregate the partial point by point data into partial match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
+We then join the partial match data with the complete data to find the score and outcome of the match.
+We do this by joining the features of tournament, year, player one, and player two.
+This final data set consists of `r nrow(gs_pbp)` observations (compared to the 5080 in the original grand slams).
+The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
 The same trend is true for the loser rank.
-This means that it is more likely for better players to have court by court data, which may effect our analysis.
+This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.
 Additionally, we found that no women's point by point data was recorded for 2015.
@@
 
-## Examining individual players
+## Modelling players individually
 
@@
 
-We fit individual linear models for these three individuals (subsetting the data to their matches only), estimating the percent of points won in a match using stepwise regression.
-The lower model is the  percent of points regressed on the opponent rank and indicator variables for court type with the FO as the reference variable.
+We fit individual linear models for these three individuals (subsetting the partial data to their matches only), estimating the percent of points won in a match using forward-backwards stepwise regression.
+The lower model is the  percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
+$$
+E[Y_{\% pts. won}|X] = \beta_0 + \beta_1 X_{opp. rank} + \beta_{2, AO}\mathcal{I}(X_{sur.} = AO)  + \beta_{2, Wim.}\mathcal{I}(X_{sur.} = Wim.) + \beta_{2, USO}\mathcal{I}(X_{sur.} = USO)  .
+$$  
 The upper model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
@@
 
-Looking at Nadal individually, the model with the lowest AIC has positive coefficients opponent rank, percent of aces, and percent of break points won.
+Looking at Nadal individually, the model with the lowest AIC has positive coefficients for opponent rank, percent of aces, and percent of break points won.
 It has negative coefficients for AO, USO, and Wim.
@@
 
-## Federer: Wimbledon extraordinaire: debunked
+## Federer: Wimbledon extraordinaire: confirmed
 
@@
 
-For Federer, the model with the lowest AIC has negative coefficients AO, FO, and USO.
+For Federer, the model with the lowest AIC has negative coefficients for AO, FO, and USO.
 vs Wim., and W/UE,  It has positive coefficients for opponent rank, percent of break points won, and all the interaction terms: court and W/UE.
@@
 
-For Williams, the model with the lowest AIC is the largest model of the three selected 22 coefficients.
-Since Williams only has 59 observations, we believe this model is overfitting and so caution against inference with this model.
-It should also be noted that no women's point by point data was recorded for 2015 so that year is excluded from this model.
-The coefficients are court, opponent rank, W/UE, average serve speed, percent aces, percent break points won, percent net points won, and interaction effects with court and W/UE, court and average service speed, court and percent aces, and court and break points won.
+For Williams, the model with the lowest AIC selected 22 coefficients, which makes it the largest model of the three.
+Since Williams only has 59 observations, we believe this model is overfitting and so caution the reader to be wary of any inference from this model.
+It should also be noted that no women's point by point data was recorded for 2015 (in which she won three grand slams) so that year is excluded from this model.
+The coefficients are surface, opponent rank, W/UE, average serve speed, percent aces, percent break points won, percent net points won, and interaction effects with court and W/UE, court and average service speed, court and percent aces, and court and break points won.
 
-The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal.
+The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal's model.
 For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wim.
@@
 For this, we do not see that Williams is dominant on the hard courts compared to the other courts.
-Rather, she has good results on all the courts.
-Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points than at the Australian Open.
+Rather, she has great results on all the courts and has the titles to prove it.
+Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points at the French Open than at the Australian Open.
 

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
height aces by tournamnet and ace ue plot

@@
 win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
 ```
@@
 
-## Tall players and aces
+# dont even know
 
@@
 
-```{r, fig.width=10, fig.height=8, fig.align='center'}
+We find that the distribution of errors and aces differ by tournament. Specifically, GAMES? in Wimbledon follow different ace and error patterns than the other tournaments.
+
+```{r, fig.height=4, fig.width=6}
+ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
+                           label = tournament, color=tournament)) + 
+  geom_jitter() +
+  labs( title = "Distribution of errors and aces differ by tournmanet",
+        x = "Total unforced errors", y = "Total aces") +
+  my_theme + scale_color_manual("Tournament", values = tournament_colors)
+
+```
+
+## Tall players and aces
+
+```{r, fig.width=10, fig.height=10}
+
 p_atp <- gs_players %>% filter(league == "ATP") %>%
-ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
-  geom_jitter() + 
-  scale_color_manual(guide=FALSE, values = win_colors, labels = c("No", "Yes")) + 
-  geom_smooth(method = "lm") + 
-  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
-                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+ggplot( aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament)) + 
+  scale_color_manual(guide=FALSE, values = tournament_colors) + 
+  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
+  geom_label_repel(data=subset(gs_players %>% filter(league == "ATP"), (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
                                 (log(ht+1) > 5.28 & log(ace+1) < 1) | 
@@
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE) + 
-  geom_label_repel(data=subset(filter(gs_players, league == "ATP"),
-                               (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+            alpha=.6, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(gs_players, (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
                                 (log(ht+1) > 5.28 & log(ace+1) < 1) | 
@@
 p_wta <- gs_players %>% filter(league == "WTA") %>%
-ggplot( aes(log(ht+1), log(ace+1), color = factor(did_win))) + 
-  geom_jitter() + 
-  scale_color_manual("Winner?", values = win_colors, labels = c("No", "Yes")) + 
-  geom_smooth(method = "lm") + 
+ggplot(aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament)) + 
+  scale_color_manual("Tournament", values = tournament_colors) + 
+  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
   geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
@@
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE) + 
+            alpha=.6, seed=1234, show.legend=FALSE) + 
   geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-                                (log(ht+1) > 5.215 & log(ace+1) < .5) | 
+                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
ideas for discussion

--- 2018-09-15 contest/surface-type/paper-tennis.Rmd
@@
 
++ HIGH LEVEL
+  + in general point distribution seems to be the same, 
+  + at the high level there are some surface differences that we have evidence for but cannot be sure either way: spainiards, 
+  + there is no home court advantage 
+  + French Open seems most different of the tournaments
+  + country does not seem to have an effect but individual does with court
++ INTERMEDIATE LEVEL
+  + Aces, Winners, and UEs are all important
+  + Taller people have more aces
+  + Wimbledon seems to have more winners and fewer UEs
++ LOW LEVEL
+  + Hierarchical model doesnt show much effect for Nadal, Federer, and Williams
+  + Individual models predicting % points won does show Nadal good on clay, fed on grass, and williams more jack of all trades
+  + we have less info about higher ranked (worse) players -> we need more data to pick out these effects
+
++ FUTURE WORK:
+  + closer look at point by point data
+  + include other masters tournaments
+  + pay more attention to year and players over time
+
 # References {#sec:refs}

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
added modeling to doc

@@
 
-## Mixed-effects models for all grand slams
+## Mixed-effects models
 
@@
 
-### Logistic regression for wins
+### Modeling wins with logistic regression
 
@@
 Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
-For these reasons, we chose to model whether a player won the match using a mixed-effects logistic regression approach, with the following predictors:
+For these reasons, we chose to model whether a player won the match using a mixed-effects logistic regression approach, with possible predictors including country of origin, tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
 
-* `ioc_fac`: Country that the player represents
-* `tournament`: Which Grand Slam the match was played at
-* `late_round`: Indicator for whether the match occured in Round of 16 or later
-* `rank`: Rank of player at the time of the match
-    + Included on log scale 
-* `opponent_rank`: Rank of opponent at the time of the match
-    + Included on log scale
-* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
-* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.
-
-We considered the following models:
-
-* No mixed effects
-    + `nocourt_logistic`: $\text{logit}(\pi_i) = \beta X$, $X$ does not include `tournament`
-        + If this model fits well, court surface does not have an effect on winning probability
-    + `base_logistic`: $\text{logit}(\pi_i) = \beta X$
-        + If this model fits better than the mixed-effects models, performance does not depend on individual-level or IOC-level 
-* IOC-level effect
-    + `country_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i C_i$
-* Player-level effect
-    + `ind_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, $\beta_{T[i]} = \alpha_i P_i$
-        * Includes both a fixed-effect term for IOC and a player-level effect for each grand slam
-    + `ind_logistic_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{T[i]}T$, where $X$ does not include IOC and $T$ represents indicators for tournament, $\beta_{T[i]} = \alpha_i P_i$
-        * Simpler than the `ind_logistic` model, since we exclude IOC effects.
-    + `ind_int_noioc`: $\text{logit}(\pi_i) = \beta X + \beta_{0[i]}$, where $X$ does not include IOC, $\beta_{0[i]} = \alpha_i$
-        * Similar to `ind_logistic_noioc`, but includes a single player-level effect rather than a player-level effect for each grand slam. 
-    + `ind_year_logistic`: $\text{logit}(\pi_i) = \beta X + \beta_{D[i]}$, where $X$ does not include IOC and $D$ represents indicators for years, $\beta_{Y[i]} = \alpha_i Y_i$
-        * Rather than including a player-level effect for grand slam, includes a player-level effect for each year in the dataset
+We considered a variety of models incorporating effects as either fixed or random.
+That is $\pi_i = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$, $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
+Based on AIC, we found the model that included a player-level term for each tournament, but excluded fixed terms for country, was the best fit for the data.
+The models are summarized in table (REF TABLE).
 
@@
 ```
-
-Based on AIC, the `ind_logistic_noioc` model performed best after accounting for the number of variables in the model.
 
@@
 
-knitr::kable(model.sums, format = "latex",
-             row.names = FALSE, col.names = c("Model", "AIC", "EDF"),
+model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef", "ind_no_ioc", "ind_intercept", "ind_year_ef")
+model.sums$fixed = c("No country, no tournament",
+                     "all",
+                     "no country, no tournament",
+                     "no tournament",
+                     "no country, no tournament",
+                     "no country",
+                     "no country, no year")
+model.sums$random = c("none",
+                      "none",
+                      "tournament (by country)",
+                      "tournament (by individual)",
+                      "tournament (by individual)",
+                      "intercept (by individual)",
+                      "year (by individual)")
+
+knitr::kable(model.sums[,c("name", "fixed", "random", "aic", "edf")], format = "latex",
+             row.names = FALSE, col.names = c("Model", "Fixed Effects", "Random Effects", "AIC", "EDF"),
              caption = "\\label{model.sums}AIC summary of the seven logistic regression models fitted.
@@
 
-If we examine the fixed effects for all seven models, and the random effects for the `ind_logistic` and `ind_logistic_noioc`, we see that the model structure chosen does not substantially change the parameter estimates.
+If we examine the fixed effects for all seven models, we see that the model structure chosen does not substantially change the parameter estimates.
 Across all models, `rank` and `opponent_rank` appear to best explain the probability of winning the match.
 Other fixed effects in the models, like whether it was a late round in the grand slam, the year, or ATP/WTF, were not found to be significantly different than zero.
+If we look at the random effects for Williams, Federer, and Nadal, however, the individual effects are not very informative (especially after taking uncertainty into consideration).
 
-```{r}
+```{r, fig.height = 3, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
 fixed.effects.plot = model.ests %>%
   filter(., !grepl("ioc", term)) %>%
-  ggplot(., aes(x = term, y = exp(estimate), col = model, ymin = exp(estimate - 2*std.error), ymax = exp(estimate + 2*std.error))) +
+  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
   geom_errorbar(width = .7, position = position_dodge()) +
-  theme_minimal() + 
+  my_theme +
   theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
-  scale_color_brewer(palette = "Dark2") + 
   ggtitle("Fixed effects") +
-  theme(legend.position = "bottom")
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "none") + 
+  xlab("") +
+  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
+  guides(col = guide_legend(nrow=2))
 
@@
   mutate(., tournament = gsub("tournament", "", tourn)) %>%
-  ggplot(., aes(x = tournament, y = exp(estimate), fill = model)) +
+  ggplot(., aes(x = tournament, y = estimate, fill = model)) +
   geom_violin() + 
-  theme_minimal() +
+  my_theme +
   scale_fill_brewer(palette = "Dark2") +
@@
   ggtitle("Random effects") +
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom") +
+  guides(col = guide_legend(nrow=3))
 
-plot_grid(fixed.effects.plot, random.effects.plot)
-```
-
-Looking at the `ind_logistic_noioc` random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
-This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
-
-```{r}
-knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
-             row.names = TRUE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
-             caption = "Correlation matrix for random effects",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
-```
-
-If we look at our big three, their individual effects are not very informative (especially after taking uncertainty into consideration).
-
-```{r}
 gs_players$name_int = as.integer(as.factor(gs_players$name))
@@
 name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
-as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+top_three_ranef = as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
   mutate(tournament = gsub("tournament", "", term)) %>%
@@
   xlab("") + 
-  ylab("Random Effect") + 
-  scale_color_brewer(palette = "Dark2") +
-  coord_flip()
+  ylab("Estimate") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  coord_flip() +
+  ggtitle("Random Effects") +
+  theme(legend.position = "bottom") + 
+  guides(col = guide_legend(nrow=2)) + 
+  theme(legend.text=element_text(size=8))
+
+grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5))
+```
+
+Looking at the random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
+
+```{r}
+knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
+             row.names = FALSE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
+             caption = "Correlation matrix for random effects",
+             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
 ```
@@
 Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
-We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, net points won, and unforced errors -- and see if individual differences are detectable.
+We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, net points won, and unforced errors -- and see if individual differences are detectable.
+We fit three additional models, each of the form $Y_i = \beta X_i + \beta_{T[i]}T$ with $\beta_{T[i]} = \alpha_i T_i$, with $X_i$ including the same variables as the model above.
 
-(some plots that show a) Serena/Federer are stronger servers, b) Federer makes fewer unforced errors at Wimbledon, c) something about Nadal)
+```{r, eval = FALSE}
+gs_partial_players$atp = gs_partial_players$Tour == "atp"
+gs_partial_players$year_fac = as.factor(gs_partial_players$year)
+gs_partial_players$late_round = gs_partial_players$round >= "R16"
+gs_partial_players$seeded = gs_partial_players$rank <= 32
+gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+```
 
+```{r}
+load("../../data/n_aces_mod.rda")
+load("../../data/n_net_mod.rda")
+load("../../data/n_ue_mod.rda")
+load("../../data/n_winners_mod.rda")
+```
+
+```{r, fig.height = 2.5, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge()) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Aces") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  coord_flip() +
+  theme(legend.position = "none")
+
+winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge()) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Winners") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  theme(axis.text.y = element_blank()) +
+  coord_flip() +
+  theme(legend.position = "none")
+
+nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge()) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Net Wins") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  theme(axis.text.y = element_blank()) +
+  coord_flip() +
+  theme(legend.position = "none")
+
+ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge()) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Unforced Errors") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  theme(axis.text.y = element_blank()) +
+  coord_flip() + 
+  theme(legend.text=element_text(size=10))
+
+grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
+
+```
+
+If we look at the random effects for Williams, Federer, and Nadal (FIGURE REF), differences are much more noticeable than in the logistic regression model.
+For instance, we would predict more aces for Williams at the Australian Open and Wimbledon than at the French Open.
+We also see that Federer's performance at Wimbledon is consistently better than other tournaments in both aces and unforced errors.
+Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.
 ## Modelling players individually

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
paper edits (AL)

@@
 
-Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open (FO), which is played on a clay surface [@bbc2018].
-In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon (Wim.), which is played on grass.
-On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open (AO) and 6 at US Open (USO)) and grass (7 at Wim.).
-These three players have different playstyles and origins, and they are legends in the sport of tennis.
-However, their success at grand slams seems to differ among the tournaments, which leads to question why that may be the case.
-
-The four grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different starting dates (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
-As a feature unique to tennis, all four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
+The four tennis grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different times of year (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
+All four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
 It is often thought that top players achieve better results on some surfaces than others.
-For example, there is the "home court advantage:" do players from the country of the grand slam tournament perform better than at other grand slams?
+For example, there is the "home court advantage" in which players from the country of the grand slam tournament are thought to perform better than at other grand slams.
 Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
 
- In this paper, we analyze the results of grand slam players from 2013-2017, and we analyze the surface effect on players at the four grand slams, first broadly at a high level and then narrowing down to player effects.
-More specifically,
+Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open, which is played on a clay surface [@bbc2018].
+In contrast, his rival Roger Federer has won the greatest proportion of his grand slam titles (8 out of 20) at Wimbledon, which is played on grass.
+On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open and 6 at US Open) and grass (7 at Wimbledon).
 
-1. Determine if results in terms of wins/losses and point distribution differ across the four surfaces
+ In this paper, we use the results of grand slam players from 2013-2017 to analyze the effect of court surface  on player performance at the four grand slams.
+We examine court surface across all participants in grand slams, and also investigate player effects.
+The goals of this work are to:
 
-2. Examine more fine grained attributes such as winners, aces, and unforced errors across the four surfaces
+1. Determine if results (in terms of wins/losses and point distributions) differ across the four surfaces
 
-3. Assess how individuals differ across the four surfaces, with special focus on Nadal, Federer, and Williams.
+2. Examine fine grained attributes such as winners, aces, and unforced errors across the four surfaces
 
-As to issue (1) quantifying the high-level effect of court surface on players, there has not been much written about with regards to tennis.
-There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
-Other sports analyses do take into account surface type such as grass vs.
-turf in soccer and football.
-Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
+3. Assess how players differ in performance across the four surfaces, with special focus on Nadal, Federer, and Williams.
 
-For issue (2) we examine the match attributes more unique to tennis such as winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
-Specifically, serving is thought to be of greater importance at Wimbledon since the ball travels faster on grass.
+There has been little work on quantifying the high-level effect of court surface in tennis (issue 1).
+Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
+[@knottenbelt2012] take court surface into account in their model, but do not compare the results of one surface to another, and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
+Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
+
+We examine the match attributes more unique to tennis to address issue 2.
+These attributes include winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
 In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].
 
-Finally, for issue (3) the player analyses, we take two approaches.
-On one hand, we use a hierarchical model with fixed and random effects taking into account player and tournament attributes using the complete data set and to fairly compare players to one another.
-This allows us to leverage tournament information to players without many matches played and to help account for noise.
-On the other hand, since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
+Finally, to assess how individual players differ in performance across surfaces, issue (3), we take two approaches.
+The first is to use a hierarchical model with fixed and random effects, taking into account player and tournament attributes using the complete data set.
+This allows us to leverage tournament information to model players with few matches played and to help account for noise.
+Our second approach is to model performance for specific players individually.
+Since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
 In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
-We also examine whether these players do have surface apparent effects.
-
-We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
-For our hierarchical model across all players, we model whether a player or lost a match based on player attributes along with opponent rank.
-For the individual models, we instead model percent of points won which can better show how dominant a player is in a match.
-For model selection, we use AIC to compare models from one another [@wasserman2004].
+We also examine whether these players do have court surface effects.
 For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.
 However, when the data becomes available, we would like to test these models on the 2018 grand slam results.
+For model selection, we use AIC to compare models from one another [@wasserman2004].
 
@@
 In [Section Data](#sec:data) we describe our grand slam tennis data.
-In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.
+In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering methods to determine how the courts differ from one another.
 In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.
 In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
-Finally in [Section Discussion](#discussion), we discuss future work and extensions or our models.
+Finally in [Section Discussion](#discussion), we summarize our results and discuss future work.
 
@@
 The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
-There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two, leagues, and 7 rounds.
+There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two leagues, and seven rounds.
 Each match has `r ncol(gs)` attributes, many of which are redundant.
@@
 Additionally, we found that no women's point by point data was recorded for 2015.
+
+```{r table, fig.pos = 'H'}
+data(gs_partial_players)
+
+top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
+matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
+  summarize(matches_played = n())
+matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
+  summarize(matches_played = sum(matches_played))
+tab <- spread(matches_played_slam, key = name, value = matches_played)
+total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
+colnames(total_df) <- colnames(tab)
+tab <- rbind(tab, total_df)
+
+knitr::kable(tab, format = "latex",
+               row.names = FALSE, col.names = c("Tournament", "Nadal", "Federer", "Williams"),
+               caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
+               booktabs = TRUE) %>% kable_styling(latex_options = "striped", "HOLD_position") %>%
+    row_spec(5, bold = TRUE, background = "lightgray")
+```
+
+In Table \ref{tab:three-gs-counts}, we display the number of matches Nadal, Federer, and Williams have played from 2013-2017.
+Over that time span, Nadal won 6 grand slams, Federer won 1, Williams won 8.
+Despite this, Federer played more total matches in Nadal.
+All three players were absent for exactly three slams during this time period due to external factors (Nadal: (1 Australian Open, 0 French Open, 1 Wimbledon, 1 US Open), Federer: (0 Australian Open, 2 French Open, 0 Wimbledon, 1 US Open), Serena: (0 AO, 1 FO, 1 Wimbledon, 1 US Open)).
+Not unexpectedly, Nadal has the most wins on clay (`r tab[2, 2]`), Federer has the most wins at Wimbledon (`r tab[4, 3]`) and Williams the most on hardcourt (`r tab[1, 4]` at Austrlian Open and `r tab[4, 4]` at US Open).
+Despite having played more matches than Nadal, Federer only has 1 grand slam to show compared to Nadal's six, which indicates that Federer made it deeper into tournaments on average but had difficulties winning the championships.
+For the WTA, Williams had her second most successful five-year span at the grand slams over this time period, winning 8.
 
@@
 
-## Mixed-effects models
+## Hierarchical models
 
-We use a mixed-effects modeling approach to examine the grand slam results from all players from 2013-2017.
-This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects.
+We use a hierarchical modeling approach to examine the grand slam results from all players from 2013-2017.
+This allows us to include fixed effects, which remain the same for each player, along with player-level individual effects.
 Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence.
-It also provides a way to examine individual player effects.
+It also provides a way to assess individual player effects.
 
@@
 Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
-For these reasons, we chose to model whether a player won the match using a mixed-effects logistic regression approach, with possible predictors including country of origin, tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
+For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible predictors including country of origin, tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
 
@@
 
-If we examine the fixed effects for all seven models, we see that the model structure chosen does not substantially change the parameter estimates.
-Across all models, `rank` and `opponent_rank` appear to best explain the probability of winning the match.
-Other fixed effects in the models, like whether it was a late round in the grand slam, the year, or ATP/WTF, were not found to be significantly different than zero.
-If we look at the random effects for Williams, Federer, and Nadal, however, the individual effects are not very informative (especially after taking uncertainty into consideration).
+If we examine the fixed effects for all seven models, we see that the parameter estimates do not substantially change across models (REF FIGURE).
+Rank and opponent rank appear to best explain the probability of winning the match.
+Other fixed effects in the models (whether it was a late round in the grand slam, the year, or ATP/WTF) were not found to be significantly different than zero.
+If we look at the random effects for Williams, Federer, and Nadal, however, the individual effects are not informative (especially after taking uncertainty into consideration).
 
@@
 
-Looking at the random effects more in-depth, we see that the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
-This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are not detectable between the two types of hard court and grass.
+More broadly, the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other court surfaces, but not between the two types of hard courts and grass.
 
@@
 
-### Linear regression for other stats
+### Linear regression for other outcomes
 
-Rank and opponent's rank clearly have the largest effect on predicting the winner.
-Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams.
-We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, net points won, and unforced errors -- and see if individual differences are detectable.
-We fit three additional models, each of the form $Y_i = \beta X_i + \beta_{T[i]}T$ with $\beta_{T[i]} = \alpha_i T_i$, with $X_i$ including the same variables as the model above.
+Instead of using `win' as the outcome variable, we'll now look at game statistics that may be more sensitive to court surface -- aces, net points won, and unforced errors -- and see if individual differences among players are detectable.
+We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
 
@@
 If we look at the random effects for Williams, Federer, and Nadal (FIGURE REF), differences are much more noticeable than in the logistic regression model.
-For instance, we would predict more aces for Williams at the Australian Open and Wimbledon than at the French Open.
+For instance, we would expect Williams to have more aces at the Australian Open and Wimbledon than at the French Open.
 We also see that Federer's performance at Wimbledon is consistently better than other tournaments in both aces and unforced errors.
 Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.
+
 ## Modelling players individually
 
-```{r table, fig.pos = 'H'}
-data(gs_partial_players)
-
-top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
-matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
-  summarize(matches_played = n())
-matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
-  summarize(matches_played = sum(matches_played))
-tab <- spread(matches_played_slam, key = name, value = matches_played)
-total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
-colnames(total_df) <- colnames(tab)
-tab <- rbind(tab, total_df)
-
-knitr::kable(tab, format = "latex",
-               row.names = FALSE, col.names = c("Tournament", "Nadal", "Federer", "Williams"),
-               caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
-               booktabs = TRUE) %>% kable_styling(latex_options = "striped", "HOLD_position") %>%
-    row_spec(5, bold = TRUE, background = "lightgray")
-```
-
-In Table \ref{tab:three-gs-counts}, we display the number of matches Nadal, Federer, and Williams have played from 2013-2017.
-Over that time span, Nadal won 6 grand slams, Federer won 1, Williams won 8.
-Despite this, Federer played more total matches in Nadal.
-All three players were absent for exactly three slams during this time period due to external factors (Nadal: (1 AO, 0 FO, 1 Wim, 1 USO), Federer: (0 AO, 2 FO, 0 Wim, 1 USO), Serena: (0 AO, 1 FO, 1 Wim, 1 USO)).
-Not unexpectedly, Nadal has the most wins on clay (`r tab[2, 2]`), Federer has the most wins at Wimbledon (`r tab[4, 3]`) and Williams the most on hardcourt (`r tab[1, 4]` at AO and `r tab[4, 4]` at USO).
-Despite having played more matches than Nadal, Federer only has 1 grand slam to show compared to Nadal's six, which indicates that Federer made it deeper into tournaments on average but had difficulties winning the championships.
-For the WTA, Williams had her second most successful five-year span at the grand slams over this time period, winning 8.
-
-We fit individual linear models for these three individuals (subsetting the partial data to their matches only), estimating the percent of points won in a match using forward-backwards stepwise regression.
+We also fit individual linear models for Williams, Federer, and Nadal (subsetting the partial data to their matches only), estimating the percent of points won in a match using forward-backwards stepwise regression.
 The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
@@
 Err.) of significant variables and their interaction terms, adjusting for the other covariates.
-Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically, AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.
+Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically at the Australian Open, we use the French Open, Wimbledon, and Australian Open, respectively, as the reference variable in the model.
 For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.
@@
 Looking at Nadal individually, the model with the lowest AIC has positive coefficients for opponent rank, percent of aces, and percent of break points won.
-It has negative coefficients for AO, USO, and Wim.
-compared to FO and for percent of net points won.
+It has negative coefficients for Australian Open, US Open, and Wimbledon compared to French Open and for percent of net points won.
 There are no interaction effects in this model.
@@
 The standardized residuals plot shown in Figure \ref{fig:individ-resids} (along with the other linear regression diagnostics plots not shown here) shows that the model is a good fit for predicting the percent of points won.
-For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)` \pm `r signif(2 * sqrt(cov_n[2,2]), 2)`$ less at the AO and $`r signif(abs(coefs_n[4]), 2)` \pm `r signif(2 * sqrt(cov_n[4,4]), 2)`$ less at Wim compared to the FO; to increase by $`r signif(abs(coefs_n[5]), 2)`\pm `r signif(2 * sqrt(cov_n[5,5]), 2)`$ for a one unit increase in opponent rank; a $`r .01 * signif(abs(coefs_n[7]), 2)`\pm `r signif(.02 * sqrt(cov_n[7,7]), 2)`$ increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)` \pm `r signif(.02 * sqrt(cov_n[8,8]), 2)`$ for a .01 increase in percent of net points won.
-As such, we see clear evidence Nadal performs better at the FO compared to the other grand slams.
+For the significant variables ($\alpha$ = 95\%), after adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)` \pm `r signif(2 * sqrt(cov_n[2,2]), 2)`$ less at the Australian Open and $`r signif(abs(coefs_n[4]), 2)` \pm `r signif(2 * sqrt(cov_n[4,4]), 2)`$ less at Wimbledon compared to the French Open; to increase by $`r signif(abs(coefs_n[5]), 2)`\pm `r signif(2 * sqrt(cov_n[5,5]), 2)`$ for a one unit increase in opponent rank; a $`r .01 * signif(abs(coefs_n[7]), 2)`\pm `r signif(.02 * sqrt(cov_n[7,7]), 2)`$ increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)` \pm `r signif(.02 * sqrt(cov_n[8,8]), 2)`$ for a .01 increase in percent of net points won.
+As such, we see clear evidence Nadal performs better at the French Open compared to the other grand slams.
 
@@
 
-For Federer, the model with the lowest AIC has negative coefficients for AO, FO, and USO.
-vs Wim., and W/UE,  It has positive coefficients for opponent rank, percent of break points won, and all the interaction terms: court and W/UE.
+For Federer, the model with the lowest AIC has negative coefficients for Australian Open, French Open, and US Open vs Wimbledon, and W/UE,  It has positive coefficients for opponent rank, percent of break points won, and all the interaction terms: court and W/UE.
 
 The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fairly good fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.
-For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)` \pm `r signif(2 * sqrt(cov_f[2,2] + cov_f[8,8] + 2 * cov_f[2,8]), 2)`$ less at AO compared to the Wim.; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[3,3] + cov_f[9,9] + 2 * cov_f[3,9]), 2)`$ less at FO compared to the Wim.; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[4,4] + cov_f[9,9] + 2 * cov_f[4,9]), 2)`$ less at USO compared to the Wim.;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[8,8] + 2 * cov_f[6,8]), 2)`$ for a one unit increase in W/UE at AO, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[9,9] + 2 * cov_f[6,9]), 2)`$ for a one unit increase in W/UE at FO, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[10,10] + 2 * cov_f[6,10]), 2)`$ for a one unit increase in W/UE at USO.
+For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)` \pm `r signif(2 * sqrt(cov_f[2,2] + cov_f[8,8] + 2 * cov_f[2,8]), 2)`$ less at Australian Open compared to Wimbledon; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[3,3] + cov_f[9,9] + 2 * cov_f[3,9]), 2)`$ less at French Open compared to the Wimbledon; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[4,4] + cov_f[9,9] + 2 * cov_f[4,9]), 2)`$ less at US Open compared to Wimbledon;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[8,8] + 2 * cov_f[6,8]), 2)`$ for a one unit increase in W/UE at Australian Open, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[9,9] + 2 * cov_f[6,9]), 2)`$ for a one unit increase in W/UE at French Open, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[10,10] + 2 * cov_f[6,10]), 2)`$ for a one unit increase in W/UE at US Open.
 This is strong evidence that Federer performs best at Wimbledon but possibly does better at the USO, depending on his W/UE ratio.
@@
 The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal's model.
-For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wim.
-compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the French Open.
+For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at French Open, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at US Open, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wimbledon compared to Australian Open; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the French Open.
 
-For this, we do not see that Williams is dominant on the hard courts compared to the other courts.
-Rather, she has great results on all the courts and has the titles to prove it.
+Based on the model results, we do not see that Williams is dominant on the hard courts compared to the other courts.
+Rather, she has great results on all the courts (and has the titles to prove it).
 Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points at the French Open than at the Australian Open.
+_Is this too speculative?
+(AL)_
 
 # Discussion {#sec:discussion}
+
+At the highest level, we find that the general point distribution seems to be the same for all court surfaces.
+There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g.
+that Spaniards perform better on clay).
+We did not find evidence for a home court advantage across all players in the grand slams.
+We also found that the French Open seems to be the most different out of the four grand slams.
+
+Using a hierarchical model approach, we did not see strong individual effects for Williams, Federer, and Nadal in predicting wins.
+We did, however, see differences for these players in hierarchical models to predict aces, net winners, and unforced errors, and that these differences were largely consistent with common tennis knowledge.
+
+Individual linear models predicting percent of points won showed that Nadal was best on clay, Federer was best on grass, and Williams was strong in all grand slams.
+
+Throughout our analyses, we often found that we were restricted by available data.
+Due to the large amount of grand slam data available, we were able to fit individual models for Federer, Nadal, and Williams that provided substantially more information about their performance on different court surfaces than the more general models which included all players.
+Ideally, these types of models would be possible for all players, not just the legends of the game.
+Point by point data from matches among lower-profile players is more likely to be missing or contain errors.
 

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
some writing fixes

@@
   - \usepackage{hyperref}
-abstract: "IDK writing is hard"
+abstract: "The Australian Open, French Open, Wimbledon, and US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.
+It is often thought that some players achieve better results at some grand slams than others and that difference in results may be attributed, at least partially, to the difference in the surfaces.
+For example three top players of all time, Rafael Nadal, Roger Federer, and Serena Williams have achieved their best results on clay, grass, and hard courts, respectively.
+In this paper, we examine the effect of surface type at grand slams from 2013-2017 for top players, adjusting for confounders such as league, competitor strength, and player attributes.
+We answer the question of court effect on player results at first a high level using a ``complete'' data with few features and end using partial data with a variety of tennis-specific features.
+Our primary model is a hierarchical model with fixed and random effects accounting for both player and tournament features.
+We show that differences in surface type become more apparent with more detailed data and that more study should be conducted in this area."
 bibliography: master.bib
@@
 
-The four tennis grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different times of year (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
-All four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
+The four tennis grand slams ( Australian Open (AO), French Open (FO),  Wimbledon (Wim.), US Open (USO)) are played in different cities (Melbourne, Paris, London, New York City) at different times of year (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players for both men (ATP) and women (WTA).
+All four slams are played on different surfaces (Plexicushion hard court, clay, grass, DecoTurf hard court).
 It is often thought that top players achieve better results on some surfaces than others.
-For example, there is the "home court advantage" in which players from the country of the grand slam tournament are thought to perform better than at other grand slams.
+For example, there is the idea of "home court advantage" in which players from the country of the grand slam tournament are thought to perform better than at other grand slams.
 Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
@@
 
-3. Assess how players differ in performance across the four surfaces, with special focus on Nadal, Federer, and Williams.
+3. Assess how players perform across the four surfaces, with focus on Nadal, Federer, and Williams.
 
@@
 Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-[@knottenbelt2012] take court surface into account in their model, but do not compare the results of one surface to another, and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
-Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
+@knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass comparead to other courts.
+Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].
 
-We examine the match attributes more unique to tennis to address issue 2.
+We examine the match attributes more unique to tennis to address issue (2).
 These attributes include winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
@@
 Our second approach is to model performance for specific players individually.
-Since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
-In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
+Since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with tennis-specific feature attributes to examine the effects individually.
+In both cases, we examine whether our models pass "common sense" tests like how the models in @thomas2013 show that commonly well known hockey players also have high status in the model.
 We also examine whether these players do have court surface effects.
@@
 However, when the data becomes available, we would like to test these models on the 2018 grand slam results.
-For model selection, we use AIC to compare models from one another [@wasserman2004].
+For model selection, we use AIC to compare models to one another [@wasserman2004].
 
@@
 We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
-Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.
+Additionally, we take into account the number of sets in a match, the court surface, and round of the tournament.
 A subset of the data is shown in Table \ref{tab:data}.
@@
 ```{r tab-data, include = TRUE, echo = FALSE}
-w_id <- which(gs$winner_name == "Serena Williams")[1:2]
-f_id <- which(gs$winner_name == "Roger Federer")[1:2]
-n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
+w_id <- which(gs$winner_name == "Serena Williams")[1]
+f_id <- which(gs$winner_name == "Roger Federer")[1]
+n_id <- which(gs$winner_name == "Rafael Nadal")[1]
 s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
@@
 The secondary data is partial point by point data for grand slam matches.
-In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.
-There is also tournament info such as surface, year, and time start.
+In this data set, each row is a point in a match with details on who won the point, serve speed, and whether a player had a forced or unforced error, winner, ace, or net point win.
+There is also tournament info such as court surface, year, and time start.
 Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.
@@
 We do this by joining the features of tournament, year, player one, and player two.
-This final data set consists of `r nrow(gs_pbp)` observations (compared to the 5080 in the original grand slams).
+This final data set consists of `r nrow(gs_partial)` observations (compared to the 5080 in the original grand slams).
 The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
@@
                row.names = FALSE, col.names = c("Tournament", "Nadal", "Federer", "Williams"),
-               caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
+               caption = "\\label{tab:three-gs-counts}\\# of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
                booktabs = TRUE) %>% kable_styling(latex_options = "striped", "HOLD_position") %>%
@@
 Despite this, Federer played more total matches in Nadal.
-All three players were absent for exactly three slams during this time period due to external factors (Nadal: (1 Australian Open, 0 French Open, 1 Wimbledon, 1 US Open), Federer: (0 Australian Open, 2 French Open, 0 Wimbledon, 1 US Open), Serena: (0 AO, 1 FO, 1 Wimbledon, 1 US Open)).
-Not unexpectedly, Nadal has the most wins on clay (`r tab[2, 2]`), Federer has the most wins at Wimbledon (`r tab[4, 3]`) and Williams the most on hardcourt (`r tab[1, 4]` at Austrlian Open and `r tab[4, 4]` at US Open).
-Despite having played more matches than Nadal, Federer only has 1 grand slam to show compared to Nadal's six, which indicates that Federer made it deeper into tournaments on average but had difficulties winning the championships.
-For the WTA, Williams had her second most successful five-year span at the grand slams over this time period, winning 8.
+All three players were absent for exactly three slams during this time period due to external factors.
+Not unexpectedly, Nadal has the most wins at FO (`r tab[2, 2]`), Federer has the most wins at Wim.
+(`r tab[4, 3]`), and Williams the most on hardcourt (`r tab[1, 4]` at AO and `r tab[4, 4]` at US Open).
 
@@
 
-We first examine the distribution of points earned per match to assess normality.
+We first examine the assumption that distribution of points comes from a normal distribution.
 The distribution of points earned is approximately normal.
@@
 
-It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
+It is commonly thought that there is a home court advantage in grand slam matches [@sb2013].
 In our data we find this to be true (i.e.

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
kayla update graphs

@@
 library(cowplot)
+library(ggpubr)
 
@@
 ## Early Data Analysis {#sec:eda}
-
 ### Examining the distribution of points earned
 
-We first examine the assumption that distribution of points comes from a normal distribution.
+We first examine the distribution of points earned per match to assess normality.
 The distribution of points earned is approximately normal.
@@
 
-```{r, fig.height=5, fig.width=12}
+```{r, fig.height=5, fig.width=10}
 
@@
 
-It is commonly thought that there is a home court advantage in grand slam matches [@sb2013].
+It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
 In our data we find this to be true (i.e.
 French players win the French open more than French players win other slams).
-But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a particular country play in "their" tournament more often than they play in other tournaments.
-We also find this to be true in our data for France, The United States, and Australia (i.e.
+But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments.
+We also find this to be true in our data for France, The United States, Australia, and Great Britain (i.e.
 the proportion of French players in the French Open is greater than the proportion of French players in other slams).
@@
 The same would be true for Australia and the US.
-But, we see that this isn't the case.
+But, we see in \ref{fig:home-court} that this isn't the case.
 After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
 
-```{r, fig.height=4, fig.width=14, fig.align='center'}
+```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
 france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
        x = "", y = "Win proportion") + 
-  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
+  scale_color_manual(" ", values=tournament_colors) + my_theme
 
@@
        x = "", y = "Win proportion") +  
-  scale_color_manual(guide=FALSE, values=tournament_colors)  + my_theme
+  scale_color_manual(" ", values=tournament_colors)  + my_theme
 
@@
 
-grid.arrange(p3, p1, p2,  p4, nrow=1, widths = c(.22, .22, .22, .34),
-             top = textGrob("Home court advantage: debunked",
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(., top = textGrob("Home court advantage: debunked",
                             gp=gpar(fontsize=20, fontfamily="serif")),
              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           # x = 0,
-                           # y = 0.5,
                            just = "left",
@@
 
-It is also commonly thought that Spaniards play better on clay.
-We are interested in whether Spaniards win the French Open more than they win in other tournaments.
-It does appear that Spaniards are winning the French Open more than they are winning other tournaments.
-But, this result is not significant.
-In addition, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, which may help explain this common misconception, although this result is not significant either.
-
-```{r, fig.height=4, fig.width=12, fig.align='center'}
+```{r}
 
@@
             rank_se = sd(rank)/sqrt(n()))
+
+prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
+                table(did_win, tournament)), 2)
+
+with(filter(gs_players, ioc == "ESP", round == "R128"),
+                table(did_win, tournament))
+
+with(filter(gs_players, ioc == "ESP", round == "R128"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
+
+```
+
+It is also commonly thought that Spaniards play better on clay.
+We are interested in whether Spaniards win the French Open more than they win in other tournaments.
+It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
+But, this result is not significant (ENTER P VALUE).
+Interestingly, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse than Spanish players in ther tournaments.
+
+```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
 
@@
   scale_color_manual(" ", values=tournament_colors) + my_theme
-# +
-#   geom_errorbar(aes(ymin = avg_ranking - 1.96*rank_se,
-#                     ymax = avg_ranking + 1.96*rank_se),
-#                 width=.2,
-#                  position=position_dodge(0.05))
 
-grid.arrange(p1, p2, nrow=1, widths = c(.44, .56),
-             top = textGrob("Spaniards perform better on Clay: debunked",
+ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
                             gp=gpar(fontsize=20, fontfamily="serif")),
@@
 
-# dont even know
+# Unforced errors and aces
 
-Why the actual EFF won't fig.width work?
+We find that the distribution of errors and aces differ by tournament.
+Specifically, GAMES?
+in Wimbledon follow different ace and error patterns than the other tournaments.
+These GAMES?
+tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces.
+It is known (SOURCE) that the ball moves faster on grass courts and so the ace result is unsuprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
 
-We find that the distribution of errors and aces differ by tournament. Specifically, GAMES? in Wimbledon follow different ace and error patterns than the other tournaments.
-
-```{r, fig.height=4, fig.width=6}
+```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
 ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
@@
   geom_jitter() +
-  labs( title = "Distribution of errors and aces differ by tournmanet",
+  labs( title = "Distribution of errors and aces differ by tournament",
         x = "Total unforced errors", y = "Total aces") +
@@
 
-```{r, fig.width=10, fig.height=10}
+Another common thought about tennis is that taller players are better servers.
+We find this to be true and see a strong positive correlation between teh variables.
+We highlight some interesting players BLAH BLAH BLAH.
+
+```{r, fig.width=10, fig.height=4.8, fig.align='center'}
+
+set.seed(8)
+
+atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
+                 "Roger Federer", "Andy Murray")
+atp_servers <- sapply(atp_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
+                 "Venus Williams", "Serena Williams", "Madison Keys")
+wta_servers <- sapply(wta_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
 
@@
   geom_jitter(aes(color = tournament)) + 
-  scale_color_manual(guide=FALSE, values = tournament_colors) + 
-  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-  geom_label_repel(data=subset(gs_players %>% filter(league == "ATP"), (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
-                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
-                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+  scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=atp_servers,
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.6, seed=1234, show.legend=FALSE) + 
-  geom_label_repel(data=subset(gs_players, (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
-                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
-                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=atp_servers,
             aes(log(ht+1), log(ace+1), label = name),
             seed=1234, fill=NA, show.legend=FALSE) +
-  labs(x = "log(height + 1)", title = "ATP")+
+  labs(x = "log(height + 1)", title = "ATP") +
   my_theme
@@
   geom_jitter(aes(color = tournament)) + 
-  scale_color_manual("Tournament", values = tournament_colors) + 
-  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
-                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-                                 (log(ace + 1) > 2.82)),
+  scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth(method = "lm", color="black") + 
+  geom_label_repel(data=wta_servers,
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.6, seed=1234, show.legend=FALSE) + 
-  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
-                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-                                 (log(ace + 1) > 2.82)),
-            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=wta_servers,
+            aes( log(ht+1), log(ace+1), label=name),
             seed=1234, fill=NA, show.legend=FALSE) +
   labs(x = "log(height + 1)", title = "WTA")+
-  my_theme + theme(legend.position="bottom")
+  my_theme
 
-grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
-             top = textGrob("Taller players are better servers: confirmed",
+# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
+#              top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") %>%
+  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
                             gp=gpar(fontsize=20, fontfamily="serif")))

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
edits

@@
 However, these additional attributes are only available for about 1/8 of the data.
-
-We aggregate the partial point by point data into partial match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
-We then join the partial match data with the complete data to find the score and outcome of the match.
-We do this by joining the features of tournament, year, player one, and player two.
-This final data set consists of `r nrow(gs_partial)` observations (compared to the 5080 in the original grand slams).
+This partial data set consists of `r nrow(gs_partial)` observations (compared to the 5080 in the original grand slams).
 The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
@@
 
-We first examine the distribution of points earned per match to assess normality.
-The distribution of points earned is approximately normal.
-This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
-As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
+We first determine whether the distribution of points comes from a normal distribution.
+We plot the distribution of points in Figure \ref{fig:distr-eda} (left), and the distribution does seem approximately normal.
+This distribution is similar across tournament, with Wim.
+differing slightly from the other grand slams.
+As expected, there are more points earned in the WTA than the ATP (Fig.
+\ref{fig:distr-eda} (right)) due to the differing number of sets played.
 Also unsurprisingly, the winners of the match tended to earn more points than the losers.
 
-```{r, fig.height=5, fig.width=10}
+```{r, fig.height=5, fig.width=10, fig.cap="\\label{fig:distr-eda} Distribution of number of points for the different grand slams (left) and distribution of points per match faceted by winner and loser for the two leagues (right)."}
 
-# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
+# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34),)
 
@@
 
-It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
-In our data we find this to be true (i.e.
+It is commonly thought that there is a home court advantage in grand slam matches [@sb2013].
+In our data, we find this to be true (e.g.
 French players win the French open more than French players win other slams).
-But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments.
-We also find this to be true in our data for France, The United States, Australia, and Great Britain (i.e.
+However, we also know that the home team is given preference for wild card bids [@usta2018] so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments.
+We also find this to be true in our data for France, the United States, Australia, and Great Britain (e.g.
 the proportion of French players in the French Open is greater than the proportion of French players in other slams).
@@
 Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments.
-If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon.
+If there really is a home court advantage, the proportion of French wins each round would be higher at the FO than the AO, the USO, and Wim..
 The same would be true for Australia and the US.
-But, we see in \ref{fig:home-court} that this isn't the case.
-After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
+However, in Figure \ref{fig:home-court} that this is not the case.
+After accounting for the number of players from each country, we do not find a home court advantage in the grand slams.
 
@@
 It is also commonly thought that Spaniards play better on clay.
-We are interested in whether Spaniards win the French Open more than they win in other tournaments.
-It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
-But, this result is not significant (ENTER P VALUE).
-Interestingly, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse than Spanish players in ther tournaments.
+We are interested in whether Spaniards win matches at the French Open more than they win in other tournaments.
+It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in Figure \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
+However, this result is not significant (ENTER P VALUE).
+The ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse in ranking but still winning more than at the other tournaments.
 
@@
 
-# Unforced errors and aces
+## Unforced errors and aces
 
-We find that the distribution of errors and aces differ by tournament.
-Specifically, GAMES?
-in Wimbledon follow different ace and error patterns than the other tournaments.
-These GAMES?
-tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces.
-It is known (SOURCE) that the ball moves faster on grass courts and so the ace result is unsuprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
+We find that the distribution of UEs and aces differ by tournament.
+In Figure \ref{fig:ue-ace} we plot the total aces in a match vs.
+the total unforced errors in a match and color by the tournament.
+Specifically, matches in Wim.
+follow different ace and error patterns than the other tournaments.
+These matches tend to cluster in the upper left hand part of the graph, meaning that at Wim.
+players are making fewer unforced errors and scoring more aces.
+It is thought that the ball moves faster on grass courts and so the ace result is unsurprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
 
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
+```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
 ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
@@
   my_theme + scale_color_manual("Tournament", values = tournament_colors)
-
-```
-
-## Tall players and aces
-
-Another common thought about tennis is that taller players are better servers.
-We find this to be true and see a strong positive correlation between teh variables.
-We highlight some interesting players BLAH BLAH BLAH.
-
-```{r, fig.width=10, fig.height=4.8, fig.align='center'}
-
-set.seed(8)
-
-atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
-                 "Roger Federer", "Andy Murray")
-atp_servers <- sapply(atp_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
-                 "Venus Williams", "Serena Williams", "Madison Keys")
-wta_servers <- sapply(wta_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-p_atp <- gs_players %>% filter(league == "ATP") %>%
-ggplot( aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament)) + 
-  scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE) + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label = name),
-            seed=1234, fill=NA, show.legend=FALSE) +
-  labs(x = "log(height + 1)", title = "ATP") +
-  my_theme
-
-p_wta <- gs_players %>% filter(league == "WTA") %>%
-ggplot(aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament)) + 
-  scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth(method = "lm", color="black") + 
-  geom_label_repel(data=wta_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE) + 
-  geom_label_repel(data=wta_servers,
-            aes( log(ht+1), log(ace+1), label=name),
-            seed=1234, fill=NA, show.legend=FALSE) +
-  labs(x = "log(height + 1)", title = "WTA")+
-  my_theme
-
-# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") %>%
-  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-                            gp=gpar(fontsize=20, fontfamily="serif")))
 
@@
 Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
-For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible predictors including country of origin, tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
+For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible covariates including country of origin (IOC), tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
 
 We considered a variety of models incorporating effects as either fixed or random.
-That is $\pi_i = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$, $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
+That is $\pi_i = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$ (i.e.
+a player win), $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
 Based on AIC, we found the model that included a player-level term for each tournament, but excluded fixed terms for country, was the best fit for the data.
-The models are summarized in table (REF TABLE).
+The models are summarized in Table \ref{tab:model.sums}.
 
@@
              row.names = FALSE, col.names = c("Model", "Fixed Effects", "Random Effects", "AIC", "EDF"),
-             caption = "\\label{model.sums}AIC summary of the seven logistic regression models fitted.
+             caption = "\\label{tab:model.sums}AIC summary of the seven logistic regression models fitted.
 Including individual effects for each grand slam, without including any country effects, leads to the best model fit according to AIC.",
@@
 
-If we examine the fixed effects for all seven models, we see that the parameter estimates do not substantially change across models (REF FIGURE).
+If we examine the fixed effects for all seven models (see Fig.
+\ref{fig:mixed-mods-ests} (left)), we see that the parameter estimates do not substantially change across models.
 Rank and opponent rank appear to best explain the probability of winning the match.
-Other fixed effects in the models (whether it was a late round in the grand slam, the year, or ATP/WTF) were not found to be significantly different than zero.
-If we look at the random effects for Williams, Federer, and Nadal, however, the individual effects are not informative (especially after taking uncertainty into consideration).
+Other fixed effects in the models (whether it was a late round in the grand slam, the year, or ATP/WTA) were not found to be significantly different than zero.
+If we look at the random effects (see Fig.
+\ref{fig:mixed-mods-ests} (right)) for Williams, Federer, and Nadal, however, the individual effects are not informative (especially after taking uncertainty into consideration).
 
@@
 
-More broadly, the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated.
-This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other court surfaces, but not between the two types of hard courts and grass.
+More broadly, the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated (see Table \ref{tab:corr}).
+This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other surfaces, but not among the hard courts and grass.
 
@@
              row.names = FALSE, col.names = c("Australian Open", "French Open", "US Open", "Wimbledon"),
-             caption = "Correlation matrix for random effects",
+             caption = "\\label{tab:corr}Correlation matrix for random effects",
              booktabs = TRUE) %>% kable_styling(latex_options = "striped")
@@
 
-Instead of using `win' as the outcome variable, we'll now look at game statistics that may be more sensitive to court surface -- aces, net points won, and unforced errors -- and see if individual differences among players are detectable.
+Instead of using `win' as the outcome variable, we now look at match statistics that may be more sensitive to court surface -- aces, net points won, and UEs -- and see if individual differences among players are detectable.
 We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
@@
 
-If we look at the random effects for Williams, Federer, and Nadal (FIGURE REF), differences are much more noticeable than in the logistic regression model.
-For instance, we would expect Williams to have more aces at the Australian Open and Wimbledon than at the French Open.
-We also see that Federer's performance at Wimbledon is consistently better than other tournaments in both aces and unforced errors.
+If we look at the random effects for Williams, Federer, and Nadal (see Fig.
+\ref{fig:mixed-mods-stats}), differences are much more noticeable than in the logistic regression model.
+For instance, we would expect Williams to have more aces at the AO and Wim.
+than at FO.
+We also see that Federer's performance at Wim.
+is consistently better than other tournaments in both aces and unforced errors.
 Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.
 
-## Modelling players individually
+## Modeling players individually
 
-We also fit individual linear models for Williams, Federer, and Nadal (subsetting the partial data to their matches only), estimating the percent of points won in a match using forward-backwards stepwise regression.
+We also fit individual linear models for Williams, Federer, and Nadal (subsetting the partial data to their matches only).
+We estimate the percent of points won in a match using forward-backwards stepwise regression.
 The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
@@
 $$  
-The upper model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
+The full model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.
 We do not display the full best fit models here, but they are available online.
@@
 Err.) of significant variables and their interaction terms, adjusting for the other covariates.
-Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically at the Australian Open, we use the French Open, Wimbledon, and Australian Open, respectively, as the reference variable in the model.
+Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically at AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.
 For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.
@@
 
-The standardized residuals plot shown in Figure \ref{fig:individ-resids} (along with the other linear regression diagnostics plots not shown here) shows that the model is a good fit for predicting the percent of points won.
  For the significant variables ($\alpha$ = 95\%), after adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)` \pm `r signif(2 * sqrt(cov_n[2,2]), 2)`$ less at the Australian Open and $`r signif(abs(coefs_n[4]), 2)` \pm `r signif(2 * sqrt(cov_n[4,4]), 2)`$ less at Wimbledon compared to the French Open; to increase by $`r signif(abs(coefs_n[5]), 2)`\pm `r signif(2 * sqrt(cov_n[5,5]), 2)`$ for a one unit increase in opponent rank; a $`r .01 * signif(abs(coefs_n[7]), 2)`\pm `r signif(.02 * sqrt(cov_n[7,7]), 2)`$ increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)` \pm `r signif(.02 * sqrt(cov_n[8,8]), 2)`$ for a .01 increase in percent of net points won.
@@
 
-The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fairly good fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.
 For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)` \pm `r signif(2 * sqrt(cov_f[2,2] + cov_f[8,8] + 2 * cov_f[2,8]), 2)`$ less at Australian Open compared to Wimbledon; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[3,3] + cov_f[9,9] + 2 * cov_f[3,9]), 2)`$ less at French Open compared to the Wimbledon; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`\pm `r signif(2 * sqrt(cov_f[4,4] + cov_f[9,9] + 2 * cov_f[4,9]), 2)`$ less at US Open compared to Wimbledon;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[8,8] + 2 * cov_f[6,8]), 2)`$ for a one unit increase in W/UE at Australian Open, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[9,9] + 2 * cov_f[6,9]), 2)`$ for a one unit increase in W/UE at French Open, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)` \pm `r signif(2 * sqrt(cov_f[6,6] + cov_f[10,10] + 2 * cov_f[6,10]), 2)`$ for a one unit increase in W/UE at US Open.
@@
 
-```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}"}
+```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}", eval = FALSE}
 
-grid.arrange(gn, gf, gw, ncol = 3, 
+g <- grid.arrange(gn, gf, gw, ncol = 3, 
               top = textGrob("Standardized Residuals vs. Predicted % Points Won",
@@
 Since Williams only has 59 observations, we believe this model is overfitting and so caution the reader to be wary of any inference from this model.
-It should also be noted that no women's point by point data was recorded for 2015 (in which she won three grand slams) so that year is excluded from this model.
-The coefficients are surface, opponent rank, W/UE, average serve speed, percent aces, percent break points won, percent net points won, and interaction effects with court and W/UE, court and average service speed, court and percent aces, and court and break points won.
+It should also be noted that no women's point by point data was recorded for 2015 (in which she won three grand slams).
 
-The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal's model.
-For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at French Open, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at US Open, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wimbledon compared to Australian Open; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the French Open.
+For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.
 
-Based on the model results, we do not see that Williams is dominant on the hard courts compared to the other courts.
-Rather, she has great results on all the courts (and has the titles to prove it).
-Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points at the French Open than at the Australian Open.
-_Is this too speculative?
-(AL)_
+Based on the model results, we do not see that Williams is dominant on the hard courts compared to the other courts.  Rather, she has respectable results on all the surfaces.
 
@@
 that Spaniards perform better on clay).
-We did not find evidence for a home court advantage across all players in the grand slams.
-We also found that the French Open seems to be the most different out of the four grand slams.
+We do not find evidence for a home court advantage across all players in the grand slams.
+We also find that the FO seems to be the most different out of the four grand slams, but Wim.
+also displays some unique characteristics with regards to low amounts of UEs.
 
-Using a hierarchical model approach, we did not see strong individual effects for Williams, Federer, and Nadal in predicting wins.
-We did, however, see differences for these players in hierarchical models to predict aces, net winners, and unforced errors, and that these differences were largely consistent with common tennis knowledge.
+Using a hierarchical model approach, we do not see strong individual effects for Williams, Federer, and Nadal in predicting wins.
+We do, however, see differences for these players in hierarchical models to predict aces, net winners, and UEs, and that these differences are largely consistent with common tennis knowledge.
 
-Individual linear models predicting percent of points won showed that Nadal was best on clay, Federer was best on grass, and Williams was strong in all grand slams.
+Individual linear models predicting percent of points won show that Nadal was best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
 
-Throughout our analyses, we often found that we were restricted by available data.
-Due to the large amount of grand slam data available, we were able to fit individual models for Federer, Nadal, and Williams that provided substantially more information about their performance on different court surfaces than the more general models which included all players.
+Throughout our analyses, we often found that we are restricted by available data.
+Due to the large amount of grand slam data available, we are able to fit individual models for Federer, Nadal, and Williams that provided substantially more information about their performance on different court surfaces than the more general models which included all players.
+Point by point data from matches among lower-profile players is more likely to be missing or contain errors.
 Ideally, these types of models would be possible for all players, not just the legends of the game.
-Point by point data from matches among lower-profile players is more likely to be missing or contain errors.
 
-+ HIGH LEVEL
-  + in general point distribution seems to be the same, 
-  + at the high level there are some surface differences that we have evidence for but cannot be sure either way: spainiards, 
-  + there is no home court advantage 
-  + French Open seems most different of the tournaments
-  + country does not seem to have an effect but individual does with court
-+ INTERMEDIATE LEVEL
-  + Aces, Winners, and UEs are all important
-  + Taller people have more aces
-  + Wimbledon seems to have more winners and fewer UEs
-+ LOW LEVEL
-  + Hierarchical model doesnt show much effect for Nadal, Federer, and Williams
-  + Individual models predicting % points won does show Nadal good on clay, fed on grass, and williams more jack of all trades
-  + we have less info about higher ranked (worse) players -> we need more data to pick out these effects
-
-+ FUTURE WORK:
-  + closer look at point by point data
-  + include other masters tournaments
-  + pay more attention to year and players over time
+Looking forward, we would like to better explore the point by point data, especially for the attributes which excluded in this analysis: rally length, minutes played, and distance covered.
+We would also like to pull in data from Masters level tournaments, which would give us a clearer view if the differences we found in this paper are due to court surface or the slam itself.
+Another way we could adjust for grand slam confounders would be to incorporate weather conditions and matches played in the last few months.
+Additionally, we would like to be able to adjust for performance over time as new competitors arrive and older players decline.
 

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
now using library!

@@
 ## To change into library(courtsports)
-devtools::load_all("../..")
+library(courtsports)
 
@@
 
-step_model_n <- model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
+step_model_n <- courtsports:::model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
 
@@
 
-step_model_f <- model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
+step_model_f <- courtsports:::model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
 
@@
 
-step_model_w <- model_individual(player_name = "Serena Williams", data = gs_partial_players,
+step_model_w <- courtsports:::model_individual(player_name = "Serena Williams", data = gs_partial_players,
                                  start_lower = TRUE, ref = "Australian Open")

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

@@
   - \usepackage{hyperref}
-abstract: "The Australian Open, French Open, Wimbledon, and US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.
+abstract: "The Australian Open, The French Open, Wimbledon, and The US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.
 It is often thought that some players achieve better results at some grand slams than others and that difference in results may be attributed, at least partially, to the difference in the surfaces.
@@
 In this paper, we examine the effect of surface type at grand slams from 2013-2017 for top players, adjusting for confounders such as league, competitor strength, and player attributes.
-We answer the question of court effect on player results at first a high level using a ``complete'' data with few features and end using partial data with a variety of tennis-specific features.
+We answer the question of court effect on player results at first a high level using ``complete'' data with few features and then using partial data with a variety of tennis-specific features.
 Our primary model is a hierarchical model with fixed and random effects accounting for both player and tournament features.
@@
 
-There has been little work on quantifying the high-level effect of court surface in tennis (issue 1).
+There has been little work on quantifying the high-level effect of court surface in tennis issue (1).
 Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-@knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass comparead to other courts.
+@knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass compared to other courts.
 Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].
@@
 The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
-There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two leagues, and seven rounds.
+There are `r nrow(gs)` observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two leagues, and seven rounds.
 Each match has `r ncol(gs)` attributes, many of which are redundant.
@@
 We plot the distribution of points in Figure \ref{fig:distr-eda} (left), and the distribution does seem approximately normal.
-This distribution is similar across tournament, with Wim.
-differing slightly from the other grand slams.
+This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
 As expected, there are more points earned in the WTA than the ATP (Fig.
@@
 If there really is a home court advantage, the proportion of French wins each round would be higher at the FO than the AO, the USO, and Wim..
-The same would be true for Australia and the US.
+The same would be true for Australia, the US, and Great Britain.
 However, in Figure \ref{fig:home-court} that this is not the case.
@@
 
-```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
+```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
 france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
 
-prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament)), 2)
+chisq_r128 <- with(filter(gs_players, ioc == "ESP", round == "R128"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
 
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament))
+chisq_r64 <- with(filter(gs_players, ioc == "ESP", round == "R64"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
 
-with(filter(gs_players, ioc == "ESP", round == "R128"),
+chisq_r32 <- with(filter(gs_players, ioc == "ESP", round == "R32"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
+
+chisq_r16 <- with(filter(gs_players, ioc == "ESP", round == "R16"),
      table(did_win, tournament)) %>%
@@
 It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in Figure \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
-However, this result is not significant (ENTER P VALUE).
-The ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse in ranking but still winning more than at the other tournaments.
+However, this result is not significant at any round (p-value = `r round(chisq_r128$p.value, 2)`, `r round(chisq_r64$p.value, 2)`, `r round(chisq_r32$p.value, 2)`, `r round(chisq_r16$p.value, 2)`).
+The ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are worse in ranking yet winning more often.
 
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
+```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}
 
@@
 ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
+grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed",
                             gp=gpar(fontsize=20, fontfamily="serif")),
@@
 the total unforced errors in a match and color by the tournament.
-Specifically, matches in Wim.
-follow different ace and error patterns than the other tournaments.
-These matches tend to cluster in the upper left hand part of the graph, meaning that at Wim.
-players are making fewer unforced errors and scoring more aces.
+Specifically, matches in Wimbledon follow different ace and error patterns than the other tournaments.
+These matches tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces.
 It is thought that the ball moves faster on grass courts and so the ace result is unsurprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
@@
 Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets).
-For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible covariates including country of origin (IOC), tournament that the match was played at, whether the match occured in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
+For these reasons, we chose to model whether a player won the match using a hierarchical logistic regression approach, with possible covariates including country of origin (IOC), tournament that the match was played at, whether the match occurred in Round of 16 or later, rank of player, rank of opposing player, year of the match, and whether it was a men's or women's match.
 
@@
 
-```{r, fig.height = 3, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
+```{r, fig.height = 3.5, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
 fixed.effects.plot = model.ests %>%
@@
   xlab("") +
+  ylab("Estimate") +
   scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
@@
 
-grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5))
+grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5),
+             top = textGrob("`ind_no_ioc' model",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
 ```
 
-More broadly, the effects for the Austrialian Open, US Open, and Wimbledon are all quite correlated (see Table \ref{tab:corr}).
+More broadly, the effects for the Australian Open, US Open, and Wimbledon are all quite correlated (see Table \ref{tab:corr}).
 This suggests that, after accounting for additional variables such as rank and opponent rank, differences in win probability are detectable between clay and the other surfaces, but not among the hard courts and grass.
@@
 
-```{r, fig.height = 2.5, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+```{r, fig.height = 3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
 aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
@@
   theme(axis.text.y = element_blank()) +
-  coord_flip() +
-  theme(legend.position = "none")
+  coord_flip()  # +
+  # theme(legend.position = "none")
 
@@
   theme(axis.text.y = element_blank()) +
-  coord_flip() +
-  theme(legend.position = "none")
+  coord_flip() # +
+  # theme(legend.position = "none")
 
@@
   theme(axis.text.y = element_blank()) +
-  coord_flip() + 
-  theme(legend.text=element_text(size=10))
+  coord_flip()# + 
+  #theme(legend.text=element_text(size=10))
 
-grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
+#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
+
+ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
+          common.legend = TRUE, legend="bottom", widths = c(.4, .3, .3)) %>%
+grid.arrange(top = textGrob("Player-level effects",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
 
@@
 than at FO.
-We also see that Federer's performance at Wim.
-is consistently better than other tournaments in both aces and unforced errors.
+We also see that Federer's performance at Wimbledon is consistently better than other tournaments in both aces and unforced errors.
 Interestingly, an increased performance on clay is not observed for Nadal in aces, net winners, or unforced errors.
@@
 Err.) of significant variables and their interaction terms, adjusting for the other covariates.
-Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specificically at AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.
+Since we postulated that Nadal is best on clay, Federer on grass, and Williams on hard court, specifically at AO, we use the FO, Wim., and AO, respectively, as the reference variable in the model.
 For Nadal and Federer we use forward-backwards stepwise regression, using AIC as the criterion, beginning with the full model.
@@
 
-For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.
+For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in opponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.
 
@@
 We do not find evidence for a home court advantage across all players in the grand slams.
-We also find that the FO seems to be the most different out of the four grand slams, but Wim.
-also displays some unique characteristics with regards to low amounts of UEs.
+We also find that the FO seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.
 

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

@@
 
-```{r, fig.height = 3.5, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
+```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
 fixed.effects.plot = model.ests %>%
@@
 
-```{r, fig.height = 3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+```{r, fig.height = 3.4, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
 aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

--- 2018-09-16 contest/surface-type/paper-tennis.Rmd
@@
 
-2. Examine fine grained attributes such as winners, aces, and unforced errors across the four surfaces
+2. Examine tennis specific attributes such as winners, aces, and unforced errors across the four surfaces
 
@@
 
-There has been little work on quantifying the high-level effect of court surface in tennis issue (1).
+There has been little work on quantifying the high-level effect of court surface in tennis (issue 1).
 Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
@@
 
-We examine the match attributes more unique to tennis to address issue (2).
+We examine the match attributes more unique to tennis to address (issue 2).
 These attributes include winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
@@
 
-Finally, to assess how individual players differ in performance across surfaces, issue (3), we take two approaches.
+Finally, to assess how individual players differ in performance across surfaces, (issue 3), we take two approaches.
 The first is to use a hierarchical model with fixed and random effects, taking into account player and tournament attributes using the complete data set.
@@
 ```{r table, fig.pos = 'H'}
-data(gs_partial_players)
-
 top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
@@
 
-```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
+```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.pos = 'H', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
 france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
 
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}
+```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}
 
@@
 
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
+```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
 ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
@@
 We considered a variety of models incorporating effects as either fixed or random.
-That is $\pi_i = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$ (i.e.
+That is $E[Y_{\% pts.
+won}|X] = \beta X$, where $E[\pi_i|X_i, \delta_i] = \beta X_i + \delta_i$ (i.e.
 a player win), $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
@@
 
-```{r, fig.height = 3.5, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
+```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left plot) and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
 fixed.effects.plot = model.ests %>%
@@
   my_theme +
-  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
   ggtitle("Fixed effects") +
   scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "none") + 
+  theme(legend.position = "bottom", legend.text=element_text(size=8),
+        # legend.margin=margin(0, 0, 0, 0),
+        legend.margin=margin(t = 0, unit='cm'),
+        legend.box.margin=margin(-10,-10,-10,-10)) + 
   xlab("") +
@@
   ggtitle("Random effects") +
-  theme(legend.position = "bottom") +
+  theme(legend.position = "bottom", legend.text=element_text(size=8)) +
   guides(col = guide_legend(nrow=3))
@@
 grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5),
-             top = textGrob("`ind_no_ioc' model",
+             top = textGrob("Parameter estimates from logistic model",
                             gp=gpar(fontsize=20, fontfamily="serif")))
@@
 
-```{r, fig.height = 3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+```{r, fig.height = 3.3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
 aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
@@
 We do not find evidence for a home court advantage across all players in the grand slams.
-We also find that the FO seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.
+We also find that the French Open seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.
 
 Using a hierarchical model approach, we do not see strong individual effects for Williams, Federer, and Nadal in predicting wins.
-We do, however, see differences for these players in hierarchical models to predict aces, net winners, and UEs, and that these differences are largely consistent with common tennis knowledge.
+We do, however, see differences for these players in hierarchical models to predict aces, net winners, and unforced errors, and that these differences are largely consistent with common tennis knowledge.
 

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

@@
 knitr::opts_chunk$set(message = FALSE)
+
 library(deuce)
@@
 The same trend is true for the loser rank.
-This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.
+This means that it is more likely for better players to have court by court data, which may affect our analysis for models when using this more detailed yet partial data.
 Additionally, we found that no women's point by point data was recorded for 2015.
@@
 
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}
+```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} Spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower than their ranking at other tournaments."}
 
@@
 
-```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
+```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wimbledon matches tend to have more aces and fewer unforced errors than the other tournaments."}
 ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
@@
 
-```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left plot) and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
+```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left) and player-level effects  for Williams, Federer and Nadal (right) under the ind_no_ioc model."}
 fixed.effects.plot = model.ests %>%
@@
 
-```{r}
+```{r, fig.pos = 'H'}
 knitr::kable(round(cor(ranef(ind_logistic_noioc)$name_int),2), format = "latex",
@@
 
-```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}", eval = FALSE}
-
-g <- grid.arrange(gn, gf, gw, ncol = 3, 
-              top = textGrob("Standardized Residuals vs. Predicted % Points Won",
-                            gp=gpar(fontsize=20, fontfamily="serif")))
-
-```
-
 For Williams, the model with the lowest AIC selected 22 coefficients, which makes it the largest model of the three.
@@
 
-Individual linear models predicting percent of points won show that Nadal was best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
+Individual linear models predicting percent of points won show that Nadal is best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
 

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
equation updates

@@
   - \usepackage{hyperref}
-abstract: "The Australian Open, The French Open, Wimbledon, and The US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.
+abstract: "The Australian Open, French Open, Wimbledon, and US Open are the four most prestigious events (grand slams) in tennis and are played on four different surfaces: Plexicushion (hard court), clay, grass, and DecoTurf (hard court), respectively.
 It is often thought that some players achieve better results at some grand slams than others and that difference in results may be attributed, at least partially, to the difference in the surfaces.
@@
 We considered a variety of models incorporating effects as either fixed or random.
-That is $E[Y_{\% pts.
-won}|X] = \beta X$, where $E[\pi_i|X_i, \delta_i] = \beta X_i + \delta_i$ (i.e.
-a player win), $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
+That is $E[\pi_i|X_i, \delta_i] = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$, $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years.
 Based on AIC, we found the model that included a player-level term for each tournament, but excluded fixed terms for country, was the best fit for the data.
@@
 Instead of using `win' as the outcome variable, we now look at match statistics that may be more sensitive to court surface -- aces, net points won, and UEs -- and see if individual differences among players are detectable.
-We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
+We fit three additional models, each of the form $E[Y_i|X_i, \delta_i] = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
 
@@
 We estimate the percent of points won in a match using forward-backwards stepwise regression.
-The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),
-$$
-E[Y_{\% pts. won}|X] = \beta_0 + \beta_1 X_{opp. rank} + \beta_{2, AO}\mathcal{I}(X_{sur.} = AO)  + \beta_{2, Wim.}\mathcal{I}(X_{sur.} = Wim.) + \beta_{2, USO}\mathcal{I}(X_{sur.} = USO)  .
-$$  
+The lower model is the percent of points regressed on the opponent rank and indicator variables for surface with one court as the reference variable (in the below equation, the reference surface as FO),$E[Y_{\% pts.
+won}|X] = \beta X$.
+
 The full model is the lower model with the additional variables of winner to unforced error ratio (W/UE), average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
Final edits

@@
 
-The four tennis grand slams ( Australian Open (AO), French Open (FO),  Wimbledon (Wim.), US Open (USO)) are played in different cities (Melbourne, Paris, London, New York City) at different times of year (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players for both men (ATP) and women (WTA).
+The four tennis grand slams (Australian Open (AO), French Open (FO),  Wimbledon (Wim.), US Open (USO)) are played in different cities (Melbourne, Paris, London, New York City) at different times of the year (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players for both men (ATP) and women (WTA).
 All four slams are played on different surfaces (Plexicushion hard court, clay, grass, DecoTurf hard court).
-It is often thought that top players achieve better results on some surfaces than others.
-For example, there is the idea of "home court advantage" in which players from the country of the grand slam tournament are thought to perform better than at other grand slams.
-Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
+It is often thought that top players perform better on certain surfaces.
+Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018].
+In addition, there is the idea of "home court advantage" in which players from the country of the grand slam tournament are thought to perform better at home than at other grand slams.
 
@@
 
-Finally, to assess how individual players differ in performance across surfaces, (issue 3), we take two approaches.
+Finally, to assess how individual players differ in performance across surfaces (issue 3), we take two approaches.
 The first is to use a hierarchical model with fixed and random effects, taking into account player and tournament attributes using the complete data set.
@@
 Few academic papers have been written about Nadal, Federer, or Williams.
-One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers [@leitner2009].
-One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].
+@leitner2009 study Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers .
+@five2015 shows how Williams has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type.
 
@@
 The rest of this paper is organized as follows.
-In [Section Data](#sec:data) we describe our grand slam tennis data.
-In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering methods to determine how the courts differ from one another.
-In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.
-In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
-Finally in [Section Discussion](#discussion), we summarize our results and discuss future work.
+In Section 2 we describe the grand slam tennis data and examine it at a high level.
+In Section 2.2 we examine the data at a high level and use clustering methods to determine how the courts differ from one another.
+In Section 3 we describe our models with (1) complete data with few features and (2) partial data with more features along with the results.
+Finally in Section 4, we summarize our results and discuss future work.
 
@@
 
-## Early Data Analysis {#sec:eda}
-### Examining the distribution of points earned
+## Examining the distribution of points earned
 
@@
 
-### Home court advantage
+## Home court advantage
 
@@
 ```
+
+\newpage
 
@@
 
-For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in opponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wi compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.
+For the significant variables ($\alpha$ = 95$\%$), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)` \pm `r signif(2 * sqrt(cov_w[5,5]), 2)`$ for a 1 unit increase in opponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[11,11] + 2 * cov_w[6,11]), 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)` \pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[12,12] + 2 * cov_w[6,12]), 2)`$ for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2) `\pm `r signif(2 * sqrt(cov_w[6,6] + cov_w[13,13] + 2 * cov_w[6,13]), 2)`$ for a one unit increase in W/UE at Wim.
+compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`\pm `r signif(.02 * sqrt(cov_w[8,8] + cov_w[17,17] + 2 * cov_w[17,8]), 2)`$ for a .01 increase in percent of aces at the FO; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)` \pm `r signif(.02 * sqrt(cov_w[9,9] + cov_w[20,20] + 2 * cov_w[20,9]), 2)`$ for a .01 increase in percent of break points won at the FO compared to AO.
 
@@
 
-At the highest level, we find that the general point distribution seems to be the same for all court surfaces.
+We find that the general point distribution seems to be the same for all court surfaces.
 There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g.
@@
 We do not find evidence for a home court advantage across all players in the grand slams.
-We also find that the French Open seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.
+We also find that the French Open seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of unforced errors.
 
@@
 
-Individual linear models predicting percent of points won show that Nadal is best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
+Individual linear models show that Nadal is best on clay, Federer is best on grass, and Williams is strong in all grand slams, which also agrees with prior conceptions.
+These models also show that features such as aces and break points are important.
 
-Throughout our analyses, we often found that we are restricted by available data.
+We often found that our analyses are restricted by available data.
 Due to the large amount of grand slam data available, we are able to fit individual models for Federer, Nadal, and Williams that provided substantially more information about their performance on different court surfaces than the more general models which included all players.
 Point by point data from matches among lower-profile players is more likely to be missing or contain errors.
-Ideally, these types of models would be possible for all players, not just the legends of the game.
+Ideally, these types of models would be possible for all players, not just those with many grand slam appearances.
 
-Looking forward, we would like to better explore the point by point data, especially for the attributes which excluded in this analysis: rally length, minutes played, and distance covered.
-We would also like to pull in data from Masters level tournaments, which would give us a clearer view if the differences we found in this paper are due to court surface or the slam itself.
-Another way we could adjust for grand slam confounders would be to incorporate weather conditions and matches played in the last few months.
+Looking forward, we would like to better explore the point by point data, especially for the attributes which were excluded in this analysis: rally length, minutes played, and distance covered.
+We would also like to include data from Masters level tournaments, which would give us a clearer view if the differences we found in this paper are due to court surface or the slam itself.
+Another way we could adjust for grand slam confounders by incorporating weather conditions and matches played in the last few months.
 Additionally, we would like to be able to adjust for performance over time as new competitors arrive and older players decline.

### 2019-02-20 contest/surface-type/paper-tennis.Rmd
final final commit
made a new graph for paper

@@
 In Section 2 we describe the grand slam tennis data and examine it at a high level.
-In Section 2.2 we examine the data at a high level and use clustering methods to determine how the courts differ from one another.
 In Section 3 we describe our models with (1) complete data with few features and (2) partial data with more features along with the results.
@@
 
+```{r nadal-graph}
+ ## NADAL COEF SIZE
+
+df_n <- data.frame(coeff = names(step_model_n$coeff), ave = step_model_n$coeff, sd = sqrt(diag(vcov(step_model_n))))
+df_n$player <- "Nadal"
+```
+
 ## Federer: Wimbledon extraordinaire: confirmed
@@
 cov_f <- vcov(step_model_f)
+```
+
+```{r federer-graph}
+
+ ## federer
+
+df_f <- data.frame(coeff = names(step_model_f$coeff), ave = step_model_f$coeff, sd = sqrt(diag(vcov(step_model_f))))
+df_f$player <- "Federer"
 ```
@@
 
+```{r williams-graph}
+ ## WilliamsCOEF SIZE
+
+df_w <- data.frame(coeff = names(step_model_w$coeff), ave = step_model_w$coeff, sd = sqrt(diag(vcov(step_model_w))))
+df_w$player <- "Williams"
+```
+
 For Williams, the model with the lowest AIC selected 22 coefficients, which makes it the largest model of the three.
@@
 Based on the model results, we do not see that Williams is dominant on the hard courts compared to the other courts.  Rather, she has respectable results on all the surfaces.
+
+```{r plot-individual-effects}
+df <- rbind(df_n, df_w, df_f)
+keep_coefs <- which(df$coeff %in% c("(Intercept)", "opponent_rank",
+                                    "pct_ace", "pct_bp", "pct_netpt",
+                                    "ave_serve_speed", "wue_ratio"))
+df <- df[keep_coefs,]
+ggplot(df, aes(x = coeff, y = ave, col = player)) +
+  geom_errorbar(aes(ymin = ave - 2 *sd, ymax = ave + 2 * sd), position = "dodge", width = .3) + 
+   my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = .4) + 
+  ggtitle("Effects on Reference Court") +
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "bottom", legend.text=element_text(size=8),
+        # legend.margin=margin(0, 0, 0, 0),
+        legend.margin=margin(t = 0, unit='cm'),
+        legend.box.margin=margin(-10,-10,-10,-10)) + 
+  xlab("") + coord_flip() + 
+  ylab("Estimate") +
+  scale_x_discrete(labels = c("Intercept", "Opp. Rank", "% Aces", "% Break Points Won", "% Net Points Won", "Ave. Serve Speed", "Winner:UE"))
+  guides(col = guide_legend(nrow=2))
+  #ggsave("~/Documents/ref-courts-eff.pdf", width = 8, height = 6)
+```
+
+```{r}
+df <- rbind(df_n, df_w, df_f)
+keep_coefs <- which(df$coeff %in% c("(Intercept)", "opponent_rank",
+                                    "pct_ace", "pct_bp", "pct_netpt",
+                                    "ave_serve_speed", "wue_ratio"))
+  ## Interaction Effects
+df2 <- df[-keep_coefs,]
+ggplot(df2, aes(x = coeff, y = ave, col = player)) +
+  geom_errorbar(aes(ymin = ave - 2 *sd, ymax = ave + 2 * sd), position = "dodge", width = .3) + 
+   my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
+  ggtitle("Interaction Effects") +
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "bottom", legend.text=element_text(size=8),
+        # legend.margin=margin(0, 0, 0, 0),
+        legend.margin=margin(t = 0, unit='cm'),
+        legend.box.margin=margin(-10,-10,-10,-10)) + 
+  xlab("") +
+  ylab("Estimate") + facet_grid(~player) + coord_flip() + 
+  #scale_x_discrete(labels = c("Intercept", "Opp. Rank", "% Aces", "% Break Points Won", "% Net Points Won", "Ave. Serve Speed", "Winner:UE"))
+  guides(col = guide_legend(nrow=2))
+```
 

### 2019-02-21 contest/graphics/graphics.Rmd
add 2/21 10:30 am version of graphics

@@
 data("gs_players")
-data("gs_partial_pbp")
+data("gs_partial")
 
@@
 
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
 ```
@@
 ```{r, fig.height=5, fig.width=12}
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
 
@@
 ```{r, fig.height=4, fig.width=12}
+
+tournament_colors <- c("#0297DB", "#b06835", "#ffd500", "#54008b")
+
 france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
        x = "", y = "Win proportion") + 
+  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme
+
+p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  geom_line() + 
+  # geom_point() +
+  labs(title = "British players",
+       x = "", y = "Win proportion") + 
   scale_color_manual(" ", values=tournament_colors) + my_theme
 
-grid.arrange(p1,p2, p3, nrow=1, widths = c(.29, .29, .42),
+grid.arrange(p3, p1,p2,  p4, nrow=1, widths = c(.22, .22, .22, .34),
              top = textGrob("Home court advantage debunked: win proportion by country",
@@
 ```{r, fig.height=5, fig.width=12}
+
+tournament_colors <- c("#0297DB", "#b06835", "#ffd500", "#54008b")
 
@@
                            gp=gpar(fontfamily="serif")))
+
+```
+
+## Follow-up
+
+```{r, fig.width=10, fig.height=10}
+
+p_atp <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament)) + 
+  scale_color_manual(guide=FALSE, values = tournament_colors) + 
+  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
+  geom_label_repel(data=subset(gs_players %>% filter(league == "ATP"), (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.6, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(gs_players, (log(ht+1) > 5.25 & log(ace+1) > 3.9) |
+                                (log(ht+1) > 5.28 & log(ace+1) < 1) | 
+                                (log(ht+1) < 5.2 & log(ace+1) > 3)),
+            aes(log(ht+1), log(ace+1), label = name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "ATP")+
+  my_theme
+
+p_wta <- gs_players %>% filter(league == "WTA") %>%
+ggplot(aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament)) + 
+  scale_color_manual("Tournament", values = tournament_colors) + 
+  geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.6, seed=1234, show.legend=FALSE) + 
+  geom_label_repel(data=subset(filter(gs_players, league == "WTA"),
+                               (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+                                (log(ht+1) > 5.225 & log(ace+1) < .5) | 
+                                (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+                                (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+                                 (log(ace + 1) > 2.82)),
+            aes(log(ht+1), log(ace+1), label=name),
+            seed=1234, fill=NA, show.legend=FALSE) +
+  labs(x = "log(height + 1)", title = "WTA")+
+  my_theme + theme(legend.position="bottom")
+
+grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
+             top = textGrob("Taller players are better servers: confirmed",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
+
+```
+
+# dont even know
+
+We find that the distribution of errors and aces differ by tournament. Specifically, GAMES? in Wimbledon follow different ace and error patterns than the other tournaments.
+
+```{r, fig.height=3, fig.width=4}
+ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
+                           label = tournament, color=tournament)) + 
+  geom_jitter() +
+  labs( title = "Distribution of errors and aces differ by tournmanet",
+        x = "Total unforced errors", y = "Total aces") +
+  my_theme + scale_color_manual("Tournament", values = tournament_colors)
+
+```
+
+# 5 matches to 3
+
+```{r}
+gs_men <- filter(gs, league == "ATP") %>% 
+  rowwise() %>%
+  mutate(winner_wins_after_3 = sum(W1 > L1,
+                                 W2 > L2,
+                                 W3 > L3),
+         winner_after_3 = if((winner_wins_after_3 >= 2) & 
+                             !is.na(winner_wins_after_3)) winner_name else
+                               if(is.na(winner_wins_after_3)) NA else loser_name,
+         winner_same = ifelse((winner_after_3 == winner_name) & !is.na(winner_after_3),
+                              "Yes", ifelse(is.na(winner_after_3), NA, "No")))
+
+table(gs_men$winner_wins_after_3, useNA = "always")
+table(is.na(gs_men$winner_after_3))
+table(gs_men$winner_same)
+
+prop.table(table(gs_men$winner_same))
+
+ggplot(gs_men) + geom_density(aes(winner_rank - loser_rank, color = factor(winner_same)))
+
+```
+
+# MDS
+
+We are interested in how the tournaments differ in terms of number of winners, number of unforced errors, and number of aces for both players within a MATCH???. We
+
+visualize the distance between tournaments
+
+```{r, fig.height=4, fig.width=10, eval=FALSE}
+# n_winners_p1, n_winners_p2,
+gs_num <- ungroup(filter(gs_partial_pbp, Tour == "atp")) %>%
+  select( n_ue_p1, n_ue_p2,
+         n_aces_p1, n_aces_p2) %>%
+  na.omit()
+
+mds <- cmdscale(dist(scale(gs_num)), k = 2) %>%
+  as.data.frame()
+
+# mds <- cmdscale(dist(gs_num), k = 2) %>%
+#   as.data.frame() 
+mds$tournament <- ungroup(filter(gs_partial_pbp, Tour == "atp")) %>%
+  select(n_winners_p1, n_winners_p2, n_ue_p1, n_ue_p2, 
+         n_aces_p1, n_aces_p2, tournament) %>% 
+  na.omit() %>% pull( tournament)
+
+p_mds <- ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + 
+  geom_jitter() +
+  labs( title = "Original") +
+  my_theme + scale_color_manual(guide=FALSE, values = tournament_colors)
+
+ggplot(gs_partial_pbp, aes(n_ue_p1 + n_ue_p2, n_aces_p1+ n_aces_p2,
+                           label = tournament, color=tournament)) + 
+  geom_jitter() +
+  labs( title = "Original") +
+  my_theme + scale_color_manual( values = tournament_colors)
+
+p_mds_zoom <- ggplot(mds, aes(V1, V2, label = tournament, color=tournament)) + 
+  geom_jitter() + xlim(c(-1, 2)) + ylim(c(-2, 1.5)) +
+  labs( title = "Zoomed") +
+  my_theme + scale_color_manual("Tournament", values = tournament_colors)
+
+grid.arrange(p_mds, p_mds_zoom, nrow=1, widths = c(.42, .58),
+             top = textGrob("MDS plot with: aces, winners, and unforced errors",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
 
@@
 
-## Follow-up
-
-```{r, fig.height=3, fig.width=4}
-ggplot(gs_players) + geom_point(aes(log(ht+1), log(ace+1), color = factor(did_win)))
-```
-
-# MDS
-
 ```{r, eval=FALSE}
@@
 ```
-
-```

### 2019-02-21 contest/graphics/graphics_for_paper.Rmd
add 2/21 10:30 am version of graphics

---
title: "Opening up the court (surface) in tennis grand slams"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{hyperref}
abstract: "IDK writing is hard"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message = FALSE)
library(deuce)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(stringr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)
library(MASS)
library(ggrepel)
library(broom)
library(lme4)
library(cowplot)
library(ggpubr)

## To change into library(courtsports)
devtools::load_all("../..")

```

```{r data-cleaning, include = FALSE}
data(gs)
data(gs_players)
data(gs_partial)
data(gs_partial_players)
```

```{r graphing-parameters, include = FALSE}
my_theme <-  theme_bw() + # White background, black and white theme
  theme(axis.text = element_text(size = 12),
        text = element_text(size = 14,
        family="serif"),
        plot.title = element_text(hjust = 0.5, size=16))

tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
league_colors <- c("#0C2340", "#902BA3")
win_colors <- c("#336699", "#339966")

# with yellow for us
tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")

```

```{r modelling, eval = FALSE}
## HIERARCHICAL MODELLING
## NOT RUN
## WARNING: Takes about 1 HOUR

gs_players$name_int = as.integer(as.factor(gs_players$name))
gs_players$hand_fac = as.factor(gs_players$hand)
gs_players$ioc_fac = as.factor(gs_players$ioc)
gs_players$atp = gs_players$league == "ATP"
gs_players$year_fac = as.factor(gs_players$year)
gs_players$late_round = gs_players$round >= "R16"
gs_players$seeded = gs_players$rank <= 32
gs_players$opponent_seeded = gs_players$opponent_rank <= 32

set.seed(091418)
nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)

```

# Introduction {#sec:intro}

Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open (FO), which is played on a clay surface [@bbc2018].  In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon (Wim.), which is played on grass.  On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open (AO) and 6 at US Open (USO)) and grass (7 at Wim.).  These three players have different playstyles and origins, and they are legends in the sport of tennis.  However, their success at grand slams seems to differ among the tournaments, which leads to question why that may be the case.

The four grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different starting dates (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.  As a feature unique to tennis, all four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).  It is often thought that top players achieve better results on some surfaces than others.  For example, there is the "home court advantage:" do players from the country of the grand slam tournament perform better than at other grand slams?  Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.

 In this paper, we analyze the results of grand slam players from 2013-2017, and we analyze the surface effect on players at the four grand slams, first broadly at a high level and then narrowing down to player effects.  More specifically,

1. Determine if results in terms of wins/losses and point distribution differ across the four surfaces

2. Examine more fine grained attributes such as winners, aces, and unforced errors across the four surfaces

3. Assess how individuals differ across the four surfaces, with special focus on Nadal, Federer, and Williams.

As to issue (1) quantifying the high-level effect of court surface on players, there has not been much written about with regards to tennis.  There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].   [@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.  Other sports analyses do take into account surface type such as grass vs. turf in soccer and football.  Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].

For issue (2) we examine the match attributes more unique to tennis such as winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).  Specifically, serving is thought to be of greater importance at Wimbledon since the ball travels faster on grass.  In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].

Finally, for issue (3) the player analyses, we take two approaches.  On one hand, we use a hierarchical model with fixed and random effects taking into account player and tournament attributes using the complete data set and to fairly compare players to one another.  This allows us to leverage tournament information to players without many matches played and to help account for noise.  On the other hand, since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.  In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.  We also examine whether these players do have surface apparent effects.

We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].  For our hierarchical model across all players, we model whether a player or lost a match based on player attributes along with opponent rank.  For the individual models, we instead model percent of points won which can better show how dominant a player is in a match.  For model selection, we use AIC to compare models from one another [@wasserman2004].  For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.  However, when the data becomes available, we would like to test these models on the 2018 grand slam results.

Few academic papers have been written about Nadal, Federer, or Williams.  One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers [@leitner2009].  One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].

Readers may object that we are looking at differences between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.  However, (1) grand slam data is the most readily available and most complete which makes it the best choice at the moment for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.

The rest of this paper is organized as follows. In [Section Data](#sec:data) we describe our grand slam tennis data.  In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.  In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.  In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.  Finally in [Section Discussion](#discussion), we discuss future work and extensions or our models.

# Data and EDA {#sec:data-eda}

## Data {#sec:data}

The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).  There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two, leagues, and 7 rounds.  Each match has `r ncol(gs)` attributes, many of which are redundant.  We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.  Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.  A subset of the data is shown in Table \ref{tab:data}.

```{r tab-data, include = TRUE, echo = FALSE}
w_id <- which(gs$winner_name == "Serena Williams")[1:2]
f_id <- which(gs$winner_name == "Roger Federer")[1:2]
n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
             caption = "\\label{tab:data}Example of the complete grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
             booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
```

The secondary data is partial point by point data for grand slam matches.  In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.  There is also tournament info such as surface, year, and time start.  Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.  However, these additional attributes are only available for about 1/8 of the data.

We aggregate the partial point by point data into partial match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.  We then join the partial match data with the complete data to find the score and outcome of the match.  We do this by joining the features of tournament, year, player one, and player two.  This final data set consists of `r nrow(gs_pbp)` observations (compared to the 5080 in the original grand slams).  The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.  The same trend is true for the loser rank.  This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.  Additionally, we found that no women's point by point data was recorded for 2015.

All data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.

## Early Data Analysis {#sec:eda}

### Examining the distribution of points earned

We first examine the distribution of points earned per match to assess normality. The distribution of points earned is approximately normal. This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams. As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played. Also unsurprisingly, the winners of the match tended to earn more points than the losers.

```{r, fig.height=5, fig.width=10}

# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))

p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
  labs(x = "Points") + my_theme +
  scale_color_manual("Tournament", values=tournament_colors)

p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
  labs(x = "Points") +
  scale_color_manual("League", values=league_colors) +
  my_theme

# grid.arrange( p2, p3, ncol=2,
#              top = textGrob("Winning points by tournament and league",
#                             gp=gpar(fontsize=20)))

p1 <- ggplot(gs_players) + geom_density(aes(pointswon, color = factor(did_win))) + 
  labs(x = "Points") +
  scale_color_manual("Winner?", values=win_colors) +
  my_theme +
  facet_wrap(~league, ncol=1)

grid.arrange(p1, p2, ncol=2,
             top = textGrob("Distribution of points per match",
                            gp=gpar(fontsize=20, fontfamily="serif")),
              layout_matrix = rbind(c(NA, 1),
                                    c(2, 1),
                                    c(NA, 1)),
             heights=c(1, 3, 1))

```

### Home court advantage

It is commonly thought that there is a home court advantage in grand slam games (SOURCE). In our data we find this to be true (i.e. French players win the French open more than French players win other slams). But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments. We also find this to be true in our data for France, The United States, Australia, and Great Britain (i.e. the proportion of French players in the French Open is greater than the proportion of French players in other slams).

Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments. If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon. The same would be true for Australia and the US. But, we see in \ref{fig:home-court} that this isn't the case. After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.

```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

us_wins <- filter(gs_players, ioc == "USA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

p1<- ggplot(france_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "French players",
       x = "", y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme

p2 <- ggplot(us_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "American players",
       x = "", y = "Win proportion") +  
  scale_color_manual(" ", values=tournament_colors)  + my_theme

p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "Australian players",
       x = "", y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme

p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "British players",
       x = "", y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme

ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
grid.arrange(., top = textGrob("Home court advantage: debunked",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           just = "left",
                           gp=gpar(fontfamily="serif")))

ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
  geom_line()

ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
  geom_point()

filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
ggplot( aes(x = ( opponent_rank - rank), y = (pointswon - (total_points - pointswon)),
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE)

p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "French players",
      y = "Win proportion",
      x = " ") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "WTA") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Australian players",
      y = "Win proportion",
      x = " ") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "WTA") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "American players",
      y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "WTA") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "English players",
      y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

# grid.arrange(p1, p2, p3, p4, nrow=2)

ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
grid.arrange(., top = textGrob("WTA",
                            gp=gpar(fontsize=30, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           just = "left",
                           gp=gpar(fontfamily="serif"))) %>%
ggsave("plots/wta.jpg", ., width = 10, height = 8)

p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "ATP") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "French players",
      y = "Win proportion",
      x = " ") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "ATP") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Australian players",
      y = "Win proportion",
      x = " ") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "ATP") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "American players",
      y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "ATP") %>%
ggplot( aes(x = ( opponent_rank - rank), y = did_win,
            color = tournament)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "English players",
      y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme +
  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
                                            labels =c("No", "Yes"),
                                            limits = c(0, 1))

ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
grid.arrange(., top = textGrob("ATP",
                            gp=gpar(fontsize=30, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           just = "left",
                           gp=gpar(fontfamily="serif"))) %>%
ggsave("plots/atp.jpg", ., width = 10, height = 8)

```

## Spaniards on clay

```{r}

spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
            avg_ranking = mean(rank),
            rank_se = sd(rank)/sqrt(n()))

prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
                table(did_win, tournament)), 2)

with(filter(gs_players, ioc == "ESP", round == "R128"),
                table(did_win, tournament))

with(filter(gs_players, ioc == "ESP", round == "R128"),
     table(did_win, tournament)) %>%
  chisq.test(correct = TRUE)

```

It is also commonly thought that Spaniards play better on clay. We are interested in whether Spaniards win the French Open more than they win in other tournaments. It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments. But, this result is not significant (ENTER P VALUE). Interestingly, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse than Spanish players in ther tournaments.

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}

my_theme <-  theme_bw() + # White background, black and white theme
  theme(axis.text = element_text(size = 24),
        text = element_text(size = 28,
        family="serif"),
        plot.title = element_text(hjust = 0.5, size=34))

p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line(size=3) + 
  labs(title = "Win proportion among Spaniards",
       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
       x = "", y = "Win proportion") + 
  scale_color_manual(guide=FALSE, values=tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)",
                                           "French Open\n(Clay)",
                                           "US Open\n(DecoTurf)",
                                           "Wimbledon\n(Grass)")) + my_theme

p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament, fill = tournament)) + 
  geom_line(size=3) + 
  labs(title = "Average ranking among Spaniards",
       x = " ",
       y = "Average rank") + 
  scale_color_manual(" ", values=tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)",
                                           "French Open\n(Clay)",
                                           "US Open\n(DecoTurf)",
                                           "Wimbledon\n(Grass)")) + my_theme +
  guides(color = guide_legend(override.aes = list(size = 7)))

ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))

ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
  ggsave("plots/spainards.jpg", ., height=7, width=16)

```

# Unforced errors and aces

We find that the distribution of errors and aces differ by tournament. Specifically, GAMES? in Wimbledon follow different ace and error patterns than the other tournaments. These GAMES? tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces. It is known (SOURCE) that the ball moves faster on grass courts and so the ace result is unsuprising, but making fewer errors is unexpected, especially because players seldom practice on grass.

```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}

my_theme <-  theme_bw() + # White background, black and white theme
  theme(axis.text = element_text(size = 24),
        text = element_text(size = 28,
        family="serif"),
        plot.title = element_text(hjust = 0.5, size=34))

ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter(size=2) +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament",
                                values = tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)\n",
                                           "French Open\n(Clay)\n",
                                           "US Open\n(DecoTurf)\n",
                                           "Wimbledon\n(Grass)\n")) + 
  geom_density_2d()

ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter(size=1.5) +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament",
                                values = tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)\n",
                                           "French Open\n(Clay)\n",
                                           "US Open\n(DecoTurf)\n",
                                           "Wimbledon\n(Grass)\n")) +
  stat_ellipse(type="t") +
  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))

ggsave("plots/errors_aces2.jpg", height = 6, width = 12)

```

## Tall players and aces

Another common thought about tennis is that taller players are better servers. We find this to be true and see a strong positive correlation between teh variables. We highlight some interesting players BLAH BLAH BLAH.

```{r, fig.width=10, fig.height=4.8, fig.align='center'}

my_theme <-  theme_bw() + # White background, black and white theme
  theme(axis.text = element_text(size = 24),
        text = element_text(size = 28,
        family="serif"),
        plot.title = element_text(hjust = 0.5, size=34))

# my_theme <-  theme_bw() + # White background, black and white theme
#   theme(axis.text = element_text(size = 26),
#         text = element_text(size = 30,
#         family="serif"),
#         plot.title = element_text(hjust = 0.5, size=36))

set.seed(8)

atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
                 "Roger Federer", "Andy Murray")
atp_servers <- sapply(atp_servers,
                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
  as.numeric() %>% gs_players[., ]

wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
                 "Venus Williams", "Serena Williams", "Madison Keys")
wta_servers <- sapply(wta_servers,
                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
  as.numeric() %>% gs_players[., ]

p_atp <- gs_players %>% filter(league == "ATP") %>%
ggplot( aes(log(ht+1), log(ace+1))) + 
  geom_jitter(aes(color = tournament), size=1.5) + 
  # scale_color_manual(" ", values = tournament_colors) + 
  geom_smooth( method = "lm", color="black") + 
  geom_label_repel(data=atp_servers,
            aes(log(ht+1), log(ace+1), label=name),
            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
  geom_label_repel(data=atp_servers,
            aes(log(ht+1), log(ace+1), label = name),
            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
  labs(x = "log(height + 1)", title = "ATP") +
  my_theme + scale_color_manual("Tournament",
                                values = tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)\n",
                                           "French Open\n(Clay)\n",
                                           "US Open\n(DecoTurf)\n",
                                           "Wimbledon\n(Grass)\n")) +
  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))

p_wta <- gs_players %>% filter(league == "WTA") %>%
ggplot(aes(log(ht+1), log(ace+1))) + 
  geom_jitter(aes(color = tournament)) + 
  # scale_color_manual(" ", values = tournament_colors) + 
  geom_smooth(method = "lm", color="black") + 
  geom_label_repel(data=wta_servers,
            aes(log(ht+1), log(ace+1), label=name),
            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
  geom_label_repel(data=wta_servers,
            aes( log(ht+1), log(ace+1), label=name),
            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
  labs(x = "log(height + 1)", title = "WTA")+
  my_theme + scale_color_manual("Tournament",
                                values = tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)\n",
                                           "French Open\n(Clay)\n",
                                           "US Open\n(DecoTurf)\n",
                                           "Wimbledon\n(Grass)\n")) +
  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))

# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
#              top = textGrob("Taller players are better servers: confirmed",
#                             gp=gpar(fontsize=20, fontfamily="serif")))

ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
#                             gp=gpar(fontsize=20, fontfamily="serif")))

# ggsave("plots/aces.jpg", height=7, width=16)
ggsave("plots/aces2.jpg", height=7, width=16)

```

```{r, fig.width=6, fig.height=6, fig.align='center'}

# p_atp <- gs_players %>% filter(league == "ATP") %>%
# ggplot( aes(log(ht+1), log(ace+1))) + 
#   geom_jitter(aes(color = tournament)) + 
#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
#   geom_smooth( method = "lm", se=FALSE, color="black") + 
#   geom_label_repel(data=atp_servers,
#             aes(log(ht+1), log(ace+1), label=name),
#             alpha=.6, seed=1234, show.legend=FALSE) + 
#   geom_label_repel(data=atp_servers,
#             aes(log(ht+1), log(ace+1), label = name),
#             seed=1234, fill=NA, show.legend=FALSE) +
#   labs(x = "log(height + 1)", title = "ATP") +
#   my_theme
# 
# # subset(filter(gs_players, league == "WTA"),
# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
# #                                  (log(ace + 1) > 2.82))
# p_wta <- gs_players %>% filter(league == "WTA") %>%
# ggplot(aes(log(ht+1), log(ace+1))) + 
#   geom_jitter(aes(color = tournament)) + 
#   scale_color_manual("Tournament", values = tournament_colors) + 
#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
#   geom_label_repel(data=wta_servers,
#             aes(log(ht+1), log(ace+1), label=name),
#             alpha=.6, seed=1234, show.legend=FALSE) + 
#   geom_label_repel(data=wta_servers,
#             aes(log(ht+1), log(ace+1), label=name),
#             seed=1234, fill=NA, show.legend=FALSE) +
#   labs(x = "log(height + 1)", title = "WTA")+
#   my_theme + theme(legend.position="bottom")
# 
# 
# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
#              top = textGrob("Taller players are better servers: confirmed",
#                             gp=gpar(fontsize=20, fontfamily="serif")))

```

### Linear regression for other outcomes

Instead of using `win' as the outcome variable, we now look at match statistics that may be more sensitive to court surface -- aces, net points won, and UEs -- and see if individual differences among players are detectable. We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.

```{r, eval = FALSE}
gs_partial_players$atp = gs_partial_players$Tour == "atp"
gs_partial_players$year_fac = as.factor(gs_partial_players$year)
gs_partial_players$late_round = gs_partial_players$round >= "R16"
gs_partial_players$seeded = gs_partial_players$rank <= 32
gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
```

```{r}
load("../../data/n_aces_mod.rda")
load("../../data/n_net_mod.rda")
load("../../data/n_ue_mod.rda")
load("../../data/n_winners_mod.rda")
```

```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}

my_theme <-  theme_bw() + # White background, black and white theme
  theme(axis.text = element_text(size = 20),
        text = element_text(size = 24,
        family="serif"),
        plot.title = element_text(hjust = 0.5, size=26))

aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval,
                ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
  ylab("Coefficient\n") + 
  xlab("") +
  ggtitle("Y = aces") + 
  scale_color_manual("", values=tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)",
                                           "French Open\n(Clay)",
                                           "US Open\n(DecoTurf)",
                                           "Wimbledon\n(Grass)")) +
  my_theme +
  coord_flip() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_hline(yintercept = 0, color = "grey") +
  guides(color = guide_legend(override.aes = list(size = 7)))

# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
#   mutate(tournament = gsub("tournament", "", term)) %>%
#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
#   geom_errorbar(height = .5, position = position_dodge()) +
#   ylab("Coefficient\n") + 
#   xlab("") +
#   ggtitle("Winners") + 
#   scale_color_manual("", values=tournament_colors,
#                                 labels = c("Australian Open\n(Plexicushion)",
#                                            "French Open\n(Clay)",
#                                            "US Open\n(DecoTurf)",
#                                            "Wimbledon\n(Grass)")) +
#   my_theme  +
#   coord_flip() +
#   theme(panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank(),
#         axis.text.y = element_blank()) +
#   geom_hline(yintercept = 0, color = "grey")

nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
  ylab("Coefficient\n") + 
  xlab("") +
  ggtitle("Y = points won at net") + 
  scale_color_manual("", values=tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)",
                                           "French Open\n(Clay)",
                                           "US Open\n(DecoTurf)",
                                           "Wimbledon\n(Grass)")) +
  my_theme  +
  coord_flip() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank()) +
  geom_hline(yintercept = 0, color = "grey") +
  guides(color = guide_legend(override.aes = list(size = 7)))

ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
  ylab("Coefficient\n") + 
  xlab("") +
  ggtitle("Y = unforced errors") + 
  scale_color_manual("", values=tournament_colors,
                                labels = c("Australian Open\n(Plexicushion)",
                                           "French Open\n(Clay)",
                                           "US Open\n(DecoTurf)",
                                           "Wimbledon\n(Grass)")) +
  my_theme  +
  coord_flip() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank()) +
  geom_hline(yintercept = 0, color = "grey") +
  guides(color = guide_legend(override.aes = list(size = 7)))

#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))

ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
ggsave("plots/player_effects.jpg", ., width=14, height=5)

```

### 2019-02-21 contest/graphics/scratch.Rmd

--- 2019-02-21 contest/graphics/graphics_for_paper.Rmd
@@
 ---
-title: "Opening up the court (surface) in tennis grand slams"
-date: '`r format(Sys.time(), "%B %d, %Y")`'
-output:
-  pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
-header-includes:
-  - \usepackage{hyperref}
-abstract: "IDK writing is hard"
+title: "scratch"
+author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
+date: "September 15, 2018"
+output: html_document
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-knitr::opts_chunk$set(warning=FALSE)
-knitr::opts_chunk$set(message = FALSE)
-library(deuce)
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(forcats)
-library(stringr)
-library(knitr)
-library(kableExtra)
-library(gridExtra)
-library(grid)
-library(MASS)
-library(ggrepel)
-library(broom)
-library(lme4)
-library(cowplot)
-library(ggpubr)
-
-## To change into library(courtsports)
-devtools::load_all("../..")
-
-```
-
-```{r data-cleaning, include = FALSE}
-data(gs)
-data(gs_players)
-data(gs_partial)
-data(gs_partial_players)
-```
-
-```{r graphing-parameters, include = FALSE}
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 12),
-        text = element_text(size = 14,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=16))
-
-tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
-league_colors <- c("#0C2340", "#902BA3")
-win_colors <- c("#336699", "#339966")
-
-# with yellow for us
-tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
-
-```
-
-```{r modelling, eval = FALSE}
-## HIERARCHICAL MODELLING
-## NOT RUN
-## WARNING: Takes about 1 HOUR
-
-gs_players$name_int = as.integer(as.factor(gs_players$name))
-gs_players$hand_fac = as.factor(gs_players$hand)
-gs_players$ioc_fac = as.factor(gs_players$ioc)
-gs_players$atp = gs_players$league == "ATP"
-gs_players$year_fac = as.factor(gs_players$year)
-gs_players$late_round = gs_players$round >= "R16"
-gs_players$seeded = gs_players$rank <= 32
-gs_players$opponent_seeded = gs_players$opponent_rank <= 32
-
-set.seed(091418)
-nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
-ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
-ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-
-```
-
-# Introduction {#sec:intro}
-
-Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open (FO), which is played on a clay surface [@bbc2018].
-In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon (Wim.), which is played on grass.
-On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open (AO) and 6 at US Open (USO)) and grass (7 at Wim.).
-These three players have different playstyles and origins, and they are legends in the sport of tennis.
-However, their success at grand slams seems to differ among the tournaments, which leads to question why that may be the case.
-
-The four grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different starting dates (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
-As a feature unique to tennis, all four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
-It is often thought that top players achieve better results on some surfaces than others.
-For example, there is the "home court advantage:" do players from the country of the grand slam tournament perform better than at other grand slams?
-Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
-
- In this paper, we analyze the results of grand slam players from 2013-2017, and we analyze the surface effect on players at the four grand slams, first broadly at a high level and then narrowing down to player effects.
-More specifically,
-
-1. Determine if results in terms of wins/losses and point distribution differ across the four surfaces
-
-2. Examine more fine grained attributes such as winners, aces, and unforced errors across the four surfaces
-
-3. Assess how individuals differ across the four surfaces, with special focus on Nadal, Federer, and Williams.
-
-As to issue (1) quantifying the high-level effect of court surface on players, there has not been much written about with regards to tennis.
-There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
-Other sports analyses do take into account surface type such as grass vs.
-turf in soccer and football.
-Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
-
-For issue (2) we examine the match attributes more unique to tennis such as winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
-Specifically, serving is thought to be of greater importance at Wimbledon since the ball travels faster on grass.
-In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].
-
-Finally, for issue (3) the player analyses, we take two approaches.
-On one hand, we use a hierarchical model with fixed and random effects taking into account player and tournament attributes using the complete data set and to fairly compare players to one another.
-This allows us to leverage tournament information to players without many matches played and to help account for noise.
-On the other hand, since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
-In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
-We also examine whether these players do have surface apparent effects.
-
-We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
-For our hierarchical model across all players, we model whether a player or lost a match based on player attributes along with opponent rank.
-For the individual models, we instead model percent of points won which can better show how dominant a player is in a match.
-For model selection, we use AIC to compare models from one another [@wasserman2004].
-For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.
-However, when the data becomes available, we would like to test these models on the 2018 grand slam results.
-
-Few academic papers have been written about Nadal, Federer, or Williams.
-One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers [@leitner2009].
-One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].
-
-Readers may object that we are looking at differences between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.
-However, (1) grand slam data is the most readily available and most complete which makes it the best choice at the moment for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.
-
-The rest of this paper is organized as follows.
-In [Section Data](#sec:data) we describe our grand slam tennis data.
-In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.
-In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.
-In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
-Finally in [Section Discussion](#discussion), we discuss future work and extensions or our models.
-
-# Data and EDA {#sec:data-eda}
-
-## Data {#sec:data}
-
-The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
-There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two, leagues, and 7 rounds.
-Each match has `r ncol(gs)` attributes, many of which are redundant.
-We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
-Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.
-A subset of the data is shown in Table \ref{tab:data}.
-
-```{r tab-data, include = TRUE, echo = FALSE}
-w_id <- which(gs$winner_name == "Serena Williams")[1:2]
-f_id <- which(gs$winner_name == "Roger Federer")[1:2]
-n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
-s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
-cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
-knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
-             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
-             caption = "\\label{tab:data}Example of the complete grand slam data.
-It includes winner and loser attributes, match attributes, and tournament attributes.
-Not all attributes are shown here.",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
-```
-
-The secondary data is partial point by point data for grand slam matches.
-In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.
-There is also tournament info such as surface, year, and time start.
-Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.
-However, these additional attributes are only available for about 1/8 of the data.
-
-We aggregate the partial point by point data into partial match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
-We then join the partial match data with the complete data to find the score and outcome of the match.
-We do this by joining the features of tournament, year, player one, and player two.
-This final data set consists of `r nrow(gs_pbp)` observations (compared to the 5080 in the original grand slams).
-The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
-The same trend is true for the loser rank.
-This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.
-Additionally, we found that no women's point by point data was recorded for 2015.
-
-All data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.
-
-## Early Data Analysis {#sec:eda}
-
-### Examining the distribution of points earned
-
-We first examine the distribution of points earned per match to assess normality.
-The distribution of points earned is approximately normal.
-This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
-As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
-Also unsurprisingly, the winners of the match tended to earn more points than the losers.
-
-```{r, fig.height=5, fig.width=10}
-
-# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
-
-p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
-  labs(x = "Points") + my_theme +
-  scale_color_manual("Tournament", values=tournament_colors)
-
-p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
-  labs(x = "Points") +
-  scale_color_manual("League", values=league_colors) +
-  my_theme
-
-# grid.arrange( p2, p3, ncol=2,
-#              top = textGrob("Winning points by tournament and league",
-#                             gp=gpar(fontsize=20)))
-
-p1 <- ggplot(gs_players) + geom_density(aes(pointswon, color = factor(did_win))) + 
-  labs(x = "Points") +
-  scale_color_manual("Winner?", values=win_colors) +
-  my_theme +
-  facet_wrap(~league, ncol=1)
-
-grid.arrange(p1, p2, ncol=2,
-             top = textGrob("Distribution of points per match",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-              layout_matrix = rbind(c(NA, 1),
-                                    c(2, 1),
-                                    c(NA, 1)),
-             heights=c(1, 3, 1))
-
-```
-
-### Home court advantage
-
-It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
-In our data we find this to be true (i.e.
-French players win the French open more than French players win other slams).
-But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments.
-We also find this to be true in our data for France, The United States, Australia, and Great Britain (i.e.
-the proportion of French players in the French Open is greater than the proportion of French players in other slams).
-
-Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments.
-If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon.
-The same would be true for Australia and the US.
-But, we see in \ref{fig:home-court} that this isn't the case.
-After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
-
-```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
-france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-us_wins <- filter(gs_players, ioc == "USA", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-p1<- ggplot(france_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
-  # geom_point() +
-  labs(title = "French players",
-       x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
-
-p2 <- ggplot(us_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
-  # geom_point() +
-  labs(title = "American players",
-       x = "", y = "Win proportion") +  
-  scale_color_manual(" ", values=tournament_colors)  + my_theme
-
-p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
-  # geom_point() +
-  labs(title = "Australian players",
-       x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
-
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
-  # geom_point() +
-  labs(title = "British players",
-       x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("Home court advantage: debunked",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_line()
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_point()
-
-filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = (pointswon - (total_points - pointswon)),
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-# grid.arrange(p1, p2, p3, p4, nrow=2)
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("WTA",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/wta.jpg", ., width = 10, height = 8)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("ATP",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/atp.jpg", ., width = 10, height = 8)
-
-```
-
-## Spaniards on clay
-
-```{r}
-
-spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
-            avg_ranking = mean(rank),
-            rank_se = sd(rank)/sqrt(n()))
-
-prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament)), 2)
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament))
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-     table(did_win, tournament)) %>%
-  chisq.test(correct = TRUE)
-
-```
-
-It is also commonly thought that Spaniards play better on clay.
-We are interested in whether Spaniards win the French Open more than they win in other tournaments.
-It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
-But, this result is not significant (ENTER P VALUE).
-Interestingly, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse than Spanish players in ther tournaments.
-
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "Win proportion among Spaniards",
-       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
-       x = "", y = "Win proportion") + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme
-
-p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament, fill = tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "Average ranking among Spaniards",
-       x = " ",
-       y = "Average rank") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           # x = 0,
-                           # y = 0.5,
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-  ggsave("plots/spainards.jpg", ., height=7, width=16)
-
-```
-
-# Unforced errors and aces
-
-We find that the distribution of errors and aces differ by tournament.
-Specifically, GAMES?
-in Wimbledon follow different ace and error patterns than the other tournaments.
-These GAMES?
-tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces.
-It is known (SOURCE) that the ball moves faster on grass courts and so the ace result is unsuprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
-
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-                           label = tournament, color=tournament)) + 
-  geom_jitter(size=2) +
-  labs( title = "Distribution of errors and aces differ by tournament",
-        x = "Total unforced errors", y = "Total aces") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) + 
-  geom_density_2d()
-
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-                           label = tournament, color=tournament)) + 
-  geom_jitter(size=1.5) +
-  labs( title = "Distribution of errors and aces differ by tournament",
-        x = "Total unforced errors", y = "Total aces") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  stat_ellipse(type="t") +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-ggsave("plots/errors_aces2.jpg", height = 6, width = 12)
-
-```
-
-## Tall players and aces
+### Tall players and aces
 
@@
 ```{r, fig.width=10, fig.height=4.8, fig.align='center'}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 26),
-#         text = element_text(size = 30,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=36))
 
@@
 ggplot( aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament), size=1.5) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_jitter(aes(color = tournament)) + 
+  scale_color_manual(" ", values = tournament_colors) + 
   geom_smooth( method = "lm", color="black") + 
@@
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+            alpha=.8, seed=1234, show.legend=FALSE) + 
   geom_label_repel(data=atp_servers,
             aes(log(ht+1), log(ace+1), label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+            seed=1234, fill=NA, show.legend=FALSE) +
   labs(x = "log(height + 1)", title = "ATP") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+  my_theme
 
@@
   geom_jitter(aes(color = tournament)) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
+  scale_color_manual(" ", values = tournament_colors) + 
   geom_smooth(method = "lm", color="black") + 
@@
             aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+            alpha=.8, seed=1234, show.legend=FALSE) + 
   geom_label_repel(data=wta_servers,
             aes( log(ht+1), log(ace+1), label=name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+            seed=1234, fill=NA, show.legend=FALSE) +
   labs(x = "log(height + 1)", title = "WTA")+
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+  my_theme
 
@@
 
-ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
-# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-# ggsave("plots/aces.jpg", height=7, width=16)
-ggsave("plots/aces2.jpg", height=7, width=16)
+ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") %>%
+  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
+                            gp=gpar(fontsize=20, fontfamily="serif")))
 
 ```
-
-```{r, fig.width=6, fig.height=6, fig.align='center'}
-
-# p_atp <- gs_players %>% filter(league == "ATP") %>%
-# ggplot( aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
-#   geom_smooth( method = "lm", se=FALSE, color="black") + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label = name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "ATP") +
-#   my_theme
-# 
-# # subset(filter(gs_players, league == "WTA"),
-# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-# #                                  (log(ace + 1) > 2.82))
-# p_wta <- gs_players %>% filter(league == "WTA") %>%
-# ggplot(aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual("Tournament", values = tournament_colors) + 
-#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "WTA")+
-#   my_theme + theme(legend.position="bottom")
-# 
-# 
-# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-```
-
-### Linear regression for other outcomes
-
-Instead of using `win' as the outcome variable, we now look at match statistics that may be more sensitive to court surface -- aces, net points won, and UEs -- and see if individual differences among players are detectable.
-We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
-
-```{r, eval = FALSE}
-gs_partial_players$atp = gs_partial_players$Tour == "atp"
-gs_partial_players$year_fac = as.factor(gs_partial_players$year)
-gs_partial_players$late_round = gs_partial_players$round >= "R16"
-gs_partial_players$seeded = gs_partial_players$rank <= 32
-gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-```
-
-```{r}
-load("../../data/n_aces_mod.rda")
-load("../../data/n_net_mod.rda")
-load("../../data/n_ue_mod.rda")
-load("../../data/n_winners_mod.rda")
-```
-
-```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 20),
-        text = element_text(size = 24,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=26))
-
-aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval,
-                ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = aces") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
-#   mutate(tournament = gsub("tournament", "", term)) %>%
-#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-#   geom_errorbar(height = .5, position = position_dodge()) +
-#   ylab("Coefficient\n") + 
-#   xlab("") +
-#   ggtitle("Winners") + 
-#   scale_color_manual("", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) +
-#   my_theme  +
-#   coord_flip() +
-#   theme(panel.grid.major = element_blank(),
-#         panel.grid.minor = element_blank(),
-#         axis.text.y = element_blank()) +
-#   geom_hline(yintercept = 0, color = "grey")
-
-nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = points won at net") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = unforced errors") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
-ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
-          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
-ggsave("plots/player_effects.jpg", ., width=14, height=5)
-
-```

### 2019-02-21 contest/graphics/make-model-figures.R
new figures for mixed effects models

@@ -1,0 +1,224 @@
+library(courtsports)
+library(tidyr)
+library(broom)
+library(lme4)
+library(ggpubr)
+library(gridExtra)
+
+load("../../data/gs_players.rda")
+load("../../data/gs_partial_players.rda")
+load("../../data/nocourt_logistic.rda")
+load("../../data/base_logistic.rda")
+load("../../data/country_logistic.rda")
+load("../../data/ind_int_logistic.rda")
+load("../../data/ind_logistic.rda")
+load("../../data/ind_logistic_noioc.rda")
+load("../../data/ind_year_logistic.rda")
+
+load("../../data/n_aces_mod.rda")
+load("../../data/n_net_mod.rda")
+load("../../data/n_ue_mod.rda")
+load("../../data/n_winners_mod.rda")
+
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+                            family="serif"),
+        plot.title = element_text(hjust = 0.5, size=16))
+
+tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
+league_colors <- c("#0C2340", "#902BA3")
+win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
+model.ests = data.frame()
+ran.effects = data.frame()
+for(mod in 1:nrow(model.sums)){
+  name = model_names[mod]
+  model.sums$aic[mod] = extractAIC(get(name))[2]
+  model.sums$edf[mod] = extractAIC(get(name))[1]
+  betas = tidy(get(name))
+  betas$model = name
+  if("group" %in% colnames(betas)){
+    model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+                         dplyr::select(., -c("group")))
+  }
+  else model.ests = rbind(model.ests, betas)
+}
+
[…]

### 2019-02-22 contest/graphics/graphics_for_paper.Rmd
make all graphs consistent for paper submission

@@
 ```{r graphing-parameters, include = FALSE}
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 12),
+#         text = element_text(size = 14,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=16))
+
 my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 12),
-        text = element_text(size = 14,
+  theme(axis.text = element_text(size = 24),
+        text = element_text(size = 28,
         family="serif"),
-        plot.title = element_text(hjust = 0.5, size=16))
+        plot.title = element_text(hjust = 0.5, size=34))
 
@@
 tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+graphic_width_long <- 16
+graphic_width_short <- 12
 
@@
 
-# Introduction {#sec:intro}
-
-Rafael Nadal is known as the "King of Clay" in tennis, having won 11 out of his current 17 grand slams titles at the French Open (FO), which is played on a clay surface [@bbc2018].
-In contrast, his rival Roger Federer has won his most grand slam titles (8 out of 20) at Wimbledon (Wim.), which is played on grass.
-On the women's side, Serena Williams, winner of 24 grand slam titles, has been dominant both on hard court (7 titles at the Australian Open (AO) and 6 at US Open (USO)) and grass (7 at Wim.).
-These three players have different playstyles and origins, and they are legends in the sport of tennis.
-However, their success at grand slams seems to differ among the tournaments, which leads to question why that may be the case.
-
-The four grand slams (AO, FO, Wim., USO) are played in different cities (Melbourne, Paris, London, New York City) at different starting dates (January, May, June, August) over the course of two weeks with 7 total rounds and 128 players.
-As a feature unique to tennis, all four slams are played on different surfaces (Plexicushion hardcourt, clay, grass, DecoTurf hardcourt).
-It is often thought that top players achieve better results on some surfaces than others.
-For example, there is the "home court advantage:" do players from the country of the grand slam tournament perform better than at other grand slams?
-Recently, Spanish players, led by Nadal, seem to have dominated the clay court scene [@tp2018] making it seem like players of a certain country may perform better on some surfaces than others.
-
- In this paper, we analyze the results of grand slam players from 2013-2017, and we analyze the surface effect on players at the four grand slams, first broadly at a high level and then narrowing down to player effects.
-More specifically,
-
-1. Determine if results in terms of wins/losses and point distribution differ across the four surfaces
-
-2. Examine more fine grained attributes such as winners, aces, and unforced errors across the four surfaces
-
-3. Assess how individuals differ across the four surfaces, with special focus on Nadal, Federer, and Williams.
-
-As to issue (1) quantifying the high-level effect of court surface on players, there has not been much written about with regards to tennis.
-There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].
-[@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another and [@pollard2007] concludes that players have more injuries on grass comparead to other courts.
-Other sports analyses do take into account surface type such as grass vs.
-turf in soccer and football.
-Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].
-
-For issue (2) we examine the match attributes more unique to tennis such as winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).
-Specifically, serving is thought to be of greater importance at Wimbledon since the ball travels faster on grass.
-In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].
-
-Finally, for issue (3) the player analyses, we take two approaches.
-On one hand, we use a hierarchical model with fixed and random effects taking into account player and tournament attributes using the complete data set and to fairly compare players to one another.
-This allows us to leverage tournament information to players without many matches played and to help account for noise.
-On the other hand, since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with feature attributes to examine the effects individually.
-In both cases, we examine whether our models pass "common sense" tests like how the models in [@thomas2013] show that commonly well known hockey players also have high status in the model.
-We also examine whether these players do have surface apparent effects.
-
-We use models that take into account both individual and group effects such as in the Gaussian-process player production basketball model or predicting individual soccer performance  [@page2013; @egidi2018].
-For our hierarchical model across all players, we model whether a player or lost a match based on player attributes along with opponent rank.
-For the individual models, we instead model percent of points won which can better show how dominant a player is in a match.
-For model selection, we use AIC to compare models from one another [@wasserman2004].
-For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.
-However, when the data becomes available, we would like to test these models on the 2018 grand slam results.
-
-Few academic papers have been written about Nadal, Federer, or Williams.
-One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon and shows that Federer was too heavily favored by bookmakers [@leitner2009].
-One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].
-
-Readers may object that we are looking at differences between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.
-However, (1) grand slam data is the most readily available and most complete which makes it the best choice at the moment for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.
-
-The rest of this paper is organized as follows.
-In [Section Data](#sec:data) we describe our grand slam tennis data.
-In [Section Early Data Analysis](#sec:eda) we examine the data at a high level and use clustering whether to determine how the courts differ from one another.
-In [Section Methods](#sec:methods) we describe our models with (1) complete data with few features and (2) partial data with more features.
-In [Section Results](#sec:results) we describe the results of our modelling and also examine the play of Nadal, Federer, and Williams.
-Finally in [Section Discussion](#discussion), we discuss future work and extensions or our models.
-
 # Data and EDA {#sec:data-eda}
@@
 ## Data {#sec:data}
-
-The primary data consists of `r nrow(gs)` matches split evenly over the four grand slams and the two leagues: ATP (men's) and WTA (women's).
-There are 5080 observations in this data set, which corresponds to a complete data set for the period of 2013-2017 for four grand slams, two, leagues, and 7 rounds.
-Each match has `r ncol(gs)` attributes, many of which are redundant.
-We focus on the following attributes for both the winner and loser of the match: games won, points won, retirement, break points faced, break points saved, aces, country of origin, and player attributes.
-Additionally, we take into account the number of sets in a match, the surface type, and round of the tournament.
-A subset of the data is shown in Table \ref{tab:data}.
 
@@
 
-The secondary data is partial point by point data for grand slam matches.
-In this data set, each row is a point in a match with details on who won the point, whether a player had a forced or unforced error, winner, ace, or net point win, and serve speed.
-There is also tournament info such as surface, year, and time start.
-Additional attributes include rally length, winner and final score of the match, retirement, and minutes played.
-However, these additional attributes are only available for about 1/8 of the data.
-
-We aggregate the partial point by point data into partial match data, summing the total number of errors, winners, aces, net points, and taking the average service speed for each player in the match.
-We then join the partial match data with the complete data to find the score and outcome of the match.
-We do this by joining the features of tournament, year, player one, and player two.
-This final data set consists of `r nrow(gs_pbp)` observations (compared to the 5080 in the original grand slams).
-The median rank of the winners for the original data is `r median(gs$winner_rank, na.rm = TRUE)`, and the median rank for the winners of the partial data is `r median(gs_partial$winner_rank)`.
-The same trend is true for the loser rank.
-This means that it is more likely for better players to have court by court data, which may effects our analysis for models when using this more detailed yet partial data.
-Additionally, we found that no women's point by point data was recorded for 2015.
-
-All data is obtained from Jeff Sackmann's open website via the `R` package `deuce` [@sackmann2018; @deuce2017].  All steps of our analysis from collection to dissemination are freely available online.
-
 ## Early Data Analysis {#sec:eda}
@@
 ### Examining the distribution of points earned
-
-We first examine the distribution of points earned per match to assess normality.
-The distribution of points earned is approximately normal.
-This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams.
-As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played.
-Also unsurprisingly, the winners of the match tended to earn more points than the losers.
 
@@
 
-p2 <- ggplot(gs_players) + geom_density(aes(pointswon, color = tournament)) + 
-  labs(x = "Points") + my_theme +
-  scale_color_manual("Tournament", values=tournament_colors)
+p2 <- ggplot(gs_players) + 
+  geom_density(aes(pointswon, ..scaled..,
+                   color=tournament, group=tournament, fill = tournament),
+                                        size = 3, alpha = 0) + 
+  scale_fill_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open (Plexicushion)",
+                                           "French Open (Clay)",
+                                           "US Open (DecoTurf)",
+                                           "Wimbledon (Grass)")) + my_theme  +
+  guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
+                              nrow=4, byrow=TRUE)) +
+    theme(legend.position="bottom") +
+    scale_color_manual(guide=FALSE, values=tournament_colors) +
+  facet_wrap(~league, ncol=1) +
+  labs(x = "Points", y = "Scaled density")
 
-p3 <- ggplot(gs_players) + geom_density(aes(pointswon, color = league))+ 
-  labs(x = "Points") +
-  scale_color_manual("League", values=league_colors) +
-  my_theme
+# 
+# ggplot(gs_players) + geom_density(aes(pointswon, color=tournament, 
+#                                             group=tournament, fill = tournament),
+#                                         size = 3, alpha = 0) + 
+#   scale_fill_manual(" ", values=tournament_colors,
+#                                 labels = c("Australian Open\n(Plexicushion)",
+#                                            "French Open\n(Clay)",
+#                                            "US Open\n(DecoTurf)",
+#                                            "Wimbledon\n(Grass)")) + my_theme  +
+#   guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
+#                               nrow=2, byrow=TRUE)) +
+#     theme(legend.position="bottom") +
+#     scale_color_manual(guide=FALSE, values=tournament_colors)
 
-# grid.arrange( p2, p3, ncol=2,
-#              top = textGrob("Winning points by tournament and league",
-#                             gp=gpar(fontsize=20)))
+p1 <- ggplot(gs_players) + 
+  geom_density(aes(pointswon, ..scaled..,
+                   color = factor(did_win),
+                   group = factor(did_win), fill = factor(did_win)),
+               size = 3, alpha = 0) + 
+  labs(x = "Points", y = " ") +
+  scale_fill_manual(" ", values=win_colors, labels = c("Lost game", "Won game")) +
+  scale_color_manual(guide = FALSE, values = win_colors) +
+  facet_wrap(~league, ncol=1) + 
+  my_theme +
+  theme(legend.position="bottom",
+        legend.spacing.y = unit(1.0, 'cm'),
+        legend.margin=margin(t = 2.2, unit='cm'))  +
+  guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
+                             nrow = 2,
+                            byrow=TRUE),
+        title.hjust = 5)
 
-p1 <- ggplot(gs_players) + geom_density(aes(pointswon, color = factor(did_win))) + 
-  labs(x = "Points") +
-  scale_color_manual("Winner?", values=win_colors) +
-  my_theme +
-  facet_wrap(~league, ncol=1)
+grid.arrange(p2, p1, ncol=2) %>%
+             # # top = textGrob("Distribution of points per match",
+             # #                gp=gpar(fontsize=20, fontfamily="serif")),
+             #  layout_matrix = rbind(c(2, 1),
+             #                        c(2, 1),
+             #                        c(NA, 1)),
+             # heights=c(1, 3, .1)) %>%
+  ggsave("plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
 
-grid.arrange(p1, p2, ncol=2,
-             top = textGrob("Distribution of points per match",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-              layout_matrix = rbind(c(NA, 1),
-                                    c(2, 1),
-                                    c(NA, 1)),
-             heights=c(1, 3, 1))
+# 
+# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+# grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
+#                             gp=gpar(fontsize=20, fontfamily="serif")),
+#              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+#                            # x = 0,
+#                            # y = 0.5,
+#                            just = "left",
+#                            gp=gpar(fontfamily="serif")))
+# 
+# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+#   ggsave("plots/spainards.jpg", ., height=7, width=16)
 
@@
 
-It is commonly thought that there is a home court advantage in grand slam games (SOURCE).
-In our data we find this to be true (i.e.
-French players win the French open more than French players win other slams).
-But, we also know that the home team is given preference for wild card bids (SOURCE) so potentially citizens of a given country play in "their" tournament more often than they play in other tournaments.
-We also find this to be true in our data for France, The United States, Australia, and Great Britain (i.e.
-the proportion of French players in the French Open is greater than the proportion of French players in other slams).
-
-Therefore, we want to see how the proportion of wins for the home country changes across the different tournaments.
-If there was really a home court advantage, the proportion of French wins each round would be higher at the French Open than the Australian Open, the US Open, and Wimbledon.
-The same would be true for Australia and the US.
-But, we see in \ref{fig:home-court} that this isn't the case.
-After accounting for the number of players from each country, we don't find a home court advantage in the grand slam.
-
-```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court}"}
-france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
+```{r, fig.height=3.5, fig.width=10, fig.align='center'}
+france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF") %>% 
+  dplyr::group_by(tournament, round) %>% 
+  dplyr::summarise(win_prop = mean(did_win),
             win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
 
-us_wins <- filter(gs_players, ioc == "USA", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
+us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF") %>% 
+  dplyr::group_by(tournament, round) %>% 
+  dplyr::summarize(win_prop = mean(did_win),
             win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
@@
   group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
+  dplyr::summarise(win_prop = mean(did_win),
             win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
@@
   group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
+  dplyr::summarise(win_prop = mean(did_win),
             win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
 
-p1<- ggplot(france_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
+p1<- ggplot(france_wins, aes(round, win_prop,
+                             color=tournament, group=tournament,
+                             alpha = tournament)) + 
+  # geom_line(size = 3) + 
   # geom_point() +
@@
        x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
+  geom_line(size=3) + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+scale_alpha_manual(guide = FALSE,
+                   values = c(0.2, 1, 0.2, 0.2))
 
-p2 <- ggplot(us_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
+p2 <- ggplot(us_wins, aes(round, win_prop,
+                          color=tournament, group=tournament,
+                          alpha = tournament)) + 
+  # geom_line() + 
   # geom_point() +
-  labs(title = "American players",
+  labs(title = "USA players",
        x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors)  + my_theme
+    geom_line(size=3) + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+scale_alpha_manual(guide = FALSE,
+                   values = c(0.2, 0.2, 1, 0.2))
 
-p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
+p3 <- ggplot(aus_wins, aes(round, win_prop,
+                           color=tournament, group=tournament, alpha = tournament)) + 
+  # geom_line() + 
   # geom_point() +
-  labs(title = "Australian players",
+  labs(title = "Aus. players",
        x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
+    geom_line(size=3) + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+scale_alpha_manual(guide = FALSE,
+                   values = c(1, 0.2, 0.2, 0.2))
 
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line() + 
-  # geom_point() +
+p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
+                           group=tournament, fill = tournament,
+                           alpha = tournament)) + 
+  geom_line(size=3) + 
   labs(title = "British players",
        x = "", y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme
+  scale_color_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme +
+  guides(color = guide_legend(override.aes = list(size = 7)))  + 
+scale_alpha_manual(guide = FALSE,
+                   values = c(0.2, 0.2, 0.2, 1))
+
+# 
+# + 
+#   geom_line() + 
+#   # geom_point() +
+#   labs(title = "British players",
+#        x = "", y = "Win proportion") + 
+#   scale_color_manual(" ", values=tournament_colors) + my_theme
 
@@
                            gp=gpar(fontfamily="serif")))
+
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(.,
+             # top = textGrob("Home court advantage: debunked",
+             #                gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           just = "left",
+                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
+  ggsave("plots/home_court.jpg", ., height=5, width=graphic_width_long)
+
+```
+
+## Spaniards on clay
+
+```{r}
+
+spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
+  group_by(tournament, round) %>% 
+  dplyr::summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
+            avg_ranking = mean(rank),
+            rank_se = sd(rank)/sqrt(n()))
+
+prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
+                table(did_win, tournament)), 2)
+
+with(filter(gs_players, ioc == "ESP", round == "R128"),
+                table(did_win, tournament))
+
+with(filter(gs_players, ioc == "ESP", round == "R128"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
+
+```
+
+```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
+
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 24),
+#         text = element_text(size = 28,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=34))
+
+p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
+  labs(title = "Win proportion among Spaniards",
+       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
+       x = "", y = "Win proportion") + 
+    geom_line(size=3) + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme
+
+p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament, fill = tournament)) + 
+  geom_line(size=3) + 
+  labs(title = "Average ranking among Spaniards",
+       x = " ",
+       y = "Average rank") + 
+  scale_color_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme +
+  guides(color = guide_legend(override.aes = list(size = 7)))
+
+ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
+                            gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           # x = 0,
+                           # y = 0.5,
+                           just = "left",
+                           gp=gpar(fontfamily="serif")))
+
+ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+  ggsave("plots/spainards.jpg", ., height=6, width=graphic_width_long)
+
+```
+
+# Unforced errors and aces
+
+```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
+
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 24),
+#         text = element_text(size = 28,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=34))
+
+# ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
+#                            label = tournament, color=tournament)) + 
+#   geom_jitter(size=2) +
+#   labs( title = "Distribution of errors and aces differ by tournament",
+#         x = "Total unforced errors", y = "Total aces") +
+#   my_theme + scale_color_manual("Tournament",
+#                                 values = tournament_colors,
+#                                 labels = c("Australian Open\n(Plexicushion)\n",
+#                                            "French Open\n(Clay)\n",
+#                                            "US Open\n(DecoTurf)\n",
+#                                            "Wimbledon\n(Grass)\n")) + 
+#   geom_density_2d()
+
+ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
+                           label = tournament, color=tournament)) + 
+  geom_jitter(size=3) +
+  labs( title = "Distribution of errors and aces differ by tournament",
+        x = "Total unforced errors", y = "Total aces") +
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  stat_ellipse(type="t", size = 2) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+ggsave("plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
+
+```
+
+## Tall players and aces
+
+```{r, fig.width=10, fig.height=4.8, fig.align='center'}
+
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 24),
+#         text = element_text(size = 28,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=34))
+
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 26),
+#         text = element_text(size = 30,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=36))
+
+set.seed(8)
+
+atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
+                 "Roger Federer", "Andy Murray")
+atp_servers <- sapply(atp_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
+                 "Venus Williams", "Serena Williams", "Madison Keys")
+wta_servers <- sapply(wta_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+p_atp <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament), size=1.5) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), log(ace+1), label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+  labs(x = "log(height + 1)", title = "ATP") +
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+p_wta <- gs_players %>% filter(league == "WTA") %>%
+ggplot(aes(log(ht+1), log(ace+1))) + 
+  geom_jitter(aes(color = tournament)) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth(method = "lm", color="black") + 
+  geom_label_repel(data=wta_servers,
+            aes(log(ht+1), log(ace+1), label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+  geom_label_repel(data=wta_servers,
+            aes( log(ht+1), log(ace+1), label=name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+  labs(x = "log(height + 1)", title = "WTA")+
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
+#              top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
+# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+# ggsave("plots/aces.jpg", height=7, width=16)
+ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
+
+```
+
+```{r, fig.width=6, fig.height=6, fig.align='center'}
+
+# p_atp <- gs_players %>% filter(league == "ATP") %>%
+# ggplot( aes(log(ht+1), log(ace+1))) + 
+#   geom_jitter(aes(color = tournament)) + 
+#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
+#   geom_smooth( method = "lm", se=FALSE, color="black") + 
+#   geom_label_repel(data=atp_servers,
+#             aes(log(ht+1), log(ace+1), label=name),
+#             alpha=.6, seed=1234, show.legend=FALSE) + 
+#   geom_label_repel(data=atp_servers,
+#             aes(log(ht+1), log(ace+1), label = name),
+#             seed=1234, fill=NA, show.legend=FALSE) +
+#   labs(x = "log(height + 1)", title = "ATP") +
+#   my_theme
+# 
+# # subset(filter(gs_players, league == "WTA"),
+# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
+# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
+# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
+# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
+# #                                  (log(ace + 1) > 2.82))
+# p_wta <- gs_players %>% filter(league == "WTA") %>%
+# ggplot(aes(log(ht+1), log(ace+1))) + 
+#   geom_jitter(aes(color = tournament)) + 
+#   scale_color_manual("Tournament", values = tournament_colors) + 
+#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
+#   geom_label_repel(data=wta_servers,
+#             aes(log(ht+1), log(ace+1), label=name),
+#             alpha=.6, seed=1234, show.legend=FALSE) + 
+#   geom_label_repel(data=wta_servers,
+#             aes(log(ht+1), log(ace+1), label=name),
+#             seed=1234, fill=NA, show.legend=FALSE) +
+#   labs(x = "log(height + 1)", title = "WTA")+
+#   my_theme + theme(legend.position="bottom")
+# 
+# 
+# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
+#              top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+```
+
+### Linear regression for other outcomes
+
+```{r, eval = FALSE}
+gs_partial_players$atp = gs_partial_players$Tour == "atp"
+gs_partial_players$year_fac = as.factor(gs_partial_players$year)
+gs_partial_players$late_round = gs_partial_players$round >= "R16"
+gs_partial_players$seeded = gs_partial_players$rank <= 32
+gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
+n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
+```
+
+```{r}
+load("../../data/n_aces_mod.rda")
+load("../../data/n_net_mod.rda")
+load("../../data/n_ue_mod.rda")
+load("../../data/n_winners_mod.rda")
+```
+
+```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+
+# my_theme <-  theme_bw() + # White background, black and white theme
+#   theme(axis.text = element_text(size = 20),
+#         text = element_text(size = 24,
+#         family="serif"),
+#         plot.title = element_text(hjust = 0.5, size=26))
+
+aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval,
+                ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
+  ylab("Coefficient\n") + 
+  xlab("") +
+  ggtitle("Y = aces") + 
+  scale_color_manual("", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) +
+  my_theme +
+  coord_flip() +
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()) +
+  geom_hline(yintercept = 0, color = "grey") +
+  guides(color = guide_legend(override.aes = list(size = 7)))
+
+# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
+#   mutate(tournament = gsub("tournament", "", term)) %>%
+#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+#   geom_errorbar(height = .5, position = position_dodge()) +
+#   ylab("Coefficient\n") + 
+#   xlab("") +
+#   ggtitle("Winners") + 
+#   scale_color_manual("", values=tournament_colors,
+#                                 labels = c("Australian Open\n(Plexicushion)",
+#                                            "French Open\n(Clay)",
+#                                            "US Open\n(DecoTurf)",
+#                                            "Wimbledon\n(Grass)")) +
+#   my_theme  +
+#   coord_flip() +
+#   theme(panel.grid.major = element_blank(),
+#         panel.grid.minor = element_blank(),
+#         axis.text.y = element_blank()) +
+#   geom_hline(yintercept = 0, color = "grey")
+
+nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
+  ylab("Coefficient\n") + 
+  xlab("") +
+  ggtitle("Y = points won at net") + 
+  scale_color_manual("", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) +
+  my_theme  +
+  coord_flip() +
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        axis.text.y = element_blank()) +
+  geom_hline(yintercept = 0, color = "grey") +
+  guides(color = guide_legend(override.aes = list(size = 7)))
+
+ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp),
+                y = condval,
+                ymin = condval- 2*condsd,
+                ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
+  ylab("Coefficient\n") + 
+  xlab("") +
+  ggtitle("Y = unforced errors") + 
+  scale_color_manual("", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) +
+  my_theme  +
+  coord_flip() +
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        axis.text.y = element_blank()) +
+  geom_hline(yintercept = 0, color = "grey") +
+  guides(color = guide_legend(override.aes = list(size = 7)))
+
+#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
+
+ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
+          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29))
+
+ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
+          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
+ggsave("plots/player_effects.jpg", ., height=6, width = graphic_width_long)
+
+```
+
+## player effects
+
+```{r nadal-model}
+
+step_model_n <- courtsports:::model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
+
+std_res_n <- rstandard(step_model_n)
+
+gn <- ggplot(lm(step_model_n$formula, data = step_model_n$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Nadal",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+coefs_n <- coefficients(step_model_n)
+cov_n <- vcov(step_model_n)
+
+ ## NADAL COEF SIZE
+
+df_n <- data.frame(coeff = names(step_model_n$coeff), ave = step_model_n$coeff, sd = sqrt(diag(vcov(step_model_n))))
+df_n$player <- "Nadal"
+```
+
+```{r federer-model}
+
+step_model_f <- courtsports:::model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
+
+std_res_f <- rstandard(step_model_f)
+
+gf <- ggplot(lm(step_model_f$formula, data = step_model_f$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Federer",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+coefs_f <- coefficients(step_model_f)
+cov_f <- vcov(step_model_f)
+
+df_f <- data.frame(coeff = names(step_model_f$coeff), ave = step_model_f$coeff, sd = sqrt(diag(vcov(step_model_f))))
+df_f$player <- "Federer"
+```
+
+```{r williams-model}
+
+step_model_w <- courtsports:::model_individual(player_name = "Serena Williams", data = gs_partial_players,
+                                 start_lower = TRUE, ref = "Australian Open")
+
+std_res_w <- rstandard(step_model_w)
+
+gw <- ggplot(lm(step_model_w$formula, data = step_model_w$data)) + geom_hline(yintercept = 0, col = "red") +
+  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
+  labs(title = "Williams",
+       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
+
+cov_w <- vcov(step_model_w)
+coefs_w <- coefficients(step_model_w)
+
+ ## WilliamsCOEF SIZE
+
+df_w <- data.frame(coeff = names(step_model_w$coeff), ave = step_model_w$coeff, sd = sqrt(diag(vcov(step_model_w))))
+df_w$player <- "Williams"
+```
+
+```{r plot-individual-effects}
+df <- rbind(df_n, df_w, df_f)
+keep_coefs <- which(df$coeff %in% c("(Intercept)", "opponent_rank",
+                                    "pct_ace", "pct_bp", "pct_netpt",
+                                    "ave_serve_speed", "wue_ratio"))
+df <- df[keep_coefs,]
+ggplot(df, aes(x = coeff, y = ave, col = player)) +
+  geom_errorbar(aes(ymin = ave - 2 *sd, ymax = ave + 2 * sd),
+                position = "dodge", width = .3, size = 2) + 
+   my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
+  geom_hline(yintercept = 0, color = "grey") + 
+  ggtitle("Fixed effects") +
+  scale_color_brewer("", palette = "Dark2") +
+  theme(legend.position = "bottom") +
+  xlab("") + 
+  ylab("Coefficient") +
+  scale_x_discrete(labels = c("Intercept", "Opp. Rank", "% Aces",
+                              "% Break Points Won", "% Net Points Won",
+                              "Ave. Serve Speed", "Winner:UE"))  +
+  guides(color = guide_legend(override.aes = list(size = 7)))  +
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+ggsave("plots/ref-courts-eff.jpg", height=6, width = graphic_width_short)
+
+  #ggsave("~/Documents/ref-courts-eff.pdf", width = 8, height = 6, height=6, width = graphic_width_long)
+```
+
+# amanda model fitting
+
+```{r}
+win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+# tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+
+model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
+model.ests = data.frame()
+ran.effects = data.frame()
+for(mod in 1:nrow(model.sums)){
+  name = model_names[mod]
+  model.sums$aic[mod] = extractAIC(get(name))[2]
+  model.sums$edf[mod] = extractAIC(get(name))[1]
+  betas = tidy(get(name))
+  betas$model = name
+  if("group" %in% colnames(betas)){
+    model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+                         dplyr::select(., -c("group")))
+  }
+  else model.ests = rbind(model.ests, betas)
+}
+
+model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef", "ind_no_ioc", "ind_intercept", "ind_year_ef")
+model.sums$fixed = c("No country, no tournament",
+                     "all",
+                     "no country, no tournament",
+                     "no tournament",
+                     "no country, no tournament",
+                     "no country",
+                     "no country, no year")
+model.sums$random = c("none",
+                      "none",
+                      "tournament (by country)",
+                      "tournament (by individual)",
+                      "tournament (by individual)",
+                      "intercept (by individual)",
+                      "year (by individual)")
+
+fixed.effects.plot = model.ests %>%
+  filter(., !grepl("ioc", term)) %>% 
+  filter(., model == 'ind_logistic_noioc') %>%
+  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
+  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
+  my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
+  ggtitle("Fixed effects") +
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "none") + 
+  xlab("") +
+  ylab("Coefficient") +
+  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "2014", "2015", "2016", "2017"))+
+  guides(col = guide_legend(nrow=2)) +
+  geom_hline(yintercept = 0, color = "grey") + 
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
+ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
+
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
+name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
+
+top_three_ranef = as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
+  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
+  ggplot(., aes(x = name, y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, color = tournament)) +
+  geom_errorbar(width = .5, position = "dodge",size = 1.5) + 
+  xlab("") + 
+  ylab("Coefficient") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  coord_flip() +
+  ggtitle("Random Effects") +
+  theme(legend.position = "bottom") + 
+  guides(col = guide_legend(nrow=2,
+                            override.aes = list(size = 7))) +
+  geom_hline(yintercept = 0, color = "grey") + 
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55) #,
+             # top = textGrob("Parameter estimates from logistic model",
+             #                gp=gpar(fontsize=20, fontfamily="serif"))
+             ) %>%
+ggsave("plots/logistic-parameter-effects.jpg",., height=7, width=graphic_width_long)
+
+```
+
+```{r}
+
+fixed.effects.plot.linear <- tidy(n_aces_mod) %>%
+  filter(., group == 'fixed') %>%
+  mutate(., model = "Aces") %>%
+  bind_rows(
+    tidy(n_ue_mod) %>%
+      filter(., group == 'fixed') %>%
+      mutate(., model = "UE")
+  ) %>%
+  bind_rows(
+    tidy(n_net_mod) %>%
+      filter(., group == 'fixed') %>%
+      mutate(., model = "Net")
+  ) %>%
+  ggplot(., aes(x = term, y = estimate, 
+                ymin = estimate - 2*std.error, 
+                ymax = estimate + 2*std.error,
+                col = model,
+                group = model)) + 
+  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
+  my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
+  ggtitle("Fixed effects") +
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "bottom") + 
+  xlab("") +
+  ylab("Coefficient") +
+  scale_x_discrete("", 
+                   breaks = waiver(), 
+                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)",
+                              "Rank (log)", "2014", "2015", "2016", "2017"))  +
+  guides(color = guide_legend(override.aes = list(size = 7)))  +
+  geom_hline(yintercept = 0, color = "grey") + 
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+ggsave("plots/linear-fixed-effects.jpg", fixed.effects.plot.linear,
+       height=6, width = graphic_width_short)
+
+```
+
+```{r}
+
+fixed.effects.plot.allmods = model.ests %>%
+  filter(., !grepl("ioc", term)) %>% 
+  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
+  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
+  my_theme +
+  # coord_flip() +
+  theme(axis.text.x=element_text(angle = 35, hjust = 1, size = 10)) +
+  ggtitle("Fixed effects") +
+  scale_color_brewer("",palette = "Dark2") +
+  theme(legend.position = "none") + 
+  xlab("") +
+  ylab("Estimate") +
+  scale_x_discrete("", 
+                   breaks = waiver(), 
+                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017")) +
+  guides(col = guide_legend(nrow=2))
+
+aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Aces") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  coord_flip() +
+  theme(legend.position = "none")
+
+nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Net Wins") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  theme(axis.text.y = element_blank()) +
+  coord_flip() # +
+# theme(legend.position = "none")
+
+ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
+  mutate(tournament = gsub("tournament", "", term)) %>%
+  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
+  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
+  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
+  ylab("Estimate") + 
+  xlab("") +
+  ggtitle("Unforced Errors") + 
+  scale_color_manual("", values=tournament_colors) +
+  my_theme +
+  theme(axis.text.y = element_blank()) +
+  coord_flip()# + 
+#theme(legend.text=element_text(size=10))
+
+#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
+
+```
+
+# extras
+
+```{r, fig.height=3.5, fig.width=10, fig.align='center', eval=FALSE}
 
@@
 ```
-
-## Spaniards on clay
-
-```{r}
-
-spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
-            avg_ranking = mean(rank),
-            rank_se = sd(rank)/sqrt(n()))
-
-prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament)), 2)
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament))
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-     table(did_win, tournament)) %>%
-  chisq.test(correct = TRUE)
-
-```
-
-It is also commonly thought that Spaniards play better on clay.
-We are interested in whether Spaniards win the French Open more than they win in other tournaments.
-It does appear that Spaniards are winning the French Open more than they are winning other tournaments as the line in \ref{fig:spain-clay} for the French Open is consistently higher than the other tournaments.
-But, this result is not significant (ENTER P VALUE).
-Interestingly, the ranking of Spaniards in the French Open is, on average, higher than other tournaments, meaning that, on average Spanish players in the French Open are, on average, worse than Spanish players in ther tournaments.
-
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "Win proportion among Spaniards",
-       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
-       x = "", y = "Win proportion") + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme
-
-p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament, fill = tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "Average ranking among Spaniards",
-       x = " ",
-       y = "Average rank") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           # x = 0,
-                           # y = 0.5,
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-  ggsave("plots/spainards.jpg", ., height=7, width=16)
-
-```
-
-# Unforced errors and aces
-
-We find that the distribution of errors and aces differ by tournament.
-Specifically, GAMES?
-in Wimbledon follow different ace and error patterns than the other tournaments.
-These GAMES?
-tend to cluster in the upper left hand part of the graph, meaning that at Wimbledon players are making fewer unforced errors and scoring more aces.
-It is known (SOURCE) that the ball moves faster on grass courts and so the ace result is unsuprising, but making fewer errors is unexpected, especially because players seldom practice on grass.
-
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-                           label = tournament, color=tournament)) + 
-  geom_jitter(size=2) +
-  labs( title = "Distribution of errors and aces differ by tournament",
-        x = "Total unforced errors", y = "Total aces") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) + 
-  geom_density_2d()
-
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-                           label = tournament, color=tournament)) + 
-  geom_jitter(size=1.5) +
-  labs( title = "Distribution of errors and aces differ by tournament",
-        x = "Total unforced errors", y = "Total aces") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  stat_ellipse(type="t") +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-ggsave("plots/errors_aces2.jpg", height = 6, width = 12)
-
-```
-
-## Tall players and aces
-
-Another common thought about tennis is that taller players are better servers.
-We find this to be true and see a strong positive correlation between teh variables.
-We highlight some interesting players BLAH BLAH BLAH.
-
-```{r, fig.width=10, fig.height=4.8, fig.align='center'}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 24),
-        text = element_text(size = 28,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 26),
-#         text = element_text(size = 30,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=36))
-
-set.seed(8)
-
-atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
-                 "Roger Federer", "Andy Murray")
-atp_servers <- sapply(atp_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
-                 "Venus Williams", "Serena Williams", "Madison Keys")
-wta_servers <- sapply(wta_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-p_atp <- gs_players %>% filter(league == "ATP") %>%
-ggplot( aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament), size=1.5) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
-  labs(x = "log(height + 1)", title = "ATP") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-p_wta <- gs_players %>% filter(league == "WTA") %>%
-ggplot(aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament)) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth(method = "lm", color="black") + 
-  geom_label_repel(data=wta_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
-  geom_label_repel(data=wta_servers,
-            aes( log(ht+1), log(ace+1), label=name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
-  labs(x = "log(height + 1)", title = "WTA")+
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
-# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-# ggsave("plots/aces.jpg", height=7, width=16)
-ggsave("plots/aces2.jpg", height=7, width=16)
-
-```
-
-```{r, fig.width=6, fig.height=6, fig.align='center'}
-
-# p_atp <- gs_players %>% filter(league == "ATP") %>%
-# ggplot( aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
-#   geom_smooth( method = "lm", se=FALSE, color="black") + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label = name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "ATP") +
-#   my_theme
-# 
-# # subset(filter(gs_players, league == "WTA"),
-# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-# #                                  (log(ace + 1) > 2.82))
-# p_wta <- gs_players %>% filter(league == "WTA") %>%
-# ggplot(aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual("Tournament", values = tournament_colors) + 
-#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "WTA")+
-#   my_theme + theme(legend.position="bottom")
-# 
-# 
-# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-```
-
-### Linear regression for other outcomes
-
-Instead of using `win' as the outcome variable, we now look at match statistics that may be more sensitive to court surface -- aces, net points won, and UEs -- and see if individual differences among players are detectable.
-We fit three additional models, each of the form $Y_i = \beta X_i + \delta_i$, where $\delta_i = \alpha_i T_i$, $T_i$ is the four grand slam tournaments, and $X_i$ includes the same variables as the model above.
-
-```{r, eval = FALSE}
-gs_partial_players$atp = gs_partial_players$Tour == "atp"
-gs_partial_players$year_fac = as.factor(gs_partial_players$year)
-gs_partial_players$late_round = gs_partial_players$round >= "R16"
-gs_partial_players$seeded = gs_partial_players$rank <= 32
-gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-```
-
-```{r}
-load("../../data/n_aces_mod.rda")
-load("../../data/n_net_mod.rda")
-load("../../data/n_ue_mod.rda")
-load("../../data/n_winners_mod.rda")
-```
-
-```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
-
-my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 20),
-        text = element_text(size = 24,
-        family="serif"),
-        plot.title = element_text(hjust = 0.5, size=26))
-
-aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval,
-                ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = aces") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
-#   mutate(tournament = gsub("tournament", "", term)) %>%
-#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-#   geom_errorbar(height = .5, position = position_dodge()) +
-#   ylab("Coefficient\n") + 
-#   xlab("") +
-#   ggtitle("Winners") + 
-#   scale_color_manual("", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) +
-#   my_theme  +
-#   coord_flip() +
-#   theme(panel.grid.major = element_blank(),
-#         panel.grid.minor = element_blank(),
-#         axis.text.y = element_blank()) +
-#   geom_hline(yintercept = 0, color = "grey")
-
-nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = points won at net") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = unforced errors") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
-ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
-          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
-ggsave("plots/player_effects.jpg", ., width=14, height=5)
-
-```

### 2019-02-26 contest/surface-type/paper-tennis.Rmd
added correlation matrix figure

@@
 ## To change into library(courtsports)
-library(courtsports)
+devtools::load_all("~/courtsports")
 
@@
 data(gs_partial_players)
+
+wim_inds <- which(gs$tournament == "Wimbledon")
+head(gs[wim_inds,])
+
+serena_inds <- which(gs_partial_players$name == "Serena Williams")
+gs_partial_players[serena_inds[1],]
+
+head(gs_partial)
+djoker_inds <- which(gs_partial_players$name == "Novak Djokovic")
+gs_partial_players[djoker_inds[1],]
 

### 2019-02-26 contest/tables/cor.R

@@ -1,0 +1,24 @@
+## SKG
+## Feb. 26, 2019
+## Correlation matrix for randome effects
+
+library(ggplot2)
+library(reshape2)
+
+
+
+mat <- matrix(c(1, 0, .94, .93,
+         0, 1, -.05, -.36,
+         .94, -.05, 1, .88,
+         .93, -.36, .88, 1), byrow = TRUE, nrow = 4)
+colnames(mat) <- c("Australian Open", "French Open", "US Open", "Wimbledon")
+rownames(mat) <- c("Australian Open", "French Open", "US Open", "Wimbledon")
+#
+df <- melt(mat)
+#
+ggplot(df, aes(x = Var1, y = Var2, fill = value)) + geom_tile() +
+     geom_tile(color = "white")+
+    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
+   midpoint = 0, limit = c(-1,1), space = "Lab", 
+   name="Correlation") + labs(x = "", y = "",
+                              title = "Correlation matrix for random effects")

### 2019-02-28 contest/graphics/graphics_for_paper.Rmd
add correlatino plots

@@
 library(ggpubr)
+library(reshape2)
 
@@
 
+our_alpha <- 0.5
+
 p1<- ggplot(france_wins, aes(round, win_prop,
                              color=tournament, group=tournament,
-                             alpha = tournament)) + 
+                             alpha = tournament,
+                             size = tournament)) + 
   # geom_line(size = 3) + 
@@
   labs(title = "French players",
-       x = "", y = "Win proportion") + 
+       x = "", y = " ") + 
   geom_line(size=3) + 
@@
 scale_alpha_manual(guide = FALSE,
-                   values = c(0.2, 1, 0.2, 0.2))
+                   values = c(our_alpha, 1, our_alpha, our_alpha)) + 
+scale_size_manual(guide = FALSE,
+                   values = c(1, 3, 1, 1))
 
@@
   labs(title = "USA players",
-       x = "", y = "Win proportion") + 
+              x = "", y = " ") + 
+       # x = "", y = "Win proportion") + 
     geom_line(size=3) + 
@@
 scale_alpha_manual(guide = FALSE,
-                   values = c(0.2, 0.2, 1, 0.2))
+                   values = c(our_alpha, our_alpha, 1, our_alpha))
 
@@
        x = "", y = "Win proportion") + 
+         # x = "", y = " ") + 
     geom_line(size=3) + 
@@
 scale_alpha_manual(guide = FALSE,
-                   values = c(1, 0.2, 0.2, 0.2))
+                   values = c(1, our_alpha, our_alpha, our_alpha))
 
@@
   labs(title = "British players",
-       x = "", y = "Win proportion") + 
+       # x = "", y = "Win proportion") + 
+       x = "", y = " ") + 
   scale_color_manual(" ", values=tournament_colors,
@@
 scale_alpha_manual(guide = FALSE,
-                   values = c(0.2, 0.2, 0.2, 1))
+                   values = c(our_alpha, our_alpha, our_alpha, 1))
 
@@
                            gp=gpar(fontfamily="serif")))
+
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(.,
+             # top = textGrob("Home court advantage: debunked",
+             #                gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           just = "left",
+                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
+  ggsave("plots/home_court.jpg", ., height=5, width=graphic_width_long)
+
+```
+
+```{r}
+
+p1<- ggplot(france_wins, aes(round, win_prop,
+                             color=tournament, group=tournament,
+                             # alpha = tournament,
+                             size = tournament,
+                             linetype = tournament)) + 
+  # geom_line(size = 3) + 
+  # geom_point() +
+  labs(title = "French players",
+       x = "", y = " ") + 
+  geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, 1, our_alpha, our_alpha)) + 
+scale_size_manual(guide = FALSE,
+                   values = c(1, 3, 1, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 1, 2, 2))
+
+p2 <- ggplot(us_wins, aes(round, win_prop,
+                          color=tournament, group=tournament,
+                          # alpha = tournament,
+                          size = tournament,
+                          linetype=tournament)) + 
+  # geom_line() + 
+  # geom_point() +
+  labs(title = "USA players",
+              x = "", y = " ") + 
+       # x = "", y = "Win proportion") + 
+    geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+  scale_size_manual(guide = FALSE,
+                   values = c(1, 1, 3, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 2, 1, 2))
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, our_alpha, 1, our_alpha))
+
+p3 <- ggplot(aus_wins, aes(round, win_prop,
+                           color=tournament, group=tournament, # alpha = tournament,
+                           size = tournament,
+                           linetype = tournament)) + 
+  # geom_line() + 
+  # geom_point() +
+  labs(title = "Aus. players",
+       x = "", y = "Win proportion") + 
+         # x = "", y = " ") + 
+    geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+  scale_size_manual(guide = FALSE,
+                   values = c(3, 1, 1, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(1, 2, 2, 2))
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(1, our_alpha, our_alpha, our_alpha))
+
+p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
+                           group=tournament, fill = tournament,
+                           # alpha = tournament,
+                           size = tournament,
+                           linetype = tournament)) + 
+  geom_line() + 
+  labs(title = "British players",
+       # x = "", y = "Win proportion") + 
+       x = "", y = " ") + 
+  scale_color_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme +
+  guides(color = guide_legend(override.aes = list(size = 7)))  + 
+  scale_size_manual(guide = FALSE,
+                   values = c(1, 1, 1, 3)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 2, 2, 1))
+
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, our_alpha, our_alpha, 1))
+
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(.,
+             # top = textGrob("Home court advantage: debunked",
+             #                gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           just = "left",
+                           gp=gpar(fontfamily="serif", fontsize = 20)))
 
@@
                            label = tournament, color=tournament)) + 
-  geom_jitter(size=3) +
+  geom_jitter(size=2) +
   labs( title = "Distribution of errors and aces differ by tournament",
@@
 
+# correlation plots
+
+```{r}
+win_colors <- c("#336699", "#339966")
+
+mat <- matrix(c(1, 0, .94, .93,
+         0, 1, -.05, -.36,
+         .94, -.05, 1, .88,
+         .93, -.36, .88, 1), byrow = TRUE, nrow = 4)
+colnames(mat) <- c("Aus. Open", "French Open", "US Open", "Wimbledon")
+rownames(mat) <- c("Aus. Open", "French Open", "US Open", "Wimbledon")
+#
+df <- melt(mat)
+#
+ggplot(df, aes(x = Var1, y = Var2, fill = value)) + geom_tile() +
+     geom_tile(color = "white")+
+    scale_fill_gradient2(low = win_colors[2], high = win_colors[1], mid = "white", 
+                         midpoint = 0, limit = c(-1,1), 
+                         space = "Lab", name="Correlation") + 
+                         # labels = c("-1.0", "-0.5", "  0.0", "  0.5", "  1.0")) + 
+  labs(x = "", y = "",
+       title = "Correlation matrix for random effects")  +
+  my_theme
+
+ggsave("plots/correlation.jpg", 
+       height=6, width = graphic_width_short)
+
+```
+
 # extras

### 2019-07-28 contest/surface-type/paper-tennis.Rmd
Numerous revisions to individual modelling and graph creation

@@
 
-library(deuce)
+#library(deuce)
 library(dplyr)

### 2019-07-28 R/model-individs.R

@@ -5,44 +5,99 @@
 #' @param data default is gs_partial_players
 #' @param ref reference court.  Default is French Open
 #' @param start_lower logical indicating whether the base model should be the lower model.
-#' @return best model fit
-model_individual <- function(player_name = "Rafael Nadal", data = gs_partial_players,
-                             ref = "French Open",
-                             start_lower = FALSE){
+#' @return list step model output, the final model on the training set,
+#' the subsetted data, the training indices, and the number of total observations in the subsetted data (n_obs), the number of training obs (n_train), and number of testing obs (n_test)
+#' @export
+model_individual <- function(player_name = "Roger Federer",
+                             data = gs_partial_players,
+                             ref = "Wimbledon",
+                             start_lower = FALSE,
+                             test_prop = 0,
+                             min_year = 2013,
+                             max_year = 2017,
+                             seed = NULL){
 
-  data_sub <- dplyr::filter(gs_partial_players, name %in% c(player_name)) %>%
-    dplyr::mutate(pct_ace = n_aces / n_netpt,
-                  pct_netpt = n_netpt_w / n_netpt,
-                  pct_bp = n_bp_w / n_bp,
-                  wue_ratio = n_winners / n_ue,
-                  pct_pts = pointswon / total_points)
+    if(!is.null(seed)){
+        set.seed(seed)
+    }
+    ### Subset data to player
+    data_sub <- dplyr::filter(data, name %in% c(player_name) &
+                                    year >= min_year &
+                                    year <= max_year
+                              ) %>%
+        dplyr::mutate(pct_ace = ifelse( n_sv > 0,
+                                       n_aces / n_sv,
+                                       0),
+                      pct_netpt = ifelse(n_netpt > 0,
+                                         n_netpt_w / n_netpt,
+                                         0),
+                      pct_bp = ifelse(n_bp > 0,
+                                      n_bp_w / n_bp,
+                                      0),
+                      wue_ratio = ifelse(n_ue > 0,
+                                         n_winners / n_ue,
+                                         100), ## magic number
+                      pct_pts = ifelse(total_points > 0,
+                                       pointswon / total_points,
+                                       0))
 
-  data_sub$court <- relevel(data_sub$tournament, ref = ref)
+    ## Make the tournament of reference
+    data_sub$court <- relevel(data_sub$tournament, ref = ref)
+    
[…]

### 2019-07-28 contest/many-players/.#most_matches.R

@@ -1,0 +1,1 @@
+shannon@archlinux.31118:1562594886

### 2019-07-28 contest/many-players/federer.R

@@ -1,0 +1,31 @@
+## Script to get Federer individual model
+## SKG July 20, 2019
+
+
+devtools::load_all("~/courtsports")
+data(gs_partial_players)
+
+player_name <- "Novak Djokovic"
+data <- gs_partial_players
+ref <- "French Open"
+start_lower <- TRUE
+min_year <- 2013
+max_year <- 2017
+seed <- 2020
+test_prop <- 0
+#
+out_list <- courtsports::model_individual(
+                              player_name = player_name,
+                              data = data,
+                              ref = ref,
+                              start_lower = start_lower,
+                              test_prop = test_prop,
+                              min_year = min_year,
+                              max_year = max_year,
+                              seed = seed)
+#
+summary(out_list$final_model)
+par(mfrow = c(2,2))
+plot(out_list$final_model)
+print(out_list$n_obs)
+

### 2019-07-28 contest/many-players/most_matches.R

@@ -1,0 +1,406 @@
+## Script to get Federer individual model
+## SKG July 20, 2019
+
+
+
+
+devtools::load_all("~/courtsports")
+library(ggplot2)
+library(forcats)
+library(dplyr)
+library(tidyr)
+library(ggcorrplot)
+library(gridExtra)
+library(knitr)
+library(broom)
+library(kableExtra)
+data(gs_partial_players)
+
+########################3
+## THeme
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+        family="serif"),
+        plot.title = element_text( size=16))
+
+tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
+league_colors <- c("#0C2340", "#902BA3")
+win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+########################################
+
+player_name <- "Roger Federer"
+data <- gs_partial_players
+ref <- "Wimbledon"
+start_lower <- TRUE
+min_year <- 2013
+max_year <- 2017
+seed <- 2020
+set.seed(seed)
+test_prop <- 0
+#
+out_list <- courtsports::model_individual(
+                              player_name = player_name,
+                              data = data,
+                              ref = ref,
+                              start_lower = start_lower,
+                              test_prop = test_prop,
+                              min_year = min_year,
+                              max_year = max_year,
+                              seed = seed)
+#
+summary(out_list$final_model)
+par(mfrow = c(2,2))
+plot(out_list$final_model)
+print(out_list$n_obs)
+df <- tidy(out_list$final_model)
+kable(df[,-4], format = "latex",
+      row.names = FALSE, col.names = c("Coef.", "Est.",
+                                       "Std. Err.", "p-value"),
+             caption = "\\label{tab:fed-ind-coef} Modeling coefficients for Federer's best fit individual model using Wimbledon as a reference court.",
+      booktabs = TRUE, digits = 4) %>% kable_styling(latex_options = "striped", "hold_position")
+
+delta_sd <- function(b, m, sigma){
[…]

### 2019-07-28 contest/tables/cor.R

@@ -22,3 +22,4 @@ ggplot(df, aes(x = Var1, y = Var2, fill = value)) + geom_tile() +
    midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Correlation") + labs(x = "", y = "",
                               title = "Correlation matrix for random effects")
+ggsave("corr-mat.pdf")

### 2019-07-29 contest/many-players/.#most_matches.R
more individual model edits

@@ -1,1 +1,0 @@
-shannon@archlinux.31118:1562594886

### 2019-07-29 contest/many-players/most_matches.R

@@ -14,6 +14,9 @@ library(gridExtra)
 library(knitr)
 library(broom)
 library(kableExtra)
+library(ggDiagnose)
+
+
 data(gs_partial_players)
 
 ########################3
@@ -62,6 +65,30 @@ kable(df[,-4], format = "latex",
                                        "Std. Err.", "p-value"),
              caption = "\\label{tab:fed-ind-coef} Modeling coefficients for Federer's best fit individual model using Wimbledon as a reference court.",
       booktabs = TRUE, digits = 4) %>% kable_styling(latex_options = "striped", "hold_position")
+car::vif(out_list$final_model)
+
+## plot diags
+g <- ggDiagnose(out_list$final_model, which = 1:6, return = TRUE) 
+ggs <- g$ggout
+ggthm <- lapply(ggs, function(gg) gg + my_theme)
+pdf("fed_diags.pdf", width = 16, height = 10)
+do.call("grid.arrange", c(ggthm, ncol = 3))
+dev.off()
+
+## outliers
+fed_df <- out_list$final_model$data
+fed_df[c(38,66,48),]
+sort(fed_df$wue, de = FALSE)
+ind <- which(fed_df$pct_bp == 0)
+fed_df[ind,]
+fed_df[which(fed_df$wue >= 5), ]
+
+mod2 <- lm(out_list$final_model$formula, data = fed_df[-c(38,66,48),])
+summary(mod2)
+
+
+## interaction effects
+
 
 delta_sd <- function(b, m, sigma){
     sbb <- sigma[1,1]
@@ -124,6 +151,12 @@ mods <- vector(mode = "list", length = L)
 data <- mods <- vector(mode = "list", length = L)
 ref <- "French Open"
 alpha <- .05
+sum_df <- data.frame(name = rep("nm", L),
+                     tour = "blah",
+                     n_obs = 0,
+                     n_cov = 0,
+                     R2 = 0,
+                     max_vif = 0, stringsAsFactors = FALSE)
 for(ii in 1:L){
     player_name <- top_players$name[ii]
     print(player_name)
@@ -158,9 +191,38 @@ for(ii in 1:L){
                      se = sqrt(diag(vcov(mod))),
                      pval = pval_j$pval)
     player_list[[ii]] <- df
+    ## summary_df
+    sum_df$name[ii] <- player_name
+    sum_df$n_obs[ii] <- out_list$n_obs
[…]

### 2019-08-20 contest/model-fitting/data-description.Rmd
paper changes

---
title: "Data description"
author: "Gallagher, Frisoli, and Luby"
date: "August 6, 2019"
output:
  pdf_document
---

```{r setup, eval = FALSE}
library(courtsports)
library(knitr)
library(kableExtra)
library(broom)
library(tidyverse)
```

```{r invis-setup, echo = FALSE}
devtools::load_all("~/courtsports")
library(knitr)
library(kableExtra)
library(broom)
library(tidyverse)
```

## `gs_players`

```{r}
data(gs_players)
dim(gs_players)
```

The data contained in `gs_players` is extracted from Jeff Sackman's tennis repository via the `R deuce` package and shown here for convenience.  It consists of *all* non-qualifying round grandslam matches between 2013-2017.  There are 127 matches in each round (which will appear twice, one for each player of reference), two leagues,  four grandslams per year, and five years.  The help files from both these sources are useful resources to exploring the data.

The covariates in this data set are the following:

- `tourney_id` unique ID for tournament round with format `{year}-{league}-{tournament type}-{slam ID}-{round}-{year}` (character)

- `tourney_name` Full name of the slam (character)

- `surface` - one of "Hard," "Clay," or "Grass" (character)

- `draw_size` - number of total players in the tournament (integer)

- `tourney_level` identifier of whether tournament is a Grand Slam (character)

- `match_num` Unique ID for the match within the tournament (integer)

- `id` unique player ID (integer)

- `seed` seed of player (integer/NA)

- `entry` whether the player was a qualifier (character)

- `name` full player name

- `hand` handedness of the player (character)

- `ht` height of player in cm (integer)

- `ioc` IOC/country of player (character)

- `age` age in years (numeric)

- `rank` rank of player at time of match (integer)

- `rank_points` points of player at time of match (integer)

- `opponent_rank` rank of opponent (integer)

- `score` final score of match (character)

- `best_of` maximum number of sets (integer)

- `round` round of the match (character)

- `minutes` match duration in minutes (numeric/NA)

- `aces` number of aces in the match by the player (integer)

- `df` number of double faults by the player in the match (integer)

- `svpt` number of service points for the player (integer)

- `1stIn` number of first serves in by the player (integer)

- `1stWon` number of first serves won (integer)

- `2ndWon` number of second serves won (integer)

- `SvGms` number of service games (integer)

- `bpSaved` number of break points saved (integer)

- `bpFaced` number of break points faced (integer)

- `Retirement` whether the player retired (forfeit) or not (logical)

- `{X}` number of games won in the first set where $X = 1, \dots, 5$ (integer)

- `TB{Y}` number of points player won in the tie breaker of set $Y$ for $Y = 1, \dots, 5$ (integer)

- `tourney_start_date` date the tournament began YYYY-MM-DD (date)

- `match_id` combination of `tourney_id` and `match_id` (character)

- `league` one of "WTA" or "ATP" (character)

`tournament` standardized tournament names (character)

- `pointswon` total number of points player won (integer)

- `gameswon` total number of games player won (integer)

- `setswon` total number of sets player won (integer)

- `total_points` total number of points player won (integer)

- `did_win` indicator of whether player won 1 = Won, 0 = Lost

Below is a summary of each of the variables.

```{r summary-gs-players, echo = FALSE}
gs_players %>% summary() %>% t() %>% 
  kable(., format = "latex",  
      booktabs = TRUE, digits = 1, 
      caption = "Summary of covariates in \\texttt{gs\\_players}.",
      escape = TRUE, linesep = "") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

## `gs_pbp`

The data contained in `gs_pbp` referes to the grandslam point-by-point data that was recorded between 2013 and 2017.  This is *crowd-sourced* data and consists of `R nrow(gs_pbp)` matches for both leagues at the four grand slams.

Each row in this data set is a match with summarized statistics such as number of winners or average serve speed.  This includes information for both players.  The covariates are

- `match_id` unique match ID (character)

- `player1` name of first player (character)

- `player2` name of second player (character)

- `tour` either "atp" or "wta" (character)

- `slam` Grand slam name (character)

- `year` (integer)

- `tournament` standardized tournament name (character)

- `ave_serve_speed_p{X}` average service speed in mph of player $X$ (numeric)

- `n_{Y}_p{X}` number of trait Y for player X where $Y$ is one of `aces` (aces), `winners` (winners), `netpt_w` (net points won), `netpt` (total net points), `bp_w` (break points won), `bp` (total break points), `ue` (unforced errors), `sv_w` (service points won), and `sv` (total number of service points)

- `aX` player names

```{r summary-gs-pbp, echo = FALSE}
gs_pbp %>% summary() %>% t() %>% 
  kable(., format = "latex",  
      booktabs = TRUE, digits = 1, 
      caption = "Summary of covariates in \\texttt{gs\\_pbp}.",
      escape = TRUE, linesep = "") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

## `gs_partial`

The data contained in `gs_partial` consists of 3066 matches and 45 covariates.  It is created by joining `gs_players` and `gs_pbp` together.  Each row in this data set corresponds to one match.

All of the covariate descriptions above and covariates are prefixed by either `w_/l_` or `winner_/loser_` to indicate who won/lost, the match.

The summary table of the covariates is displayed below.

```{r summary-gs-partial, echo = FALSE}
gs_partial %>% summary() %>% t() %>% 
  kable(., format = "latex",  
      booktabs = TRUE, digits = 1, 
      caption = "Summary of covariates in \\texttt{gs\\_partial}.",
      escape = TRUE, linesep = "") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

## `gs_partial_players`

The data contained in `gs_partial_players` contains 6132 observations and 33 covariates.  Each row corresponds to one match with one player.  It is created by taking `gs_partial` and separately looking at the covariates for each player in the match.  All the covariates are described above.

The summary table of the covariates is displayed below.

```{r summary-gs-partial-players, echo = FALSE}
gs_partial_players %>% summary() %>% t() %>% 
  kable(., format = "latex",  
      booktabs = TRUE, digits = 1, 
      caption = "Summary of covariates in \\texttt{gs\\_partial\\_players}.",
      escape = TRUE, linesep = "") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

### 2019-08-20 contest/many-players/most_matches.R

@@ -383,6 +383,13 @@ pred_df <- gs_partial_players %>% group_by(name) %>%
     filter(name %in% top_players$name) %>%
     summarize_if(is.numeric, quantile, na.rm = TRUE, probs = quant)
 
+## rank difference
+gs_partial_players %>%
+    filter(name %in% top_players$name)  %>%
+    group_by(name) %>% 
+    summarize( diff = (max(rank, na.rm = TRUE) - min(rank, na.rm = TRUE))) %>%
+    filter(diff >= 50)
+
 pred_list <- vector(mode = "list", length = L)
 for(ll in 1:L){
     ii <- which(top_players$name == as.character(unique(player_df$name))[ll])
@@ -465,9 +472,9 @@ ggsave("individual-models-results-mod-sub.pdf", width = 10, height = 14)
 ## diags for federer DONE
 ## and interpretation of model DONE
 ## table of # of observations, # of covariates, R^2, and VIF DONE
-## rewrite interps of cool graphs
-## add how this work will transform tennis forever
+## rewrite interps of cool graphs DONE
+## add how this work will transform tennis forever DONE
 ## Explain about different styles of play and emphasize court differences in
-## discussion/intro
-## variable discussion and statistics ugh
+## discussion/intro DONE
+## variable discussion and statistics ugh DONE
 ## R package maintenace// vignette mostly to get EDA, hiearchical model, and individual models

### 2019-08-20 contest/graphics/graphics_for_paper.Rmd

@@
         family="serif"),
-        plot.title = element_text(hjust = 0.5, size=34))
+        plot.title = element_text(hjust = 0.5, size=34),
+        plot.subtitle = element_text(hjust = 0.5))
 
@@
 
+dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% pull(name) %>% unique() %>% length()
+dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% nrow()
+
+p1<- ggplot(france_wins, aes(round, win_prop,
+                             color=tournament, group=tournament,
+                             # alpha = tournament,
+                             size = tournament,
+                             linetype = tournament)) + 
+  # geom_line(size = 3) + 
+  # geom_point() +
+  labs(title = "French players",
+       subtitle = "53 players, n=828",
+       x = "", y = " ") + 
+  geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, 1, our_alpha, our_alpha)) + 
+scale_size_manual(guide = FALSE,
+                   values = c(1, 3, 1, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 1, 2, 2)) + ylim(c(0, 1))
+
+p2 <- ggplot(us_wins, aes(round, win_prop,
+                          color=tournament, group=tournament,
+                          # alpha = tournament,
+                          size = tournament,
+                          linetype=tournament)) + 
+  # geom_line() + 
+  # geom_point() +
+  labs(title = "USA players",
+       subtitle = "81 players, n=1046",
+              x = "", y = " ") + 
+       # x = "", y = "Win proportion") + 
+    geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+  scale_size_manual(guide = FALSE,
+                   values = c(1, 1, 3, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 2, 1, 2)) + ylim(c(0, 1))
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, our_alpha, 1, our_alpha))
+
+p3 <- ggplot(aus_wins, aes(round, win_prop,
+                           color=tournament, group=tournament, # alpha = tournament,
+                           size = tournament,
+                           linetype = tournament)) + 
+  # geom_line() + 
+  # geom_point() +
+  labs(title = "Aus. players",
+       subtitle = "40 players, n=376",
+       x = "", y = "Win proportion") + 
+         # x = "", y = " ") + 
+    geom_line() + 
+  scale_color_manual(guide=FALSE, values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme   + 
+  scale_size_manual(guide = FALSE,
+                   values = c(3, 1, 1, 1)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(1, 2, 2, 2)) + ylim(c(0, 1))
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(1, our_alpha, our_alpha, our_alpha))
+
+p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
+                           group=tournament, fill = tournament,
+                           # alpha = tournament,
+                           size = tournament,
+                           linetype = tournament)) + 
+  geom_line() + 
+  labs(title = "British players",
+       subtitle = "21 players, n=249",
+       # x = "", y = "Win proportion") + 
+       x = "", y = " ") + 
+  scale_color_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme +
+  guides(color = guide_legend(override.aes = list(size = 7)))  + 
+  scale_size_manual(guide = FALSE,
+                   values = c(1, 1, 1, 3)) + 
+  scale_linetype_manual(guide = FALSE,
+                   values = c(2, 2, 2, 1)) + ylim(c(0, 1))
+
+# scale_alpha_manual(guide = FALSE,
+#                    values = c(our_alpha, our_alpha, our_alpha, 1))
+
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(.,
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           just = "left",
+                           gp=gpar(fontfamily="serif", fontsize = 20)))
+
+ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
+grid.arrange(.,
+             # top = textGrob("Home court advantage: debunked",
+             #                gp=gpar(fontsize=20, fontfamily="serif")),
+             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
+                           just = "left",
+                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
+  ggsave("plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
+
+```
+
+```{r}
+# home court only with those in all 4
+
+all4 <- 
+  gs_players %>% 
+  group_by(name) %>% 
+  summarize(blah = length(unique(tournament))) %>%
+  filter(blah == 4) %>%
+  pull(name)
+
+france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF", name %in% all4) %>% 
+  dplyr::group_by(tournament, round) %>% 
+  dplyr::summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF", name %in% all4) %>% 
+  dplyr::group_by(tournament, round) %>% 
+  dplyr::summarize(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+aus_wins <- filter(gs_players, ioc == "AUS", round < "QF", name %in% all4) %>% 
+  group_by(tournament, round) %>% 
+  dplyr::summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
+gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF", name %in% all4) %>% 
+  group_by(tournament, round) %>% 
+  dplyr::summarise(win_prop = mean(did_win),
+            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
+
 p1<- ggplot(france_wins, aes(round, win_prop,
@@
   scale_linetype_manual(guide = FALSE,
-                   values = c(2, 1, 2, 2))
+                   values = c(2, 1, 2, 2)) + ylim(c(0, 1))
 
@@
   scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 1, 2))
+                   values = c(2, 2, 1, 2)) + ylim(c(0, 1))
 # scale_alpha_manual(guide = FALSE,
@@
   scale_linetype_manual(guide = FALSE,
-                   values = c(1, 2, 2, 2))
+                   values = c(1, 2, 2, 2)) + ylim(c(0, 1))
 # scale_alpha_manual(guide = FALSE,
@@
   scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 2, 1))
+                   values = c(2, 2, 2, 1)) + ylim(c(0, 1))
 
@@
 grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
@@
 grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
+             # top = textGrob("Only those who made all 4",
              #                gp=gpar(fontsize=20, fontfamily="serif")),
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court.jpg", ., height=5, width=graphic_width_long)
+  ggsave("plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
 

### 2019-08-21 contest/graphics/graphics_for_paper.Rmd
update aces to ace percentage

@@
 ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
+
+```
+
+```{r, fig.width=10, fig.height=4.8, fig.align='center'}
+
+set.seed(8)
+
+atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
+                 "Roger Federer", "Andy Murray")
+atp_servers <- sapply(atp_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
+                 "Venus Williams", "Serena Williams", "Madison Keys")
+wta_servers <- sapply(wta_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+p_atp_percent <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), ace/svpt)) + 
+  geom_jitter(aes(color = tournament), size=1.5) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), ace/svpt, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), ace/svpt, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+  labs(x = "log(height + 1)", y = "ace% = aces / serves", title = "ATP") +
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+p_wta_percent <- gs_players %>% filter(league == "WTA") %>%
+ggplot(aes(log(ht+1), ace/svpt)) + 
+  geom_jitter(aes(color = tournament)) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth(method = "lm", color="black") + 
+  geom_label_repel(data=wta_servers,
+            aes(log(ht+1), ace/svpt, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
+  geom_label_repel(data=wta_servers,
+            aes( log(ht+1), ace/svpt, label=name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
+  labs(x = "log(height + 1)", y = "ace%  = aces / serves", title = "WTA")+
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
+#              top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom") 
+# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+# ggsave("plots/aces.jpg", height=7, width=16)
+ggsave("plots/aces2_percent.jpg", height=7, width=graphic_width_long)
 

### 2019-08-21 contest/graphics/graphics_for_paper.Rmd
update code for p-values for chi squared tests spainiards

--- 2019-08-20 contest/graphics/graphics_for_paper.Rmd
@@
 
+################################################################################
+# TESTS FOR PAPER
+################################################################################
 with(filter(gs_players, ioc == "ESP", round == "R128"),
@@
   chisq.test(correct = TRUE)
-
+with(filter(gs_players, ioc == "ESP", round == "R64"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
+with(filter(gs_players, ioc == "ESP", round == "R32"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
+with(filter(gs_players, ioc == "ESP", round == "R16"),
+     table(did_win, tournament)) %>%
+  chisq.test(correct = TRUE)
 ```
@@
 ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
+
+```
+
+```{r, fig.width=10, fig.height=4.8, fig.align='center'}
+
+set.seed(8)
+
+atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
+                 "Roger Federer", "Andy Murray")
+atp_servers <- sapply(atp_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
+                 "Venus Williams", "Serena Williams", "Madison Keys")
+wta_servers <- sapply(wta_servers,
+                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
+  as.numeric() %>% gs_players[., ]
+
+p_atp_percent <- gs_players %>% filter(league == "ATP") %>%
+ggplot( aes(log(ht+1), ace/svpt)) + 
+  geom_jitter(aes(color = tournament), size=1.5) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), ace/svpt, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 7) + 
+  geom_label_repel(data=atp_servers,
+            aes(log(ht+1), ace/svpt, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 7) +
+  labs(x = "log(height + 1)", y = "ace% = aces / serves", title = "ATP") +
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+p_wta_percent <- gs_players %>% filter(league == "WTA") %>%
+ggplot(aes(log(ht+1), ace/svpt)) + 
+  geom_jitter(aes(color = tournament)) + 
+  # scale_color_manual(" ", values = tournament_colors) + 
+  geom_smooth(method = "lm", color="black") + 
+  geom_label_repel(data=wta_servers,
+            aes(log(ht+1), ace/svpt, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 7) + 
+  geom_label_repel(data=wta_servers,
+            aes( log(ht+1), ace/svpt, label=name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 7) +
+  labs(x = "log(height + 1)", y = "ace%  = aces / serves", title = "WTA")+
+  my_theme + scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
+#              top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom") 
+# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
+#                             gp=gpar(fontsize=20, fontfamily="serif")))
+
+# ggsave("plots/aces.jpg", height=7, width=16)
+ggsave("plots/aces2_percent.jpg", height=8, width=graphic_width_long)
 

### 2019-08-27 paper_results/Data/data-description.Rmd
Formatting readme

--- 2019-08-20 contest/model-fitting/data-description.Rmd
@@
 author: "Gallagher, Frisoli, and Luby"
-date: "August 6, 2019"
+date: "September 5, 2019"
 output:
-  pdf_document
+  html_document
 ---
@@
 
-```{r invis-setup, echo = FALSE}
+```{r invis-setup, echo = FALSE, warning = FALSE, message = FALSE}
 devtools::load_all("~/courtsports")
@@
 gs_players %>% summary() %>% t() %>% 
-  kable(., format = "latex",  
+  kable(., format = "html",  
       booktabs = TRUE, digits = 1, 
-      caption = "Summary of covariates in \\texttt{gs\\_players}.",
+      caption = "Summary of covariates in gs_players.",
       escape = TRUE, linesep = "") %>%
-  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
 ```
@@
 gs_pbp %>% summary() %>% t() %>% 
-  kable(., format = "latex",  
+  kable(., format = "html",  
       booktabs = TRUE, digits = 1, 
-      caption = "Summary of covariates in \\texttt{gs\\_pbp}.",
+      caption = "Summary of covariates in gs_pbp.",
       escape = TRUE, linesep = "") %>%
-  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
 ```
@@
 gs_partial %>% summary() %>% t() %>% 
-  kable(., format = "latex",  
+  kable(., format = "html",  
       booktabs = TRUE, digits = 1, 
-      caption = "Summary of covariates in \\texttt{gs\\_partial}.",
+      caption = "Summary of covariates in gs_partial.",
       escape = TRUE, linesep = "") %>%
-  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) #%>%
+  ##kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
 ```
@@
 gs_partial_players %>% summary() %>% t() %>% 
-  kable(., format = "latex",  
+  kable(., format = "html",  
       booktabs = TRUE, digits = 1, 
-      caption = "Summary of covariates in \\texttt{gs\\_partial\\_players}.",
+      caption = "Summary of covariates in gs_partial_players}.",
       escape = TRUE, linesep = "") %>%
-  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
+ ## kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
 ```

### 2019-08-27 paper_results/EDA/eda.Rmd
Updated readme

--- 2018-09-07 contest/eda/eda.Rmd
@@
 title: "EDA"
-author: "The Competent Ones"
-date: "August 4, 2018"
+author: ""
+date: "Sept. 5, 2019"
 output: 
@@
 library(knitr)
-# devtools::load_all("..") # Loads courtsports package
+#library(courtsports) ## UNCOMMENT if running
+#devtools::load_all("..") # Loads courtsports package
+
 ```

### 2019-08-27 paper_results/EDA/makeGSF.R

@@ -1,0 +1,16 @@
+library(deuce)
+library(dplyr)
+
+data("point_by_point")
+pbp = tbl_df(point_by_point)
+pbp$wim = grepl("Wimbledon", pbp$tny_name)
+pbp$french = grepl("FrenchOpen", pbp$tny_name)
+pbp$aus = grepl("AustralianOpen", pbp$tny_name)
+pbp$us = grepl("USOpen", pbp$tny_name)
+pbp$qual = grepl("Qualifying", pbp$tny_name)
+
+# Create data for Grand Slam finals
+gsf = select(pbp, c("tny_name", "wim", "french", "us", "aus", "qual")) %>%
+  filter(., wim | french | us | aus & !qual)
+
+save(gsf, file = "GrandSlamFinals.RData")

### 2019-08-27 paper_results/Hierarchical/fit-and-plot.R

@@ -1,0 +1,337 @@
+library(courtsports)
+library(tidyverse)
+library(broom)
+library(lme4)
+
+## Define theme and colors
+
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+                            family="serif"),
+        plot.title = element_text(hjust = 0.5, size=16))
+
+win_colors <- c("#336699", "#339966")
+
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+## Define variables for modeling
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+gs_players$name_fac = as.factor(gs_players$name)
+gs_players$hand_fac = as.factor(gs_players$hand)
+gs_players$ioc_fac = as.factor(gs_players$ioc)
+gs_players$atp = gs_players$league == "ATP"
+gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
+
+gs_partial_players$name_int = as.integer(as.factor(gs_partial_players$name))
+gs_partial_players$name_fac = as.factor(gs_partial_players$name)
+gs_partial_players$ioc_fac = as.factor(gs_partial_players$ioc)
+gs_partial_players$atp = gs_partial_players$Tour == "atp"
+gs_partial_players$year_fac = as.factor(gs_partial_players$year)
+gs_partial_players$late_round = gs_partial_players$round >= "R16"
+gs_partial_players$seeded = gs_partial_players$rank <= 32
+gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
+
+## Fit models used in paper
+set.seed(091418)
+ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial", nAGQ =0)
+n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_partial_players)
[…]

### 2019-08-27 paper_results/Individual/most_matches.R

@@ -1,0 +1,480 @@
+## Script to get Federer individual model
+## SKG July 20, 2019
+
+
+
+
+library("courtsports")
+library(ggplot2)
+library(forcats)
+library(dplyr)
+library(tidyr)
+library(ggcorrplot)
+library(gridExtra)
+library(knitr)
+library(broom)
+library(kableExtra)
+library(ggDiagnose)
+
+
+data(gs_partial_players)
+
+########################3
+## THeme
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+        family="serif"),
+        plot.title = element_text( size=16))
+
+tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
+league_colors <- c("#0C2340", "#902BA3")
+win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+########################################
+
+player_name <- "Roger Federer"
+data <- gs_partial_players
+ref <- "Wimbledon"
+start_lower <- TRUE
+min_year <- 2013
+max_year <- 2017
+seed <- 2020
+set.seed(seed)
+test_prop <- 0
+#
+out_list <- courtsports::model_individual(
+                              player_name = player_name,
+                              data = data,
+                              ref = ref,
+                              start_lower = start_lower,
+                              test_prop = test_prop,
+                              min_year = min_year,
+                              max_year = max_year,
+                              seed = seed)
+#
+summary(out_list$final_model)
+par(mfrow = c(2,2))
+plot(out_list$final_model)
+print(out_list$n_obs)
+df <- tidy(out_list$final_model)
+kable(df[,-4], format = "latex",
+      row.names = FALSE, col.names = c("Coef.", "Est.",
+                                       "Std. Err.", "p-value"),
+             caption = "\\label{tab:fed-ind-coef} Modeling coefficients for Federer's best fit individual model using Wimbledon as a reference court.",
[…]

### 2019-09-02 paper_results/Hierarchical/fit-and-plot.R
regression tables for appendix

@@ -3,6 +3,9 @@ library(tidyverse)
 library(broom)
 library(lme4)
 
+# Need development version of tidyr (pivot_wider and pivot_longer)
+# devtools::install_github("tidyverse/tidyr")
+
 ## Define theme and colors
 
 my_theme <-  theme_bw() + # White background, black and white theme
@@ -37,11 +40,13 @@ gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
 
 ## Fit models used in paper
 set.seed(091418)
-ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial", nAGQ =0)
-n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_partial_players)
-n_winners_mod = lme4::lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_partial_players)
-n_net_mod = lme4::lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_partial_players)
+ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_players, family = "binomial", nAGQ =0)
+n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp +
+                          (0 + tournament |name_fac), data = gs_partial_players)
+n_net_mod = lme4::lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp +
+                         (0 + tournament |name_fac), data = gs_partial_players, control = lmerControl(optimizer = "bobyqa"))
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac),
[…]

### 2019-09-05 paper_results/EDA/graphics_eda.Rmd
add graphs for the EDA section and for the complete data section

--- 2019-08-21 contest/graphics/graphics_for_paper.Rmd
@@
 title: "Opening up the court (surface) in tennis grand slams"
-date: '`r format(Sys.time(), "%B %d, %Y")`'
+date: ' '
 output:
   pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
+    toc:  true
 header-includes:
   - \usepackage{hyperref}
-abstract: "IDK writing is hard"
 ---
@@
 devtools::load_all("../..")
-
-```
-
-```{r data-cleaning, include = FALSE}
-data(gs)
-data(gs_players)
-data(gs_partial)
-data(gs_partial_players)
 ```
@@
 ```{r graphing-parameters, include = FALSE}
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 12),
-#         text = element_text(size = 14,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=16))
-
 my_theme <-  theme_bw() + # White background, black and white theme
@@
 
+# SHOULD WE WRITE/SAVE THE GRAPHICS LOCALLY??
+save_graph <- FALSE
 ```
 
-```{r modelling, eval = FALSE}
-## HIERARCHICAL MODELLING
-## NOT RUN
-## WARNING: Takes about 1 HOUR
-
-gs_players$name_int = as.integer(as.factor(gs_players$name))
-gs_players$hand_fac = as.factor(gs_players$hand)
-gs_players$ioc_fac = as.factor(gs_players$ioc)
-gs_players$atp = gs_players$league == "ATP"
-gs_players$year_fac = as.factor(gs_players$year)
-gs_players$late_round = gs_players$round >= "R16"
-gs_players$seeded = gs_players$rank <= 32
-gs_players$opponent_seeded = gs_players$opponent_rank <= 32
-
-set.seed(091418)
-nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
-ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
-ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-
+```{r data-cleaning, include = FALSE}
+data(gs)
+data(gs_players)
+data(gs_partial)
+data(gs_partial_players)
 ```
 
-# Data and EDA {#sec:data-eda}
+# EDA
 
-## Data {#sec:data}
+## Distribution of points earned
 
-```{r tab-data, include = TRUE, echo = FALSE}
-w_id <- which(gs$winner_name == "Serena Williams")[1:2]
-f_id <- which(gs$winner_name == "Roger Federer")[1:2]
-n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
-s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
-cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
-knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
-             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
-             caption = "\\label{tab:data}Example of the complete grand slam data.
-It includes winner and loser attributes, match attributes, and tournament attributes.
-Not all attributes are shown here.",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
-```
-
-## Early Data Analysis {#sec:eda}
-
-### Examining the distribution of points earned
-
-```{r, fig.height=5, fig.width=10}
-
-# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
-
+```{r, fig.height=8, fig.width=16}
 p2 <- ggplot(gs_players) + 
@@
   labs(x = "Points", y = "Scaled density")
-
-# 
-# ggplot(gs_players) + geom_density(aes(pointswon, color=tournament, 
-#                                             group=tournament, fill = tournament),
-#                                         size = 3, alpha = 0) + 
-#   scale_fill_manual(" ", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) + my_theme  +
-#   guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
-#                               nrow=2, byrow=TRUE)) +
-#     theme(legend.position="bottom") +
-#     scale_color_manual(guide=FALSE, values=tournament_colors)
 
@@
 
+grid.arrange(p2, p1, ncol=2)
+
+if(save_graph){
   grid.arrange(p2, p1, ncol=2) %>%
-             # # top = textGrob("Distribution of points per match",
-             # #                gp=gpar(fontsize=20, fontfamily="serif")),
-             #  layout_matrix = rbind(c(2, 1),
-             #                        c(2, 1),
-             #                        c(NA, 1)),
-             # heights=c(1, 3, .1)) %>%
   ggsave("plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
-
-# 
-# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-# grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-#                             gp=gpar(fontsize=20, fontfamily="serif")),
-#              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-#                            # x = 0,
-#                            # y = 0.5,
-#                            just = "left",
-#                            gp=gpar(fontfamily="serif")))
-# 
-# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-#   ggsave("plots/spainards.jpg", ., height=7, width=16)
-
+}
 ```
 
-### Home court advantage
+## Home court advantage
 
-```{r, fig.height=3.5, fig.width=10, fig.align='center'}
+```{r}
 france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF") %>% 
@@
 
-our_alpha <- 0.5
+```
 
+```{r, echo=TRUE}
+dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% 
+  pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>%
+  pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% 
+  pull(name) %>% unique() %>% length()
+
+dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% nrow()
+dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% 
+  pull(name) %>% unique() %>% length()
+```
+
+```{r, fig.height=5.5, fig.width=16, fig.align='center'}
 p1<- ggplot(france_wins, aes(round, win_prop,
                              color=tournament, group=tournament,
-                             alpha = tournament,
-                             size = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
-  labs(title = "French players",
-       x = "", y = " ") + 
-  geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, 1, our_alpha, our_alpha)) + 
-scale_size_manual(guide = FALSE,
-                   values = c(1, 3, 1, 1))
-
-p2 <- ggplot(us_wins, aes(round, win_prop,
-                          color=tournament, group=tournament,
-                          alpha = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "USA players",
-              x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
-    geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, our_alpha, 1, our_alpha))
-
-p3 <- ggplot(aus_wins, aes(round, win_prop,
-                           color=tournament, group=tournament, alpha = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "Aus. players",
-       x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
-    geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(1, our_alpha, our_alpha, our_alpha))
-
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
-                           group=tournament, fill = tournament,
-                           alpha = tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "British players",
-       # x = "", y = "Win proportion") + 
-       x = "", y = " ") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))  + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, our_alpha, our_alpha, 1))
-
-# 
-# + 
-#   geom_line() + 
-#   # geom_point() +
-#   labs(title = "British players",
-#        x = "", y = "Win proportion") + 
-#   scale_color_manual(" ", values=tournament_colors) + my_theme
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("Home court advantage: debunked",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court.jpg", ., height=5, width=graphic_width_long)
-
-```
-
-```{r}
-
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% pull(name) %>% unique() %>% length()
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% nrow()
-
-p1<- ggplot(france_wins, aes(round, win_prop,
-                             color=tournament, group=tournament,
-                             # alpha = tournament,
                              size = tournament,
                              linetype = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
   labs(title = "French players",
@@
                                            "Wimbledon\n(Grass)")) + my_theme   + 
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, 1, our_alpha, our_alpha)) + 
 scale_size_manual(guide = FALSE,
@@
                           color=tournament, group=tournament,
-                          # alpha = tournament,
                           size = tournament,
                           linetype=tournament)) + 
-  # geom_line() + 
-  # geom_point() +
   labs(title = "USA players",
@@
               x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
     geom_line() + 
@@
                    values = c(2, 2, 1, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, 1, our_alpha))
 
@@
                            linetype = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
   labs(title = "Aus. players",
@@
        x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
     geom_line() + 
@@
                    values = c(1, 2, 2, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(1, our_alpha, our_alpha, our_alpha))
 
@@
                            group=tournament, fill = tournament,
-                           # alpha = tournament,
                            size = tournament,
@@
        subtitle = "21 players, n=249",
-       # x = "", y = "Win proportion") + 
        x = "", y = " ") + 
@@
 
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, our_alpha, 1))
-
 ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) 
-
+if(save_graph){
 ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
 grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
@@
   ggsave("plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
-
+}
 ```
 
-```{r}
-# home court only with those in all 4
+### Subset: players who played in all 4 tournaments
 
+```{r, fig.height=5, fig.width=16, fig.align='center'}
 all4 <- 
@@
                              color=tournament, group=tournament,
-                             # alpha = tournament,
                              size = tournament,
                              linetype = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
   labs(title = "French players",
@@
                           color=tournament, group=tournament,
-                          # alpha = tournament,
                           size = tournament,
                           linetype=tournament)) + 
-  # geom_line() + 
-  # geom_point() +
   labs(title = "USA players",
               x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
     geom_line() + 
@@
                            linetype = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
   labs(title = "Aus. players",
        x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
     geom_line() + 
@@
   labs(title = "British players",
-       # x = "", y = "Win proportion") + 
        x = "", y = " ") + 
@@
 
+if(save_graph){
 ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
 grid.arrange(.,
-             # top = textGrob("Only those who made all 4",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
@@
   ggsave("plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
-
+}
 ```
@@
 ```{r}
-
 spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
@@
 
+```
+
+### Chi-squared tests
+
+```{r}
 ################################################################################
@@
 
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
+```{r, fig.height=6, fig.width=16, fig.align='center'}
 p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
   labs(title = "Win proportion among Spaniards",
-       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
        x = "", y = "Win proportion") + 
@@
 
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           # x = 0,
-                           # y = 0.5,
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
+ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom")
 
+if(save_graph){
 ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
   ggsave("plots/spainards.jpg", ., height=6, width=graphic_width_long)
+}
 
@@
 
-# Unforced errors and aces
+## Unforced errors and aces
 
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
-# ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-#                            label = tournament, color=tournament)) + 
-#   geom_jitter(size=2) +
-#   labs( title = "Distribution of errors and aces differ by tournament",
-#         x = "Total unforced errors", y = "Total aces") +
-#   my_theme + scale_color_manual("Tournament",
-#                                 values = tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)\n",
-#                                            "French Open\n(Clay)\n",
-#                                            "US Open\n(DecoTurf)\n",
-#                                            "Wimbledon\n(Grass)\n")) + 
-#   geom_density_2d()
-
+```{r, fig.width=12, fig.height=6, fig.align='center'}
 ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
@@
 
+if(save_graph){
 ggsave("plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
-
+}
 ```
@@
 
-```{r, fig.width=10, fig.height=4.8, fig.align='center'}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 26),
-#         text = element_text(size = 30,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=36))
-
+```{r, fig.width=16, fig.height=7, fig.align='center'}
 set.seed(8)
@@
 
-# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
+ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom")
 
-ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
-# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-# ggsave("plots/aces.jpg", height=7, width=16)
+if(save_graph){
 ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
-
+}
 ```
 
-```{r, fig.width=10, fig.height=4.8, fig.align='center'}
+#### Using ace percentage
 
+```{r, fig.width=16, fig.height=8, fig.align='center'}
 set.seed(8)
@@
 
-# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
+ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom")
 
-ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom") 
-# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-# ggsave("plots/aces.jpg", height=7, width=16)
+if(save_graph){
 ggsave("plots/aces2_percent.jpg", height=8, width=graphic_width_long)
-
+}
 ```
-
-```{r, fig.width=6, fig.height=6, fig.align='center'}
-
-# p_atp <- gs_players %>% filter(league == "ATP") %>%
-# ggplot( aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
-#   geom_smooth( method = "lm", se=FALSE, color="black") + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label = name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "ATP") +
-#   my_theme
-# 
-# # subset(filter(gs_players, league == "WTA"),
-# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-# #                                  (log(ace + 1) > 2.82))
-# p_wta <- gs_players %>% filter(league == "WTA") %>%
-# ggplot(aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual("Tournament", values = tournament_colors) + 
-#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "WTA")+
-#   my_theme + theme(legend.position="bottom")
-# 
-# 
-# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-```
-
-### Linear regression for other outcomes
-
-```{r, eval = FALSE}
-gs_partial_players$atp = gs_partial_players$Tour == "atp"
-gs_partial_players$year_fac = as.factor(gs_partial_players$year)
-gs_partial_players$late_round = gs_partial_players$round >= "R16"
-gs_partial_players$seeded = gs_partial_players$rank <= 32
-gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-```
-
-```{r}
-load("../../data/n_aces_mod.rda")
-load("../../data/n_net_mod.rda")
-load("../../data/n_ue_mod.rda")
-load("../../data/n_winners_mod.rda")
-```
-
-```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 20),
-#         text = element_text(size = 24,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=26))
-
-aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval,
-                ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = aces") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
-#   mutate(tournament = gsub("tournament", "", term)) %>%
-#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-#   geom_errorbar(height = .5, position = position_dodge()) +
-#   ylab("Coefficient\n") + 
-#   xlab("") +
-#   ggtitle("Winners") + 
-#   scale_color_manual("", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) +
-#   my_theme  +
-#   coord_flip() +
-#   theme(panel.grid.major = element_blank(),
-#         panel.grid.minor = element_blank(),
-#         axis.text.y = element_blank()) +
-#   geom_hline(yintercept = 0, color = "grey")
-
-nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = points won at net") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp),
-                y = condval,
-                ymin = condval- 2*condsd,
-                ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(width=.5), size =2) +
-  ylab("Coefficient\n") + 
-  xlab("") +
-  ggtitle("Y = unforced errors") + 
-  scale_color_manual("", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) +
-  my_theme  +
-  coord_flip() +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.text.y = element_blank()) +
-  geom_hline(yintercept = 0, color = "grey") +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
-ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
-          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29))
-
-ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
-          common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
-ggsave("plots/player_effects.jpg", ., height=6, width = graphic_width_long)
-
-```
-
-## player effects
-
-```{r nadal-model}
-
-step_model_n <- courtsports:::model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
-
-std_res_n <- rstandard(step_model_n)
-
-gn <- ggplot(lm(step_model_n$formula, data = step_model_n$data)) + geom_hline(yintercept = 0, col = "red") +
-  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
-  labs(title = "Nadal",
-       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
-
-coefs_n <- coefficients(step_model_n)
-cov_n <- vcov(step_model_n)
-
- ## NADAL COEF SIZE
-
-df_n <- data.frame(coeff = names(step_model_n$coeff), ave = step_model_n$coeff, sd = sqrt(diag(vcov(step_model_n))))
-df_n$player <- "Nadal"
-```
-
-```{r federer-model}
-
-step_model_f <- courtsports:::model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
-
-std_res_f <- rstandard(step_model_f)
-
-gf <- ggplot(lm(step_model_f$formula, data = step_model_f$data)) + geom_hline(yintercept = 0, col = "red") +
-  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
-  labs(title = "Federer",
-       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
-
-coefs_f <- coefficients(step_model_f)
-cov_f <- vcov(step_model_f)
-
-df_f <- data.frame(coeff = names(step_model_f$coeff), ave = step_model_f$coeff, sd = sqrt(diag(vcov(step_model_f))))
-df_f$player <- "Federer"
-```
-
-```{r williams-model}
-
-step_model_w <- courtsports:::model_individual(player_name = "Serena Williams", data = gs_partial_players,
-                                 start_lower = TRUE, ref = "Australian Open")
-
-std_res_w <- rstandard(step_model_w)
-
-gw <- ggplot(lm(step_model_w$formula, data = step_model_w$data)) + geom_hline(yintercept = 0, col = "red") +
-  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
-  labs(title = "Williams",
-       x = "Predicted Percent of Points Won", y = "Standardized Residuals")
-
-cov_w <- vcov(step_model_w)
-coefs_w <- coefficients(step_model_w)
-
- ## WilliamsCOEF SIZE
-
-df_w <- data.frame(coeff = names(step_model_w$coeff), ave = step_model_w$coeff, sd = sqrt(diag(vcov(step_model_w))))
-df_w$player <- "Williams"
-```
-
-```{r plot-individual-effects}
-df <- rbind(df_n, df_w, df_f)
-keep_coefs <- which(df$coeff %in% c("(Intercept)", "opponent_rank",
-                                    "pct_ace", "pct_bp", "pct_netpt",
-                                    "ave_serve_speed", "wue_ratio"))
-df <- df[keep_coefs,]
-ggplot(df, aes(x = coeff, y = ave, col = player)) +
-  geom_errorbar(aes(ymin = ave - 2 *sd, ymax = ave + 2 * sd),
-                position = "dodge", width = .3, size = 2) + 
-   my_theme +
-  # coord_flip() +
-  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
-  geom_hline(yintercept = 0, color = "grey") + 
-  ggtitle("Fixed effects") +
-  scale_color_brewer("", palette = "Dark2") +
-  theme(legend.position = "bottom") +
-  xlab("") + 
-  ylab("Coefficient") +
-  scale_x_discrete(labels = c("Intercept", "Opp. Rank", "% Aces",
-                              "% Break Points Won", "% Net Points Won",
-                              "Ave. Serve Speed", "Winner:UE"))  +
-  guides(color = guide_legend(override.aes = list(size = 7)))  +
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-ggsave("plots/ref-courts-eff.jpg", height=6, width = graphic_width_short)
-
-  #ggsave("~/Documents/ref-courts-eff.pdf", width = 8, height = 6, height=6, width = graphic_width_long)
-```
-
-# amanda model fitting
-
-```{r}
-win_colors <- c("#336699", "#339966")
-
-# with yellow for us
-# tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
-
-model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
-                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
-
-model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
-model.ests = data.frame()
-ran.effects = data.frame()
-for(mod in 1:nrow(model.sums)){
-  name = model_names[mod]
-  model.sums$aic[mod] = extractAIC(get(name))[2]
-  model.sums$edf[mod] = extractAIC(get(name))[1]
-  betas = tidy(get(name))
-  betas$model = name
-  if("group" %in% colnames(betas)){
-    model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
-                         dplyr::select(., -c("group")))
-  }
-  else model.ests = rbind(model.ests, betas)
-}
-
-model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef", "ind_no_ioc", "ind_intercept", "ind_year_ef")
-model.sums$fixed = c("No country, no tournament",
-                     "all",
-                     "no country, no tournament",
-                     "no tournament",
-                     "no country, no tournament",
-                     "no country",
-                     "no country, no year")
-model.sums$random = c("none",
-                      "none",
-                      "tournament (by country)",
-                      "tournament (by individual)",
-                      "tournament (by individual)",
-                      "intercept (by individual)",
-                      "year (by individual)")
-
-fixed.effects.plot = model.ests %>%
-  filter(., !grepl("ioc", term)) %>% 
-  filter(., model == 'ind_logistic_noioc') %>%
-  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
-  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
-  my_theme +
-  # coord_flip() +
-  theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
-  ggtitle("Fixed effects") +
-  scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "none") + 
-  xlab("") +
-  ylab("Coefficient") +
-  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "2014", "2015", "2016", "2017"))+
-  guides(col = guide_legend(nrow=2)) +
-  geom_hline(yintercept = 0, color = "grey") + 
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
-ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
-
-gs_players$name_int = as.integer(as.factor(gs_players$name))
-name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
-name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
-
-top_three_ranef = as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
-  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
-  ggplot(., aes(x = name, y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, color = tournament)) +
-  geom_errorbar(width = .5, position = "dodge",size = 1.5) + 
-  xlab("") + 
-  ylab("Coefficient") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  coord_flip() +
-  ggtitle("Random Effects") +
-  theme(legend.position = "bottom") + 
-  guides(col = guide_legend(nrow=2,
-                            override.aes = list(size = 7))) +
-  geom_hline(yintercept = 0, color = "grey") + 
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55) #,
-             # top = textGrob("Parameter estimates from logistic model",
-             #                gp=gpar(fontsize=20, fontfamily="serif"))
-             ) %>%
-ggsave("plots/logistic-parameter-effects.jpg",., height=7, width=graphic_width_long)
-
-```
-
-```{r}
-
-fixed.effects.plot.linear <- tidy(n_aces_mod) %>%
-  filter(., group == 'fixed') %>%
-  mutate(., model = "Aces") %>%
-  bind_rows(
-    tidy(n_ue_mod) %>%
-      filter(., group == 'fixed') %>%
-      mutate(., model = "UE")
-  ) %>%
-  bind_rows(
-    tidy(n_net_mod) %>%
-      filter(., group == 'fixed') %>%
-      mutate(., model = "Net")
-  ) %>%
-  ggplot(., aes(x = term, y = estimate, 
-                ymin = estimate - 2*std.error, 
-                ymax = estimate + 2*std.error,
-                col = model,
-                group = model)) + 
-  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
-  my_theme +
-  # coord_flip() +
-  theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
-  ggtitle("Fixed effects") +
-  scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "bottom") + 
-  xlab("") +
-  ylab("Coefficient") +
-  scale_x_discrete("", 
-                   breaks = waiver(), 
-                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)",
-                              "Rank (log)", "2014", "2015", "2016", "2017"))  +
-  guides(color = guide_legend(override.aes = list(size = 7)))  +
-  geom_hline(yintercept = 0, color = "grey") + 
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-ggsave("plots/linear-fixed-effects.jpg", fixed.effects.plot.linear,
-       height=6, width = graphic_width_short)
-
-```
-
-```{r}
-
-fixed.effects.plot.allmods = model.ests %>%
-  filter(., !grepl("ioc", term)) %>% 
-  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
-  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
-  my_theme +
-  # coord_flip() +
-  theme(axis.text.x=element_text(angle = 35, hjust = 1, size = 10)) +
-  ggtitle("Fixed effects") +
-  scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "none") + 
-  xlab("") +
-  ylab("Estimate") +
-  scale_x_discrete("", 
-                   breaks = waiver(), 
-                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017")) +
-  guides(col = guide_legend(nrow=2))
-
-aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Aces") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  coord_flip() +
-  theme(legend.position = "none")
-
-nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Net Wins") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  theme(axis.text.y = element_blank()) +
-  coord_flip() # +
-# theme(legend.position = "none")
-
-ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Unforced Errors") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  theme(axis.text.y = element_blank()) +
-  coord_flip()# + 
-#theme(legend.text=element_text(size=10))
-
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
-```
-
-# correlation plots
-
-```{r}
-win_colors <- c("#336699", "#339966")
-
-mat <- matrix(c(1, 0, .94, .93,
-         0, 1, -.05, -.36,
-         .94, -.05, 1, .88,
-         .93, -.36, .88, 1), byrow = TRUE, nrow = 4)
-colnames(mat) <- c("Aus. Open", "French Open", "US Open", "Wimbledon")
-rownames(mat) <- c("Aus. Open", "French Open", "US Open", "Wimbledon")
-#
-df <- melt(mat)
-#
-ggplot(df, aes(x = Var1, y = Var2, fill = value)) + geom_tile() +
-     geom_tile(color = "white")+
-    scale_fill_gradient2(low = win_colors[2], high = win_colors[1], mid = "white", 
-                         midpoint = 0, limit = c(-1,1), 
-                         space = "Lab", name="Correlation") + 
-                         # labels = c("-1.0", "-0.5", "  0.0", "  0.5", "  1.0")) + 
-  labs(x = "", y = "",
-       title = "Correlation matrix for random effects")  +
-  my_theme
-
-ggsave("plots/correlation.jpg", 
-       height=6, width = graphic_width_short)
-
-```
-
-# extras
-
-```{r, fig.height=3.5, fig.width=10, fig.align='center', eval=FALSE}
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_line()
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_point()
-
-filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = (pointswon - (total_points - pointswon)),
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-# grid.arrange(p1, p2, p3, p4, nrow=2)
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("WTA",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/wta.jpg", ., width = 10, height = 8)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("ATP",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/atp.jpg", ., width = 10, height = 8)
-
-```

### 2019-09-05 paper_results/EDA/graphics_model_complete_data.Rmd
add graphs for the EDA section and for the complete data section

--- 2019-08-20 contest/graphics/graphics_for_paper.Rmd
@@
 
+# SHOULD WE WRITE/SAVE THE GRAPHICS LOCALLY??
+save_graph <- FALSE
 ```
 
-```{r modelling, eval = FALSE}
-## HIERARCHICAL MODELLING
-## NOT RUN
-## WARNING: Takes about 1 HOUR
-
-gs_players$name_int = as.integer(as.factor(gs_players$name))
-gs_players$hand_fac = as.factor(gs_players$hand)
-gs_players$ioc_fac = as.factor(gs_players$ioc)
-gs_players$atp = gs_players$league == "ATP"
-gs_players$year_fac = as.factor(gs_players$year)
-gs_players$late_round = gs_players$round >= "R16"
-gs_players$seeded = gs_players$rank <= 32
-gs_players$opponent_seeded = gs_players$opponent_rank <= 32
-
-set.seed(091418)
-nocourt_logistic = glm(did_win ~ ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-base_logistic = glm(did_win ~ ioc_fac +  tournament + late_round + log(rank) + log(opponent_rank) + year_fac + atp, data = gs_players, family = "binomial")
-country_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |ioc_fac), data = gs_players, family = "binomial")
-ind_logistic = glmer(did_win ~  ioc_fac + late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_logistic_noioc = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_int), data = gs_players, family = "binomial")
-ind_int_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + tournament + (1|name_int), data = gs_players, family = "binomial")
-ind_year_logistic = glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + atp + tournament + (0+year_fac|name_int), data = gs_players, family = "binomial")
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-
+```{r data-cleaning, include = FALSE}
+data(gs)
+data(gs_players)
+data(gs_partial)
+data(gs_partial_players)
 ```
 
-# Data and EDA {#sec:data-eda}
-
-## Data {#sec:data}
-
-```{r tab-data, include = TRUE, echo = FALSE}
-w_id <- which(gs$winner_name == "Serena Williams")[1:2]
-f_id <- which(gs$winner_name == "Roger Federer")[1:2]
-n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
-s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
-cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
-knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
-             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
-             caption = "\\label{tab:data}Example of the complete grand slam data.
-It includes winner and loser attributes, match attributes, and tournament attributes.
-Not all attributes are shown here.",
-             booktabs = TRUE) %>% kable_styling(latex_options = "striped", "hold_position")
-```
-
-## Early Data Analysis {#sec:eda}
-
-### Examining the distribution of points earned
-
-```{r, fig.height=5, fig.width=10}
-
-# grid.arrange(p1, p2, p3, ncol=3, widths=c(.25, .41, .34))
-
-p2 <- ggplot(gs_players) + 
-  geom_density(aes(pointswon, ..scaled..,
-                   color=tournament, group=tournament, fill = tournament),
-                                        size = 3, alpha = 0) + 
-  scale_fill_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open (Plexicushion)",
-                                           "French Open (Clay)",
-                                           "US Open (DecoTurf)",
-                                           "Wimbledon (Grass)")) + my_theme  +
-  guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
-                              nrow=4, byrow=TRUE)) +
-    theme(legend.position="bottom") +
-    scale_color_manual(guide=FALSE, values=tournament_colors) +
-  facet_wrap(~league, ncol=1) +
-  labs(x = "Points", y = "Scaled density")
-
-# 
-# ggplot(gs_players) + geom_density(aes(pointswon, color=tournament, 
-#                                             group=tournament, fill = tournament),
-#                                         size = 3, alpha = 0) + 
-#   scale_fill_manual(" ", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) + my_theme  +
-#   guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
-#                               nrow=2, byrow=TRUE)) +
-#     theme(legend.position="bottom") +
-#     scale_color_manual(guide=FALSE, values=tournament_colors)
-
-p1 <- ggplot(gs_players) + 
-  geom_density(aes(pointswon, ..scaled..,
-                   color = factor(did_win),
-                   group = factor(did_win), fill = factor(did_win)),
-               size = 3, alpha = 0) + 
-  labs(x = "Points", y = " ") +
-  scale_fill_manual(" ", values=win_colors, labels = c("Lost game", "Won game")) +
-  scale_color_manual(guide = FALSE, values = win_colors) +
-  facet_wrap(~league, ncol=1) + 
-  my_theme +
-  theme(legend.position="bottom",
-        legend.spacing.y = unit(1.0, 'cm'),
-        legend.margin=margin(t = 2.2, unit='cm'))  +
-  guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
-                             nrow = 2,
-                            byrow=TRUE),
-        title.hjust = 5)
-
-grid.arrange(p2, p1, ncol=2) %>%
-             # # top = textGrob("Distribution of points per match",
-             # #                gp=gpar(fontsize=20, fontfamily="serif")),
-             #  layout_matrix = rbind(c(2, 1),
-             #                        c(2, 1),
-             #                        c(NA, 1)),
-             # heights=c(1, 3, .1)) %>%
-  ggsave("plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
-
-# 
-# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-# grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-#                             gp=gpar(fontsize=20, fontfamily="serif")),
-#              bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-#                            # x = 0,
-#                            # y = 0.5,
-#                            just = "left",
-#                            gp=gpar(fontfamily="serif")))
-# 
-# ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-#   ggsave("plots/spainards.jpg", ., height=7, width=16)
-
-```
-
-### Home court advantage
-
-```{r, fig.height=3.5, fig.width=10, fig.align='center'}
-france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF") %>% 
-  dplyr::group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF") %>% 
-  dplyr::group_by(tournament, round) %>% 
-  dplyr::summarize(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-our_alpha <- 0.5
-
-p1<- ggplot(france_wins, aes(round, win_prop,
-                             color=tournament, group=tournament,
-                             alpha = tournament,
-                             size = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
-  labs(title = "French players",
-       x = "", y = " ") + 
-  geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, 1, our_alpha, our_alpha)) + 
-scale_size_manual(guide = FALSE,
-                   values = c(1, 3, 1, 1))
-
-p2 <- ggplot(us_wins, aes(round, win_prop,
-                          color=tournament, group=tournament,
-                          alpha = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "USA players",
-              x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
-    geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, our_alpha, 1, our_alpha))
-
-p3 <- ggplot(aus_wins, aes(round, win_prop,
-                           color=tournament, group=tournament, alpha = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "Aus. players",
-       x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
-    geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(1, our_alpha, our_alpha, our_alpha))
-
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
-                           group=tournament, fill = tournament,
-                           alpha = tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "British players",
-       # x = "", y = "Win proportion") + 
-       x = "", y = " ") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))  + 
-scale_alpha_manual(guide = FALSE,
-                   values = c(our_alpha, our_alpha, our_alpha, 1))
-
-# 
-# + 
-#   geom_line() + 
-#   # geom_point() +
-#   labs(title = "British players",
-#        x = "", y = "Win proportion") + 
-#   scale_color_manual(" ", values=tournament_colors) + my_theme
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("Home court advantage: debunked",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court.jpg", ., height=5, width=graphic_width_long)
-
-```
-
-```{r}
-
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% pull(name) %>% unique() %>% length()
-
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% pull(name) %>% unique() %>% length()
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% nrow()
-
-p1<- ggplot(france_wins, aes(round, win_prop,
-                             color=tournament, group=tournament,
-                             # alpha = tournament,
-                             size = tournament,
-                             linetype = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
-  labs(title = "French players",
-       subtitle = "53 players, n=828",
-       x = "", y = " ") + 
-  geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, 1, our_alpha, our_alpha)) + 
-scale_size_manual(guide = FALSE,
-                   values = c(1, 3, 1, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 1, 2, 2)) + ylim(c(0, 1))
-
-p2 <- ggplot(us_wins, aes(round, win_prop,
-                          color=tournament, group=tournament,
-                          # alpha = tournament,
-                          size = tournament,
-                          linetype=tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "USA players",
-       subtitle = "81 players, n=1046",
-              x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
-    geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-  scale_size_manual(guide = FALSE,
-                   values = c(1, 1, 3, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 1, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, 1, our_alpha))
-
-p3 <- ggplot(aus_wins, aes(round, win_prop,
-                           color=tournament, group=tournament, # alpha = tournament,
-                           size = tournament,
-                           linetype = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "Aus. players",
-       subtitle = "40 players, n=376",
-       x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
-    geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-  scale_size_manual(guide = FALSE,
-                   values = c(3, 1, 1, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(1, 2, 2, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(1, our_alpha, our_alpha, our_alpha))
-
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
-                           group=tournament, fill = tournament,
-                           # alpha = tournament,
-                           size = tournament,
-                           linetype = tournament)) + 
-  geom_line() + 
-  labs(title = "British players",
-       subtitle = "21 players, n=249",
-       # x = "", y = "Win proportion") + 
-       x = "", y = " ") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))  + 
-  scale_size_manual(guide = FALSE,
-                   values = c(1, 1, 1, 3)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 2, 1)) + ylim(c(0, 1))
-
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, our_alpha, 1))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20)))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             # top = textGrob("Home court advantage: debunked",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
-
-```
-
-```{r}
-# home court only with those in all 4
-
-all4 <- 
-  gs_players %>% 
-  group_by(name) %>% 
-  summarize(blah = length(unique(tournament))) %>%
-  filter(blah == 4) %>%
-  pull(name)
-
-france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF", name %in% all4) %>% 
-  dplyr::group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF", name %in% all4) %>% 
-  dplyr::group_by(tournament, round) %>% 
-  dplyr::summarize(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-aus_wins <- filter(gs_players, ioc == "AUS", round < "QF", name %in% all4) %>% 
-  group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF", name %in% all4) %>% 
-  group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))
-
-p1<- ggplot(france_wins, aes(round, win_prop,
-                             color=tournament, group=tournament,
-                             # alpha = tournament,
-                             size = tournament,
-                             linetype = tournament)) + 
-  # geom_line(size = 3) + 
-  # geom_point() +
-  labs(title = "French players",
-       x = "", y = " ") + 
-  geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, 1, our_alpha, our_alpha)) + 
-scale_size_manual(guide = FALSE,
-                   values = c(1, 3, 1, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 1, 2, 2)) + ylim(c(0, 1))
-
-p2 <- ggplot(us_wins, aes(round, win_prop,
-                          color=tournament, group=tournament,
-                          # alpha = tournament,
-                          size = tournament,
-                          linetype=tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "USA players",
-              x = "", y = " ") + 
-       # x = "", y = "Win proportion") + 
-    geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-  scale_size_manual(guide = FALSE,
-                   values = c(1, 1, 3, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 1, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, 1, our_alpha))
-
-p3 <- ggplot(aus_wins, aes(round, win_prop,
-                           color=tournament, group=tournament, # alpha = tournament,
-                           size = tournament,
-                           linetype = tournament)) + 
-  # geom_line() + 
-  # geom_point() +
-  labs(title = "Aus. players",
-       x = "", y = "Win proportion") + 
-         # x = "", y = " ") + 
-    geom_line() + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme   + 
-  scale_size_manual(guide = FALSE,
-                   values = c(3, 1, 1, 1)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(1, 2, 2, 2)) + ylim(c(0, 1))
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(1, our_alpha, our_alpha, our_alpha))
-
-p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament,
-                           group=tournament, fill = tournament,
-                           # alpha = tournament,
-                           size = tournament,
-                           linetype = tournament)) + 
-  geom_line() + 
-  labs(title = "British players",
-       # x = "", y = "Win proportion") + 
-       x = "", y = " ") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))  + 
-  scale_size_manual(guide = FALSE,
-                   values = c(1, 1, 1, 3)) + 
-  scale_linetype_manual(guide = FALSE,
-                   values = c(2, 2, 2, 1)) + ylim(c(0, 1))
-
-# scale_alpha_manual(guide = FALSE,
-#                    values = c(our_alpha, our_alpha, our_alpha, 1))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20)))
-
-ggarrange(p3, p1, p2,  p4, nrow=1, ncol=4, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(.,
-             # top = textGrob("Only those who made all 4",
-             #                gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
-
-```
-
-## Spaniards on clay
-
-```{r}
-
-spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
-  group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
-            win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
-            avg_ranking = mean(rank),
-            rank_se = sd(rank)/sqrt(n()))
-
-prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament)), 2)
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-                table(did_win, tournament))
-
-with(filter(gs_players, ioc == "ESP", round == "R128"),
-     table(did_win, tournament)) %>%
-  chisq.test(correct = TRUE)
-
-```
-
-```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay}"}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
-p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  labs(title = "Win proportion among Spaniards",
-       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
-       x = "", y = "Win proportion") + 
-    geom_line(size=3) + 
-  scale_color_manual(guide=FALSE, values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme
-
-p2 <- ggplot(spain_wins, aes(round, avg_ranking, color=tournament, group=tournament, fill = tournament)) + 
-  geom_line(size=3) + 
-  labs(title = "Average ranking among Spaniards",
-       x = " ",
-       y = "Average rank") + 
-  scale_color_manual(" ", values=tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)",
-                                           "French Open\n(Clay)",
-                                           "US Open\n(DecoTurf)",
-                                           "Wimbledon\n(Grass)")) + my_theme +
-  guides(color = guide_legend(override.aes = list(size = 7)))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(top = textGrob("Spaniards perform better on Clay: confirmed?",
-                            gp=gpar(fontsize=20, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           # x = 0,
-                           # y = 0.5,
-                           just = "left",
-                           gp=gpar(fontfamily="serif")))
-
-ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-  ggsave("plots/spainards.jpg", ., height=6, width=graphic_width_long)
-
-```
-
-# Unforced errors and aces
-
-```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center'}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
-# ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-#                            label = tournament, color=tournament)) + 
-#   geom_jitter(size=2) +
-#   labs( title = "Distribution of errors and aces differ by tournament",
-#         x = "Total unforced errors", y = "Total aces") +
-#   my_theme + scale_color_manual("Tournament",
-#                                 values = tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)\n",
-#                                            "French Open\n(Clay)\n",
-#                                            "US Open\n(DecoTurf)\n",
-#                                            "Wimbledon\n(Grass)\n")) + 
-#   geom_density_2d()
-
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
-                           label = tournament, color=tournament)) + 
-  geom_jitter(size=2) +
-  labs( title = "Distribution of errors and aces differ by tournament",
-        x = "Total unforced errors", y = "Total aces") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  stat_ellipse(type="t", size = 2) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-ggsave("plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
-
-```
-
-## Tall players and aces
-
-```{r, fig.width=10, fig.height=4.8, fig.align='center'}
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 24),
-#         text = element_text(size = 28,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=34))
-
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 26),
-#         text = element_text(size = 30,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=36))
-
-set.seed(8)
-
-atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
-                 "Roger Federer", "Andy Murray")
-atp_servers <- sapply(atp_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
-                 "Venus Williams", "Serena Williams", "Madison Keys")
-wta_servers <- sapply(wta_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
-
-p_atp <- gs_players %>% filter(league == "ATP") %>%
-ggplot( aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament), size=1.5) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
-  geom_label_repel(data=atp_servers,
-            aes(log(ht+1), log(ace+1), label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
-  labs(x = "log(height + 1)", title = "ATP") +
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-p_wta <- gs_players %>% filter(league == "WTA") %>%
-ggplot(aes(log(ht+1), log(ace+1))) + 
-  geom_jitter(aes(color = tournament)) + 
-  # scale_color_manual(" ", values = tournament_colors) + 
-  geom_smooth(method = "lm", color="black") + 
-  geom_label_repel(data=wta_servers,
-            aes(log(ht+1), log(ace+1), label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 8) + 
-  geom_label_repel(data=wta_servers,
-            aes( log(ht+1), log(ace+1), label=name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 8) +
-  labs(x = "log(height + 1)", title = "WTA")+
-  my_theme + scale_color_manual("Tournament",
-                                values = tournament_colors,
-                                labels = c("Australian Open\n(Plexicushion)\n",
-                                           "French Open\n(Clay)\n",
-                                           "US Open\n(DecoTurf)\n",
-                                           "Wimbledon\n(Grass)\n")) +
-  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
-
-# grid.arrange(p_atp, p_wta, nrow=1, widths=c(.4, .6),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-ggarrange(p_atp, p_wta, common.legend = TRUE, legend="bottom") 
-# %>%  grid.arrange(top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-# ggsave("plots/aces.jpg", height=7, width=16)
-ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
-
-```
-
-```{r, fig.width=6, fig.height=6, fig.align='center'}
-
-# p_atp <- gs_players %>% filter(league == "ATP") %>%
-# ggplot( aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual(guide=FALSE, values = tournament_colors) + 
-#   geom_smooth( method = "lm", se=FALSE, color="black") + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=atp_servers,
-#             aes(log(ht+1), log(ace+1), label = name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "ATP") +
-#   my_theme
-# 
-# # subset(filter(gs_players, league == "WTA"),
-# #                                (log(ht+1) > 5.22 & log(ace+1) > 2.6) |
-# #                                 (log(ht+1) > 5.225 & log(ace+1) < .5) | 
-# #                                 (log(ht+1) < 5.13 & log(ace+1) > 2.1) | 
-# #                                 (log(ht+1) < 5.09 & log(ace+1) > 1.5) |
-# #                                  (log(ace + 1) > 2.82))
-# p_wta <- gs_players %>% filter(league == "WTA") %>%
-# ggplot(aes(log(ht+1), log(ace+1))) + 
-#   geom_jitter(aes(color = tournament)) + 
-#   scale_color_manual("Tournament", values = tournament_colors) + 
-#   geom_smooth(aes(color = tournament), method = "lm", se=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             alpha=.6, seed=1234, show.legend=FALSE) + 
-#   geom_label_repel(data=wta_servers,
-#             aes(log(ht+1), log(ace+1), label=name),
-#             seed=1234, fill=NA, show.legend=FALSE) +
-#   labs(x = "log(height + 1)", title = "WTA")+
-#   my_theme + theme(legend.position="bottom")
-# 
-# 
-# grid.arrange(p_atp, p_wta, nrow=2, heights = c(.46, .54),
-#              top = textGrob("Taller players are better servers: confirmed",
-#                             gp=gpar(fontsize=20, fontfamily="serif")))
-
-```
-
-### Linear regression for other outcomes
-
-```{r, eval = FALSE}
-gs_partial_players$atp = gs_partial_players$Tour == "atp"
-gs_partial_players$year_fac = as.factor(gs_partial_players$year)
-gs_partial_players$late_round = gs_partial_players$round >= "R16"
-gs_partial_players$seeded = gs_partial_players$rank <= 32
-gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
-n_aces_mod = lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_winners_mod = lmer(n_winners ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_net_mod = lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name), data = gs_partial_players)
-```
+## Linear regression for other outcomes
 
@@
 
-```{r, fig.height = 3.3, fig.width=10,  fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
+### Player effects
 
-# my_theme <-  theme_bw() + # White background, black and white theme
-#   theme(axis.text = element_text(size = 20),
-#         text = element_text(size = 24,
-#         family="serif"),
-#         plot.title = element_text(hjust = 0.5, size=26))
-
+```{r, fig.height = 6, fig.width=16}
 aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
@@
   guides(color = guide_legend(override.aes = list(size = 7)))
-
-# winners_ranef = as.data.frame(ranef(n_winners_mod, condVar = TRUE)) %>%
-#   mutate(tournament = gsub("tournament", "", term)) %>%
-#   filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-#   ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-#   geom_errorbar(height = .5, position = position_dodge()) +
-#   ylab("Coefficient\n") + 
-#   xlab("") +
-#   ggtitle("Winners") + 
-#   scale_color_manual("", values=tournament_colors,
-#                                 labels = c("Australian Open\n(Plexicushion)",
-#                                            "French Open\n(Clay)",
-#                                            "US Open\n(DecoTurf)",
-#                                            "Wimbledon\n(Grass)")) +
-#   my_theme  +
-#   coord_flip() +
-#   theme(panel.grid.major = element_blank(),
-#         panel.grid.minor = element_blank(),
-#         axis.text.y = element_blank()) +
-#   geom_hline(yintercept = 0, color = "grey")
 
@@
 
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
 ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
@@
 
+if(save_graph){
 ggarrange(aces_ranef, nets_ranef, ue_ranef, nrow=1, ncol=3,
@@
 ggsave("plots/player_effects.jpg", ., height=6, width = graphic_width_long)
-
+}
 ```
 
-## player effects
-
 ```{r nadal-model}
-
-step_model_n <- courtsports:::model_individual(player_name = "Rafael Nadal", data = gs_partial_players)
+step_model_n <- courtsports::model_individual(player_name = "Rafael Nadal",
+                                              data = gs_partial_players)$final_model
 
@@
 
-gn <- ggplot(lm(step_model_n$formula, data = step_model_n$data)) + geom_hline(yintercept = 0, col = "red") +
+gn <- ggplot(lm(step_model_n$formula, 
+                data = step_model_n$data)) + 
+  geom_hline(yintercept = 0, col = "red") +
   geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
@@
 
-df_n <- data.frame(coeff = names(step_model_n$coeff), ave = step_model_n$coeff, sd = sqrt(diag(vcov(step_model_n))))
+df_n <- data.frame(coeff = names(step_model_n$coeff),
+                   ave = step_model_n$coeff, 
+                   sd = sqrt(diag(vcov(step_model_n))))
 df_n$player <- "Nadal"
@@
 ```{r federer-model}
+step_model_f <- courtsports::model_individual(player_name = "Roger Federer", 
+                                              data = gs_partial_players,
+                                              ref = "Wimbledon")$final_model
 
-step_model_f <- courtsports:::model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")
-
-std_res_f <- rstandard(step_model_f)
-
-gf <- ggplot(lm(step_model_f$formula, data = step_model_f$data)) + geom_hline(yintercept = 0, col = "red") +
+gf <- ggplot(lm(step_model_f$formula,
+                data = step_model_f$data)) + 
+  geom_hline(yintercept = 0, col = "red") +
   geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
@@
 
-df_f <- data.frame(coeff = names(step_model_f$coeff), ave = step_model_f$coeff, sd = sqrt(diag(vcov(step_model_f))))
+df_f <- data.frame(coeff = names(step_model_f$coeff),
+                   ave = step_model_f$coeff, 
+                   sd = sqrt(diag(vcov(step_model_f))))
 df_f$player <- "Federer"
@@
 ```{r williams-model}
-
-step_model_w <- courtsports:::model_individual(player_name = "Serena Williams", data = gs_partial_players,
-                                 start_lower = TRUE, ref = "Australian Open")
+step_model_w <- courtsports:::model_individual(player_name = "Serena Williams",
+                                               data = gs_partial_players,
+                                 start_lower = TRUE, ref = "Australian Open")$final_model
 
@@
 
-gw <- ggplot(lm(step_model_w$formula, data = step_model_w$data)) + geom_hline(yintercept = 0, col = "red") +
+gw <- ggplot(lm(step_model_w$formula,
+                data = step_model_w$data)) + 
+  geom_hline(yintercept = 0, col = "red") +
   geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
@@
  ## WilliamsCOEF SIZE
-
-df_w <- data.frame(coeff = names(step_model_w$coeff), ave = step_model_w$coeff, sd = sqrt(diag(vcov(step_model_w))))
+df_w <- data.frame(coeff = names(step_model_w$coeff),
+                   ave = step_model_w$coeff, 
+                   sd = sqrt(diag(vcov(step_model_w))))
 df_w$player <- "Williams"
@@
 
-```{r plot-individual-effects}
+## Reference court effects
+
+```{r plot-individual-effects, fig.width=12, fig.height=6}
 df <- rbind(df_n, df_w, df_f)
@@
 df <- df[keep_coefs,]
+
 ggplot(df, aes(x = coeff, y = ave, col = player)) +
@@
 
+if(save_graph){
 ggsave("plots/ref-courts-eff.jpg", height=6, width = graphic_width_short)
-
-  #ggsave("~/Documents/ref-courts-eff.pdf", width = 8, height = 6, height=6, width = graphic_width_long)
+}
 ```
 
-# amanda model fitting
+## Logistic parameter effects
 
-```{r}
+```{r, fig.height=7, fig.width=16, fig.align='center'}
 win_colors <- c("#336699", "#339966")
-
-# with yellow for us
-# tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
 
@@
 
-model.sums = data.frame(model = model_names, aic = rep(NA, length(model_names)), edf = rep(NA,length(model_names)))
+model.sums = data.frame(model = model_names, 
+                        aic = rep(NA, length(model_names)), 
+                        edf = rep(NA,length(model_names)))
 model.ests = data.frame()
@@
 
-model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef", "ind_no_ioc", "ind_intercept", "ind_year_ef")
+model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef",
+                    "ind_no_ioc", "ind_intercept", "ind_year_ef")
 model.sums$fixed = c("No country, no tournament",
@@
   filter(., model == 'ind_logistic_noioc') %>%
-  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
+  ggplot(., aes(x = term,
+                y = estimate,
+                col = model,
+                ymin = estimate - 2*std.error,
+                ymax = estimate + 2*std.error)) +
   geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
   my_theme +
-  # coord_flip() +
   theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
@@
   ylab("Coefficient") +
-  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "2014", "2015", "2016", "2017"))+
+  scale_x_discrete("", breaks = waiver(), 
+                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)",
+                              "Rank (log)", "2014", "2015", "2016", "2017")) +
   guides(col = guide_legend(nrow=2)) +
@@
 ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
-ran_tourn$model = c(rep("ind_logistic", nrow(ranef(ind_logistic)$name_int)), rep("ind_logistic_noioc", nrow(ranef(ind_logistic)$name_int)))
+ran_tourn$model = c(rep("ind_logistic", 
+                        nrow(ranef(ind_logistic)$name_int)), 
+                    rep("ind_logistic_noioc",
+                        nrow(ranef(ind_logistic)$name_int)))
 
@@
 name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
-name_lookup = name_lookup[order(as.integer(name_lookup[,2])),]
+name_lookup = name_lookup[order(as.integer(name_lookup[,2])), ]
 
@@
   filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
-  ggplot(., aes(x = name, y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, color = tournament)) +
+  ggplot(., aes(x = name, 
+                y = condval, 
+                ymin = condval- 2*condsd, 
+                ymax = condval + 2*condsd, 
+                color = tournament)) +
   geom_errorbar(width = .5, position = "dodge",size = 1.5) + 
@@
 
-grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55) #,
-             # top = textGrob("Parameter estimates from logistic model",
-             #                gp=gpar(fontsize=20, fontfamily="serif"))
-             ) %>%
+grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55))
+
+if(save_graph){
+grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55)) %>%
 ggsave("plots/logistic-parameter-effects.jpg",., height=7, width=graphic_width_long)
-
+}
 ```
 
-```{r}
+### Linear fixed effects
 
+```{r, fig.height=6, fig.width=12, fig.align='center'}
 fixed.effects.plot.linear <- tidy(n_aces_mod) %>%
@@
 
+if(save_graph){
 ggsave("plots/linear-fixed-effects.jpg", fixed.effects.plot.linear,
        height=6, width = graphic_width_short)
-
+}
 ```
 
-```{r}
+## Correlation plots
 
-fixed.effects.plot.allmods = model.ests %>%
-  filter(., !grepl("ioc", term)) %>% 
-  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
-  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
-  my_theme +
-  # coord_flip() +
-  theme(axis.text.x=element_text(angle = 35, hjust = 1, size = 10)) +
-  ggtitle("Fixed effects") +
-  scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "none") + 
-  xlab("") +
-  ylab("Estimate") +
-  scale_x_discrete("", 
-                   breaks = waiver(), 
-                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017")) +
-  guides(col = guide_legend(nrow=2))
-
-aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Aces") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  coord_flip() +
-  theme(legend.position = "none")
-
-nets_ranef = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Net Wins") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  theme(axis.text.y = element_blank()) +
-  coord_flip() # +
-# theme(legend.position = "none")
-
-ue_ranef = as.data.frame(ranef(n_ue_mod, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
-  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
-  geom_errorbar(height = .5, position = position_dodge(), size = 1.5) +
-  ylab("Estimate") + 
-  xlab("") +
-  ggtitle("Unforced Errors") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  theme(axis.text.y = element_blank()) +
-  coord_flip()# + 
-#theme(legend.text=element_text(size=10))
-
-#grid.arrange(aces_ranef, nets_ranef, ue_ranef, nrow = 1, widths = c(.38, .22, .45))
-
-```
-
-# correlation plots
-
-```{r}
+```{r, fig.height=6, fig.width=12, fig.align='center'}
 win_colors <- c("#336699", "#339966")
@@
                          space = "Lab", name="Correlation") + 
-                         # labels = c("-1.0", "-0.5", "  0.0", "  0.5", "  1.0")) + 
   labs(x = "", y = "",
@@
 
+if(save_graph){
 ggsave("plots/correlation.jpg", 
        height=6, width = graphic_width_short)
-
+}
 ```
-
-# extras
-
-```{r, fig.height=3.5, fig.width=10, fig.align='center', eval=FALSE}
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_line()
-
-ggplot(gs_players, aes(round, pointswon/(rank - opponent_rank), group=tournament)) + 
-  geom_point()
-
-filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = (pointswon - (total_points - pointswon)),
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "WTA") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1300, 800))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-# grid.arrange(p1, p2, p3, p4, nrow=2)
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("WTA",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/wta.jpg", ., width = 10, height = 8)
-
-p1 <- filter(gs_players, ioc == "FRA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "French players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p2 <-filter(gs_players, ioc == "AUS", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "Australian players",
-      y = "Win proportion",
-      x = " ") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p3 <-filter(gs_players, ioc == "USA", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "American players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-p4 <-filter(gs_players, ioc == "GBR", round < "QF", league == "ATP") %>%
-ggplot( aes(x = ( opponent_rank - rank), y = did_win,
-            color = tournament)) + 
-  geom_point() +
-  geom_smooth(method = "lm", se=FALSE) +
-  labs(title = "English players",
-      y = "Win proportion") + 
-  scale_color_manual(" ", values=tournament_colors) + my_theme +
-  xlim(c(-1500, 1500))  + scale_y_continuous(breaks = c(0, 1),
-                                            labels =c("No", "Yes"),
-                                            limits = c(0, 1))
-
-ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, common.legend = TRUE, legend="bottom") %>%
-grid.arrange(., top = textGrob("ATP",
-                            gp=gpar(fontsize=30, fontfamily="serif")),
-             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
-                           just = "left",
-                           gp=gpar(fontfamily="serif"))) %>%
-ggsave("plots/atp.jpg", ., width = 10, height = 8)
-
-```

### 2019-09-05 paper_results/Data/data-description.Rmd
Updated data description
re-organize and clean files

@@
 title: "Data description"
-author: "Gallagher, Frisoli, and Luby"
+author: ""
 date: "September 5, 2019"
@@
 
-```{r setup, eval = FALSE}
+```{r invis-setup, echo = FALSE, warning = FALSE, message = FALSE}
+# devtools::install_github("shannong19/courtsports")
 library(courtsports)
-library(knitr)
-library(kableExtra)
-library(broom)
-library(tidyverse)
-```
-
-```{r invis-setup, echo = FALSE, warning = FALSE, message = FALSE}
-devtools::load_all("~/courtsports")
 library(knitr)

### 2019-09-05 paper_results/EDA/eda.Rmd
small edits
re-organize and clean files

--- 2018-09-07 contest/eda/eda.Rmd
@@
 title: "EDA"
-author: "The Competent Ones"
-date: "August 4, 2018"
+author: ""
+date: " "
 output: 
@@
 library(devtools)
+# devtools::install_github("skoval/deuce")
 library(deuce)
 library(knitr)
-# devtools::load_all("..") # Loads courtsports package
+# devtools::install_github("shannong19/courtsports")
+library(courtsports)
 ```

### 2019-09-05 paper_results/EDA/graphics_eda.Rmd
re-organize and clean files

@@
 knitr::opts_chunk$set(message = FALSE)
+
+# devtools::install_github("skoval/deuce")
 library(deuce)
@@
 library(reshape2)
-
-## To change into library(courtsports)
-devtools::load_all("../..")
+# devtools::install_github("shannong19/courtsports")
+library(courtsports)
 ```
@@
 
-tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
 league_colors <- c("#0C2340", "#902BA3")
 win_colors <- c("#336699", "#339966")
-
-# with yellow for us
+# with yellow for USA
 tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
@@
   grid.arrange(p2, p1, ncol=2) %>%
-  ggsave("plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
+  ggsave("../plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
 }
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
+  ggsave("../plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
 }
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
+  ggsave("../plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
 }
@@
 ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-  ggsave("plots/spainards.jpg", ., height=6, width=graphic_width_long)
+  ggsave("../plots/spainards.jpg", ., height=6, width=graphic_width_long)
 }
@@
 if(save_graph){
-ggsave("plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
+ggsave("../plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
 }
@@
 if(save_graph){
-ggsave("plots/aces2.jpg", height=7, width=graphic_width_long)
+ggsave("../plots/aces2.jpg", height=7, width=graphic_width_long)
 }
@@
 if(save_graph){
-ggsave("plots/aces2_percent.jpg", height=8, width=graphic_width_long)
+ggsave("../plots/aces2_percent.jpg", height=8, width=graphic_width_long)
 }

### 2019-09-05 paper_results/Hierarchical/graphics_model_complete_data.Rmd
re-organize and clean files

--- 2019-09-05 paper_results/EDA/graphics_model_complete_data.Rmd
@@
 
-tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
 league_colors <- c("#0C2340", "#902BA3")
 win_colors <- c("#336699", "#339966")
-
 # with yellow for us
@@
           common.legend = TRUE, legend="bottom", widths = c(.42, .29, .29)) %>%
-ggsave("plots/player_effects.jpg", ., height=6, width = graphic_width_long)
+ggsave("../plots/player_effects.jpg", ., height=6, width = graphic_width_long)
 }
@@
 if(save_graph){
-ggsave("plots/ref-courts-eff.jpg", height=6, width = graphic_width_short)
+ggsave("../plots/ref-courts-eff.jpg", height=6, width = graphic_width_short)
 }
@@
 ```{r, fig.height=7, fig.width=16, fig.align='center'}
-win_colors <- c("#336699", "#339966")
-
-model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
-                "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
-
-model.sums = data.frame(model = model_names, 
-                        aic = rep(NA, length(model_names)), 
-                        edf = rep(NA,length(model_names)))
-model.ests = data.frame()
-ran.effects = data.frame()
-for(mod in 1:nrow(model.sums)){
-  name = model_names[mod]
-  model.sums$aic[mod] = extractAIC(get(name))[2]
-  model.sums$edf[mod] = extractAIC(get(name))[1]
-  betas = tidy(get(name))
-  betas$model = name
-  if("group" %in% colnames(betas)){
-    model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
-                         dplyr::select(., -c("group")))
-  }
-  else model.ests = rbind(model.ests, betas)
-}
-
-model.sums$name = c("no_court", "no_random_ef", "country_ef", "ind_ef",
-                    "ind_no_ioc", "ind_intercept", "ind_year_ef")
-model.sums$fixed = c("No country, no tournament",
-                     "all",
-                     "no country, no tournament",
-                     "no tournament",
-                     "no country, no tournament",
-                     "no country",
-                     "no country, no year")
-model.sums$random = c("none",
-                      "none",
-                      "tournament (by country)",
-                      "tournament (by individual)",
-                      "tournament (by individual)",
-                      "intercept (by individual)",
-                      "year (by individual)")
-
-fixed.effects.plot = model.ests %>%
-  filter(., !grepl("ioc", term)) %>% 
-  filter(., model == 'ind_logistic_noioc') %>%
-  ggplot(., aes(x = term,
-                y = estimate,
-                col = model,
-                ymin = estimate - 2*std.error,
-                ymax = estimate + 2*std.error)) +
-  geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
-  my_theme +
-  theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
-  ggtitle("Fixed effects") +
-  scale_color_brewer("",palette = "Dark2") +
-  theme(legend.position = "none") + 
-  xlab("") +
-  ylab("Coefficient") +
-  scale_x_discrete("", breaks = waiver(), 
-                   labels = c("Intercept", "ATP", "Late round", "Op Rank (log)",
-                              "Rank (log)", "2014", "2015", "2016", "2017")) +
-  guides(col = guide_legend(nrow=2)) +
-  geom_hline(yintercept = 0, color = "grey") + 
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
-ran_tourn$model = c(rep("ind_logistic", 
-                        nrow(ranef(ind_logistic)$name_int)), 
-                    rep("ind_logistic_noioc",
-                        nrow(ranef(ind_logistic)$name_int)))
-
-gs_players$name_int = as.integer(as.factor(gs_players$name))
-name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
-name_lookup = name_lookup[order(as.integer(name_lookup[,2])), ]
-
-top_three_ranef = as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
-  mutate(tournament = gsub("tournament", "", term)) %>%
-  mutate(name = name_lookup[,1][as.integer(grp)]) %>% 
-  filter(., name == "Serena Williams" | name == "Rafael Nadal" | name == "Roger Federer") %>%
-  ggplot(., aes(x = name, 
-                y = condval, 
-                ymin = condval- 2*condsd, 
-                ymax = condval + 2*condsd, 
-                color = tournament)) +
-  geom_errorbar(width = .5, position = "dodge",size = 1.5) + 
-  xlab("") + 
-  ylab("Coefficient") + 
-  scale_color_manual("", values=tournament_colors) +
-  my_theme +
-  coord_flip() +
-  ggtitle("Random Effects") +
-  theme(legend.position = "bottom") + 
-  guides(col = guide_legend(nrow=2,
-                            override.aes = list(size = 7))) +
-  geom_hline(yintercept = 0, color = "grey") + 
-  theme(panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
-
-grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55))
-
-if(save_graph){
-grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55)) %>%
-ggsave("plots/logistic-parameter-effects.jpg",., height=7, width=graphic_width_long)
-}
+# win_colors <- c("#336699", "#339966")
+# 
+# model_names = c("nocourt_logistic", "base_logistic", "country_logistic", "ind_logistic", 
+#                 "ind_logistic_noioc", "ind_int_logistic", "ind_year_logistic")
+# 
+# model.sums = data.frame(model = model_names, 
+#                         aic = rep(NA, length(model_names)), 
+#                         edf = rep(NA,length(model_names)))
+# model.ests = data.frame()
+# ran.effects = data.frame()
+# for(mod in 1:nrow(model.sums)){
+#   our_name = model_names[mod]
+#   model.sums$aic[mod] = extractAIC(get(our_name))[2]
+#   model.sums$edf[mod] = extractAIC(get(our_name))[1]
+#   betas = tidy(get(our_name))
+#   betas$model = our_name
+#   if("group" %in% colnames(betas)){
+#     model.ests = rbind(model.ests, filter(betas, group == "fixed") %>%
+#                          dplyr::select(., -c("group")))
+#   } else model.ests = rbind(model.ests, betas)
+# }
+# 
+# model.sums$our_name = c("no_court", "no_random_ef", "country_ef", "ind_ef",
+#                     "ind_no_ioc", "ind_intercept", "ind_year_ef")
+# model.sums$fixed = c("No country, no tournament",
+#                      "all",
+#                      "no country, no tournament",
+#                      "no tournament",
+#                      "no country, no tournament",
+#                      "no country",
+#                      "no country, no year")
+# model.sums$random = c("none",
+#                       "none",
+#                       "tournament (by country)",
+#                       "tournament (by individual)",
+#                       "tournament (by individual)",
+#                       "intercept (by individual)",
+#                       "year (by individual)")
+# 
+# fixed.effects.plot = model.ests %>%
+#   filter(., !grepl("ioc", term)) %>% 
+#   filter(., model == 'ind_logistic_noioc') %>%
+#   ggplot(., aes(x = term,
+#                 y = estimate,
+#                 col = model,
+#                 ymin = estimate - 2*std.error,
+#                 ymax = estimate + 2*std.error)) +
+#   geom_errorbar(width = .7, position = position_dodge(), size = 1.5) +
+#   my_theme +
+#   theme(axis.text.x=element_text(angle = 35, hjust = 1)) +
+#   ggtitle("Fixed effects") +
+#   scale_color_brewer("",palette = "Dark2") +
+#   theme(legend.position = "none") + 
+#   xlab("") +
+#   ylab("Coefficient") +
+#   scale_x_discrete("", breaks = waiver(), 
+#                    labels = c("Intercept", "ATP", "Late round", "Op Rank (log)",
+#                               "Rank (log)", "2014", "2015", "2016", "2017")) +
+#   guides(col = guide_legend(nrow=2)) +
+#   geom_hline(yintercept = 0, color = "grey") + 
+#   theme(panel.grid.major = element_blank(),
+#         panel.grid.minor = element_blank())
+# 
+# ran_tourn = rbind(ranef(ind_logistic)$name_int, ranef(ind_logistic_noioc)$name_int)
+# ran_tourn$model = c(rep("ind_logistic", 
+#                         nrow(ranef(ind_logistic)$name_int)), 
+#                     rep("ind_logistic_noioc",
+#                         nrow(ranef(ind_logistic)$name_int)))
+# 
+# gs_players$name_int = as.integer(as.factor(gs_players$our_name))
+# name_lookup = unique(cbind(gs_players$name, gs_players$name_int))
+# name_lookup = name_lookup[order(as.integer(name_lookup[,2])), ]
+# 
+# top_three_ranef = as.data.frame(ranef(ind_logistic_noioc, condVar = TRUE)) %>%
+#   mutate(tournament = gsub("tournament", "", term)) %>%
+#   mutate(our_name = name_lookup[,1][as.integer(grp)]) %>% 
+#   filter(., our_name == "Serena Williams" | our_name == "Rafael Nadal" | our_name == "Roger Federer") %>%
+#   ggplot(., aes(x = our_name, 
+#                 y = condval, 
+#                 ymin = condval- 2*condsd, 
+#                 ymax = condval + 2*condsd, 
+#                 color = tournament)) +
+#   geom_errorbar(width = .5, position = "dodge",size = 1.5) + 
+#   xlab("") + 
+#   ylab("Coefficient") + 
+#   scale_color_manual("", values=tournament_colors) +
+#   my_theme +
+#   coord_flip() +
+#   ggtitle("Random Effects") +
+#   theme(legend.position = "bottom") + 
+#   guides(col = guide_legend(nrow=2,
+#                             override.aes = list(size = 7))) +
+#   geom_hline(yintercept = 0, color = "grey") + 
+#   theme(panel.grid.major = element_blank(),
+#         panel.grid.minor = element_blank())
+# 
+# grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55))
+# 
+# if(save_graph){
+# grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.45, .55)) %>%
+# ggsave("../plots/logistic-parameter-effects.jpg",., height=7, width=graphic_width_long)
+# }
 ```
@@
 if(save_graph){
-ggsave("plots/linear-fixed-effects.jpg", fixed.effects.plot.linear,
+ggsave("../plots/linear-fixed-effects.jpg", fixed.effects.plot.linear,
        height=6, width = graphic_width_short)
@@
 if(save_graph){
-ggsave("plots/correlation.jpg", 
+ggsave("../plots/correlation.jpg", 
        height=6, width = graphic_width_short)

### 2019-09-05 paper_results/Data/makeGSF.R

@@ -1,3 +1,4 @@
+# devtools::install_github("skoval/deuce")
 library(deuce)
 library(dplyr)
 

### 2019-09-05 paper_results/Hierarchical/fit-and-plot.R

@@ -2,6 +2,7 @@ library(courtsports)
 library(tidyverse)
 library(broom)
 library(lme4)
+library(dplyr)
 
 # Need development version of tidyr (pivot_wider and pivot_longer)
 # devtools::install_github("tidyverse/tidyr")
@@ -9,14 +10,23 @@ library(lme4)
 ## Define theme and colors
 
 my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 12),
-        text = element_text(size = 14,
+  theme(axis.text = element_text(size = 24),
+        text = element_text(size = 28,
                             family="serif"),
-        plot.title = element_text(hjust = 0.5, size=16))
+        plot.title = element_text(hjust = 0.5, size=34),
+        plot.subtitle = element_text(hjust = 0.5))
 
+league_colors <- c("#0C2340", "#902BA3")
 win_colors <- c("#336699", "#339966")
 
+# with yellow for us
 tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+graphic_width_long <- 16
+graphic_width_short <- 12
+
+# write/save graphs locally?
+save_graph <- FALSE
 
 ## Define variables for modeling
 gs_players$name_int = as.integer(as.factor(gs_players$name))
@@ -40,19 +50,30 @@ gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
 
 ## Fit models used in paper
 set.seed(091418)
-ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac), data = gs_players, family = "binomial", nAGQ =0)
-n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp +
-                          (0 + tournament |name_fac), data = gs_partial_players)
-n_net_mod = lme4::lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp +
-                         (0 + tournament |name_fac), data = gs_partial_players, control = lmerControl(optimizer = "bobyqa"))
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + year_fac + atp + (0 + tournament |name_fac),
+ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) +
[…]

### 2019-09-05 paper_results/Individual/most_matches.R

@@ -1,8 +1,8 @@
 ## Script to get Federer individual model
 ## SKG July 20, 2019
 
-
-
+# devtools::install_github("skoval/deuce")
+# devtools::install_github("shannong19/courtsports")
 
 library("courtsports")
 library(ggplot2)
@@ -14,18 +14,20 @@ library(gridExtra)
 library(knitr)
 library(broom)
 library(kableExtra)
+# devtools::install_github("benjaminleroy/ggDiagnose")
 library(ggDiagnose)
 
 
 data(gs_partial_players)
 
-########################3
+########################################################################
 ## THeme
 my_theme <-  theme_bw() + # White background, black and white theme
-  theme(axis.text = element_text(size = 12),
-        text = element_text(size = 14,
-        family="serif"),
-        plot.title = element_text( size=16))
+  theme(axis.text = element_text(size = 24),
+        text = element_text(size = 28,
+                            family="serif"),
+        plot.title = element_text(hjust = 0.5, size=34),
+        plot.subtitle = element_text(hjust = 0.5))
 
 tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
 league_colors <- c("#0C2340", "#902BA3")
@@ -33,7 +35,13 @@ win_colors <- c("#336699", "#339966")
 
 # with yellow for us
 tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
-########################################
+
+graphic_width_long <- 16
+graphic_width_short <- 12
+
+# write/save graphs locally?
+save_graph <- FALSE
+###########################################################################
 
 player_name <- "Roger Federer"
 data <- gs_partial_players
@@ -144,7 +152,7 @@ print(top_players, n = 45)
 
 
 ## Build a list of data frames
-devtools::load_all("~/courtsports")
+# devtools::load_all("~/courtsports")
 L <- nrow(top_players)
 player_list <- vector(mode = "list", length = L)
 mods <- vector(mode = "list", length = L)
@@ -262,8 +270,9 @@ ggplot(data = na.omit(gdf[, -which(colnames(gdf) == "pos_bonf")]),
          x = "Court and interaction") +
     my_theme
 
-ggsave("individual-models-coef-signs-sub.pdf", width = 8, height = 10)
-
[…]

### 2019-09-05 paper_results/Individual/most_matches.R
change select to dplyr::select due to errors

@@ -327,7 +327,7 @@ ggsave("../plots/individual-models-cor-mat-atp-sub.pdf", width = 8, height = 8)
 
 ## Women's
 df_wide_w <- df_wide %>% filter(tour == "wta")
-cor <- df_wide_w %>% select(-c("tour", "name", "(Intercept)")) %>% cor()
+cor <- df_wide_w %>% dplyr::select(-c("tour", "name", "(Intercept)")) %>% cor()
 ggcorrplot(cor, hc.order = FALSE, type = "lower",
            outline.col = "black") + my_theme +
     theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
@@ -349,7 +349,7 @@ for(jj in 1:length(tours)){
     for(ii in 1:length(slam_abb)){
         slam_vars <- grep(slam_abb[ii], colnames(df_wide), value = TRUE)
         tour <- tours[jj]
-        cor <- df_wide_w %>% select(slam_vars) %>% cor()
+        cor <- df_wide_w %>% dplyr::select(slam_vars) %>% cor()
         g_list[[length(slam_abb) * (jj -1) + ii]] <- ggcorrplot(cor, hc.order = FALSE, type = "lower",
                                              outline.col = "black") + my_theme +
             theme(axis.text.x = element_text(angle = 90, vjust = 0)) +

### 2019-09-05 paper_results/Individual/most_matches.R
ggplot settings

@@ -29,6 +29,12 @@ my_theme <-  theme_bw() + # White background, black and white theme
         plot.title = element_text(hjust = 0.5, size=34),
         plot.subtitle = element_text(hjust = 0.5))
 
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 12),
+        text = element_text(size = 14,
+                            family="serif"),
+        plot.title = element_text( size=16))
+
 tournament_colors <- c("#0297DB", "#b06835", "#0C2340", "#54008b")
 league_colors <- c("#0C2340", "#902BA3")
 win_colors <- c("#336699", "#339966")
@@ -79,7 +85,7 @@ car::vif(out_list$final_model)
 g <- ggDiagnose(out_list$final_model, which = 1:6, return = TRUE)
 ggs <- g$ggout
 ggthm <- lapply(ggs, function(gg) gg + my_theme)
-pdf("fed_diags.pdf", width = 16, height = 10)
+pdf("../plots/fed_diags.pdf", width = 16, height = 10)
 do.call("grid.arrange", c(ggthm, ncol = 3))
 dev.off()
 
@@ -361,7 +367,7 @@ for(jj in 1:length(tours)){
     }
 
 if(save_graph){
-pdf("cor-slams-sub.pdf", width = 16, height = 10)
+pdf("../plots/cor-slams-sub.pdf", width = 16, height = 10)
 do.call("grid.arrange", c(g_list, ncol = 3))
 dev.off()
 }

### 2019-09-05 paper_results/Hierarchical/fit-and-plot.R
delete old stuff

@@ -51,19 +51,19 @@ gs_partial_players$opponent_seeded = gs_partial_players$opponent_rank <= 32
 ## Fit models used in paper
 set.seed(091418)
 ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) +
-                                   year_fac + atp + (0 + tournament |name_fac), 
+                                   year_fac + atp + (0 + tournament |name_fac),
                                  data = gs_players, family = "binomial", nAGQ =0)
 
-n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) + 
+n_aces_mod = lme4::lmer(n_aces ~ late_round + log(rank) + log(opponent_rank) +
                           year_fac + atp +
-                          (0 + tournament |name_fac), 
+                          (0 + tournament |name_fac),
                         data = gs_partial_players)
 
 n_net_mod = lme4::lmer(n_netpt_w ~ late_round + log(rank) + log(opponent_rank) +
-                         year_fac + atp + (0 + tournament |name_fac), 
+                         year_fac + atp + (0 + tournament |name_fac),
                        data = gs_partial_players, control = lmerControl(optimizer = "bobyqa"))
 
-n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) + 
+n_ue_mod = lmer(n_ue ~ late_round + log(rank) + log(opponent_rank) +
                   year_fac + atp + (0 + tournament |name_fac),
                 data = gs_partial_players, control = lmerControl(optimizer = "bobyqa"))
 
@@ -195,9 +195,9 @@ ranef_net_many = as.data.frame(ranef(n_net_mod, condVar = TRUE)) %>%
   mutate(tournament = gsub("tournament", "", term)) %>%
   filter(grp %in% more_players) %>%
   ggplot(., aes(x = grp,
-                y = condval, 
-                ymin = condval- 2*condsd, 
-                ymax = condval + 2*condsd, 
+                y = condval,
+                ymin = condval- 2*condsd,
+                ymax = condval + 2*condsd,
                 color = tournament)) +
   geom_errorbar(width = .5, position = "dodge",size = 1.5) +
   xlab("") +
@@ -327,9 +327,3 @@ tidy(n_ue_mod) %>%
   dplyr::select(-group) %>%
   knitr::kable(format = "latex", digits = 2)
 
-iln_fast = ind_logistic_noioc
[…]

### 2020-08-01 paper_results/Hierarchical/fit-and-plot-revision2.R
update models, graphs PRE NEW DATA

@@ -1,0 +1,394 @@
+library(courtsports)
+library(tidyverse)
+library(broom)
+library(lme4)
+library(dplyr)
+
+my_theme <-  theme_bw() + # White background, black and white theme
+  theme(axis.text = element_text(size = 24),
+        text = element_text(size = 28,
+                            family="serif"),
+        plot.title = element_text(hjust = 0.5, size=34),
+        plot.subtitle = element_text(hjust = 0.5))
+
+league_colors <- c("#0C2340", "#902BA3")
+win_colors <- c("#336699", "#339966")
+
+# with yellow for us
+tournament_colors <- c("#0297DB", "#b06835", "#ffe500", "#54008b")
+
+graphic_width_long <- 16
+graphic_width_short <- 12
+
+# write/save graphs locally?
+save_graph <- FALSE
+
+## Define variables for modeling
+gs_players$name_int = as.integer(as.factor(gs_players$name))
+gs_players$name_fac = as.factor(gs_players$name)
+gs_players$hand_fac = as.factor(gs_players$hand)
+gs_players$ioc_fac = as.factor(gs_players$ioc)
+gs_players$atp = gs_players$league == "ATP"
+gs_players$year_fac = as.factor(gs_players$year)
+gs_players$late_round = gs_players$round >= "R16"
+gs_players$seeded = gs_players$rank <= 32
+gs_players$opponent_seeded = gs_players$opponent_rank <= 32
+
+gs_partial_players %>%
+  mutate(
+    name_int = as.integer(as.factor(name)),
+    name_fac = as.factor(name),
+    ioc_fac = as.factor(ioc),
+    atp = (Tour == "atp"),
+    year_fac = as.factor(year),
+    late_round = round >= "R16",
+    seeded = rank <= 32,
+    opponent_seeded = opponent_rank <= 32,
+    ace_pct = n_aces/total_points
+  ) -> gs_partial_players
+
+## Fit models used in paper
+set.seed(091418)
+ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent_rank) +
+                                   year_fac + atp + (0 + tournament |name_fac),
+                                 data = gs_players, family = "binomial", nAGQ =0)
+
+logistic_newmod = lme4::glmer(did_win ~ 0 + late_round + log(rank) + log(opponent_rank) +
[…]

### 2020-08-02 paper_results/EDA/graphics_eda.Rmd
kayla update figure 5 to try new things with height vs aces

@@
 ```
+
+#### Using ace percentage.... one dot per tournament (colored)
+
+```{r, fig.width=16, fig.height=8, fig.align='center'}
+set.seed(8)
+
+maxish <- function(x){
+  xx <- na.omit(unique(x))
+  return(ifelse(length(xx) > 1, max(xx), xx))
+}
+
+atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal",
+                 "Roger Federer", "Andy Murray")
+
+wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki",
+                 "Venus Williams", "Serena Williams", "Madison Keys")
+
+atp_ace_ht_tour <-   gs_players %>% 
+  filter(league == "ATP") %>%
+  dplyr::group_by(name, tournament) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ace_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+wta_ace_ht_tour <-   gs_players %>% 
+  filter(league == "WTA") %>%
+  dplyr::group_by(name, tournament) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ace_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+# atp_servers <- sapply(ungroup(atp_ace_ht_tour),
+#                       function(x) return(sample(which(ungroup(atp_ace_ht_tour)$name == x), 2))) %>%
+#   as.numeric() %>% ungroup(atp_ace_ht_tour)[., ]
+# 
+# wta_servers <- sapply(wta_ace_ht_tour,
+#                       function(x) return(sample(which(wta_ace_ht_tour$name == x), 2))) %>%
+#   as.numeric() %>% gs_players[., ]
+
+p_atp_percent <- 
+atp_ace_ht_tour %>% # could also try median
+ggplot(aes(ht, med_ace_perc)) + 
+  geom_jitter(aes(color = tournament), size=1.5) + 
+  # scale_color_manual(" ", values = tournament_colors) +
+  geom_smooth(aes(color = tournament), method = "lm") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht_tour, name %in% atp_servers),
+            aes(ht, med_ace_perc, label=name, color = tournament),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht_tour, name %in% atp_servers),
+            aes(ht, med_ace_perc, label = name, color = tournament),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
+  my_theme +
+  scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+# +
+#   guides(color = guide_legend(override.aes = list(size = 7, shape=15))) + 
+facet_wrap(~tournament) +
+theme(legend.position="none")
+
+p_atp_percent
+
+if(save_graph){
+ggsave("../plots/ace_ht_facet_atp.jpg", height=8, width=graphic_width_short)
+}
+
+p_wta_percent <- 
+wta_ace_ht_tour %>% # could also try median
+ggplot(aes(ht, med_ace_perc)) + 
+  geom_jitter(aes(color = tournament), size=1.5) + 
+  # scale_color_manual(" ", values = tournament_colors) +
+  geom_smooth(aes(color = tournament), method = "lm") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht_tour, name %in% wta_servers),
+            aes(ht, med_ace_perc, label=name, color = tournament),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = rep(c(-.1, .06, 0.06, 0.08, -0.1), 4),
+            nudge_x = rep(c(-5, 0, 0, -7, 0), 4)) +
+  geom_label_repel(data=dplyr::filter(wta_ace_ht_tour, name %in% wta_servers),
+            aes(ht, med_ace_perc, label = name, color = tournament),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = rep(c(-.1, .06, 0.06, 0.08, -0.1), 4),
+            nudge_x = rep(c(-5, 0, 0, -7, 0), 4)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "WTA") +
+  my_theme +
+  scale_color_manual("Tournament",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+# +
+#   guides(color = guide_legend(override.aes = list(size = 7, shape=15))) + 
+facet_wrap(~tournament) +
+theme(legend.position="none") + 
+  ylim(c(-0.02, 0.2))
+
+p_wta_percent
+
+if(save_graph){
+ggsave("../plots/ace_ht_facet_wta.jpg", height=8, width=graphic_width_short)
+}
+
+```
+
+<!-- #### Using ace percentage.... one dot per tournament (black white) -->
+
+<!-- ```{r, fig.width=16, fig.height=8, fig.align='center'} -->
+<!-- set.seed(8) -->
+
+<!-- maxish <- function(x){ -->
+<!--   xx <- na.omit(unique(x)) -->
+<!--   return(ifelse(length(xx) > 1, max(xx), xx)) -->
+<!-- } -->
+
+<!-- atp_servers <- c("Ivo Karlovic", "John Isner", "Rafael Nadal", -->
+<!--                  "Roger Federer", "Andy Murray") -->
+
+<!-- wta_servers <- c("Karolina Pliskova", "Caroline Wozniacki", -->
+<!--                  "Venus Williams", "Serena Williams", "Madison Keys") -->
+
+<!-- atp_ace_ht_tour <-   gs_players %>%  -->
+<!--   filter(league == "ATP") %>% -->
+<!--   dplyr::group_by(name, tournament) %>% -->
+<!--   dplyr::summarise(ht = maxish(ht), -->
+<!--                    avg_ace_perc = mean(ace/svpt, na.rm=TRUE), -->
+<!--                    med_ace_perc = median(ace/svpt, na.rm=TRUE), -->
+<!--                    count_n = n()) -->
+
+<!-- wta_ace_ht_tour <-   gs_players %>%  -->
+<!--   filter(league == "WTA") %>% -->
+<!--   dplyr::group_by(name, tournament) %>% -->
+<!--   dplyr::summarise(ht = maxish(ht), -->
+<!--                    avg_ace_perc = mean(ace/svpt, na.rm=TRUE), -->
+<!--                    med_ace_perc = median(ace/svpt, na.rm=TRUE), -->
+<!--                    count_n = n()) -->
+
+<!-- atp_servers <- sapply(atp_servers, -->
+<!--                       function(x) return(sample(which(gs_players$name == x), 2))) %>% -->
+<!--   as.numeric() %>%  -->
+<!--   gs_players[., ]  %>% -->
+<!--   dplyr::mutate(med_ace_perc = median(ace/svpt, na.rm=TRUE)) -->
+
+<!-- wta_servers <- sapply(wta_servers, -->
+<!--                       function(x) return(sample(which(gs_players$name == x), 2))) %>% -->
+<!--   as.numeric() %>% gs_players[., ] %>% -->
+<!--   dplyr::mutate(med_ace_perc = median(ace/svpt, na.rm=TRUE)) -->
+
+<!-- p_atp_percent <-  -->
+<!-- atp_ace_ht_tour %>% # could also try median -->
+<!-- ggplot(aes(ht, med_ace_perc)) +  -->
+<!--   geom_jitter( size=1.5) +  -->
+<!--   # scale_color_manual(" ", values = tournament_colors) + -->
+<!--   geom_smooth( method = "lm", color = "black") +  -->
+<!--   geom_label_repel(data=atp_servers, -->
+<!--             aes(ht, med_ace_perc, label=name), -->
+<!--             alpha=.8, seed=1234, show.legend=FALSE, size = 7, -->
+<!--             nudge_y = c(-.075, 0, 0, -0.05,0.02 ), -->
+<!--             nudge_x = c(5, 5, -5, -3, 0)) +  -->
+<!--   geom_label_repel(data=atp_servers, -->
+<!--             aes(ht, med_ace_perc, label = name), -->
+<!--             seed=1234, fill=NA, show.legend=FALSE, size = 7, -->
+<!--             nudge_y = c(-.075, 0, 0, -0.05,0.02 ), -->
+<!--             nudge_x = c(5, 5, -5, -3, 0)) + -->
+<!--   labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") + -->
+<!--    my_theme # + -->
+<!-- #   scale_color_manual("Tournament", -->
+<!-- #                                 values = tournament_colors, -->
+<!-- #                                 labels = c("Australian Open\n(Plexicushion)\n", -->
+<!-- #                                            "French Open\n(Clay)\n", -->
+<!-- #                                            "US Open\n(DecoTurf)\n", -->
+<!-- #                                            "Wimbledon\n(Grass)\n")) + -->
+<!-- # # + -->
+<!-- # #   guides(color = guide_legend(override.aes = list(size = 7, shape=15))) +  -->
+<!-- # facet_wrap(~tournament) + -->
+<!-- # theme(legend.position="none") -->
+
+<!-- p_atp_percent -->
+
+<!-- p_wta_percent <-  -->
+<!-- wta_ace_ht_tour %>% # could also try median -->
+<!-- ggplot(aes(ht, med_ace_perc)) +  -->
+<!--   geom_jitter(size=1.5) +  -->
+<!--   # scale_color_manual(" ", values = tournament_colors) + -->
+<!--   geom_smooth(method = "lm", color = "black") +  -->
+<!--   geom_label_repel(data=wta_servers, -->
+<!--             aes(ht, med_ace_perc, label=name), -->
+<!--             seed=1234, fill=NA, show.legend=FALSE, size = 5) + -->
+<!--             # nudge_y = rep(c(-.1, .06, 0.06, 0.08, -0.1), 4), -->
+<!--             # nudge_x = rep(c(-5, 0, 0, -7, 0), 4)) + -->
+<!--   geom_label_repel(data=wta_servers, -->
+<!--             aes(ht, med_ace_perc, label = name), -->
+<!--             seed=1234, fill=NA, show.legend=FALSE, size = 5) + -->
+<!--             # nudge_y = c(-.1, .06, 0.06, 0.08, -0.1), -->
+<!--             # nudge_x = c(-5, 0, 0, -7, 0)) + -->
+<!--   labs(x = "Height in cm", y = "Median ace% per match", title = "WTA") + -->
+<!--   my_theme # + -->
+<!-- #   scale_color_manual("Tournament", -->
+<!-- #                                 values = tournament_colors, -->
+<!-- #                                 labels = c("Australian Open\n(Plexicushion)\n", -->
+<!-- #                                            "French Open\n(Clay)\n", -->
+<!-- #                                            "US Open\n(DecoTurf)\n", -->
+<!-- #                                            "Wimbledon\n(Grass)\n")) + -->
+<!-- # # + -->
+<!-- # #   guides(color = guide_legend(override.aes = list(size = 7, shape=15))) +  -->
+<!-- # facet_wrap(~tournament) + -->
+<!-- # theme(legend.position="none") +  -->
+<!-- #   ylim(c(-0.02, 0.2)) -->
+
+<!-- p_wta_percent -->
+
+<!-- ggarrange(p_atp_percent, NULL, p_wta_percent, common.legend = TRUE, legend="bottom", widths = c(1, 0.08, 1), nrow = 1)  -->
+
+<!-- if(save_graph){ -->
+<!-- ggsave("../plots/aces_median_percent_BW2.jpg", height=8, width=graphic_width_long) -->
+<!-- } -->
+
+<!-- ``` -->
+
+#### Using ace percentage.... one dot per player - black white
+
+```{r, fig.width=16, fig.height=8, fig.align='center'}
+set.seed(8)
+
+atp_ace_ht <-   gs_players %>% 
+  filter(league == "ATP") %>%
+  dplyr::group_by(name) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+wta_ace_ht <-   gs_players %>% 
+  filter(league == "WTA") %>%
+  dplyr::group_by(name) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+p_atp_percent <- atp_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
+  my_theme 
+# + 
+#   ylim(c(0, 0.25))
+p_atp_percent
+
+p_wta_percent <- wta_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "WTA") +
+  my_theme
+# + 
+#   ylim(c(0, 0.25))
+
+p_wta_percent
+
+ggarrange(p_atp_percent, NULL, p_wta_percent, common.legend = TRUE, legend="bottom", widths = c(1, 0.08, 1), nrow = 1)
+
+if(save_graph){
+ggsave("../plots/aces_median_percent_BW2.jpg", height=8, width=graphic_width_long)
+}
+
+p_atp_percent <- atp_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
+  my_theme +
+  ylim(c(0, 0.25))
+
+p_wta_percent <- wta_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) +
+  labs(x = "Height in cm", y = " ", title = "WTA") +
+  my_theme +
+  ylim(c(0, 0.25))
+
+ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom",
+          nrow=1)
+
+if(save_graph){
+ggsave("../plots/aces_median_percent_BW1.jpg", height=8, width=graphic_width_long)
+}
+
+```

### 2020-08-02 R/clean-data.R
made new data sets for data up through 2019.

@@ -157,9 +157,16 @@ summarize_pbp <- function(pbp,
 
     ## Omit non-complete columns
     na_col_inds <- colSums(is.na(pbp))
-    complete_cols <- which(na_col_inds == 0)
+    complete_cols <- which(na_col_inds <= 1000)
     pbp <- pbp %>% dplyr::select(names(pbp[complete_cols])) %>%
-        filter(year >= start_year& year <= end_year)
+        dplyr::filter(year >= start_year& year <= end_year)
+    ## Of these, which rows have NAs?
+    na_row_inds <- which(rowSums(is.na(pbp)) > 0)
+    ## Find matches associated with NAs
+    missing_match_ids <- unique(pbp$match_id[na_row_inds])
+    ## Exclude entire matches which have any missing values of above relevant columns
+    pbp <- pbp %>% dplyr::filter(!(match_id %in% missing_match_ids))
+    
 
     ## Change tournament to match the names of gs data
     pbp$tournament <- factor(pbp$slam) %>%
@@ -171,7 +178,8 @@ summarize_pbp <- function(pbp,
 
 
     ## Group by unique matches
-    pbp_matches <- pbp %>% group_by(match_id, player1, player2, Tour,
+
+    pbp_matches <- pbp %>% group_by(match_id, player1, player2, 
                                     slam, year, tournament) %>%
         dplyr::summarize(ave_serve_speed_p1 = mean(Speed_MPH[PointServer == 1]),
                          ave_serve_speed_p2 = mean(Speed_MPH[PointServer == 2]),

### 2020-08-02 data-raw/make-gs-data_updated.R
made new data sets for data up through 2019.

@@ -1,0 +1,72 @@
+## Script to make data used in our analysis, specifically
+##
+## 1. gs.rda
+## 2. gs_players.rda
+## 3. gs_pbp.rda
+## 4. gs_partial.rda
+
+devtools::load_all("../") ## Load court sports functions
+library(deuce)
+library(dplyr)
+
+
+## Load WTA and ATP matches from deuce package
+data(atp_matches)
+data(wta_matches)
+
+## Clean data
+gs_v2 <- clean_matches_data(atp_matches = atp_matches,
+                         wta_matches = wta_matches,
+                         start_year = 2013,
+                         end_year = 2020,
+                         include_quals = FALSE,
+                         level = "Grand Slams")
+
+
+
+## Extract sets, points, and games
+gs_v2 <- extract_sgp(gs_v2)
+
+usethis::use_data(gs_v2, overwrite = TRUE)
+
+## Make into player data
+gs_players_v2 <- matches_to_player_data(gs_v2,
+                                     w_opp_vars = c("loser_rank"),
+                                     l_opp_vars = c("winner_rank"),
+                                     opp_var_names = c("opponent_rank")
+)
+
+usethis::use_data(gs_players_v2, overwrite = TRUE)
+
+###############################################
+
+### pbp data
+data(gs_point_by_point)
+
+
+
+gs_pbp_v2 <- as.data.frame(summarize_pbp(gs_point_by_point,
+                        start_year = 2013,
+                        end_year = 2020))
+
+usethis::use_data(gs_pbp_v2, overwrite = TRUE)
+
+data(gs_v2)
+
+## Join gs_pbp and gs together
+
+gs_join <- join_gs_pbp(gs_pbp_v2, gs_v2)
+
+gs_partial_v2 <- as.data.frame(combine_fields(gs_join))
+
+
+head(gs_partial[, c("winner_name", "loser_name", "player1", "player2", "w_n_sv")])
+
+usethis::use_data(gs_partial_v2, overwrite = TRUE)
+
+## gs_partial_players
+devtools::load_all("../") ## Load court sports functions
+data(gs_partial_v2)
+gs_partial_players_v2 <- matches_to_player_data(gs_partial_v2)
+
+usethis::use_data(gs_partial_players_v2, overwrite = TRUE)

### 2020-08-02 paper_results/EDA/graphics_eda.Rmd
kayla fixes graphs, updates with v2 data

@@
 ```{r data-cleaning, include = FALSE}
-data(gs)
-data(gs_players)
-data(gs_partial)
-data(gs_partial_players)
+# data(gs)
+# data(gs_players)
+# data(gs_partial)
+# data(gs_partial_players)
+
+data(gs_v2)
+data(gs_players_v2)
+data(gs_partial_v2)
+data(gs_partial_players_v2)
 ```
@@
 ```{r, fig.height=8, fig.width=16}
-p2 <- ggplot(gs_players) + 
+p2 <- ggplot(gs_players_v2) + 
   geom_density(aes(pointswon, ..scaled..,
@@
 
-p1 <- ggplot(gs_players) + 
+p1 <- ggplot(gs_players_v2) + 
   geom_density(aes(pointswon, ..scaled..,
@@
   labs(x = "Points", y = " ") +
-  scale_fill_manual(" ", values=win_colors, labels = c("Lost game", "Won game")) +
+  scale_fill_manual(" ", values=win_colors, labels = c("Lost match", "Won match")) +
   scale_color_manual(guide = FALSE, values = win_colors) +
@@
         legend.spacing.y = unit(1.0, 'cm'),
-        legend.margin=margin(t = 2.2, unit='cm'))  +
+        legend.margin=ggplot2::margin(t = 2.2, unit='cm'))  +
   guides(fill = guide_legend(override.aes = list(size = 0, alpha = 1),
@@
   grid.arrange(p2, p1, ncol=2) %>%
-  ggsave("../plots/points_per_match.jpg", ., height=8, width=graphic_width_long)
+  ggsave("../plots/points_per_match_v2.jpg", ., height=8, width=graphic_width_long)
 }
@@
 ```{r}
-france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF") %>% 
+france_wins <- dplyr::filter(gs_players_v2, ioc == "FRA", round < "QF") %>% 
   dplyr::group_by(tournament, round) %>% 
@@
 
-us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF") %>% 
+us_wins <- dplyr::filter(gs_players_v2, ioc == "USA", round < "QF") %>% 
   dplyr::group_by(tournament, round) %>% 
@@
 
-aus_wins <- filter(gs_players, ioc == "AUS", round < "QF") %>% 
+aus_wins <- filter(gs_players_v2, ioc == "AUS", round < "QF") %>% 
   group_by(tournament, round) %>% 
@@
 
-gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
+gbr_wins <- filter(gs_players_v2, ioc == "GBR", round < "QF") %>% 
   group_by(tournament, round) %>% 
@@
 ```{r, echo=TRUE}
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "USA", round < "QF")  %>% 
+dplyr::filter(gs_players_v2, ioc == "USA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players_v2, ioc == "USA", round < "QF")  %>% 
   pull(name) %>% unique() %>% length()
 
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "FRA", round < "QF")  %>%
+dplyr::filter(gs_players_v2, ioc == "FRA", round < "QF")  %>% nrow()
+dplyr::filter(gs_players_v2, ioc == "FRA", round < "QF")  %>%
   pull(name) %>% unique() %>% length()
 
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "AUS", round < "QF")  %>% 
+dplyr::filter(gs_players_v2, ioc == "AUS", round < "QF")  %>% nrow()
+dplyr::filter(gs_players_v2, ioc == "AUS", round < "QF")  %>% 
   pull(name) %>% unique() %>% length()
 
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% nrow()
-dplyr::filter(gs_players, ioc == "GBR", round < "QF")  %>% 
+dplyr::filter(gs_players_v2, ioc == "GBR", round < "QF")  %>% nrow()
+dplyr::filter(gs_players_v2, ioc == "GBR", round < "QF")  %>% 
   pull(name) %>% unique() %>% length()
@@
   labs(title = "French players",
-       subtitle = "53 players, n=828",
+       # subtitle = "53 players, n=828",
+       subtitle = "66 players, n=1119",
        x = "", y = " ") + 
@@
   labs(title = "USA players",
-       subtitle = "81 players, n=1046",
+       subtitle = "94 players, n=1512",
+      # subtitle = "81 players, n=1046",
               x = "", y = " ") + 
@@
   labs(title = "Aus. players",
-       subtitle = "40 players, n=376",
+       subtitle = "47 players, n=576",
+              # subtitle = "40 players, n=376",
        x = "", y = "Win proportion") + 
@@
   labs(title = "British players",
-       subtitle = "21 players, n=249",
+       subtitle = "26 players, n=322",
+              # subtitle = "21 players, n=249",
        x = "", y = " ") + 
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("../plots/home_court.jpg", ., height=5.5, width=graphic_width_long)
+  ggsave("../plots/home_court_v2.jpg", ., height=5.5, width=graphic_width_long)
 }
@@
 all4 <- 
-  gs_players %>% 
-  group_by(name) %>% 
-  summarize(blah = length(unique(tournament))) %>%
-  filter(blah == 4) %>%
-  pull(name)
+  gs_players_v2 %>% 
+  dplyr::group_by(name) %>% 
+  dplyr::summarize(tourn_n = length(unique(tournament))) %>%
+  dplyr::filter(tourn_n == 4) %>%
+  dplyr::pull(name)
 
-france_wins <- dplyr::filter(gs_players, ioc == "FRA", round < "QF", name %in% all4) %>% 
+france_wins <- dplyr::filter(gs_players_v2, ioc == "FRA", round < "QF", name %in% all4) %>% 
   dplyr::group_by(tournament, round) %>% 
@@
 
-us_wins <- dplyr::filter(gs_players, ioc == "USA", round < "QF", name %in% all4) %>% 
+us_wins <- dplyr::filter(gs_players_v2, ioc == "USA", round < "QF", name %in% all4) %>% 
   dplyr::group_by(tournament, round) %>% 
@@
 
-aus_wins <- filter(gs_players, ioc == "AUS", round < "QF", name %in% all4) %>% 
+aus_wins <- filter(gs_players_v2, ioc == "AUS", round < "QF", name %in% all4) %>% 
   group_by(tournament, round) %>% 
@@
 
-gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF", name %in% all4) %>% 
+gbr_wins <- filter(gs_players_v2, ioc == "GBR", round < "QF", name %in% all4) %>% 
   group_by(tournament, round) %>% 
@@
                            gp=gpar(fontfamily="serif", fontsize = 20))) %>%
-  ggsave("../plots/home_court_all4.jpg", ., height=5, width=graphic_width_long)
+  ggsave("../plots/home_court_all4_v2.jpg", ., height=5, width=graphic_width_long)
 }
@@
 ```{r}
-spain_wins <- filter(gs_players, ioc == "ESP", round < "QF") %>% 
+spain_wins <- filter(gs_players_v2, ioc == "ESP", round < "QF") %>% 
   group_by(tournament, round) %>% 
-  dplyr::summarise(win_prop = mean(did_win),
+  dplyr::summarise(win_prop = mean(did_win, na.rm=TRUE),
             win_prop_se = sqrt(win_prop*(1-win_prop)/n()),
             avg_ranking = mean(rank),
-            rank_se = sd(rank)/sqrt(n()))
+            rank_se = sd(rank)/sqrt(n()),
+            n_size = n())
 
-prop.table(with(filter(gs_players, ioc == "ESP", round == "R128"),
+prop.table(with(filter(gs_players_v2, ioc == "ESP", round == "R128"),
                 table(did_win, tournament)), 2)
 
-with(filter(gs_players, ioc == "ESP", round == "R128"),
+with(filter(gs_players_v2, ioc == "ESP", round == "R128"),
                 table(did_win, tournament))
@@
 ################################################################################
-with(filter(gs_players, ioc == "ESP", round == "R128"),
+with(filter(gs_players_v2, ioc == "ESP", round == "R128"),
      table(did_win, tournament)) %>%
   chisq.test(correct = TRUE)
-with(filter(gs_players, ioc == "ESP", round == "R64"),
+with(filter(gs_players_v2, ioc == "ESP", round == "R64"),
      table(did_win, tournament)) %>%
   chisq.test(correct = TRUE)
-with(filter(gs_players, ioc == "ESP", round == "R32"),
+with(filter(gs_players_v2, ioc == "ESP", round == "R32"),
      table(did_win, tournament)) %>%
   chisq.test(correct = TRUE)
-with(filter(gs_players, ioc == "ESP", round == "R16"),
+with(filter(gs_players_v2, ioc == "ESP", round == "R16"),
      table(did_win, tournament)) %>%
@@
 ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
-  ggsave("../plots/spainards.jpg", ., height=6, width=graphic_width_long)
+  ggsave("../plots/spainards_v2.jpg", ., height=6, width=graphic_width_long)
 }
@@
 ```{r, fig.width=12, fig.height=6, fig.align='center'}
-ggplot(gs_partial, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
+ggplot(gs_partial_v2, aes(w_n_ue + l_n_ue, w_n_aces + l_n_aces,
                            label = tournament, color=tournament)) + 
-  geom_jitter(size=2) +
+  geom_jitter(size=1.8) + # used to be size=2 in v1
   labs( title = "Distribution of errors and aces differ by tournament",
@@
 if(save_graph){
-ggsave("../plots/errors_aces2.jpg", height = 6, width = graphic_width_short)
+ggsave("../plots/errors_aces2_v2.jpg", height = 6, width = graphic_width_short)
 }
@@
 atp_servers <- sapply(atp_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
+                      function(x) return(sample(which(gs_players_v2$name == x), 2))) %>%
+  as.numeric() %>% gs_players_v2[., ]
 
@@
 wta_servers <- sapply(wta_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
+                      function(x) return(sample(which(gs_players_v2$name == x), 2))) %>%
+  as.numeric() %>% gs_players_v2[., ]
 
-p_atp <- gs_players %>% filter(league == "ATP") %>%
+p_atp <- gs_players_v2 %>% filter(league == "ATP") %>%
 ggplot( aes(log(ht+1), log(ace+1))) + 
@@
 
-p_wta <- gs_players %>% filter(league == "WTA") %>%
+p_wta <- gs_players_v2 %>% filter(league == "WTA") %>%
 ggplot(aes(log(ht+1), log(ace+1))) + 
@@
 if(save_graph){
-ggsave("../plots/aces2.jpg", height=7, width=graphic_width_long)
+ggsave("../plots/aces2_v2.jpg", height=7, width=graphic_width_long)
 }
@@
 wta_servers <- sapply(wta_servers,
-                      function(x) return(sample(which(gs_players$name == x), 2))) %>%
-  as.numeric() %>% gs_players[., ]
+                      function(x) return(sample(which(gs_players_v2$name == x), 2))) %>%
+  as.numeric() %>% gs_players_v2[., ]
 
-p_atp_percent <- gs_players %>% filter(league == "ATP") %>%
+p_atp_percent <- gs_players_v2 %>% filter(league == "ATP") %>%
 ggplot( aes(log(ht+1), ace/svpt)) + 
@@
 
-p_wta_percent <- gs_players %>% filter(league == "WTA") %>%
+p_wta_percent <- gs_players_v2 %>% filter(league == "WTA") %>%
 ggplot(aes(log(ht+1), ace/svpt)) + 
@@
 if(save_graph){
-ggsave("../plots/aces2_percent.jpg", height=8, width=graphic_width_long)
+ggsave("../plots/aces2_percent_v2.jpg", height=8, width=graphic_width_long)
 }
@@
 
-atp_ace_ht_tour <-   gs_players %>% 
+atp_ace_ht_tour <-   gs_players_v2 %>% 
   filter(league == "ATP") %>%
@@
 
-wta_ace_ht_tour <-   gs_players %>% 
+wta_ace_ht_tour <-   gs_players_v2 %>% 
   filter(league == "WTA") %>%
@@
 #                       function(x) return(sample(which(wta_ace_ht_tour$name == x), 2))) %>%
-#   as.numeric() %>% gs_players[., ]
+#   as.numeric() %>% gs_players_v2[., ]
 
@@
 if(save_graph){
-ggsave("../plots/ace_ht_facet_atp.jpg", height=8, width=graphic_width_short)
+ggsave("../plots/ace_ht_facet_atp_v2.jpg", height=8, width=graphic_width_short)
 }
@@
 if(save_graph){
-ggsave("../plots/ace_ht_facet_wta.jpg", height=8, width=graphic_width_short)
+ggsave("../plots/ace_ht_facet_wta_v2.jpg", height=8, width=graphic_width_short)
 }
+
+```
+
+#### Using ace percentage.... one dot per player - black white
+
+```{r, fig.width=16, fig.height=8, fig.align='center'}
+set.seed(8)
+
+atp_ace_ht <-   gs_players_v2 %>% 
+  filter(league == "ATP") %>%
+  dplyr::group_by(name) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+wta_ace_ht <-   gs_players_v2 %>% 
+  filter(league == "WTA") %>%
+  dplyr::group_by(name) %>%
+  dplyr::summarise(ht = maxish(ht),
+                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
+                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
+                   count_n = n())
+
+p_atp_percent <- atp_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5, width = 0.8) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 6,
+            nudge_y = c(-.025, 0, 0, -0.05,0.02),
+            nudge_x = c(6, -2, -2, -3, -1)) +
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 6,
+            nudge_y = c(-.025, 0, 0, -0.05,0.02),
+            nudge_x = c(6, -2, -2, -3, -1)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
+  my_theme 
+# + 
+#   ylim(c(0, 0.25))
+p_atp_percent
+
+p_wta_percent <- wta_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5, width=.6) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 6,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-3, 5, -3, 0, 0)) + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 6,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-3, 5, -3, 0, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "WTA") +
+  my_theme
+# + 
+#   ylim(c(0, 0.25))
+
+p_wta_percent
+
+ggarrange(p_atp_percent, NULL, p_wta_percent, common.legend = TRUE, legend="bottom", widths = c(1, 0.08, 1), nrow = 1)
+
+if(save_graph){
+ggsave("../plots/aces_median_percent_BW2_v2.jpg", height=8, width=graphic_width_long)
+}
+
+p_atp_percent <- atp_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5, width = 0.8) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
+            nudge_x = c(5, 5, -5, -3, 0)) +
+  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
+  my_theme +
+  ylim(c(0, 0.25))
+
+p_wta_percent <- wta_ace_ht %>% # could also try median
+ggplot( aes(ht, med_ace_perc)) + 
+  geom_jitter(size=1.5, width = 0.6) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, med_ace_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 5,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_x = c(-1, 5, 5, 0, 0)) +
+  labs(x = "Height in cm", y = " ", title = "WTA") +
+  my_theme +
+  ylim(c(0, 0.25))
+
+# ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom",
+#           nrow=1) 
+# 
+# if(save_graph){
+# ggsave("../plots/aces_median_percent_BW1_v2.jpg", height=8, width=graphic_width_long)
+# }
 
@@
 
-<!-- atp_ace_ht_tour <-   gs_players %>%  -->
+<!-- atp_ace_ht_tour <-   gs_players_v2 %>%  -->
 <!--   filter(league == "ATP") %>% -->
@@
 
-<!-- wta_ace_ht_tour <-   gs_players %>%  -->
+<!-- wta_ace_ht_tour <-   gs_players_v2 %>%  -->
 <!--   filter(league == "WTA") %>% -->
@@
 <!-- atp_servers <- sapply(atp_servers, -->
-<!--                       function(x) return(sample(which(gs_players$name == x), 2))) %>% -->
+<!--                       function(x) return(sample(which(gs_players_v2$name == x), 2))) %>% -->
 <!--   as.numeric() %>%  -->
-<!--   gs_players[., ]  %>% -->
+<!--   gs_players_v2[., ]  %>% -->
 <!--   dplyr::mutate(med_ace_perc = median(ace/svpt, na.rm=TRUE)) -->
@@
 <!-- wta_servers <- sapply(wta_servers, -->
-<!--                       function(x) return(sample(which(gs_players$name == x), 2))) %>% -->
-<!--   as.numeric() %>% gs_players[., ] %>% -->
+<!--                       function(x) return(sample(which(gs_players_v2$name == x), 2))) %>% -->
+<!--   as.numeric() %>% gs_players_v2[., ] %>% -->
 <!--   dplyr::mutate(med_ace_perc = median(ace/svpt, na.rm=TRUE)) -->
@@
 <!-- if(save_graph){ -->
-<!-- ggsave("../plots/aces_median_percent_BW2.jpg", height=8, width=graphic_width_long) -->
+<!-- ggsave("../plots/aces_median_percent_BW2_v2.jpg", height=8, width=graphic_width_long) -->
 <!-- } -->
@@
 <!-- ``` -->
-
-#### Using ace percentage.... one dot per player - black white
-
-```{r, fig.width=16, fig.height=8, fig.align='center'}
-set.seed(8)
-
-atp_ace_ht <-   gs_players %>% 
-  filter(league == "ATP") %>%
-  dplyr::group_by(name) %>%
-  dplyr::summarise(ht = maxish(ht),
-                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
-                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
-                   count_n = n())
-
-wta_ace_ht <-   gs_players %>% 
-  filter(league == "WTA") %>%
-  dplyr::group_by(name) %>%
-  dplyr::summarise(ht = maxish(ht),
-                   avg_ave_perc = mean(ace/svpt, na.rm=TRUE),
-                   med_ace_perc = median(ace/svpt, na.rm=TRUE),
-                   count_n = n())
-
-p_atp_percent <- atp_ace_ht %>% # could also try median
-ggplot( aes(ht, med_ace_perc)) + 
-  geom_jitter(size=1.5) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
-            aes(ht, med_ace_perc, label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
-            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
-            nudge_x = c(5, 5, -5, -3, 0)) +
-  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
-            aes(ht, med_ace_perc, label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 5,
-            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
-            nudge_x = c(5, 5, -5, -3, 0)) +
-  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
-  my_theme 
-# + 
-#   ylim(c(0, 0.25))
-p_atp_percent
-
-p_wta_percent <- wta_ace_ht %>% # could also try median
-ggplot( aes(ht, med_ace_perc)) + 
-  geom_jitter(size=1.5) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
-            aes(ht, med_ace_perc, label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
-            nudge_x = c(-1, 5, 5, 0, 0)) + 
-  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
-            aes(ht, med_ace_perc, label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
-            nudge_x = c(-1, 5, 5, 0, 0)) +
-  labs(x = "Height in cm", y = "Median ace% per match", title = "WTA") +
-  my_theme
-# + 
-#   ylim(c(0, 0.25))
-
-p_wta_percent
-
-ggarrange(p_atp_percent, NULL, p_wta_percent, common.legend = TRUE, legend="bottom", widths = c(1, 0.08, 1), nrow = 1)
-
-if(save_graph){
-ggsave("../plots/aces_median_percent_BW2.jpg", height=8, width=graphic_width_long)
-}
-
-p_atp_percent <- atp_ace_ht %>% # could also try median
-ggplot( aes(ht, med_ace_perc)) + 
-  geom_jitter(size=1.5) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
-            aes(ht, med_ace_perc, label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
-            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
-            nudge_x = c(5, 5, -5, -3, 0)) +
-  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
-            aes(ht, med_ace_perc, label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 5,
-            nudge_y = c(-.075, 0, 0, -0.05,0.02 ),
-            nudge_x = c(5, 5, -5, -3, 0)) +
-  labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
-  my_theme +
-  ylim(c(0, 0.25))
-
-p_wta_percent <- wta_ace_ht %>% # could also try median
-ggplot( aes(ht, med_ace_perc)) + 
-  geom_jitter(size=1.5) + 
-  geom_smooth( method = "lm", color="black") + 
-  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
-            aes(ht, med_ace_perc, label=name),
-            alpha=.8, seed=1234, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
-            nudge_x = c(-1, 5, 5, 0, 0)) + 
-  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
-            aes(ht, med_ace_perc, label = name),
-            seed=1234, fill=NA, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
-            nudge_x = c(-1, 5, 5, 0, 0)) +
-  labs(x = "Height in cm", y = " ", title = "WTA") +
-  my_theme +
-  ylim(c(0, 0.25))
-
-ggarrange(p_atp_percent, p_wta_percent, common.legend = TRUE, legend="bottom",
-          nrow=1)
-
-if(save_graph){
-ggsave("../plots/aces_median_percent_BW1.jpg", height=8, width=graphic_width_long)
-}
-
-```

### 2020-08-02 paper_results/EDA/graphics_eda.Rmd
more graph updates WOO

@@
             rank_se = sd(rank)/sqrt(n()),
-            n_size = n())
+            n_size = n(),
+            med_ranking = median(rank, na.rm = TRUE))
 
@@
 }
+
+ggplot(spain_wins, aes(round, med_ranking, color=tournament, group=tournament, fill = tournament)) + 
+  geom_line(size=3) + 
+  labs(title = "Median ranking among Spaniards",
+       x = " ",
+       y = "Median rank") + 
+  scale_color_manual(" ", values=tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)",
+                                           "French Open\n(Clay)",
+                                           "US Open\n(DecoTurf)",
+                                           "Wimbledon\n(Grass)")) + my_theme +
+  guides(color = guide_legend(override.aes = list(size = 7)))
 
@@
             alpha=.8, seed=1234, show.legend=FALSE, size = 6,
-            nudge_y = c(-.025, 0, 0, -0.05,0.02),
-            nudge_x = c(6, -2, -2, -3, -1)) +
+            nudge_y = c(-.025, 0, -.01, -0.05,0.02),
+            nudge_x = c(6, -2, 0, -3, -1)) +
   geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
@@
             seed=1234, fill=NA, show.legend=FALSE, size = 6,
-            nudge_y = c(-.025, 0, 0, -0.05,0.02),
-            nudge_x = c(6, -2, -2, -3, -1)) +
+            nudge_y = c(-.025, 0, -0.01, -0.05,0.02),
+            nudge_x = c(6, -2, 0, -3, -1)) +
   labs(x = "Height in cm", y = "Median ace% per match", title = "ATP") +
@@
             alpha=.8, seed=1234, show.legend=FALSE, size = 6,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.03 ),
             nudge_x = c(-3, 5, -3, 0, 0)) + 
@@
             seed=1234, fill=NA, show.legend=FALSE, size = 6,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.03 ),
             nudge_x = c(-3, 5, -3, 0, 0)) +
@@
             alpha=.8, seed=1234, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.01),
             nudge_x = c(-1, 5, 5, 0, 0)) + 
@@
             seed=1234, fill=NA, show.legend=FALSE, size = 5,
-            nudge_y = c(-.02, .01, 0.01, -0.01, -0.02 ),
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.01),
             nudge_x = c(-1, 5, 5, 0, 0)) +
@@
 # }
+
+p_atp_percent <- atp_ace_ht %>% # could also try median
+ggplot( aes(ht, avg_ave_perc)) + 
+  geom_jitter(size=1.5, width = 0.8) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, avg_ave_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 6,
+            nudge_y = c(-.025, 0, -.01, -0.05,0.02),
+            nudge_x = c(6, -2, 0, -3, -1)) +
+  geom_label_repel(data=dplyr::filter(atp_ace_ht, name %in% atp_servers),
+            aes(ht, avg_ave_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 6,
+            nudge_y = c(-.025, 0, -0.01, -0.05,0.02),
+            nudge_x = c(6, -2, 0, -3, -1)) +
+  labs(x = "Height in cm", y = "Mean ace% per match", title = "ATP") +
+  my_theme 
+# + 
+#   ylim(c(0, 0.25))
+p_atp_percent
+
+p_wta_percent <- wta_ace_ht %>% # could also try median
+ggplot( aes(ht, avg_ave_perc)) + 
+  geom_jitter(size=1.5, width=.6) + 
+  geom_smooth( method = "lm", color="black") + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, avg_ave_perc, label=name),
+            alpha=.8, seed=1234, show.legend=FALSE, size = 6,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.03 ),
+            nudge_x = c(-3, 5, -3, 0, 0)) + 
+  geom_label_repel(data=dplyr::filter(wta_ace_ht, name %in% wta_servers),
+            aes(ht, avg_ave_perc, label = name),
+            seed=1234, fill=NA, show.legend=FALSE, size = 6,
+            nudge_y = c(-.02, .01, 0.01, -0.01, -0.03 ),
+            nudge_x = c(-3, 5, -3, 0, 0)) +
+  labs(x = "Height in cm", y = "Mean ace% per match", title = "WTA") +
+  my_theme
+# + 
+#   ylim(c(0, 0.25))
+
+p_wta_percent
+
+ggarrange(p_atp_percent, NULL, p_wta_percent, common.legend = TRUE, legend="bottom", widths = c(1, 0.08, 1), nrow = 1)
+
+if(save_graph){
+ggsave("../plots/aces_mean_percent_v2.jpg", height=8, width=graphic_width_long)
+}
 

### 2020-09-17 paper_results/Hierarchical/fit-and-plot-revision2.R
plz be last commit

@@ -61,19 +61,18 @@ ind_logistic_noioc = lme4::glmer(did_win ~ late_round + log(rank) + log(opponent
                                    year_fac + atp + (0 + tournament |name_fac),
                                  data = gs_players, family = "binomial", nAGQ =0)
 
-logistic_newmod = lme4::glmer(did_win ~ 0 + late_round + log(rank) + log(opponent_rank) +
-                                    scale(age) + (0 + tournament |name_fac),
-                                 data = gs_players, family = "binomial", nAGQ =0)
+logistic_newmod = lme4::glmer(did_win ~ 0 + log(rank) + log(opponent_rank) + scale(age) + (0 + tournament |name_fac),
+                              data = gs_players, family = "binomial", nAGQ =0)
 
-aces_new = lme4::glmer(cbind(n_aces, total_points - n_aces) ~ late_round + log(rank) + log(opponent_rank) + scale(age) + atp +
+aces_new = lme4::glmer(cbind(n_aces, total_points - n_aces) ~ late_round + log(rank) + log(opponent_rank) + scale(age) +  atp +
                         (0 + tournament |name_fac),
                       data = gs_partial_players, family = "binomial", nAGQ = 0)
 
-nets_new = lme4::glmer(cbind(n_netpt_w, total_points - n_netpt_w) ~ late_round + log(rank) + log(opponent_rank) + scale(age) + atp +
+nets_new = lme4::glmer(cbind(n_netpt_w, total_points - n_netpt_w) ~ late_round + log(rank) + log(opponent_rank) + scale(age) +  atp +
                          (0 + tournament |name_fac),
                        data = gs_partial_players, family = "binomial", nAGQ = 0)
 
-ue_new = lme4::glmer(cbind(n_ue, total_points - n_ue) ~ late_round + log(rank) + log(opponent_rank) + scale(age) + atp +
+ue_new = lme4::glmer(cbind(n_ue, total_points - n_ue) ~ late_round + log(rank) + log(opponent_rank) + scale(age) +  atp +
                         (0 + tournament |name_fac),
                        data = gs_partial_players, family = "binomial", nAGQ = 0)
 
@@ -125,6 +124,7 @@ fixed.effects.plot = tidy(logistic_newmod) %>%
                  ymax = estimate + 2*std.error)) +
[…]

### 2020-09-19 R/clean-data.R
updated graphics and figures for indiivdual models

@@ -224,7 +224,7 @@ join_gs_pbp <- function(gs_pbp, gs){
                               loser_rank, winner_age, winner_ioc, Retirement,
                               w_pointswon, w_gameswon, w_setswon,
                               l_pointswon, l_gameswon, l_setswon,
-                              round) %>%
+                              round, league) %>%
         summarize(a1 = sort(c(winner_name, loser_name))[1],
                   a2 = sort(c(winner_name, loser_name), decreasing = TRUE)[1])
 

### 2020-09-19 data-raw/make-gs-data_updated.R

@@ -1,3 +1,4 @@
+## Updated in August 2020 to get data up through 2019
 ## Script to make data used in our analysis, specifically
 ##
 ## 1. gs.rda

### 2020-09-19 paper_results/Individual/.#most_matches.R

@@ -1,0 +1,1 @@
+shannon@sgarch.118036:1600447784

### 2020-09-19 paper_results/Individual/rafa-fed-serena.R

@@ -1,0 +1,25 @@
+## August 21, 2020
+## Script to get number of matches of these three between 2013-2019
+
+
+devtools::load_all()
+library(dplyr)
+data(gs_players_v2)
+    
+goat <- gs_players_v2 %>%
+    filter(name %in% c("Serena Williams",
+                       "Roger Federer",
+                       "Rafael Nadal"))
+
+tab <- goat %>% group_by(name,
+                  tournament) %>%
+    summarize(n_played = n())
+
+tab %>% 
+    print(n = 100)
+
+tab %>% group_by(name) %>%
+    summarize(n = sum(n_played))
+
+data(wta_matches)
+table(gs_partial_v2$league, gs_partial_v2$year)

### 2020-09-19 paper_results/Individual/.#most_matches.R
updated graphics and figures for indiivdual models

@@ -1,1 +1,0 @@
-shannon@sgarch.118036:1600447784

### 2020-09-19 paper_results/Individual/most_matches.R

@@ -124,6 +124,8 @@ coefs[aus_vars[2]]
 
 
 ## plot diags
+my_theme <-  theme_bw(base_size = 20) + # White background, black and white theme
+  theme(text = element_text(family="serif"))
 g <- ggDiagnose(out_list$final_model, which = 1:6, return = TRUE)
 ggs <- g$ggout
 ggthm <- lapply(ggs, function(gg) gg + my_theme)

### 2020-09-20 paper_results/EDA/graphics_eda.Rmd
change spaniards to spanish players

@@
 p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
-  labs(title = "Win proportion among Spaniards",
+  labs(title = "Spanish Player Performance",
        x = "", y = "Win proportion") + 
@@
   geom_line(size=3) + 
-  labs(title = "Average ranking among Spaniards",
+  labs(title = "Ranking of Spanish Players",
        x = " ",
@@
 
-ggplot(spain_wins, aes(round, med_ranking, color=tournament, group=tournament, fill = tournament)) + 
+p2 <- ggplot(spain_wins, aes(round, med_ranking, color=tournament, group=tournament, fill = tournament)) + 
   geom_line(size=3) + 
-  labs(title = "Median ranking among Spaniards",
+  labs(title = "Spanish Player Rankings",
        x = " ",
@@
   guides(color = guide_legend(override.aes = list(size = 7)))
+
+if(save_graph){
+ggarrange( p1, p2, nrow=1, ncol=2, common.legend = TRUE, legend="bottom") %>%
+  ggsave("../plots/spainards_v2_med.jpg", ., height=6, width=graphic_width_long)
+}
 
@@
 ggsave("../plots/errors_aces2_v2.jpg", height = 6, width = graphic_width_short)
+}
+
+which((gs_partial_v2$w_n_ue + gs_partial_v2$l_n_ue)/(gs_partial_v2$w_pointswon + gs_partial_v2$l_pointswon) > 1)
+
+gs_partial_v2[-c(1181), ] %>% 
+ggplot(aes((w_n_ue + l_n_ue)/(w_pointswon + l_pointswon),
+                          (w_n_aces + l_n_aces)/(w_pointswon + l_pointswon),
+                           label = tournament, color=tournament)) + 
+  geom_jitter(size=1.8) + # used to be size=2 in v1
+  labs( title = "Distribution of errors and aces by slam",
+        x = "UE% = unforced errors / total points", y = "ace% = aces / total points") +
+  my_theme + scale_color_manual(" ",
+                                values = tournament_colors,
+                                labels = c("Australian Open\n(Plexicushion)\n",
+                                           "French Open\n(Clay)\n",
+                                           "US Open\n(DecoTurf)\n",
+                                           "Wimbledon\n(Grass)\n")) +
+  stat_ellipse(type="t", size = 2) +
+  guides(color = guide_legend(override.aes = list(size = 7, shape=15)))
+
+if(save_graph){
+ggsave("../plots/errors_aces2_v2_perc.jpg", height = 6, width = graphic_width_short)
 }
